### 2021-11-27 getdata.Rmd
first commit

---
title: "Get population and temperature data"
author: "Göran Broström"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(eha)
library(skum)
options(show.signif.stars = FALSE)
g <- function(x) (log(1 + 365 * x))^3
```

# Population data

We use the observation file rather than the  population file, so we can get 
correct parish information.

```{r readobs}
bb <- obs$id[obs$starttyp == 2]
births <- obs[obs$id %in% bb, 
    c("id", "sex", "birthdate", "nofrs", "sluttyp", "socStatus", "hisclass", 
      "enter", "exit", "ort", "ortnmn")]
births$event <- as.numeric(births$sluttyp == 2)
births <- age.window(births, c(0, 1))
births <- births[order(births$id, births$enter), ]
##indx <- tapply(births$id, births$id)
##births$mexit <- tapply(births$exit, births$id, max)[indx]
##births$mevent <- tapply(births$event, births$id, max)[indx]
##births <- births[!duplicated(births$id), ]
### Deaths:
did <- unique(births$id[births$event == 1])
deaths <- births[births$id %in% did, ]
indx <- tapply(deaths$id, deaths$id)
deaths$menter <- tapply(deaths$enter, deaths$id, min)[indx]
deaths <- deaths[!duplicated(deaths$id), ]
deaths$enter <- deaths$menter
deaths$menter <- NULL
deaths$enter <- 0 ## satisfied....
### Survivors:
survs <- births[!(births$id %in% did), ]
indx <- tapply(survs$id, survs$id)
survs$mexit <- tapply(survs$exit, survs$id, max)[indx]
survs <- survs[!duplicated(survs$id), ]
survs$exit <- survs$mexit
survs$mexit <- NULL
survs$enter <- 0 ## Ok ...
### Now, give birth to NEW births:
births <- rbind(deaths, survs)
births <- births[births$birthdate > 1895, ]

indx <- match(births$id, pers$id)
births$ab <- pers$ab[indx]
##births$df <- pers$df[indx]
births$illeg <- factor(births$ab == 2, labels = c("no", "yes"))
births$ab <- NULL
births$sluttyp <- NULL
births$parity <- pers$paritet_g[indx]
summary(births)                                
```

We put on 'subreg':

```{r subreg}
births$subreg <- factor(births$nofrs)
levels(births$subreg) <- c(rep("ume", 2), rep("bjur", 4), rep("sten", 3))
## Remove "sten":
births <- births[births$subreg != "sten", ]
births$subreg <- factor(births$subreg) # Remove level "sten"
```

# Neonatal mortality, an overview

```{r neo, fig.cap = "Neonatal mortality by year, all regions."}
neo <- age.window(births, c(0, 1 / 13))
surv <- as.matrix(neo[, c("enter", "exit", "event", "birthdate")])
pall <- perstat(surv, period = 1900:1951)$intensity
plot(1900:1950, pall, type = "b", ylim = c(0, 0.5), col = "blue", xlab = "Year",
     ylab = "Mortality")
abline(h = 0)
abline(h = 0.3, lty = 3)
```

## By region

```{r neoby, fig.cap = "Neonatal mortality by year  and region."}
usurv <- as.matrix(neo[neo$subreg == "ume", c("enter", "exit", "event", "birthdate")])
pume <-  perstat(usurv, period = 1900:1951)$intensity
plot(1900:1950, pall, type = "b", ylim = c(0, 0.7), col = "blue", xlab = "Year",
     ylab = "Mortality")
bsurv <- as.matrix(neo[neo$subreg == "bjur", c("enter", "exit", "event", "birthdate")])
pjur <-  perstat(bsurv, period = 1900:1951)$intensity
lines(1900:1950, pjur, type = "b", ylim = c(0, 0.5), col = "red")
##ssurv <- as.matrix(neo[neo$subreg == "sten", c("enter", "exit", "event", "birthdate")])
##sjur <-  perstat(ssurv, period = 1900:1951)$intensity
##lines(1900:1950, sjur, type = "b", ylim = c(0, 0.5), col = "black")
abline(h = 0)
abline(h = 0.3, lty = 3)
```

# Post-neonatal mortality, an overview

```{r pnneo, fig.cap = "Post-eonatal mortality by year, all regions."}
pneo <- age.window(births, c(1 / 13, 1))
surv <- as.matrix(pneo[, c("enter", "exit", "event", "birthdate")])
pall <- perstat(surv, period = 1900:1951)$intensity
plot(1900:1950, pall, type = "b", ylim = c(0, 0.1), col = "blue",
     xlab = "Year", ylab = "Mortality")
abline(h = 0)
abline(h = 0.3, lty = 3)
```

## By region

```{r pneoby, fig.cap = "Post-neonatal mortality by year and regions."}
usurv <- as.matrix(pneo[pneo$subreg == "ume", c("enter", "exit", "event", "birthdate")])
pume <-  perstat(usurv, period = 1900:1951)$intensity
plot(1900:1950, pall, type = "b", ylim = c(0, 0.1), col = "blue",
     xlab = "Year", ylab = "Mortality")
bsurv <- as.matrix(pneo[pneo$subreg == "bjur", c("enter", "exit", "event", "birthdate")])
pjur <-  perstat(bsurv, period = 1900:1951)$intensity
lines(1900:1950, pjur, type = "b", ylim = c(0, 0.5), col = "red")
##ssurv <- as.matrix(pneo[pneo$subreg == "sten", c("enter", "exit", "event", "birthdate")])
##sjur <-  perstat(ssurv, period = 1900:1951)$intensity
##lines(1900:1950, sjur, type = "b", ylim = c(0, 0.5), col = "black")
abline(h = 0)
abline(h = 0.3, lty = 3)
```

# Temperature data

We have two stations, *Umeå* and *Bjuröklubb*.

```{r gettemp}
source("R/tempdat.R")
umetemp <- tempdat("ume", c(1895, 1950))
bjurtemp <- tempdat("bjur", c(1895, 1950))
##stentemp <- tempdat("~/Forskning/Data/sten_temp.csv", c(1899, 1952))
saveRDS(umetemp, file = "data/umetemp.rds")
saveRDS(bjurtemp, file = "data/bjurtemp.rds")
```

# Make communal

```{r makecommunal}
##source("R/make_comm.R")
lagg <- 3 / 365  ## NOTE!!
## Ume:
bume <- births[births$subreg == "ume", ]
bume <- make.communal(bume, umetemp["heat"], start = 1895, period = 1/52, lag = lagg)
bume <- bume[bume$enter < bume$exit, ]
bume <- make.communal(bume, umetemp["cold"], start = 1895, period = 1 / 52, lag = lagg)
bume <- bume[bume$enter < bume$exit, ]
bume <- make.communal(bume, umetemp["week"], start = 1895, period = 1 / 52, lag = lagg)
bume <- bume[bume$enter < bume$exit, ]
### Ske:
bskel <- births[births$subreg == "bjur", ]
bskel <- make.communal(bskel, bjurtemp["heat"], start = 1895, period = 1/52, lag = lagg)
bskel <- bskel[bskel$enter < bskel$exit, ]
bskel <- make.communal(bskel, bjurtemp["cold"], start = 1895, period = 1 / 52, lag = lagg)
bskel <- bskel[bskel$enter < bskel$exit, ]
bskel <- make.communal(bskel, bjurtemp["week"], start = 1895, period = 1 / 52, lag = lagg)
bskel <- bskel[bskel$enter < bskel$exit, ]
##bsten <- births[births$subreg == "sten", ]
##bsten <- make_comm(bsten, stentemp, start = 1899, period = 1/52, lag = lagg)
bume$region <- "ume"
bskel$region <- "ske"
infdat <- rbind(bume, bskel)
infdat <- infdat[order(infdat$id, infdat$enter), c("region", "id", "sex", "enter", "exit", "event", 
                                                   "birthdate", "socStatus", "hisclass", "illeg",
                                                   "parity", "heat", "cold", "week")]
saveRDS(infdat, file = "data/infdat.rds")
```

Done!

### 2021-11-27 getdata2.Rmd
first commit

--- 2021-11-27 getdata.Rmd
@@
 ---
-title: "Get population and temperature data"
+title: "Get population and temperature data, version 2"
 author: "Göran Broström"
@@
 
+# Introduction
+
+The only (so far) difference compared to the article "getdata.Rmd" is that time 
+here is measured in *days* instead of in a fraction of a year.
+
 # Population data
@@
 births <- obs[obs$id %in% bb, 
-    c("id", "sex", "birthdate", "nofrs", "sluttyp", "socStatus", "hisclass", 
-      "enter", "exit", "ort", "ortnmn")]
+    c("id", "sex", "foddat", "startdat", "slutdat", "nofrs", "sluttyp", "socStatus", "hisclass", 
+      "ort", "ortnmn")]
 births$event <- as.numeric(births$sluttyp == 2)
-births <- age.window(births, c(0, 1))
+## New start in days...:
+births$enter <- as.numeric(with(births, difftime(startdat, foddat, unit = "days")))
+births$exit <- as.numeric(with(births, difftime(slutdat, foddat, unit = "days")))
+births <- age.window(births, c(0, 365))
 births <- births[order(births$id, births$enter), ]
@@
 births <- rbind(deaths, survs)
-births <- births[births$birthdate > 1895, ]
+births <- births[births$foddat > as.Date("1894-12-31"), ]
 
@@
 ```{r neo, fig.cap = "Neonatal mortality by year, all regions."}
-neo <- age.window(births, c(0, 1 / 13))
+neo <- births
+neo$exit <- neo$exit / 365
+neo$birthdate <- toTime(neo$foddat)
+neo <- age.window(neo, c(0, 28 / 365))
+##neo$exit <- 365 * neo$exit
 surv <- as.matrix(neo[, c("enter", "exit", "event", "birthdate")])

### 2021-11-27 neo.Rmd
first commit

---
title: "Neonatal models"
author: "Göran Broström"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
bibliography: bio.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage{a4wide}    
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
abstract: "The effect of extreme temperatures on neonatal mortality in the Umeå region 1895-1950 is studied as an example of how to do it with tabular, anonymised data."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment = NA)
```

# Introduction

How to analyse table data with the same statistical models, e.e., *Cox regression*
as you are used to with individual, interval based data, is shown here. You need the 
**R** package *eha* [@eha; @ehar2] for easy conversion of individual data to tabular
suitable for survival analysis.

# Example: Neonatal mortality in Umeå/Skellefteå, 1895--1950

We use a data set for studying infant mortality from the *Demographic Database*,
Umeå University.

## Read data

Start by reading the prepared file and select a few columns, remove zero-length
spells, and put on reasonable row names:

```{r readdata}
infdat <- readRDS("data/infdat.rds")[, c("id", "enter", "exit", "event", 
                                         "sex", "socStatus", "illeg",  
                                         "parity", "cold", "week")]
## remove zero-length spells:
infdat <- infdat[infdat$exit > infdat$enter + 1.e-6, ] 
rownames(infdat) <- 1:NROW(infdat)
NROW(infdat)
length(unique(infdat$id))
head(infdat)
```

Then, cut the spells to fit into the first four weeks of life. Use the function 
*age.window* from the package *eha*.

```{r agecut}
library(eha)
neo <- age.window(infdat, c(0, 28 / 365))
```

## Transform time unit 'years' to 'days' and "round", down and up:

Note that the original data file measures time in *years*. For neonatal mortality,
a more convenient time unit is *days*, so we convert and round.

```{r transform}
neo$enter <- round(365 * neo$enter)
neo$exit <- round(365 * neo$exit)
neo <- neo[neo$exit > neo$enter, ] # Remove zero-length spells

dim(neo)
head(neo)
```

## Cleaning data

Since this is just an example showing how to analyse given data, we clean data in 
order to avoid discussions about missing values and other disturbances.

```{r clean}
neo <- neo[neo$parity > 0, ] # Remove missing = -1
neo$parity <- cut(neo$parity, c(0, 1.5, 3.5, 18.5), labels = c("1", "2-3", "4+"))
neo <- neo[!is.na(neo$socStatus), ] # Remove records with missing values
```

Next, *tabulate* data with the aid of the function *toTpch*, found in *eha*:

```{r tabulate}
tneo <- toTpch(Surv(enter, exit, event) ~  
                   sex + socStatus + illeg + parity + cold, 
               data = neo, cuts = 0:28)
dim(tneo)
head(tneo)
summary(tneo)
saveRDS(tneo, file = "data/tneo.rds")
```

Note how it works: For each combination of *socStatus*, *illeg*, *parity* *cold*
(*cold* is temperature above minimum average), and *age*, the number of deaths (*event*)
and the *exposure* time (in days) are calculated. Note that the tabulated data frame is 
a tenth in size compared to the original file, and it is also *anonymised*.

So, it should not be any problem with public distribution of the tabulated data.

# The Cox regression model

On the original data frame, we can run an ordinary Cox regression, see Table 
\@ref(tab:coxreg).

```{r coxreg, results = 'asis'}
(system.time(fit.cr <- coxreg(Surv(enter, exit, event) ~ 
                                  sex + socStatus + illeg + parity + cold,
                 data = neo))) 
dr <- drop1(fit.cr, test = "Chisq")
ltx(fit.cr, dr = dr, label = "tab:coxreg", caption = "Cox regression.")
```

# The tabulated piecewise constant hazards (pch) model

On the tabulated data frame, we get, see Table \@ref(tab:tpchreg).

```{r tpchreg, results = 'asis'}
(system.time(fit.pch <- tpchreg(oe(event, exposure) ~ 
                                    sex + socStatus + illeg + parity + cold,
                                time = age, data = tneo)))
dr2 <- drop1(fit.pch, test = "Chisq")
ltx(fit.pch, dr = dr2, label = "tab:tpchreg", 
    caption = "Piecewise constant proportional hazards model.")
```

Almost identical results compared to Table \@ref(tab:coxreg). We can compare 
baseline cumulative hazards plots (Figure \@ref(fig:baseplots)):

```{r baseplots, fig.cap = "Comparison.", fig.height = 6}
op <- par(mfrow = c(2, 1), las = 1)
plot(fit.cr, main = "Cox regression")
plot(fit.pch, fn = "cum", main = "Pch regression")
par(op)
```

 Almost identical, disregarding differences in smoothness. An advantage with the 
 pch model is that we have access to an estimate of the baseline hazard function,
 see Figure \@ref(fig:haz).

```{r haz, fig.cap = "Baseline hazard function.",  fig.height = 4}
par(las = 1)
plot(fit.pch, fn = "haz")
abline(h = 0)
```

# "Logistic regression"

If, for some reason, it is desirable to use logistic regression as a proxy for Cox regression,
it is straightforward with the tabulated data.

```{r logistic}
fit.l <- glm(cbind(event, exposure - event) ~ 
                 sex + socStatus + illeg + parity + cold + age,
             family = binomial(link = "cloglog"), data = tneo)
coefficients(fit.l)[1:6]
drop1(fit.l, test = "Chisq")
```

Note the *cbind* and *link = "cloglog"*! Results are quite similar to previous ones, 
but estimation and plotting of baseline distributions are tedious.

However, starting with the non-tabulated data frame, everything is straightforward using
*eha::coxreg* with *method = "ml"* ("maximum likelihood"). Recommended for
the logistic regression approach with discrete time data.

### 2021-11-27 paper.Rmd
first commit

---
title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
classoption: titlepage
bibliography: bio.bib
titlepage: yes
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
abstract: "The effect of extreme temperatures on infant mortality in the Umeå region 1895-1950 is studied in a biometric analysis setting."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=4,comment=NA)
library(knitr)
library(eha)
g <- function(x) (log(1 + 365 * x))^3
ginv <- function(y) (1 / 365) * (exp(y^(1/3)) - 1)
```

# Introduction

The impact of ambient temperature variations on infant mortality is studied for 
a northern Sweden coastal area, the Umeå and Skellefteå regions, during the 
first half of the twentieth century. Two recent papers [@junkka; @karlsson] studied 
neonatal mortality and climate variations in the same geographical area during the 
years 1880--1950. Climate and mortality in general is a research area that has 
generated great interest over the last years, see @tbgb10.

The effect of seasonal variation and the occurrence of extreme monthly temperatures
is studied and interacted with sex, social class, and legitimacy. Studies are performed 
separatelyfor neonatal  and post-neonatal mortality, and for winter and summer seasons, 
and in a joint analysis, the classification into endogenous and exogenous factors will be discussed.

# The Bourgeouis-Pichat model and the approximation

A variation of the Bourgeouis-Pichat (B-P) biometric analysis [@bp51a;
@bp51b] of infant mortality is suggested. B-P suggests that mortality
during the last eleven months of infancy follows a *uniform* distribution
given a specific transformation of time (age), the 
*log-cube ("BP") transformation*. We show the BP transformation is closely approximated by
an *exponential distribution*, 
that is, the *cumulative hazards function* is linear.
 The advantage is that estimation of exogeneous and
endogeneous infant mortality is easily performed with standard survival
analysis programs.

Central in the *Bourgeouis-Pichat (B-P)* model is the *log-cube* time transform $g$:

\begin{equation*}\label{eq:g}
g(t) = \log^3(1 + 365 t), \quad 0 \le t \le 1,
\end{equation*}

where $t$ is age measured in *years*. 
Its graph is shown in Figure \@ref(fig:figgraph).

```{r figgraph, fig.cap = "The log-cube transform g."}
##g <- function(x) (log(1 + x))^3  # Note: days!
x <- seq(0, 1, length = 1000)
y <- g(x)
plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, col = "blue")
xat <- c(0, 28 / 365, 1 / 2, 1)
yat <- g(xat)
axis(1, at = xat, labels = round(365 * xat, 3))
axis(2, at = yat, labels = round(yat, 1), las = 2)
box()
lines(c(30, 30), c(0, g(30)), lty = 2, col = "darkgreen")
lines(c(0, 30), c(g(30), g(30)), lty = 2, col = "darkgreen")
abline(h = 0, v = 0)
```

The B-P model specifies that, on the transformed age scale, the *exogeneous*
mortality distribution, conditional on death before age one,  is *uniform*, that is, the 
survivor function of $T$ is *linear*.

\begin{equation}
S(t) = 1 - \alpha t, \, 0 < t < g(1).
\end{equation}

Furthermore, it is assumed that the *endogeneous* component is non-zero only during the first
month of life.

It seems to fit infant mortality data very well, but one drawback with it is that the 
uniform distribution is not easy to work with in a survival analysis context.

However, we show that the BP transformation is well approximated by a model where the 
*cumulative hazards function* is linear instead, that is, an *exponential distribution*,
easy to use in standard survival regression analysis, often allowing for massive data reduction
through the statistical *sufficency pinciple*.

# Data

## Infant Mortality

```{r readdata}
infdat <- readRDS("data/infdat.rds")
```

From the Umeå-Skellefteå region, we follow all persons born in the 
region 1895--1950 until death, reaching the age of 1, or leaving follow-up for whatever 
reason, whichever comes first.

Figure \@ref(fig:mortstat) shows the average monthly crude infant mortality, and a clear seasonal pattern is visible.

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950."}
tid <- toTpch(Surv(enter, exit, event) ~ week, data =  infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
     xlab = "Week", ylab = "IMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1)
abline(h = c(0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
box()
```

In accordance with the Bourgeouis-Picat ideas, we group the analyses into two parts, 
*neonatal* (first 28 days) och *postneonatal* mortality (days 29--365).

### Neonatal mortality

The average monthly neonatal mortality is shown in Figure \@ref(fig:neodata).

```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950."}
neo <- age.window(infdat, c(0, 28 / 365))
tneo <- toTpch(Surv(enter, exit, event) ~ week, data = neo, cuts = c(0, 28 / 365))
y <- tneo$event / tneo$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
     xlab = "Week", ylab = "IMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1)
abline(h = c(0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
box()
```

The seasonal pattern is similar to the one we found above for infant mortality.

## Postneonatal mortality

The average monthly postneonatal mortality is shown in Figure \@ref(fig:pneodata).

```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950."}
pneo <- age.window(infdat, c(28 / 365, 1))
tpneo <- toTpch(Surv(enter, exit, event) ~ week, data = pneo, cuts = c(28 / 365, 1))
y <- tpneo$event / tpneo$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
     xlab = "Week", ylab = "IMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1)
abline(h = c(0.04, 0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
box()
```

The seasonal pattern is once again similar to the one we found for infant mortality.

## Temperature

```{r readtemp}
umetemp <- readRDS("data/umetemp.rds")
bjurtemp <- readRDS("data/bjurtemp.rds")
```

Temperature data are collected from two weather stations, *Umeå* and *Bjuröklubb*
(used with population data from Skellefteå coastal area). Both stations deliver
daily temperature data covering our time period, usually three measures per day, morning,
noon, and evening. Weekly averages are calculated year by year, and deviations from 
the averages of the weekly averages are used as a time-varying *communal covariate*. See Figure 
\@ref(fig:monavgtemp) for the monthly distributions.

```{r monavgtemp, fig.cap="Weekly max and min temperature averages, 1895--1950."}
oldpar <- par(mfrow= c(1, 2))
## Umeå:
umaxavg <- umetemp$emaxtemp[1:52]
uminavg <- umetemp$emintemp[1:52]
plot(1:52, umaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)")
text(17, 27, "Umeå")
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = c(-20, -10, 10, 20), lty = 3)
lines(1:52, uminavg, type = "b", col = "blue")
text(8, 11, "max", col = "red")
text(20, -9, "min", col = "blue")
box()
## Skellefteå:
bmaxavg <- bjurtemp$emaxtemp[1:52]
bminavg <- bjurtemp$emintemp[1:52]
plot(1:52, bmaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)")
text(17, 27, "Skellefteå")
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = c(-20, -10, 10, 20), lty = 3)
lines(1:52, bminavg, type = "b", col = "blue")
text(8, 11, "max", col = "red")
text(20, -9, "min", col = "blue")
box()
##
par(oldpar)
```

### 2021-11-27 R/tbl.R

@@ -1,0 +1,8 @@
+tbl <- function(tt, caption = "", fs = 11, rownames = FALSE, align = NULL){
+    library(kableExtra)
+    kbl(tt, caption = caption, booktabs = TRUE, row.names = rownames, 
+        format = "latex", align = align) %>%
+        row_spec(row = 0, bold = TRUE) %>%
+        kable_styling(full_width = FALSE, font_size = fs,
+                    position = "center")
+}

### 2021-11-27 R/tempdat.R

@@ -1,0 +1,45 @@
+tempdat <- function(filename = "ume", period = c(1895, 1950)){
+
+    filename <- switch(filename, ume = "~/Forskning/Data/ume_temp.csv",
+                       bjur = "~/Forskning/Data/bjur_temp.csv",
+                       sten = "~/Forskning/Data/sten_temp.csv")
+    temp <- read.table(filename, 
+                       header = TRUE,
+                       skip = 10, sep = ";")[, 1:3]
+    ## We need only the first three columns
+    names(temp) <- c("date", "time", "temp")
+    temp$date <- as.Date(temp$date)
+    temp$year <- as.numeric(format(temp$date, "%Y"))
+    temp$month <- as.numeric(format(temp$date, "%m"))
+    temp <- temp[order(temp$date, temp$time), ]
+    temp$quarter <- quarters(temp$date)
+    temp <- temp[temp$year %in% period[1]:period[2], ]
+    indx <- tapply(temp$date, temp$date)
+    temp$mintemp <- tapply(temp$temp, temp$dat, min)[indx]
+    temp$maxtemp <- tapply(temp$temp, temp$dat, max)[indx]
+    temp <- temp[!duplicated(temp$date), c("date", "year", "month", "quarter", "mintemp", "maxtemp")]
+    ## Create "week", 1:52 each year, week 52 with 8(9) days:
+    get_week <- function(days){ # days = vector of 365 or 366, length of year
+        n <- length(days)
+        tmp <- c(rep(1:52, each = 7), 52, 52)
+        tmp[1:n]
+    }
+    temp$week <- as.numeric(unlist(tapply(temp$year, temp$year, get_week)))
+    ## Calculate weekly extremes of min amd max temp
+    wmin <- aggregate(mintemp ~ week + year, data = temp, FUN = min)
+    wmax <- aggregate(maxtemp ~ week + year, data = temp, FUN = max)
+    ##indx <- tapply(temp$week, temp$week)
+    ##wmintemp <- with(temp, tapply(mintemp, list(week, year), min))
+    ##wmaxtemp <- with(temp, tapply(maxtemp, list(week, year), max))[indx]
+######
+    temp <- wmin
+    temp$maxtemp <- wmax$maxtemp
+    ##
+    indx <- tapply(temp$week, temp$week)
+    temp$emintemp <- tapply(temp$mintemp, temp$week, mean)[indx]
[…]

### 2021-11-29 neo.Rmd
modifications and additions

@@
 summary(tneo)
-saveRDS(tneo, file = "data/tneo.rds")
+saveRDS(tneo, file = "t_data/tneo.rds")
 ```
@@
 Note how it works: For each combination of *socStatus*, *illeg*, *parity* *cold*
-(*cold* is temperature above minimum average), and *age*, the number of deaths (*event*)
+(*cold* is temperature above average minimum), and *age*, the number of deaths (*event*)
 and the *exposure* time (in days) are calculated. Note that the tabulated data frame is 
-a tenth in size compared to the original file, and it is also *anonymised*.
+a tenth in size compared to the original file, and it is also *anonymous*.
 
@@
 the logistic regression approach with discrete time data.
+
+# Effects of time transformation and tabular data.
+
+As an example of time transformation, we study the *log-cube* transformation used by 
+Bourgeouis-Pichat [@bp51a; @bp51b] in his study of endogeneous and exogeneous 
+factors explaining infant mortality. It is given by
+
+\begin{equation}
+g(t) = \bigl(log(1 + t)\bigr)^3, \quad 0 \le t \le 365,
+\end{equation}
+
+where $t$ is measured in days.
+
+So what happens if data are time-transformed before tabulation? Let us see:
+First the transformation of the data frame *neo*:
+
+```{r gtrans}
+g <- function(t) (log(1 + t))^3
+gneo <- neo
+gneo$enter <- g(neo$enter)
+gneo$exit <- g(neo$exit)
+head(gneo)
+```
+
+Next tabulation: We need to transform the chosen cut points, if we want comparability, so
+
+```{r tgneo}
+tgneo <- toTpch(Surv(enter, exit, event) ~ 
+                    sex + socStatus + illeg + parity + cold, 
+               data = gneo, cuts = g(0:28))
+head(tgneo)
+#summary(tgneo)
+```
+
+Note the clumsy levels of the factor *age*. Not important. Let us try a "piecewise constant ..."
+
+```{r tpchreg_g, results = 'asis'}
+(system.time(fit.pch_g <- tpchreg(oe(event, exposure) ~ 
+                                    sex + socStatus + illeg + parity + cold,
+                                time = age, data = tgneo)))
+dr2_g <- drop1(fit.pch_g, test = "Chisq")
+ltx(fit.pch_g, dr = dr2_g, label = "tab:tpchreg_g", 
+    caption = "Piecewise constant proportional hazards model with time transformation.")
+```
+
+Compare the Tables \@ref(tab:tpchreg) and \@ref(tab:tpchreg_g). Identical except for *TTR*
+(Total Time at Risk). And graphs will be stretcehd.
+
+```{r haz_g, fig.cap = "Baseline hazard function, transformed time.",  fig.height = 4}
+par(las = 1)
+plot(fit.pch_g, fn = "haz")
+abline(h = 0)
+```
+
+We can spot how the transformation drives the the hazard function to *constant* 
+over time, say after $ g(t) = 10$, that is, after around 8 days.

### 2021-11-29 paper.Rmd
modifications and additions

@@
 
-In accordance with the Bourgeouis-Picat ideas, we group the analyses into two parts, 
+In accordance with the Bourgeouis-Pichat ideas, we group the analyses into two parts, 
 *neonatal* (first 28 days) och *postneonatal* mortality (days 29--365).

### 2021-11-29 getdata2.Rmd
firtst commit

@@
                                                    "parity", "heat", "cold", "week")]
-saveRDS(infdat, file = "data/infdat.rds")
+saveRDS(infdat, file = "data/infdat2.rds")
 ```

### 2021-11-29 R/mkham.R

@@ -1,0 +1,24 @@
+fun <- function(beta, enter, exit, event){
+    shape <- numeric(2)
+    scale <- 1 / beta[3]
+    shape[1] <- exp(beta[1])
+    shape[2] <- exp(beta[2])
+    p1 <- sum(event * hmakeham(exit, shape = shape, scale = scale, log = TRUE))
+    p2 <- sum(Hmakeham(exit, shape = shape, scale = scale))
+    p3 <- sum(Hmakeham(enter, shape = shape, scale = scale))
+    p1 - p2 + p3
+}
+
+maxlog <- function(enter, exit, event){
+    beta <- numeric(3)
+    beta[3] <- 1
+    res <- optim(beta, fun, enter = enter, exit = exit, event = event,
+                 control = list(fnscale = -1, trace = 2))
+    res
+}
+
+testmk <- function(){
+    x <- readRDS("data/infdat.rds")
+    fit <- maxlog(infdat$enter, infdat$exit, infdat$event)
+    fit
+}

### 2021-11-29 R/gm.R
first commit

@@ -1,0 +1,67 @@
+## Gompertz-Makeham functions
+
+hgm <- function(t, a, log = FALSE){
+    ## Hazard function:
+    ## a is a vector of length 3:
+    ## hgm(t) = exp(a[1]) + exp(a[2] + a[3] * t)
+    ret <- exp(a[1]) + exp(a[2] + a[3] * t)
+    if (log) ret <- log(ret)
+    ret
+}
+
+hgm1 <- function(t, a){
+    ## Derivative wrt a[1] of hgm(log = FALSE)
+    exp(a[1])
+}
+
+hgm2 <- function(t, a){
+    ## Derivative wrt a[2].
+    exp(a[2] + a[3] * t)
+}
+
+hgm3 <- function(t, a){
+    ## Derivative wrt a[3].
+    t * exp(a[2] + a[3] * t)
+}
+    
+Hgm <- function(t, a){
+    ## Cumulative hazards function,
+    ## has length 3.
+    t * exp(a[1]) - exp(a[2]) * expm1(a[3] * t) / a[3]
+}
+
+Hgm1 <- function(t, a){
+    t * exp(a[1])
+}
+
+Hgm2 <- function(t, a){
+    -exp(a[2]) * expm1(a[3] * t) / a[3]
+}
+
+Hgm3 <- function(t, a){
+    p1 <- expm1(a[3] * t) / a[3]
+    p2 <- t * exp(a[3] * t)
+    exp(a[2]) * (p1 - p2) / a[3]
+}
+
+loggm <- function(a, enter, exit, event){
+    ## log likelihood function
+    ## a is vector of length 3 (parameter)
+    ## enter, exit, event are equal-length vectors (data).
+    p1 <- sum(event * hgm(exit, a, log = TRUE))
+    p2 <- sum(Hgm(enter, a) - Hgm(exit, a))
+    p1 + p2
+}
+
+dloggm <- function(a, enter, exit, event){
+    ## Score function(s)
+
+    sco <- numeric(3)
+    sco[1] <- sum(event * hgm1(exit, a) / hgm(exit, a) +
+                  Hgm1(enter, a) - Hgm1(exit, a))
+    sco[2] <- sum(event * hgm2(exit, a) / hgm(exit, a) +
+                  Hgm2(enter, a) - Hgm2(exit, a))
+    sco[3] <- sum(event * hgm3(exit, a) / hgm(exit, a) +
+                  Hgm3(enter, a) - Hgm3(exit, a))
+    sco
+}

### 2021-12-01 getdata.Rmd
correct data

@@
 births$event <- as.numeric(births$sluttyp == 2)
+## NEW (and necessary):
+extend <- with(births, (exit <= enter) & (event == 1))
+births$exit[extend] <- births$enter[extend] + 1 / (4 * 365)
+## NOW, remove all zero-length intervals:
+births <- births[births$exit > births$enter, ]
+## End NEW, now OK to select:
 births <- age.window(births, c(0, 1))
@@
 lagg <- 3 / 365  ## NOTE!!
+##lagg <- 0
 ## Ume:

### 2021-12-01 neo.Rmd
correct data

@@
 
-```{r tpchreg_g, results = 'asis'}
-(system.time(fit.pch_g <- tpchreg(oe(event, exposure) ~ 
+```{r tpchreg.g, results = 'asis'}
+(system.time(fit.pch.g <- tpchreg(oe(event, exposure) ~ 
                                     sex + socStatus + illeg + parity + cold,
                                 time = age, data = tgneo)))
-dr2_g <- drop1(fit.pch_g, test = "Chisq")
-ltx(fit.pch_g, dr = dr2_g, label = "tab:tpchreg_g", 
+dr2.g <- drop1(fit.pch.g, test = "Chisq")
+ltx(fit.pch.g, dr = dr2.g, label = "tab:tpchreg.g", 
     caption = "Piecewise constant proportional hazards model with time transformation.")
@@
 
-Compare the Tables \@ref(tab:tpchreg) and \@ref(tab:tpchreg_g). Identical except for *TTR*
-(Total Time at Risk). And graphs will be stretcehd.
+Compare the Tables \@ref(tab:tpchreg) and \@ref(tab:tpchreg.g). Identical except for *TTR*
+(Total Time at Risk). And graphs will be stretced.
 
-```{r haz_g, fig.cap = "Baseline hazard function, transformed time.",  fig.height = 4}
+```{r haz.g, fig.cap = "Baseline hazard function, transformed time.",  fig.height = 4}
 par(las = 1)
-plot(fit.pch_g, fn = "haz")
+plot(fit.pch.g, fn = "haz")
 abline(h = 0)

### 2021-12-01 R/mkham.R
fix dataframe name error

@@ -19,6 +19,6 @@ maxlog <- function(enter, exit, event){
 
 testmk <- function(){
     x <- readRDS("data/infdat.rds")
-    fit <- maxlog(infdat$enter, infdat$exit, infdat$event)
+    fit <- maxlog(infdat$enter, x$exit, x$event)
     fit
 }

### 2021-12-01 R/check_obs.R
first commit

@@ -1,0 +1,14 @@
+check_obs <- function(){
+    library(skum)
+    ## Check of age at death, in days.
+    qid <- obs$id[obs$starttyp == 2]
+    oj <- obs[obs$id %in% qid, c("id", "startdat", "slutdat", "foddat", "sluttyp")]
+    oj <- oj[oj$foddat > as.Date("1894-12-31") & oj$foddat < as.Date("1951-01-01"), ]
+    oj$event <- as.integer(oj$sluttyp == 2)
+    oj$sluttyp <- NULL
+    oj$enter <- as.double(difftime(oj$startdat, oj$foddat, units = "days"))
+    oj$exit <- as.double(difftime(oj$slutdat, oj$foddat, units = "days"))
+    oj <- age.window(oj, c(-1, 28))
+    
+    oj
+}

### 2021-12-01 neo.Rmd
look for errors (found but not fixed)

@@
 length(unique(infdat$id))
-head(infdat)
+head(infdat) ## OK!
 ```
@@
 neo <- age.window(infdat, c(0, 28 / 365))
+## events = 2693
 ```
@@
 ```{r transform}
-neo$enter <- round(365 * neo$enter)
-neo$exit <- round(365 * neo$exit)
-neo <- neo[neo$exit > neo$enter, ] # Remove zero-length spells
-
+##neo$enter <- round(365 * neo$enter)
+##neo$exit <- round(365 * neo$exit)
+##neo <- neo[neo$exit > neo$enter, ] # Remove zero-length spells
+neo$enter <- 365 * neo$enter
+neo$exit <- 365 * neo$exit
 dim(neo)
-head(neo)
+head(neo) ## OK
 ```
@@
 neo$parity <- cut(neo$parity, c(0, 1.5, 3.5, 18.5), labels = c("1", "2-3", "4+"))
-neo <- neo[!is.na(neo$socStatus), ] # Remove records with missing values
+##neo <- neo[!is.na(neo$socStatus), ] # Remove records with missing values
+## Instead, replace missing socStatus with "low"
+neo$socStatus[is.na(neo$socStatus)] <- "low"
 ```

### 2021-12-03 getdata.Rmd
add Introduction and fix bug
modifications ...

@@
 ```
+
+# Introduction
+
+Note thet we now have corrected the mistake with zero-length intervals.
 
@@
 source("R/tempdat.R")
-umetemp <- tempdat("ume", c(1895, 1950))
-bjurtemp <- tempdat("bjur", c(1895, 1950))
+temp_start <- 1894
+temp_end <- 1951
+t_ivl <- c(temp_start, temp_end)
+umetemp <- tempdat("ume", t_ivl) # Changed interval!
+bjurtemp <- tempdat("bjur", t_ivl) # Ditto!
 ##stentemp <- tempdat("~/Forskning/Data/sten_temp.csv", c(1899, 1952))
@@
 bume <- births[births$subreg == "ume", ]
-bume <- make.communal(bume, umetemp["heat"], start = 1895, period = 1/52, lag = lagg)
+bume <- make.communal(bume, umetemp["heat"], start = temp_start, period = 1/52, lag = lagg)
 bume <- bume[bume$enter < bume$exit, ]
-bume <- make.communal(bume, umetemp["cold"], start = 1895, period = 1 / 52, lag = lagg)
+bume <- make.communal(bume, umetemp["cold"], start = temp_start, period = 1 / 52, lag = lagg)
 bume <- bume[bume$enter < bume$exit, ]
-bume <- make.communal(bume, umetemp["week"], start = 1895, period = 1 / 52, lag = lagg)
+bume <- make.communal(bume, umetemp["week"], start = temp_start, period = 1 / 52, lag = lagg)
 bume <- bume[bume$enter < bume$exit, ]
@@
 bskel <- births[births$subreg == "bjur", ]
-bskel <- make.communal(bskel, bjurtemp["heat"], start = 1895, period = 1/52, lag = lagg)
+bskel <- make.communal(bskel, bjurtemp["heat"], start = temp_start, period = 1/52, lag = lagg)
 bskel <- bskel[bskel$enter < bskel$exit, ]
-bskel <- make.communal(bskel, bjurtemp["cold"], start = 1895, period = 1 / 52, lag = lagg)
+bskel <- make.communal(bskel, bjurtemp["cold"], start = temp_start, period = 1 / 52, lag = lagg)
 bskel <- bskel[bskel$enter < bskel$exit, ]
-bskel <- make.communal(bskel, bjurtemp["week"], start = 1895, period = 1 / 52, lag = lagg)
+bskel <- make.communal(bskel, bjurtemp["week"], start = temp_start, period = 1 / 52, lag = lagg)
 bskel <- bskel[bskel$enter < bskel$exit, ]

### 2021-12-03 R/check_obs.R

@@ -2,13 +2,22 @@ check_obs <- function(){
     library(skum)
     ## Check of age at death, in days.
     qid <- obs$id[obs$starttyp == 2]
-    oj <- obs[obs$id %in% qid, c("id", "startdat", "slutdat", "foddat", "sluttyp")]
-    oj <- oj[oj$foddat > as.Date("1894-12-31") & oj$foddat < as.Date("1951-01-01"), ]
+    oj <- obs[obs$id %in% qid, 
+              c("id", "sex", "startdat", "slutdat", "foddat", "sluttyp")]
+    oj <- oj[oj$foddat > as.Date("1894-12-31") & 
+                 oj$foddat < as.Date("1951-01-01"), ]
     oj$event <- as.integer(oj$sluttyp == 2)
     oj$sluttyp <- NULL
-    oj$enter <- as.double(difftime(oj$startdat, oj$foddat, units = "days"))
-    oj$exit <- as.double(difftime(oj$slutdat, oj$foddat, units = "days"))
-    oj <- age.window(oj, c(-1, 28))
-    
-    oj
+    oj$denter <- as.double(difftime(oj$startdat, oj$foddat, units = "days"))
+    oj$dexit <- as.double(difftime(oj$slutdat, oj$foddat, units = "days")) + 1
+    oj <- age.window(oj, c(-1, 28), surv = c("denter", "dexit", "event"))
+    ume <- readRDS("data/umetemp.rds")
+    oj$exit <- oj$dexit / 365
+    oj$enter <- oj$denter / 365
+    oj$birthdate <- toTime(oj$foddat)
+    ##oj1 <- make.communal(oj, umetemp["week"], start = 1894.9999, period = 1 / 52)
+    source("R/mkomm.R")
+    oj1 <- mkomm(oj, ume2[, c("year", "week", "cold")], start = 1894, period = 1 / 52)
+    ##list(oj = oj, oj1 = oj1)
+    oj1
 }

### 2021-12-03 R/tempdat.R

@@ -1,4 +1,4 @@
-tempdat <- function(filename = "ume", period = c(1895, 1950)){
+tempdat <- function(filename = "ume", period = c(1894, 1951)){
 
     filename <- switch(filename, ume = "~/Forskning/Data/ume_temp.csv",
                        bjur = "~/Forskning/Data/bjur_temp.csv",

### 2021-12-03 R/mkomm.R
new or modif? So what!

@@ -1,0 +1,152 @@
+mkomm <- function(dat, com.dat, communal = TRUE, start, period = 1, lag = 0, 
+             surv = c("enter", "exit", "event", "birthdate"), tol = 1e-04, 
+             fortran = TRUE) 
+{
+    if (!is.data.frame(dat)) 
+        stop("dat must be a data frame")
+    if (!is.data.frame(com.dat)) 
+        stop("com.dat must be a data frame")
+    if (length(surv) != 4) 
+        stop("surv must have length 4")
+    fixed.names <- names(dat)
+    surv.indices <- match(surv, fixed.names)
+    if (length(which(is.na(surv.indices)))) {
+        x <- which(is.na(surv.indices))
+        stop(paste(surv[x], " is not a name in the fixed data frame."))
+    }
+    com.names <- names(com.dat)
+    if (length(x <- which(!is.na(match(com.names, c(surv, fixed.names)))))) 
+        stop(paste(com.names[x], "are names in fixed data frame."))
+    nn <- nrow(dat)
+    n.years <- nrow(com.dat)
+    n.com <- NCOL(com.dat)
+    ##if (n.com > 1) stop("Only one communal covariate at a time!")
+    cuts <- start + c(0, (1:n.years) * period) - lag
+    ##cat("cuts = ", cuts, "\n")
+    beg.per <- cuts[1]
+    end.per <- cuts[n.years + 1]
+    cat("beg.per = ", beg.per, "end.per = ", end.per, "\n")
+    iv.length <- period
+    if (communal){ ## Added 26 March 2004 ##
+        spell.tot <- sum(dat[[surv.indices[2]]] - dat[[surv.indices[1]]])
+        dat <- cal.window(dat, c(beg.per, end.per), surv)
+        
+        if (sum(dat[[surv.indices[2]]] - dat[[surv.indices[1]]]) < 
+            spell.tot) 
+            warning("Spells are cut")
+    }#######################################
+    nn <- nrow(dat)
+    if (!communal) {
+        get.per <- function(dates){
+            pmin(pmax(1, ceiling((dates - beg.per)/iv.length)), n.years)
+        }
+        dates <- dat[[surv.indices[4]]]
+        
+        ppp <- get.per(dates)
+        
+        yy <- matrix(0, ncol = n.com, nrow = nn)
+        for (i in 1:n.com) {
+        ##yy[, i] <- com.dat[ppp, i]
+	    yy[, i] <- com.dat[[i]][ppp] # Changed 25 Sep 2017
+        }
[…]

### 2021-12-07 getdata2.Rmd
modifications positively ...

--- 2021-11-27 getdata2.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = TRUE)
 library(knitr)
@@
       "ort", "ortnmn")]
-births$event <- as.numeric(births$sluttyp == 2)
+births$event <- (births$sluttyp == 2)
 ## New start in days...:
@@
 births$exit <- as.numeric(with(births, difftime(slutdat, foddat, unit = "days")))
+##
+expand <- (births$startdat == births$slutdat) & births$event
+births$exit[expand] <- births$enter[expand] + 0.25 # 6 hours!
+##
+weq <- (births$startdat == births$slutdat) & (!births$event)
+births <- births[!weq, ]
+##
 births <- age.window(births, c(0, 365))
@@
 ### Deaths:
-did <- unique(births$id[births$event == 1])
+did <- unique(births$id[births$event])
 deaths <- births[births$id %in% did, ]
@@
 births$parity <- pers$paritet_g[indx]
+## Add:
+births$birthdate <- toTime(births$foddat)
+##
+births$exit <- births$exit / 365 ## Measure time in years!!
 summary(births)                                
@@
 neo <- births
-neo$exit <- neo$exit / 365
-neo$birthdate <- toTime(neo$foddat)
+##
+##neo$birthdate <- toTime(neo$foddat)
 neo <- age.window(neo, c(0, 28 / 365))
@@
 
-```{r pnneo, fig.cap = "Post-eonatal mortality by year, all regions."}
-pneo <- age.window(births, c(1 / 13, 1))
+```{r pneo, fig.cap = "Post-neonatal mortality by year, all regions."}
+pneo <- births
+##
+##pneo$birthdate <- toTime(pneo$foddat)
+pneo <- age.window(pneo, c(28 / 365, 1))
 surv <- as.matrix(pneo[, c("enter", "exit", "event", "birthdate")])
@@
 source("R/tempdat.R")
-umetemp <- tempdat("ume", c(1895, 1950))
-bjurtemp <- tempdat("bjur", c(1895, 1950))
+temp_start <- 1894
+temp_end <- 1951
+temp_ivl <- c(temp_start, temp_end)
+##
+umetemp <- tempdat("ume", temp_ivl)
+bjurtemp <- tempdat("bjur", temp_ivl)
 ##stentemp <- tempdat("~/Forskning/Data/sten_temp.csv", c(1899, 1952))
@@
 lagg <- 3 / 365  ## NOTE!!
+tempvars <-  c("heat", "aver", "cold", "emintemp", "emeantemp", "emaxtemp", 
+               "week", "year") 
 ## Ume:
 bume <- births[births$subreg == "ume", ]
-bume <- make.communal(bume, umetemp["heat"], start = 1895, period = 1/52, lag = lagg)
-bume <- bume[bume$enter < bume$exit, ]
-bume <- make.communal(bume, umetemp["cold"], start = 1895, period = 1 / 52, lag = lagg)
-bume <- bume[bume$enter < bume$exit, ]
-bume <- make.communal(bume, umetemp["week"], start = 1895, period = 1 / 52, lag = lagg)
-bume <- bume[bume$enter < bume$exit, ]
+bume <- make.communal(bume, umetemp[, tempvars], 
+                      start = temp_start, period = 1/52, lag = lagg)
+
 ### Ske:
 bskel <- births[births$subreg == "bjur", ]
-bskel <- make.communal(bskel, bjurtemp["heat"], start = 1895, period = 1/52, lag = lagg)
-bskel <- bskel[bskel$enter < bskel$exit, ]
-bskel <- make.communal(bskel, bjurtemp["cold"], start = 1895, period = 1 / 52, lag = lagg)
-bskel <- bskel[bskel$enter < bskel$exit, ]
-bskel <- make.communal(bskel, bjurtemp["week"], start = 1895, period = 1 / 52, lag = lagg)
-bskel <- bskel[bskel$enter < bskel$exit, ]
-##bsten <- births[births$subreg == "sten", ]
-##bsten <- make_comm(bsten, stentemp, start = 1899, period = 1/52, lag = lagg)
+bskel <- make.communal(bskel, bjurtemp[, tempvars], 
+                       start = temp_start, period = 1/52, lag = lagg)
+
 bume$region <- "ume"
@@
                                                    "birthdate", "socStatus", "hisclass", "illeg",
-                                                   "parity", "heat", "cold", "week")]
-saveRDS(infdat, file = "data/infdat.rds")
+                                                   "parity", tempvars)]
+saveRDS(infdat, file = "data/infdat2_year.rds")
 ```

### 2021-12-07 R/tempdat.R

@@ -17,7 +17,9 @@ tempdat <- function(filename = "ume", period = c(1894, 1951)){
     indx <- tapply(temp$date, temp$date)
     temp$mintemp <- tapply(temp$temp, temp$dat, min)[indx]
     temp$maxtemp <- tapply(temp$temp, temp$dat, max)[indx]
-    temp <- temp[!duplicated(temp$date), c("date", "year", "month", "quarter", "mintemp", "maxtemp")]
+    temp$meantemp <- tapply(temp$temp, temp$dat, mean)[indx]
+    temp <- temp[!duplicated(temp$date), 
+                 c("date", "year", "month", "quarter", "mintemp", "maxtemp", "meantemp")]
     ## Create "week", 1:52 each year, week 52 with 8(9) days:
     get_week <- function(days){ # days = vector of 365 or 366, length of year
         n <- length(days)
@@ -25,21 +27,25 @@ tempdat <- function(filename = "ume", period = c(1894, 1951)){
         tmp[1:n]
     }
     temp$week <- as.numeric(unlist(tapply(temp$year, temp$year, get_week)))
-    ## Calculate weekly extremes of min amd max temp
+    ## Calculate weekly extremes of min and max temp
     wmin <- aggregate(mintemp ~ week + year, data = temp, FUN = min)
     wmax <- aggregate(maxtemp ~ week + year, data = temp, FUN = max)
+    wmean <- aggregate(meantemp ~ week + year, data = temp, FUN = mean)
     ##indx <- tapply(temp$week, temp$week)
     ##wmintemp <- with(temp, tapply(mintemp, list(week, year), min))
     ##wmaxtemp <- with(temp, tapply(maxtemp, list(week, year), max))[indx]
 ######
     temp <- wmin
     temp$maxtemp <- wmax$maxtemp
+    temp$meantemp <- round(wmean$meantemp, 1)
     ##
     indx <- tapply(temp$week, temp$week)
     temp$emintemp <- tapply(temp$mintemp, temp$week, mean)[indx]
     temp$emaxtemp <- tapply(temp$maxtemp, temp$week, mean)[indx]
+    temp$emeantemp <- tapply(temp$meantemp, temp$week, mean)[indx]
     temp$heat <- round(temp$maxtemp - temp$emaxtemp)
     temp$cold <- round(temp$mintemp - temp$emintemp)
+    temp$aver <- round(temp$meantemp - temp$emeantemp)
     ##
     temp
 }

### 2021-12-07 Test/bysex.Rmd
first commit

---
title: "Female advantage in infant survival, Umeå and Skellefteå 1901-1950"
author: "Göran Broström"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
bibliography: ../bio.bib
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
abstract: "Is the female advantage in infant survival constant over time? Investigation of Umeå and Skellefteå 1901-1950. The answer is 'yes'."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
dt <- function(x, y) as.numeric(difftime(x, y, unit = "days")) # wrapper of 'difftime'
```

# Introduction

The following presentation relates to infants *born* in the parish(es) we investigate.
They are the parishes with codes 82750, 82780 (`ume`) and 82980, 82981, 82983, 82988 (`ske`).

The **R** code generating the output is given throughout, it gets a little bit messy, maybe.
I am using the **R** [@cran] package *eha* [@ehar2; @eha] for the analyses.

It is all about creating tables of number of *deaths* and total *exposure* time by *sex* and *year*.
And running *Poisson regression* on the tables.

# Get data

Data are extracted from *DDB-uttaget U15014* [@u15014], which is transformed into an **R** package *skum*, available on request.

```{r readdata}
library(skum)
births.id <- obs$id[obs$starttyp == 2] # ID of Born in parish
births <- obs[obs$id %in% births.id, # Records of born in parish
              c("id", "sex", "foddat", "startdat", "slutdat", 
                "sluttyp", "nofrs")]
births$enter <- with(births, dt(startdat, foddat)) # Tid i dagar
births$exit <- with(births, dt(slutdat, foddat))
births$event <- births$sluttyp == 2 # event == death
## Fix deaths at foddat: (Problem: age at death? 6 hours?)
fix <- (births$startdat >= births$slutdat) & births$event
births$exit[fix] <- births$enter[fix] + 6 / 24 # Add 6 hours
births <- births[births$enter < births$exit, ] # Remove zero-length intervals
infants <- age.window(births, c(0, 365)) # Cut at age 365 days
infants$birthdate <- toTime(infants$foddat)
## Remove 'intermediates':
infants <- infants[order(infants$id, -infants$exit), ]
infants <- infants[!duplicated(infants$id), ]
infants$enter <- 0 # All born in parish!
infants$sluttyp <- infants$startdat <- infants$slutdat <- NULL # Not needed any more.
infants$exit <- infants$exit / 365 # Measure age in years.
infants <- cal.window(infants, c(1901, 1951)) # Select the years 1901-1950.
infants <- infants[order(infants$birthdate), ] # Sort by birthdate
## Define coastal areas 'ume' and 'ske':
infants$region <- factor(infants$nofrs)
levels(infants$region) <- c(rep("ume", 2), rep("ske", 4), 
                            rep("inland", 3))
infants$nofrs <- NULL # Not needed any more
## Remove region 'inland':
infants <- infants[infants$region != "inland", ]
infants$region <- factor(infants$region) # Removes unused level 'inland'
source("../R/tbl.R")
x <- head(infants, 5) # First 5 lines of 'infants', store in 'x'
tbl(x, caption = "First lines of the data frame 'infants', Umeå and Skellefteå.",
    fs = 11) # fs = 'font size'
```

```{r lastlines}
y <- tail(infants, 5) # Last 5 lines of 'infants'.
tbl(y, caption = "Last lines of the data frame 'infants'.", fs = 11)
```

# Table by year and sex

See Table \@ref(tab:tabs) for the result of the following code.

```{r tabs}
## Period statistics (period = year)
girls <- with(infants[infants$sex == "female", ], 
              perstat(cbind(enter, exit, event, birthdate), 
                      period = 1901:1951, age = c(0, 1)))
boys <- with(infants[infants$sex == "male", ], 
              perstat(cbind(enter, exit, event, birthdate), 
                      period = 1901:1951, age = c(0, 1)))
## Rearrange:
boy <- data.frame(year = 1901:1950, sex = rep("boys", 50),
                  deaths = as.numeric(boys$events), 
                  exposure = as.numeric(boys$exposure))

girl <- data.frame(year = 1901:1950, sex = rep("girls", 50),
                   deaths = as.numeric(girls$events), 
                   exposure = as.numeric(girls$exposure))
## Put together
infant <- rbind(boy, girl)
infant <- infant[order(infant$year, infant$sex), ]
infant$sex <- as.factor(infant$sex)
infant$year <- as.factor(infant$year)
z <- head(infant, 5)
tbl(z, caption = "Number of deaths and total exposure time (years) by sex and year. Early years.")
```

```{r lateyrs}
w <- tail(infant, 5)
tbl(w, caption = "Number of deaths and total exposure time (years) by sex and year. Late years.")
```

# Simple analysis

Is the female survival advantage constant over 50 years 1901--1950? A simple *Poisson regression*
gives the answer:

```{r poisreg}
infant$age <- "0-1"
fit <- tpchreg(oe(deaths, exposure) ~ sex * year, time = age, data = infant)
drop1(fit, test = "Chisq") # Likelihood Ratio Test
```

Obviously, there is no evidence of that the female advantage should vary with time.
The overall femal advantage is found by the multiplicative model, see Table \@ref(tab:mult).

```{r mult, results = 'asis'}
fit <- tpchreg(oe(deaths, exposure) ~ sex + as.numeric(year),
               time = age, data = infant)
x <- summary(fit)
hr <-  fit$coefficients[1]
eupper <- round(exp(hr + 1.96 * sqrt(fit$var[1,1])), 3)
elower <- round(exp(hr - 1.96 * sqrt(fit$var[1,1])), 3)
ehr <- round(exp(hr), 3)
ltx(fit, caption = "Infant mortality by year and sex.", label = "tab:mult")
```

So, the female infant mortality is estimated to `r 100 * ehr` per cent of the male, 
with 95% confidence interval (`r format(100 * elower, nsmall = 1)`, `r 100 * eupper`) per cent.

# Graph by year

```{r byyear, fig.cap = "Female mortality in relation to male (blue squares) by year with 95 per cent pointwise confidence intervals."}
par(las = 1)
hr <- numeric(50)
up <- numeric(50)
low <- numeric(50)
i <- 0
for (year in 1901:1950){
    i <- i + 1
    me <- infant[infant$year == year, ]
    ff <- tpchreg(oe(deaths, exposure) ~ sex, time = age,
                  data = me)
    beta <- ff$coefficients[1]
    sd <- sqrt(ff$var[1, 1])
    hr[i] <- exp(beta)
    up[i] <- exp(beta + 1.96 * sd)
    low[i] <- exp(beta - 1.96 * sd)
}
plot(1901:1950, hr, type = "p", ylim = c(0.3, 2.3), 
     xlab = "Year", ylab = "Hazard ratio (per cent)", 
     col = "blue", axes = FALSE, cex = 5, pch = ".")
axis(1, at = c(1901, 1913, 1925, 1938, 1950))
axis(2, at = c(0.5, ehr, 1, 1.5, 2), 
     labels = c(0.5, ehr, 1, 1.5, 2) * 100)
box()
for (i in 1:50){
    lines(c(1900 + i, 1900 + i), c(low[i], up[i]), 
          lty = 3, col = "darkgreen")
    points(c(1900 + i, 1900 + i), c(low[i], up[i]), lty = 3, 
           col = "darkgreen", pch = 5, cex = 0.3)
}
abline(h = 1)
abline(h = ehr, lty = 2, col = "blue")
abline(h = c(eupper, elower), lty = 3, col = "blue")
```

The numbers of deaths decrease rapidly with time, see Figure \@ref(fig:derap),
which explains the increasing lenghts of the confidence intervals with time.

```{r derap, fig.cap = "Infant mortality by year, 1901-1950."}
deaths <- with(infant, tapply(deaths, year, sum))
exposures <- with(infant, tapply(exposure, year, sum))
intensity <- deaths / exposures
aver <- sum(deaths) / sum(exposures)
plot(1901:1950, intensity, type = "b", axes = FALSE, 
     ylim = c(0, 0.13),  xlab = "Year", 
     ylab = "Mortality, per cent", col = "blue")
axis(1, at = c(1901, 1913, 1925, 1938, 1950))
axis(2, at = seq(0.00, 0.12, by = 0.02), 
     labels = seq(0, 12, by = 2), las = 1)
box()
abline(h = 0)
abline(h = aver, col = "blue", lty = 2)
text(1942, 0.08, "average", col = "blue")
```

# Conclusion

Nothing much happens, the variation over the years we can see in the relative 
difference in mortality between boys and girls is obviously purely random, no
trend is visible. That is, the female advantage is strong and persistent.

One obvious observation, on the other hand, is that infant mortality
decreases markedly over the first half of the 20th century.

### 2021-12-07 Test/testcomm.R

@@ -1,0 +1,33 @@
+testcomm <- function(){
+    ##lagg <- 3 / 365  ## NOTE!!
+    lagg <- 0
+    temp_start <- 1894
+    ## Ume:
+    umetemp <- readRDS("../data/umetemp.rds")
+    ##bu10 <- readRDS("bu10.rds")
+    bumeo <- readRDS("../data/births.rds")
+    bume <- make.communal(bumeo, umetemp["week"], 
+                          start = temp_start, period = 1 / 52, lag = lagg)
+    ##bume <- bume[bume$enter < bume$exit, ]
+    bume <- make.communal(bume, umetemp["year"], 
+                          start = temp_start, period = 1 / 52, lag = lagg)
+    ##bume <- bume[bume$enter < bume$exit, ]
+    ##
+    bume <- make.communal(bume, umetemp["heat"], 
+                          start = temp_start, period = 1/52, lag = lagg)
+    ##bume <- bume[bume$enter < bume$exit, ]
+    bume <- make.communal(bume, umetemp["cold"], 
+                          start = temp_start, period = 1 / 52, lag = lagg)
+    ##bume <- bume[bume$enter < bume$exit, ]
+    ## mkomm:
+    cat("mkomm:\n")
+    source("../R/mkomm.R")
+    b2 <- mkomm(bumeo, umetemp[, c("week", "year", "heat", "cold")], 
+                start = temp_start, period = 1 / 52, lag = lagg)
+    ## Handräkning:
+    ##endt <- seq(0, 1, by = 53)
+    
+    ##
+    cat("Diff = ", with(bume, sum(exit - enter)) - with(b2, sum(exit - enter)), "\n")
+    list(bume = bume, b2 = b2)
+}

### 2021-12-09 getdata2.Rmd
changes ...

@@
 The only (so far) difference compared to the article "getdata.Rmd" is that time 
-here is measured in *days* instead of in a fraction of a year.
+here is measured in *days* instead of in a fraction of a year. *However*, "at the end", time is transformed back to units in years. The main reason is that the function `make.communal`
+assumes time measured in years, it is probably the variable `birthdate` that is the culprit.
+I will look into that in due course.
 
@@
 neo <- births
+par(las = 1)
 ##
@@
 pall <- perstat(surv, period = 1900:1951)$intensity
-plot(1900:1950, pall, type = "b", ylim = c(0, 0.5), col = "blue", xlab = "Year",
+plot(1900:1950, pall, type = "b", ylim = c(0, 0.7), col = "blue", xlab = "Year",
      ylab = "Mortality")
 abline(h = 0)
-abline(h = 0.3, lty = 3)
+abline(h = c(0.1, 0.3, 0.5) , lty = 3)
 ```

### 2021-12-09 neo2.Rmd
first commit

---
title: "Neonatal models"
author: "Göran Broström"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
bibliography: bio.bib
titlepage: no
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage{a4wide}    
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
abstract: "The effect of various covariates  on neonatal mortality in the Umeå region 1895--1950 is studied as an example of how to do it with tabular, anonymised data."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,comment = NA)
```

# Introduction

How to analyse table data with the same statistical models, for instance *Cox regression*,
as you are used to with individual data, is shown here. It helps if you have the 
**R** package *eha* [@eha; @ehar2] for easy conversion of individual data to tabular
suitable for survival analysis.

# Example: Neonatal mortality in Umeå/Skellefteå, 1895--1950

We use a data set for studying infant mortality from the *Demographic Database*,
Umeå University.

## Read data

Start by reading the raw data file and select a few columns: See Table \@ref(tab:clean).

```{r readdata}
infdat <- readRDS("data/infdat.rds")[, c("id", "enter", "exit", "event", 
                                         "sex", "socStatus", "illeg",  
                                         "parity")]
## Do not remove zero-length spells:
## <- infdat[infdat$exit > infdat$enter + 1.e-6, ] 
rownames(infdat) <- 1:NROW(infdat)
nrrawd <- NROW(infdat)
##length(unique(infdat$id))
x <- head(infdat) ## OK!
```

```{r agecut}
library(eha)
neo <- age.window(infdat, c(0, 28 / 365))
## events = 2693
```

```{r transform}
source("R/tbl.R")
##neo$enter <- round(365 * neo$enter)
##neo$exit <- round(365 * neo$exit)
##neo <- neo[neo$exit > neo$enter, ] # Remove zero-length spells
neo$enter <- round(365 * neo$enter)
neo$exit <- round(365 * neo$exit)
zero <- neo$exit <= 0 & neo$event == 1
neo$exit[zero] <- 6 / 24
neo <- neo[neo$exit > neo$enter, ]
dneo <- dim(neo)

```

```{r clean}
neo <- neo[neo$parity > 0, ] # Remove missing = -1
neo$parity <- cut(neo$parity, c(0, 1.5, 3.5, 18.5), labels = c("1", "2-3", "4+"))
neo$socStatus[is.na(neo$socStatus)] <- "low"
x <- head(neo) ## OK
tbl(x, caption = "Raw individual data.")
```

Next, *tabulate* data with the aid of the function *toTpch*, found in *eha*, see Table \@ref(tab:tabulate).

```{r tabulate}
tneo <- toTpch(Surv(enter, exit, event) ~  
                   sex + socStatus + illeg + parity, 
               data = neo, cuts = 0:28)
DIMT <- dim(tneo)
X <- head(tneo)
##summary(tneo)
tbl(X, caption = "Same data, tabular form.")
saveRDS(tneo, file = "t_data/tneo.rds")
```

Note how it works: For each combination of *socStatus*, *illeg*, *parity*,
 and *age*, the number of deaths (*event*)
and the *exposure* time (in days) are calculated. The tabulated data frame is 
0.17 per cent in size compared to the original file, and it is also *anonymous*.

The age interval is cut into 28 pieces, "0-1", "1-2", etc., so the underlying statistical survival model is that of a *piecevise constant hazards*.

So, it should not be any problem with public distribution of the tabulated data, but how about the analyses of the two data sets, will there be any differences in results? Not really, see what follows.

# The Cox regression model

On the original data frame, we can run an ordinary Cox regression, see Table 
\@ref(tab:coxreg) for results.

```{r coxreg, results = 'asis'}
fit.cr <- coxreg(Surv(enter, exit, event) ~ 
                                  sex + socStatus + illeg + parity,
                 data = neo) 
dr <- drop1(fit.cr, test = "Chisq")
ltx(fit.cr, dr = dr, label = "tab:coxreg", caption = "Cox regression.")
```

# The tabulated piecewise constant hazards (pch) model

On the tabulated data frame, we get, see Table \@ref(tab:tpchreg).

```{r tpchreg, results = 'asis'}
fit.pch <- tpchreg(oe(event, exposure) ~ 
                                    sex + socStatus + illeg + parity,
                                time = age, data = tneo)
dr2 <- drop1(fit.pch, test = "Chisq")
ltx(fit.pch, dr = dr2, label = "tab:tpchreg", 
    caption = "Piecewise constant proportional hazards regression.")
```

Almost identical results compared to Table \@ref(tab:coxreg). We can compare 
baseline cumulative hazards plots (Figure \@ref(fig:baseplots)):

```{r baseplots, fig.cap = "Comparison.", fig.height = 6}
op <- par(mfrow = c(2, 1), las = 1)
plot(fit.cr, main = "Cox regression", xlab = "Age (days)", ylab = "Cumulative hazard")
plot(fit.pch, fn = "cum", main = "Pch regression", xlab = "Age (days)")
abline(h = 0)
par(op)
```

 Almost identical, disregarding differences in smoothness. An advantage with the 
 pch model is that we have access to an estimate of the baseline *hazard* function,
 see Figure \@ref(fig:haz).

```{r haz, fig.cap = "Baseline hazard function.",  fig.height = 4}
par(las = 1)
plot(fit.pch, fn = "haz")
abline(h = 0)
```

# Conclusion

There is almost never any disadvantage in analyzing tabular data instead of the original individual-based data, rather the opposite: No loss of information, and the table is *anonymous* and may be freely distributed and published. Further, the table is often just a small fraction in size
compared to the original (Table 1 consists of 391009 rows, Table 2 of 672 rows), resulting in analysis time being much shorter.

### 2021-12-14 getdata2.Rmd
modifications ...

@@
 options(show.signif.stars = FALSE)
-g <- function(x) (log(1 + 365 * x))^3
+g <- function(x) (log(1 + x))^3 # x is time in days! #### Note ###
 ```
@@
 The only (so far) difference compared to the article "getdata.Rmd" is that time 
-here is measured in *days* instead of in a fraction of a year. *However*, "at the end", time is transformed back to units in years. The main reason is that the function `make.communal`
+here is measured in *days* instead of in a fraction of a year. *However*, "at the end", 
+time is transformed back to units in years. The main reason is that the function `make.communal`
 assumes time measured in years, it is probably the variable `birthdate` that is the culprit.

### 2021-12-14 neo2.Rmd
modifications ...

@@
 as you are used to with individual data, is shown here. It helps if you have the 
-**R** package *eha* [@eha; @ehar2] for easy conversion of individual data to tabular
+**R** package *eha* [@eha; @ehar2] for easy conversion of individual data to tabular data
 suitable for survival analysis.
@@
 x <- head(neo) ## OK
-tbl(x, caption = "Raw individual data.")
+tbl(x, caption = "Raw individual data, age in days. First few rows.")
 ```
@@
 ##summary(tneo)
-tbl(X, caption = "Same data, tabular form.")
+tbl(X, caption = "Same data, tabular form, age in days. First few rows.")
 saveRDS(tneo, file = "t_data/tneo.rds")
@@
 
-The age interval is cut into 28 pieces, "0-1", "1-2", etc., so the underlying statistical survival model is that of a *piecevise constant hazards*.
+The age interval $(0, 28]$ is cut into 28 pieces, (0-1], (1-2], etc., so the underlying statistical survival model is that of a *piecevise constant hazards*.
 

### 2021-12-18 getdata3.Rmd
first commit

---
title: "getdata3.Rmd"
author: "Göran Broström"
date: "12/17/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(skum) # Also loads 'eha'
```

# Introduction

Here we start with skum::pers instead of skum::obs, and age is always in days,
with stillbirth age 0.099 and born alive but dead on birthday at age 0.25 days.

Further, we skip the use of eha::make.communal in  favour of *match*. Reason: 
eha::make.communal assumes age is measured in *years*.

# Read skum::pers

```{r readpers}
per <- pers[pers$frsbobtyp == 2 | pers$df == 1, ]
per <- per[per$foddat > 18950000, ]
dn <- per$doddat < 1
per$doddat[dn] <- per$foddat[dn] + 40000
## Fix foddat:
yr <- per$foddat %% 10000 == 0
per$foddat[yr] <- per$foddat[yr] + 101
mon <- per$foddat %% 100 == 0
per$foddat[mon] <- per$foddat[mon] + 1
## Fix doddat:
yr <- per$doddat %% 10000 == 0
per$doddat[yr] <- per$doddat[yr] + 101
mon <- per$doddat %% 100 == 0
per$doddat[mon] <- per$doddat[mon] + 1
##
per$birthdate <- as.Date(as.character(per$foddat), format = "%Y%m%d")
per$deathdate <- as.Date(as.character(per$doddat), format = "%Y%m%d")
nabir <- is.na(per$birthdate)
per$foddat[nabir] <- per$foddat[nabir] - 2
per$doddat[nabir] <- per$doddat[nabir] - 2
##
per$birthdate <- as.Date(as.character(per$foddat), format = "%Y%m%d")
per$exit <- as.numeric(difftime(per$deathdate, per$birthdate))
per$deathdate <- as.Date(as.character(per$doddat), format = "%Y%m%d")
## Fix doddat:
per$doddat[is.na(per$deathdate)] <- per$doddat[is.na(per$deathdate)] - 1
per$deathdate <- as.Date(as.character(per$doddat), format = "%Y%m%d")
##
per$exit <- as.numeric(difftime(per$deathdate, per$birthdate, units = "days"))
oj <- per$exit < 0
per$doddat[oj] <- per$doddat[oj] + 10000
per$deathdate <-  as.Date(as.character(per$doddat), format = "%Y%m%d")
per$exit <- as.numeric(difftime(per$deathdate, per$birthdate, units = "days"))
per$event <- per$exit <= 365
per$exit[per$exit == 0] <- 0.25
per$exit[per$exit > 365] <- 365
```

### 2021-12-18 getdata3.Rmd
continuing ...

@@
 
-Here we start with skum::pers instead of skum::obs, and age is always in days,
+Here we start with *skum::pers* instead of *skum::obs*, and age is always in days,
 with stillbirth age 0.099 and born alive but dead on birthday at age 0.25 days.
 
-Further, we skip the use of eha::make.communal in  favour of *match*. Reason: 
+Further, we eventually skip the use of eha::make.communal in  favor of *match*. Reason: 
 eha::make.communal assumes age is measured in *years*.
@@
 ```
+
+Now we need *nofrs* for newborn, found in the *observation* file(s), so
+
+```{r fromobs}
+bobs <- obs[obs$starttyp == 2 & obs$id %in% per$id, ]
+
+```

### 2021-12-19 getdata3.Rmd
continuing ..

@@
 bobs <- obs[obs$starttyp == 2 & obs$id %in% per$id, ]
-
+indx <- match(per$id, bobs$id)
+per$nofrs <- bobs$nofrs[indx]
+per <- per[!is.na(per$nofrs), ]
+per$subreg <- factor(per$nofrs)
+levels(per$subreg) <- c(rep("ume", 2), 
+                        rep("bjur", 4), 
+                        rep("sten", 3))
+per <- per[(per$subreg != "sten") & per$df == 0, ]
+per$subreg <- factor(per$subreg)
 ```

### 2021-12-20 getdata2.Rmd
modifications ...

@@
 ### Deaths:
-did <- unique(births$id[births$event])
+did <- unique(births$id[births$event == 1]) # NOTE: 'age.window' changes event = logical to numeric!
 deaths <- births[births$id %in% did, ]
@@
                                                    "parity", tempvars)]
-saveRDS(infdat, file = "data/infdat2_year.rds")
+saveRDS(infdat, file = "data/infdat2_year2.rds")
 ```

### 2021-12-20 getdata3.Rmd
modifications ...

--- 2021-12-18 getdata3.Rmd
@@
 bobs <- obs[obs$starttyp == 2 & obs$id %in% per$id, ]
-
+indx <- match(per$id, bobs$id)
+per$nofrs <- bobs$nofrs[indx]
+per <- per[!is.na(per$nofrs), ]
+per$subreg <- factor(per$nofrs)
+levels(per$subreg) <- c(rep("ume", 2), 
+                        rep("bjur", 4), 
+                        rep("sten", 3))
+per <- per[(per$subreg != "sten") & per$df == 0, ]
+per$subreg <- factor(per$subreg)
+saveRDS(per, file = "data/infdat3.rds")
 ```

### 2021-12-20 R/readobs.R
first commit

@@ -1,0 +1,65 @@
+readobs <- function(){
+    ## Equal to Code Chunk 'readobs' in getdata2.Rmd
+    ## Bug tracking!
+    library(skum)
+    ###
+    bb <- obs$id[obs$starttyp == 2]
+    births <- obs[obs$id %in% bb, 
+                  c("id", "sex", "foddat", "startdat", "slutdat", "nofrs", "sluttyp", "socStatus", "hisclass", 
+                    "ort", "ortnmn")]
+    births$event <- (births$sluttyp == 2)
+    ## New start in days...:
+    births$enter <- as.numeric(with(births, difftime(startdat, foddat, unit = "days")))
+    births$exit <- as.numeric(with(births, difftime(slutdat, foddat, unit = "days")))
+    ##
+    expand <- (births$startdat == births$slutdat) & births$event
+    births$exit[expand] <- births$enter[expand] + 0.25 # 6 hours!
+    ##
+    weq <- (births$startdat == births$slutdat) & (!births$event)
+
+    births <- births[!weq, ]
+    ##
+    births <- age.window(births, c(0, 365))
+    births <- births[order(births$id, births$enter), ]
+    ##return(births)
+    ################################### 1st check OK!!!!
+    ##indx <- tapply(births$id, births$id)
+    ##births$mexit <- tapply(births$exit, births$id, max)[indx]
+    ##births$mevent <- tapply(births$event, births$id, max)[indx]
+    ##births <- births[!duplicated(births$id), ]
+    ### Deaths:
+    did <- unique(births$id[births$event == 1]) # NOTE: 'age.window' changes event = logical to numeric!
+    deaths <- births[births$id %in% did, ]
+    indx <- tapply(deaths$id, deaths$id)
+    deaths$menter <- tapply(deaths$enter, deaths$id, min)[indx]
+    return(deaths)
+    ################################# 2nd check
+    deaths <- deaths[!duplicated(deaths$id), ]
+    deaths$enter <- deaths$menter
+    deaths$menter <- NULL
+    deaths$enter <- 0 ## satisfied....
+    ### Survivors:
+    survs <- births[!(births$id %in% did), ]
+    indx <- tapply(survs$id, survs$id)
+    survs$mexit <- tapply(survs$exit, survs$id, max)[indx]
+    survs <- survs[!duplicated(survs$id), ]
+    survs$exit <- survs$mexit
+    survs$mexit <- NULL
[…]

### 2021-12-21 R/readobs.R
adding ...

@@ -32,10 +32,14 @@ readobs <- function(){
     deaths <- births[births$id %in% did, ]
     indx <- tapply(deaths$id, deaths$id)
     deaths$menter <- tapply(deaths$enter, deaths$id, min)[indx]
-    return(deaths)
+    ##return(deaths)
     ################################# 2nd check
+    ## NEW: Sort before unique:
+    deaths <- deaths[order(deaths$exit, decreasing = TRUE), ]
     deaths <- deaths[!duplicated(deaths$id), ]
     deaths$enter <- deaths$menter
+    return(deaths)
+    ################################## 3rd check
     deaths$menter <- NULL
     deaths$enter <- 0 ## satisfied....
     ### Survivors:

### 2021-12-21 R/readobs.R
modified, but probably abandoned

@@ -2,15 +2,30 @@ readobs <- function(){
     ## Equal to Code Chunk 'readobs' in getdata2.Rmd
     ## Bug tracking!
     library(skum)
-    ###
+### Separate check of consistency pers <--> obs:
+    bo <- obs[obs$starttyp == 2, c("id", "startdat", "foddat")]
+    bo <- bo[order(bo$id), ]
+    bp <- pers[pers$id %in% bo$id, c("id", "frsbobtyp", "foddat")]
+    ##bp <- pers[pers$frsbobtyp == 2, c("id", "foddat")]
+    bp$startdat <- as.Date(as.character(bp$foddat), format = "%Y%m%d")
+    ##bp$slutdat <- as.Date(as.character(bp$doddat), format = "%Y%m%d")
+    bp <- bp[order(bp$id), ]
+    ## Remove NA i startdat i bp
+    
+    return(list(bo = bo, bp = bp))
+### End separate
+    
     bb <- obs$id[obs$starttyp == 2]
     births <- obs[obs$id %in% bb, 
-                  c("id", "sex", "foddat", "startdat", "slutdat", "nofrs", "sluttyp", "socStatus", "hisclass", 
+                  c("id", "sex", "foddat", "startdat", "slutdat",
+                    "nofrs", "sluttyp", "socStatus", "hisclass", 
                     "ort", "ortnmn")]
     births$event <- (births$sluttyp == 2)
     ## New start in days...:
-    births$enter <- as.numeric(with(births, difftime(startdat, foddat, unit = "days")))
-    births$exit <- as.numeric(with(births, difftime(slutdat, foddat, unit = "days")))
+    births$enter <- as.numeric(with(births,
+                                    difftime(startdat, foddat, unit = "days")))
+    births$exit <- as.numeric(with(births,
+                                   difftime(slutdat, foddat, unit = "days")))
     ##
     expand <- (births$startdat == births$slutdat) & births$event
     births$exit[expand] <- births$enter[expand] + 0.25 # 6 hours!

### 2021-12-21 getdata3.Rmd
modified, and probably the way to go (via 'pers')

--- 2021-12-18 getdata3.Rmd
@@
 author: "Göran Broström"
-date: "12/17/2021"
+date: "`r Sys.Date()`"
 output:
@@
 
-Here we start with *skum::pers* instead of *skum::obs*, and age is always in days,
-with stillbirth age 0.099 and born alive but dead on birthday at age 0.25 days.
+Now (December 22, 2021) I have come to the conclusion that data from the data frame 
+*persons* is the best and easiest start.
+
+So we start with *skum::persons* instead of *skum::obs*, and age is to begin with in days,
+with stillbirths age *removed* and born alive but dead on birthday at age 0.25 days.
+
+For user of *eha::make.communal*, we need to temporarily change time unit to years.
 
 Further, we eventually skip the use of eha::make.communal in  favor of *match*. Reason: 
-eha::make.communal assumes age is measured in *years*.
+eha::make.communal assumes age is measured in *years*. But this is no promise yet.
 
@@
 ```{r readpers}
-per <- pers[pers$frsbobtyp == 2 | pers$df == 1, ]
-per <- per[per$foddat > 18950000, ]
+per <- pers[(pers$frsbobtyp == 2) & (pers$df == 0), ]
+per$frsbobtyp <- per$df <- per$f2id <- per$m2id <- NULL
+per$dodors <- per$frsbobdatmin <- per$frsbobdatmax <- per$utdatmax <- per$utdatmin <- NULL
+per <- per[(per$foddat > 18950000), ]
 dn <- per$doddat < 1
@@
 yr <- per$foddat %% 10000 == 0
-per$foddat[yr] <- per$foddat[yr] + 101
+per$foddat[yr] <- per$foddat[yr] + 101 # As early as possible
 mon <- per$foddat %% 100 == 0
-per$foddat[mon] <- per$foddat[mon] + 1
+per$foddat[mon] <- per$foddat[mon] + 1 # Ditto
 ## Fix doddat:
 yr <- per$doddat %% 10000 == 0
-per$doddat[yr] <- per$doddat[yr] + 101
+per$doddat[yr] <- per$doddat[yr] + 1231 # As late as possible
 mon <- per$doddat %% 100 == 0
-per$doddat[mon] <- per$doddat[mon] + 1
+per$doddat[mon] <- per$doddat[mon] + 28 # Ditto (guard for February ...)
 ##
@@
 bobs <- obs[obs$starttyp == 2 & obs$id %in% per$id, ]
-
+indx <- match(per$id, bobs$id)
+per$nofrs <- bobs$nofrs[indx]
+per <- per[!is.na(per$nofrs), ]
+per$subreg <- factor(per$nofrs)
+levels(per$subreg) <- c(rep("ume", 2), 
+                        rep("bjur", 4), 
+                        rep("sten", 3))
+per <- per[(per$subreg != "sten"), ]
+per$subreg <- factor(per$subreg)
+per$enter <- 0
+saveRDS(per, file = "data/infdat3.rds")
 ```

### 2021-12-22 getdata3.Rmd
further development based on skum::pers

@@
 per$enter <- 0
-saveRDS(per, file = "data/infdat3.rds")
+per$sex <- factor(per$kon, labels = c("boy", "girl"))
+
+##saveRDS(per, file = "data/infdat3.rds")
 ```
+
+# Put on temperature data
+
+```{r makecommunal}
+temp_start <- 1894
+temp_end <- 1951
+temp_ivl <- c(temp_start, temp_end)
+##
+umetemp <- readRDS("data/umetemp.rds")
+bjurtemp <- readRDS("data/bjurtemp.rds")
+##
+
+lagg <- 3 / 365  ## NOTE!!
+tempvars <-  c("heat", "aver", "cold", "emintemp", "emeantemp", "emaxtemp", 
+               "week", "year") 
+## Ume:
+bume <- per[per$subreg == "ume", ]
+bume <- make.communal(bume, umetemp[, tempvars], 
+                      start = temp_start, period = 1/52, lag = lagg)
+
+### Ske:
+bskel <- per[per$subreg == "bjur", ]
+bskel <- make.communal(bskel, bjurtemp[, tempvars], 
+                       start = temp_start, period = 1/52, lag = lagg)
+
+bume$region <- "ume"
+bskel$region <- "ske"
+infdat <- rbind(bume, bskel)
+infdat <- infdat[order(infdat$id, infdat$enter), c("region", "id", "sex", "enter", "exit", "event", 
+                                                   "birthdate", "socStatus", "hisclass", "illeg",
+                                                   "parity", tempvars)]
+saveRDS(infdat, file = "data/infdat2_year2.rds")
+
+```

### 2021-12-22 paper.Rmd
modifications positively ...
further development based on skum::pers

@@
 library(eha)
-g <- function(x) (log(1 + 365 * x))^3
-ginv <- function(y) (1 / 365) * (exp(y^(1/3)) - 1)
+g <- function(x) (log(1 + x))^3 # x in days.
+ginv <- function(y) exp(y^(1/3)) - 1
 ```
@@
 \begin{equation*}\label{eq:g}
-g(t) = \log^3(1 + 365 t), \quad 0 \le t \le 1,
+g(t) = \log^3(1 + t), \quad 0 \le t \le 365,
 \end{equation*}
 
-where $t$ is age measured in *years*. 
+where $t$ is age measured in *days*. 
 Its graph is shown in Figure \@ref(fig:figgraph).
@@
 ##g <- function(x) (log(1 + x))^3  # Note: days!
-x <- seq(0, 1, length = 1000)
+x <- seq(0, 365, length = 1000)
 y <- g(x)
 plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, col = "blue")
-xat <- c(0, 28 / 365, 1 / 2, 1)
+xat <- c(0, 28, 365)
 yat <- g(xat)
-axis(1, at = xat, labels = round(365 * xat, 3))
-axis(2, at = yat, labels = round(yat, 1), las = 2)
+axis(1, at = xat)
+axis(2, at = yat, labels = round(yat, 1), las = 1)
 box()
@@
 abline(h = 0, v = 0)
+abline(a = 0, b = 1, col = "red")
 ```
@@
 The B-P model specifies that, on the transformed age scale, the *exogeneous*
-mortality distribution, conditional on death before age one,  is *uniform*, that is, the 
+mortality distribution, conditional on death before age 365 days,  is *uniform*, that is, the 
 survivor function of $T$ is *linear*.
@@
 ```{r readdata}
-infdat <- readRDS("data/infdat.rds")
+infdat <- readRDS("data/infdat3.rds")
 ```
@@
 From the Umeå-Skellefteå region, we follow all persons born in the 
-region 1895--1950 until death, reaching the age of 1, or leaving follow-up for whatever 
-reason, whichever comes first.
+region 1895--1950 until death, reaching the age of 1 year, or leaving follow-up for whatever 
+reason, whichever comes first. The first few lines (infants) are
+
+```{r infdat3}
+source("R/tbl.R")
+sinf <- infdat[, c("region", "kon", "birthdate", "deathdate", "exit", "event")]
+x <- head(infdat, 5)
+tbl(x, caption = "A few infants in the data set.")
+```
+
+## The Bourgeois-Pichat procedure
+
+## Add weather data
 
@@
 ```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950."}
+
 tid <- toTpch(Surv(enter, exit, event) ~ week, data =  infdat, cuts = c(0, 1))

### 2021-12-28 getdata3.Rmd
add some personal variables

--- 2021-12-21 getdata3.Rmd
@@
 indx <- match(per$id, bobs$id)
+per$socStatus <- bobs$socStatus[indx]
+per$hisclass <- bobs$hisclass[indx]
 per$nofrs <- bobs$nofrs[indx]
@@
 per$enter <- 0
+per$sex <- factor(per$kon, labels = c("boy", "girl"))
+per$birthdate <- toTime(per$birthdate)
+per$illeg <- as.numeric(per$ab == 2)
+per$parity <- per$paritet_g
 saveRDS(per, file = "data/infdat3.rds")
 ```
+
+# Put on temperature data
+
+Here we must change time unit to *year*.
+
+```{r makecommunal}
+per$enter <- per$enter / 365
+per$exit <- per$exit / 365
+## per$birthdate <- toTime(per$birthdate) #already done!
+temp_start <- 1894
+temp_end <- 1951
+temp_ivl <- c(temp_start, temp_end)
+##
+umetemp <- readRDS("data/umetemp.rds")
+bjurtemp <- readRDS("data/bjurtemp.rds")
+##
+
+lagg <- 3 / 365  ## NOTE!!
+tempvars <-  c("heat", "aver", "cold", "emintemp", "emeantemp", "emaxtemp", 
+               "week", "year") 
+## Ume:
+bume <- per[per$subreg == "ume", ]
+bume <- make.communal(bume, umetemp[, tempvars], 
+                      start = temp_start, period = 1/52, lag = lagg)
+
+### Ske:
+bskel <- per[per$subreg == "bjur", ]
+bskel <- make.communal(bskel, bjurtemp[, tempvars], 
+                       start = temp_start, period = 1/52, lag = lagg)
+
+bume$region <- "ume"
+bskel$region <- "ske"
+infdat <- rbind(bume, bskel)
+infdat <- infdat[order(infdat$id, infdat$enter), c("region", "id", "sex", "enter", "exit", "event", 
+                                                   "birthdate", "socStatus", "hisclass", "illeg",
+                                                   "parity", tempvars)]
+saveRDS(infdat, file = "data/infdat3_year3.rds")
+
+```

### 2021-12-30 getdata3.Rmd
reduce variable list

@@
 per$parity <- per$paritet_g
+per$paritet_g <- per$paritet_k <- per$doddat <- per$frsbosdat <- NULL
+per$fid <- per$fboort <- per$ab <- per$fyrkfd <- per$fyrkind <- NULL
+per$region <- per$kon <- per$fodhfrs <- per$frsbobdat <- per$frsbostyp <- NULL
+per$fbonmn <- per$fboind <- per$fyrkrel <- per$fyrketxt <- NULL
+per$foddat <- as.Date(as.character(per$foddat), format = "%Y%m%d")
+per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", 
+               "subreg", "socStatus", "enter", "exit", "event")]
 saveRDS(per, file = "data/infdat3.rds")
@@
 infdat <- rbind(bume, bskel)
-infdat <- infdat[order(infdat$id, infdat$enter), c("region", "id", "sex", "enter", "exit", "event", 
-                                                   "birthdate", "socStatus", "hisclass", "illeg",
+infdat <- infdat[order(infdat$id, infdat$enter), c("subreg", "id", "sex", "enter", "exit", "event", 
+                                                   "birthdate", "socStatus", "illeg",
                                                    "parity", tempvars)]

### 2022-01-01 paper.Rmd
developing ...

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=4,comment=NA)
+knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=5,comment=NA)
 library(knitr)
@@
 source("R/tbl.R")
-sinf <- infdat[, c("region", "kon", "birthdate", "deathdate", "exit", "event")]
+sinf <- infdat[, c("subreg", "sex", "birthdate", "exit", "event")]
 x <- head(infdat, 5)
@@
 
+The BP procedure consists of two steps, to begin with:
+
+1.   Apply the log-cube transformation to the age data, and
+
+2.   plot the cumulative sum of the number of deaths vs. age.
+
+We plot without the first step in the left hand panel and with it in the
+right hand panel, see Figure \@ref(fig:ggplot2).
+
+```{r ggplot2, fig.cap = "BB plots with and without time transform.", fig.height = 7}
+nbths <- NROW(infdat)
+oldpar <- par(mfrow = c(2, 1))
+sur <- with(infdat, Surv(enter, exit, event))
+rs <- risksets(sur)
+ss <- cumsum(rs$n.events)
+## Natural:
+xt <- rs$risktimes
+att <- c(0, 28, 183, 365)
+plot(xt, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
+     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
+     main = "Untransformed", ylab = "Cum. deaths")
+abline(h = 0)
+axis(2, las = 1)
+axis(1, at = att, labels = att)
+box()
+## g-transformed:
+plot(g(xt), ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
+     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
+     main = "After g-transform", ylab = "Cum. deaths")
+abline(h = 0)
+axis(2, las = 1)
+axis(1, at = g(att), labels = att)
+box()
+##
+par(oldpar)
+```
+
 ## Add weather data
@@
 ```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950."}
-
+infdat <- readRDS("data/infdat3_year3.rds")
 tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))

### 2022-01-03 paper.Rmd
continuations ...
enhance g-transform plot

@@
 
-We plot without the first step in the left hand panel and with it in the
-right hand panel, see Figure \@ref(fig:ggplot2).
+We plot without the first step in the top panel and with it in the
+bottom panel, see Figure \@ref(fig:ggplot2).
 
-```{r ggplot2, fig.cap = "BB plots with and without time transform.", fig.height = 7}
+```{r ggplot2, fig.cap = "BB plots with and without time transform.", fig.height = 5}
 nbths <- NROW(infdat)
-oldpar <- par(mfrow = c(2, 1))
+##oldpar <- par(mfrow = c(2, 1))
 sur <- with(infdat, Surv(enter, exit, event))
@@
      axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
-     main = "Untransformed", ylab = "Cum. deaths")
+     ylab = "Cum. deaths")
 abline(h = 0)
@@
 ## g-transformed:
-plot(g(xt), ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
-     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
+lines(g(xt) / g(365) * 365, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
+     col = "red", lwd = 2, xlab = "Age (days)",
      main = "After g-transform", ylab = "Cum. deaths")
-abline(h = 0)
-axis(2, las = 1)
-axis(1, at = g(att), labels = att)
+##abline(h = 0)
+##axis(2, las = 1)
+axis(3, at = g(att) / g(365) * 365, labels = att, col = "red")
 box()
+abline(v = 0)
+abline(v = g(c(28, 183, 365)) / g(365) * 365, col = "red", lty = 3)
+abline(v = c(28, 183, 365), col = "blue", lty = 3)
 ##
-par(oldpar)
+legend("bottomright", legend = c("Natural", "g-transform"), col = c("blue", "red"), lty = 1)
+abline(h = (ss / nbths) [28], lty = 3 )
+abline(h = (ss / nbths) [180], lty = 3 )
+##par(oldpar)
 ```

### 2022-01-07 paper.Rmd
as usual

@@
 plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, col = "blue")
-xat <- c(0, 28, 365)
+xat <- c(0, 28, 183, 365)
 yat <- g(xat)
@@
 box()
-lines(c(30, 30), c(0, g(30)), lty = 2, col = "darkgreen")
-lines(c(0, 30), c(g(30), g(30)), lty = 2, col = "darkgreen")
+lines(c(28, 28), c(0, g(28)), lty = 2, col = "darkgreen")
+lines(c(0, 28), c(g(28), g(28)), lty = 2, col = "darkgreen")
+##
+lines(c(183, 183), c(0, g(183)), lty = 2, col = "darkgreen")
+lines(c(0, 183), c(g(183), g(183)), lty = 2, col = "darkgreen")
+##
+lines(c(365, 365), c(0, g(365)), lty = 2, col = "darkgreen")
+lines(c(0, 365), c(g(365), g(365)), lty = 2, col = "darkgreen")
 abline(h = 0, v = 0)
-abline(a = 0, b = 1, col = "red")
+##abline(a = 0, b = 1, col = "red")
 ```
@@
 region 1895--1950 until death, reaching the age of 1 year, or leaving follow-up for whatever 
-reason, whichever comes first. The first few lines (infants) are
+reason, whichever comes first. The first few lines (infants) are shown in Table 
+\@ref(tab:infdat3).
 
@@
 
-We plot without the first step in the top panel and with it in the
-bottom panel, see Figure \@ref(fig:ggplot2).
+The result is shown in Figure \@ref(fig:ggplot2), with the original, untransformed
+data together with the transform, stretched to 365.
 
-```{r ggplot2, fig.cap = "BB plots with and without time transform.", fig.height = 5}
+```{r ggplot2, fig.cap = "BB plots with and without time transform, Umeå-Skellefteå 1895-1950.", fig.height = 5}
 nbths <- NROW(infdat)
@@
 ```
+
+The plot, disregarding the transform, is just a graph of the empirical *cumulative distribution function (CDF)*,
+that is, one minus the *Kaplan-Meier estimate* of the *survivor function*.
+
+## Post-neonatal mortality
+
+We begin by studying mortality after age 28 days on the g-scale, BP predicts that it is approximately *uniform*. 
+The uniform distribution, however, is not very useful in a survival analysis context, so we will
+investigate a close alternative, the *exponential * distribution. Modifying the BP plot by using successive 
+divison by *risk set size* instead of sample size should result in an approximately linear graph if the 
+exponential assumption is valid.
+
+```{r pneo, fig.cap = "Relation between cumulative hazards and CDF for transformed post-neonatal mortality data."}
+ginfdat <- age.window(infdat, c(28, 365))
+ginfdat$enter <- g(ginfdat$enter)
+ginfdat$exit <- g(ginfdat$exit)
+grs <- risksets(with(ginfdat, Surv(enter, exit, event)), members = FALSE)
+fit <- coxreg(Surv(enter, exit, event) ~ 1, data = ginfdat)
+ser <- plot(fit, fn = "sur", fig = FALSE)
+##plot(fit, col = "blue", axes = FALSE)
+cum <- fit$hazards[[1]]
+x <- cum[, 1]
+ch <- cumsum(cum[, 2])
+plot(x, ch, type = "s", axes = FALSE, col = "blue", xlab = "Age (days)", ylab = "")
+axis(1, at = c(g(28), g(183), g(365)), labels = c(28, 183, 365))
+axis(2, las = 1)
+abline(h = 0)
+box()
+x <- ser[[1]][[1]][, 1]
+y <- ser[[1]][[1]][, 2]
+lines(x, 1 - y, col = "red", lty = 2)
+legend("topleft", legend = c("Cumulative hazards", "CDF (Bourgeouis-Pichat)"), col = c("blue", "red"),
+       lty = c(1, 2))
+```
+
+As is evident from Figure \@ref(fig:pneo), the difference between the the two curves is
+negligible, and both are well approximated by a straight line, so we will from now on work with the 
+assumption that the transformed post-neonatal data are *exponentially distributed*.
+There is also a mathematical argument motivating the similarity: The cdf for an exponentially 
+distributed random variable $T$ is
+
+\begin{equation*}
+F_T(t) = 1 - \exp(-\lambda t), \quad t, \lambda > 0, 
+\end{equation*}
+
+and the first order approximation of this cdf is
+
+\begin{equation*}
+F_T(t) \approx \lambda t, \quad t, \lambda > 0,
+\end{equation*}
+
+which is the *cumulative hazards* function of $T$,
+and the approximation is good for small values of $\lambda t$.
 

### 2022-01-09 paper.Rmd
more on neonatal

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=5,comment=NA)
+knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=4,comment=NA)
 library(knitr)
@@
 
-## The Bourgeois-Pichat procedure
+### The Bourgeois-Pichat procedure
 
@@
 
-## Post-neonatal mortality
+### Post-neonatal mortality
 
@@
 and the approximation is good for small values of $\lambda t$.
+
+So, the *hazard function* (which is the slope of the lines in Figure \@ref(fig:pneo))
+is simply estimated by the ratio of the number of deaths to the total time of exposure:
+
+```{r simpest, echo = TRUE}
+posthaz <- with(ginfdat, sum(event) / sum(exit - enter))
+```
+
+which evaluates to `r format(posthaz, digits = 3, scientific = TRUE)`. A small number, but remember
+that the time unit is defined on the scale defined by the function $g$ (close to *day*).
+
+### Neonatal mortality
+
+Is the sum of two components, *external* and *internal* causes. According to the *B-P*
+theory, the external component is equal to the post-neonatal mortality, which contains 
+only an external component.
+
+We create the neonatal data set by censor all lives at the age 28:
+
+```{r neon, echo = TRUE}
+neo <- age.window(infdat, c(0, 28))
+neo$enter <- g(neo$enter)
+neo$exit <- g(neo$exit)
+```
+
+and then fit a piecewise constant hazards model with cutpoints at (0, 1, 2, 4, 8, 12, 18, 24, 30, 35, 39), see 
+Figure \@ref(fig:pchneo).
+
+```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 4}
+fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = c(0,  1, 2, 4, 8, 12, 18, 24, 30, 35, 39))
+plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)")
+abline(h = 0, v = 0)
+abline(h = posthaz, col = "red", lty = 2)
+```
+
+Let us take a closer look at what happens at the end of the neonatal inerval:
+
+```{r pchneo2, fig.cap = "Neonatal hazard function with external component (dashed), right hand tail.", fig.height = 4}
+##fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = c(0,  1, 2, 4, 8, 12, 18, 24, 30, 39))
+plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", xlim = c(15, 39))
+abline(h = 0, v = 0)
+abline(h = posthaz, col = "red", lty = 2)
+```
+
+A smooth transition to the post-neonatal mortality.
 

### 2022-01-09 paper.Rmd
rearranging

@@
     number_sections: yes
-    toc: no
-    toc_depth: 2
+    toc: yes
+    toc_depth: 3
   bookdown::word_document2:
@@
 
-# The Bourgeouis-Pichat model and the approximation
+# The Bourgeouis-Pichat model
 
@@
 easy to use in standard survival regression analysis, often allowing for massive data reduction
-through the statistical *sufficency pinciple*.
-
-# Data
+through the statistical *sufficency principle*.
 
@@
 
-## Add weather data
+### Putting it together
+
+# Data
+
+## Infant data
+
+In the previous chapter we introduced data on infant mortality from the Umeå-Skellefteå
+region 1895--1950, and here we introduce covariates of interest in the forthcoming  analyses, 
+see Table
+
+## Weather data
 
@@
 
-In accordance with the Bourgeouis-Pichat ideas, we group the analyses into two parts, 
+In accordance with the ideas of Bourgeouis-Pichat, we group the analyses into two parts, 
 *neonatal* (first 28 days) och *postneonatal* mortality (days 29--365).

### 2022-01-11 getdata3.Rmd
add socBranch to output

--- 2021-12-28 getdata3.Rmd
@@
 per$socStatus <- bobs$socStatus[indx]
+per$socBranch <- bobs$socBranch[indx]
 per$hisclass <- bobs$hisclass[indx]
@@
 per$parity <- per$paritet_g
+per$paritet_g <- per$paritet_k <- per$doddat <- per$frsbosdat <- NULL
+per$fid <- per$fboort <- per$ab <- per$fyrkfd <- per$fyrkind <- NULL
+per$region <- per$kon <- per$fodhfrs <- per$frsbobdat <- per$frsbostyp <- NULL
+per$fbonmn <- per$fboind <- per$fyrkrel <- per$fyrketxt <- NULL
+per$foddat <- as.Date(as.character(per$foddat), format = "%Y%m%d")
+per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", 
+               "subreg", "socBranch", "socStatus", "enter", "exit", "event")]
 saveRDS(per, file = "data/infdat3.rds")
@@
 infdat <- rbind(bume, bskel)
-infdat <- infdat[order(infdat$id, infdat$enter), c("region", "id", "sex", "enter", "exit", "event", 
-                                                   "birthdate", "socStatus", "hisclass", "illeg",
+infdat <- infdat[order(infdat$id, infdat$enter), c("subreg", "id", "sex", "enter", "exit", "event", 
+                                                   "birthdate", "socBranch", "socStatus", "illeg",
                                                    "parity", tempvars)]

### 2022-01-11 paper.Rmd
rearranging ...

@@
 g <- function(x) (log(1 + x))^3 # x in days.
+dg <- function(x) (3 * (log(1 + x))^2) / (1 + x)
 ginv <- function(y) exp(y^(1/3)) - 1
@@
 y <- g(x)
-plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, col = "blue")
+plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, 
+     col = "blue")
 xat <- c(0, 28, 183, 365)
@@
 
+If `event == FALSE`, then the infant died at age (in days) given by `exit`, otherwise
+the infant survived (at least) 365 days and follow-up is lost at `exit == 365`. For the
+time being, no covariates are involved. So, in Table \@ref(tab:infdat3), infants No. 1--4 survived, 
+while infant No. 5 died at age 162 days.
+
 ```{r infdat3}
 source("R/tbl.R")
-sinf <- infdat[, c("subreg", "sex", "birthdate", "exit", "event")]
-x <- head(infdat, 5)
-tbl(x, caption = "A few infants in the data set.")
+sinf <- infdat[, c("enter", "exit", "event")]
+x <- head(sinf, 4)
+y <- sinf[sinf$event == 1, ][1, ]
+z <- rbind(x, y)
+z$id <- 1:5
+z <- z[, c("id", "enter", "exit", "event")]
+tbl(z, caption = "A few infants in the data set.")
 ```
@@
 
-```{r ggplot2, fig.cap = "BB plots with and without time transform, Umeå-Skellefteå 1895-1950.", fig.height = 5}
+```{r ggplot2, fig.cap = "BB plots with and without time transform, Umeå-Skellefteå 1895-1950."}
 nbths <- NROW(infdat)
@@
 
-We begin by studying mortality after age 28 days on the g-scale, BP predicts that it is approximately *uniform*. 
+We begin by studying mortality after age 28 days on the $g$-scale, *B-P* predicts that it is approximately *uniform*. 
 The uniform distribution, however, is not very useful in a survival analysis context, so we will
-investigate a close alternative, the *exponential * distribution. Modifying the BP plot by using successive 
-divison by *risk set size* instead of sample size should result in an approximately linear graph if the 
+investigate a close alternative, the *exponential * distribution. Modifying the *B-P* plot by using successive 
+divisions by *risk set size* instead of sample size should result in an approximately linear graph if the 
 exponential assumption is valid.
@@
 
-and the first order approximation of this cdf is
+and the first order approximation of this expression is
 
@@
 
-and then fit a piecewise constant hazards model with cutpoints at (0, 1, 2, 4, 8, 12, 18, 24, 30, 35, 39), see 
+and then fit a piecewise constant hazards model with cutpoints at $g(0, 1, 2, 4, 7, 11, 16, 22, 28)$, see 
 Figure \@ref(fig:pchneo).
@@
 ```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 4}
-fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = c(0,  1, 2, 4, 8, 12, 18, 24, 30, 35, 39))
-plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)")
+neocuts <- g(c(0, 2, 4, 7, 11, 16, 22, 28))
+fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
+plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", axes = FALSE)
+axis(1)
+axis(2, las = 1)
+box()
 abline(h = 0, v = 0)
@@
 
-Let us take a closer look at what happens at the end of the neonatal inerval:
+Let us take a closer look at what happens at the end of the neonatal interval, see 
+Figure \@ref(fig:pchneo2).
 
 ```{r pchneo2, fig.cap = "Neonatal hazard function with external component (dashed), right hand tail.", fig.height = 4}
-##fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = c(0,  1, 2, 4, 8, 12, 18, 24, 30, 39))
-plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", xlim = c(15, 39))
+##fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
+par(las = 1)
+plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", xlim = c(5, 39))
 abline(h = 0, v = 0)
@@
 
-### Putting it together
+### Putting all together
+
+Looking at the full infant mortality interval on the $g$-scale, we have a piecewise 
+constant hazards model with cutpoints at $g(0, 1, 2, 4, 7, 11, 16, 22, 28)$, see 
+Figure \@ref(fig:together) (where the value close to zero is truncated).
+
+```{r together, fig.cap = "Hazard function, infant mortality, transformed time."}
+allcuts <- c(neocuts, g(365))
+ginfdat <- infdat
+ginfdat$exit <- g(ginfdat$exit)
+fitall <- pchreg(Surv(enter, exit, event) ~ 1, data = ginfdat, cuts = allcuts)
+plot(fitall, fn = "haz", col = "blue", main = "", xlab = "g(Days)", axes = FALSE,
+     ylim = c(0, 0.002))
+pcuts <- g(c(0, 28, 183, 365))
+axis(1, at = pcuts, labels = round(pcuts))
+axis(2, las = 1)
+box()
+abline(h = 0, v = 0)
+```
+
+Now we would like to see the content of Figure \@ref(fig:together) on the natural time 
+scale. If $Y = g(T)$, we have the relation between their respective hazard functions
+
+\begin{equation*}
+h_T(t) = \frac{d}{dt} g(t) h_Y\bigl(g^{-1}(t)\bigr), \quad 0 \le t \le 365,
+\end{equation*}
+
+and with $h_Y$ shown in Figure \@ref(fig:together), we get $h_T$ in Figure 
+\@ref(fig:allnorm).
+
+```{r allnorm, fig.cap = "Hazard function, infant mortality, natural time scale."}
+y <- hazards(fitall, cum = FALSE)$x
+hy <- hazards(fitall, cum = FALSE)$y
+te <- ginv(y)
+hte <- dg(te) * hy
+plot(te, hte, type = "s", col = "blue", xlab = "Days", ylab = "", axes = FALSE,
+     ylim = c(0, 0.002))
+axis(1, at = ginv(pcuts))
+axis(2, las = 1)
+box()
+abline(h = 0, v = 0)
+```
 
 # Data
-
-## Infant data
-
-In the previous chapter we introduced data on infant mortality from the Umeå-Skellefteå
-region 1895--1950, and here we introduce covariates of interest in the forthcoming  analyses, 
-see Table
 
@@
 *neonatal* (first 28 days) och *postneonatal* mortality (days 29--365).
+
+## Infant data
+
+In the previous chapter we introduced data on infant mortality from the Umeå-Skellefteå
+region 1895--1950, and here we introduce covariates of interest in the forthcoming  analyses, 
+see Table
 

### 2022-01-11 paper.Rmd
a4wide

@@
 - \usepackage{graphicx}
+- \usepackage{a4wide}
 abstract: "The effect of extreme temperatures on infant mortality in the Umeå region 1895-1950 is studied in a biometric analysis setting."

### 2022-01-11 paper.Rmd
expanding

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=4,comment=NA)
+knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3,comment=NA)
 library(knitr)
@@
 is studied and interacted with sex, social class, and legitimacy. Studies are performed 
-separatelyfor neonatal  and post-neonatal mortality, and for winter and summer seasons, 
+separately for neonatal  and postneonatal mortality, and for winter and summer seasons, 
 and in a joint analysis, the classification into endogenous and exogenous factors will be discussed.
+
+One important reason for studying neonatal and postneonatal mortality separately is the 
+findings by Bourgeouis-Pichat [@bp51a;  @bp51b] about endogeneous and exogeneous mortality and the log-cube transform,
+so we start by recollecting his results and formulate them in a modern statistical setting.
 
@@
 
-A variation of the Bourgeouis-Pichat (B-P) biometric analysis [@bp51a;
-@bp51b] of infant mortality is suggested. B-P suggests that mortality
-during the last eleven months of infancy follows a *uniform* distribution
+Bourgeouis-Picat (*B-P*) suggests that mortality
+during the last eleven months (28 days) of infancy follows a *uniform* distribution
 given a specific transformation of time (age), the 
-*log-cube ("BP") transformation*. We show the BP transformation is closely approximated by
-an *exponential distribution*, 
-that is, the *cumulative hazards function* is linear.
- The advantage is that estimation of exogeneous and
-endogeneous infant mortality is easily performed with standard survival
-analysis programs.
-
-Central in the *Bourgeouis-Pichat (B-P)* model is the *log-cube* time transform $g$:
+*log-cube transformation* $g$.
 
@@
 
-```{r figgraph, fig.cap = "The log-cube transform g."}
+```{r figgraph, fig.cap = "The log-cube transform g.", fig.height = 4}
 ##g <- function(x) (log(1 + x))^3  # Note: days!
@@
 
-The B-P model specifies that, on the transformed age scale, the *exogeneous*
+The *B-P* model specifies that, on the transformed age scale, the *exogeneous*
 mortality distribution, conditional on death before age 365 days,  is *uniform*, that is, the 
-survivor function of $T$ is *linear*.
+survivor function of $T$ is *linear*,
 
 \begin{equation}
-S(t) = 1 - \alpha t, \, 0 < t < g(1).
+S(t) = 1 - \alpha t, \, 0 \le t \le g(365).
 \end{equation}
@@
 
-It seems to fit infant mortality data very well, but one drawback with it is that the 
-uniform distribution is not easy to work with in a survival analysis context.
+Next we apply the model to our infant mortality data from the Umeå-Skellefteå region.
 
-However, we show that the BP transformation is well approximated by a model where the 
-*cumulative hazards function* is linear instead, that is, an *exponential distribution*,
-easy to use in standard survival regression analysis, often allowing for massive data reduction
-through the statistical *sufficency principle*.
-
-## Infant Mortality
+## Infant mortality, a first look
 
@@
 reason, whichever comes first. The first few lines (infants) are shown in Table 
-\@ref(tab:infdat3).
+\@ref(tab:infdat3). Note that covariates are left out for the time being.
 
-If `event == FALSE`, then the infant died at age (in days) given by `exit`, otherwise
-the infant survived (at least) 365 days and follow-up is lost at `exit == 365`. For the
-time being, no covariates are involved. So, in Table \@ref(tab:infdat3), infants No. 1--4 survived, 
-while infant No. 5 died at age 162 days.
+If `event == TRUE`, then the infant died at age (in days) given by `exit`, otherwise
+the infant survived (at least) 365 days and follow-up is lost at `exit == 365`. 
+So, in Table \@ref(tab:infdat3), infants No. 1--4 survived, 
+while infant No. 5 died at age 162 days. We assume here that all infants are followed from birth,
+that is, `enter == 0` in all cases (no left truncation).
 
@@
 
-The BP procedure consists of two steps, to begin with:
+The BP procedure consists of three steps:
 
-1.   Apply the log-cube transformation to the age data, and
+1.   Order the data set ascending by `exit`,
 
-2.   plot the cumulative sum of the number of deaths vs. age.
+2.   apply the log-cube transformation to the age data, and
 
-The result is shown in Figure \@ref(fig:ggplot2), with the original, untransformed
-data together with the transform, stretched to 365.
+3.   plot the cumulative sum of the number of deaths vs. age.
+
+The result is shown in Figure \@ref(fig:ggplot2), together with the original
+data.  The transform data is stretced  stretched to 365 for easier comparison.
 
@@
 ## g-transformed:
-lines(g(xt) / g(365) * 365, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
-     col = "red", lwd = 2, xlab = "Age (days)",
-     main = "After g-transform", ylab = "Cum. deaths")
+lines(g(xt) / g(365) * 365, ss / nbths, 
+     col = "red", lwd = 2, xlab = "Age (days)", lty = 2)
 ##abline(h = 0)
 ##axis(2, las = 1)
-axis(3, at = g(att) / g(365) * 365, labels = att, col = "red")
+axis(3, at = g(att) / g(365) * 365, labels = att, col = "red", lty = 2)
 box()
@@
 
-The plot, disregarding the transform, is just a graph of the empirical *cumulative distribution function (CDF)*,
-that is, one minus the *Kaplan-Meier estimate* of the *survivor function*.
+The curves are the empirical *cumulative distribution function (CDF)* expressed 
+on the two scales, that is, one minus the *Kaplan-Meier [@km58] estimates* of the 
+*survivor functions*.
 
-### Post-neonatal mortality
+### Postneonatal mortality
 
@@
 investigate a close alternative, the *exponential * distribution. Modifying the *B-P* plot by using successive 
-divisions by *risk set size* instead of sample size should result in an approximately linear graph if the 
+divisions by current *risk set size* instead of original sample size should result in an approximately linear graph if the 
 exponential assumption is valid.
 
-```{r pneo, fig.cap = "Relation between cumulative hazards and CDF for transformed post-neonatal mortality data."}
+```{r pneo, fig.cap = "Relation between cumulative hazards and CDF for transformed postneonatal mortality data."}
 ginfdat <- age.window(infdat, c(28, 365))
@@
 negligible, and both are well approximated by a straight line, so we will from now on work with the 
-assumption that the transformed post-neonatal data are *exponentially distributed*.
+assumption that the transformed postneonatal data are *exponentially distributed*.
 There is also a mathematical argument motivating the similarity: The cdf for an exponentially 
@@
 So, the *hazard function* (which is the slope of the lines in Figure \@ref(fig:pneo))
-is simply estimated by the ratio of the number of deaths to the total time of exposure:
+is simply estimated by the ratio of the *number of deaths* to the *total time of exposure*
+(an *occurrence/exposure* rate):
 
@@
 Is the sum of two components, *external* and *internal* causes. According to the *B-P*
-theory, the external component is equal to the post-neonatal mortality, which contains 
+theory, the external component is equal to the postneonatal mortality, which contains 
 only an external component.
@@
 
-and then fit a piecewise constant hazards model with cutpoints at $g(0, 1, 2, 4, 7, 11, 16, 22, 28)$, see 
+and then fit a piecewise constant hazards model with cutpoints at $g(0, 2, 4, 7, 11, 16, 22, 28)$, see 
 Figure \@ref(fig:pchneo).
 
-```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 4}
+```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 3}
 neocuts <- g(c(0, 2, 4, 7, 11, 16, 22, 28))
@@
 Let us take a closer look at what happens at the end of the neonatal interval, see 
-Figure \@ref(fig:pchneo2).
+Figure \@ref(fig:pchneo2). It shows a smooth transition towards the postneonatal 
+(external) mortality.
 
-```{r pchneo2, fig.cap = "Neonatal hazard function with external component (dashed), right hand tail.", fig.height = 4}
+```{r pchneo2, fig.cap = "Neonatal hazard function with external component (dashed), right hand tail.", fig.height = 3}
 ##fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
@@
 ```
-
-A smooth transition to the post-neonatal mortality.
 
@@
 and with $h_Y$ shown in Figure \@ref(fig:together), we get $h_T$ in Figure 
-\@ref(fig:allnorm).
+\@ref(fig:allnorm) **(must be checked!)**.
 
 ```{r allnorm, fig.cap = "Hazard function, infant mortality, natural time scale."}
-y <- hazards(fitall, cum = FALSE)$x
-hy <- hazards(fitall, cum = FALSE)$y
+y <- hazards(fitall, cum = FALSE, n.points = 3000)$x
+hy <- hazards(fitall, cum = FALSE, n.points = 3000)$y
 te <- ginv(y)
@@
      ylim = c(0, 0.002)) 
+     ##xlim = c(0, 28))
 axis(1, at = ginv(pcuts))
@@
 
-## Weather data
+## Infant mortality, summary graphs
 
@@
 y <- tid$event / tid$exposure
-plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
+plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
      xlab = "Week", ylab = "IMR")
 axis(1, at = c(1, 10, 20, 30, 40, 52))
-axis(2, las = 1)
-abline(h = c(0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
+axis(2, las = 1, at = c(0, 0.03, 0.06, 0.09))
+abline(h = c(0.05, 0.07, 0.09, 0.11), lty = 3)
+abline(h = 0)
 box()
 ```
-
-In accordance with the ideas of Bourgeouis-Pichat, we group the analyses into two parts, 
-*neonatal* (first 28 days) och *postneonatal* mortality (days 29--365).
-
-## Infant data
-
-In the previous chapter we introduced data on infant mortality from the Umeå-Skellefteå
-region 1895--1950, and here we introduce covariates of interest in the forthcoming  analyses, 
-see Table
-
-### Neonatal mortality
 
@@
 y <- tneo$event / tneo$exposure
-plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
-     xlab = "Week", ylab = "IMR")
+plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
+     xlab = "Week", ylab = "NMR")
 axis(1, at = c(1, 10, 20, 30, 40, 52))
-axis(2, las = 1)
-abline(h = c(0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
+axis(2, las = 1, at = c(0, 0.2, 0.4, 0.6))
+abline(h = c(0.3, 0.4, 0.5, 0.6), lty = 3)
+abline(h = 0)
 box()
@@
 The seasonal pattern is similar to the one we found above for infant mortality.
-
-## Postneonatal mortality
 
@@
 y <- tpneo$event / tpneo$exposure
-plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(min(y), max(y)),
-     xlab = "Week", ylab = "IMR")
+plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
+     xlab = "Week", ylab = "PNMR")
 axis(1, at = c(1, 10, 20, 30, 40, 52))
-axis(2, las = 1)
-abline(h = c(0.04, 0.05, 0.06, 0.07, 0.08, 0.09), lty = 3)
+axis(2, at = c(0, 0.02, 0.04, 0.06), las = 1)
+abline(h = c(0.03, 0.04, 0.05, 0.06, 0.07), lty = 3)
+abline(h = 0)
 box()
@@
 The seasonal pattern is once again similar to the one we found for infant mortality.
+Next, the decline over the years in Figures \@ref(fig:longtrend) and \@ref(fig:ntrend).
+
+```{r longtrend, fig.cap = "Crude IMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
+trend <- toTpch(Surv(enter, exit, event) ~ year, data = infdat, cuts = c(0, 1))
+y <- trend$event / trend$exposure
+par(las = 1)
+plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
+     ylab = "IMR")
+abline(h = 0)
+```
+
+```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
+ndat <- age.window(infdat, c(0, 28 / 365))
+tnda <- toTpch(Surv(enter, exit, event) ~ year, data = ndat, cuts = c(0, 28 / 365))
+y <- tnda$event / tnda$exposure
+par(las = 1)
+plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
+     ylab = "NMR")
+abline(h = 0)
+```
 

### 2022-01-12 paper.Rmd
add final analysis section

@@
 - \usepackage{a4wide}
-abstract: "The effect of extreme temperatures on infant mortality in the Umeå region 1895-1950 is studied in a biometric analysis setting."
+- \usepackage{endfloat}
+- \renewcommand{\efloatseparator}{\mbox{}}
+abstract: "The effect of extreme temperatures on infant mortality in the Umeå region 1895-1950 is studied in a biometric analysis setting.
+More precisely, the effect of climate and weather, measured by temperature, is investigated.
+It turns out that climate is more important than weather, low average temperatures are more important than temporary dips in temperature.
+The investigated geographical region is situated too far north for bad effects of high temperatures to show."
 ---
@@
 ```
+
+# Infant mortality, climate, and weather
+
+The analyses are performed on the log-cube scale with proportional hazards modelling. Note that
+the property of proportional hazards are preserved under a strictly monotone increasing time transform.
+However, the estimates of baseline distribution characteristics will change.
+
+We start by performing separate analyses for neonatal and postneonatal mortality. Then we make comparisons.
+
+## Postneonatal mortality
+
+This is the simplest part, because *B-P* predicts that the baseline distribution (given the log-cube
+transformation) is *exponential*, that is, the hazard function is *constant*
+
+```{r postanal, results='asis'}
+infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
+pneo <- age.window(infdat, c(28 / 365, 1))
+##
+pneo$enter <- g(pneo$enter * 365)
+pneo$exit <- g(pneo$exit * 365)
+##
+pneo$emintemp <- round(pneo$emintemp)
+##
+tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + cold + emintemp + year + week, 
+                data = pneo, cuts = g(c(28, 365)))
+##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
+  ##           data = pneo, shape = 1)
+##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
+  ##           data = pneo, cuts = g(c(28, 365)))
+fit.tpch <- tpchreg(oe(event, exposure) ~ sex + socBranch + socStatus + illeg + parity + I(year - 1923) + cold + emintemp, 
+                    data = tpneo, time = age)
+dr <- drop1(fit.tpch, test = "Chisq")
+ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality", label = "tab:postanal")
+```
+
+The result in Table \@ref(tab:postanal) shows that *climate* (`emintemp`) is more important than 
+*weather* (`cold`). Moreover, no signs of interaction between weather or climate and the rest of 
+covariates (not shown).
+
+## Neonatal mortality
+
+It turns out that here we have some significant interactions, with `sex` involved.
+Maybe surprising. So we run separate analyses for the sexes.
+
+### Girls
+
+```{r neoanal, results = 'asis'}
+neo <- age.window(infdat, c(0, 28 /365))
+neo$exit <- g(neo$exit * 365)
+neo$enter <- g(neo$enter * 365)
+fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + I(year - 1923) + socBranch  + cold + emintemp,
+                 data = neo[neo$sex == "girl", ])
+dr <- drop1(fit.ph, test = "Chisq")
+ltx(fit.ph, dr = dr, caption = "Neonatal mortality, girls", label = "tab:neonatal")
+```
+
+Same here, *climate* is more important than *weather*.
+
+### Boys
+
+```{r neoanalboys, results = 'asis'}
+fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + I(year - 1923) + socBranch  + cold + emintemp,
+                 data = neo[neo$sex == "boy", ])
+dr <- drop1(fit.ph, test = "Chisq")
+ltx(fit.ph, dr = dr, caption = "Neonatal mortality, boys", label = "tab:neonatal")
+```
+
+Here, however, boys are sensitive to temperatures below average, in addition to the 
+sensitivity for cold climate.
+
+# Conclusion
+
+Remains to be written. Feels like two papers, one about Bourgeouis-Pichat and one
+about climate and weather. Maybe discuss exogeneity/endogeneity?

### 2022-01-13 paper.Rmd
endfloat, figuresonly

@@
 - \usepackage{a4wide}
-- \usepackage{endfloat}
+- \usepackage[figuresonly]{endfloat}
 - \renewcommand{\efloatseparator}{\mbox{}}

### 2022-01-14 paper.Rmd
use lag=0 (no lag)

@@
 This is the simplest part, because *B-P* predicts that the baseline distribution (given the log-cube
-transformation) is *exponential*, that is, the hazard function is *constant*
+transformation) is *exponential*, that is, the hazard function is *constant*.
 
@@
 infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
+levels(infdat$subreg) <- c("ume", "ske")
+infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
+infdat$season <- as.factor(infdat$season)
 pneo <- age.window(infdat, c(28 / 365, 1))
@@
 ##
-tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + cold + emintemp + year + week, 
+tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + 
+                    cold + emintemp + year + season + subreg, 
                 data = pneo, cuts = g(c(28, 365)))
@@
   ##           data = pneo, cuts = g(c(28, 365)))
-fit.tpch <- tpchreg(oe(event, exposure) ~ sex + socBranch + socStatus + illeg + parity + I(year - 1923) + cold + emintemp, 
+fit.tpch <- tpchreg(oe(event, exposure) ~ sex + socBranch + socStatus + illeg + parity + I(year - 1923) + 
+                        cold + emintemp + season + subreg, 
                     data = tpneo, time = age)
 dr <- drop1(fit.tpch, test = "Chisq")
-ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality", label = "tab:postanal")
+ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality.", label = "tab:postanal")
 ```
@@
 neo$enter <- g(neo$enter * 365)
-fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + I(year - 1923) + socBranch  + cold + emintemp,
+fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + I(year - 1923) + socBranch  + cold + emintemp + subreg + season,
                  data = neo[neo$sex == "girl", ])
 dr <- drop1(fit.ph, test = "Chisq")
-ltx(fit.ph, dr = dr, caption = "Neonatal mortality, girls", label = "tab:neonatal")
+ltx(fit.ph, dr = dr, caption = "Neonatal mortality, girls.", label = "tab:neonatal")
 ```
@@
 ```{r neoanalboys, results = 'asis'}
-fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + I(year - 1923) + socBranch  + cold + emintemp,
+fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + I(year - 1923) + socBranch  + cold + emintemp + subreg + season,
                  data = neo[neo$sex == "boy", ])
 dr <- drop1(fit.ph, test = "Chisq")
-ltx(fit.ph, dr = dr, caption = "Neonatal mortality, boys", label = "tab:neonatal")
+ltx(fit.ph, dr = dr, caption = "Neonatal mortality, boys.", label = "tab:neonatalboys")
+```
+
+Here, however, boys are sensitive to temperatures below average, in addition to the 
+sensitivity for cold climate.
+
+## Postneonatal mortality, winter and spring
+
+This is the simplest part, because *B-P* predicts that the baseline distribution (given the log-cube
+transformation) is *exponential*, that is, the hazard function is *constant*.
+
+```{r postanalw, results='asis'}
+##infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
+##levels(infdat$subreg) <- c("ume", "ske")
+##infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
+##infdat$season <- as.factor(infdat$season)
+pneo <- age.window(infdat[infdat$season %in% c("winter", "spring"), ], c(28 / 365, 1))
+##
+pneo$enter <- g(pneo$enter * 365)
+pneo$exit <- g(pneo$exit * 365)
+##
+pneo$emintemp <- round(pneo$emintemp)
+##
+tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + 
+                    cold + emintemp + year + subreg, 
+                data = pneo, cuts = g(c(28, 365)))
+##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
+  ##           data = pneo, shape = 1)
+##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
+  ##           data = pneo, cuts = g(c(28, 365)))
+fit.tpch <- tpchreg(oe(event, exposure) ~ sex + socBranch + socStatus + illeg + parity + I(year - 1923) + 
+                        cold + emintemp + subreg, 
+                    data = tpneo, time = age)
+dr <- drop1(fit.tpch, test = "Chisq")
+ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.", label = "tab:postanalw")
+```
+
+The result in Table \@ref(tab:postanalw) shows that *climate* (`emintemp`) is more important than 
+*weather* (`cold`). Moreover, no signs of interaction between weather or climate and the rest of 
+covariates (not shown).
+
+## Neonatal mortality, winter and spring
+
+It turns out that here we have some significant interactions, with `sex` involved.
+Maybe surprising. So we run separate analyses for the sexes.
+
+### Girls
+
+```{r neoanalw, results = 'asis'}
+neo <- age.window(infdat[infdat$season %in% c("winter", "spring"), ], c(0, 28 /365))
+neo$exit <- g(neo$exit * 365)
+neo$enter <- g(neo$enter * 365)
+fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + I(year - 1923) + socBranch  + cold + emintemp + subreg,
+                 data = neo[neo$sex == "girl", ])
+dr <- drop1(fit.ph, test = "Chisq")
+ltx(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, girls.", label = "tab:neonatalw")
+```
+
+Same here, *climate* is more important than *weather*.
+
+### Boys
+
+```{r neoanalboysw, results = 'asis'}
+fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + I(year - 1923) + socBranch  + cold + emintemp + subreg,
+                 data = neo[neo$sex == "boy", ])
+dr <- drop1(fit.ph, test = "Chisq")
+ltx(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, boys.", label = "tab:neonatalboysw")
 ```

### 2022-01-14 paper.Rmd
introduce period

--- 2022-01-13 paper.Rmd
@@
 infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
+infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951))
+levels(infdat$subreg) <- c("ume", "ske")
+infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
+infdat$season <- as.factor(infdat$season)
 pneo <- age.window(infdat, c(28 / 365, 1))
@@
 ##
-tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + cold + emintemp + year + week, 
+tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + period +
+                    cold + emintemp + season + subreg, 
                 data = pneo, cuts = g(c(28, 365)))
@@
   ##           data = pneo, cuts = g(c(28, 365)))
-fit.tpch <- tpchreg(oe(event, exposure) ~ sex + socBranch + socStatus + illeg + parity + I(year - 1923) + cold + emintemp, 
+fit.tpch <- tpchreg(oe(event, exposure) ~ sex + socBranch + socStatus + illeg + parity +
+                        period + cold + emintemp + season + subreg, 
                     data = tpneo, time = age)
 dr <- drop1(fit.tpch, test = "Chisq")
-ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality", label = "tab:postanal")
+ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality.", label = "tab:postanal")
 ```
@@
 neo$enter <- g(neo$enter * 365)
-fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + I(year - 1923) + socBranch  + cold + emintemp,
+fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg + season,
                  data = neo[neo$sex == "girl", ])
 dr <- drop1(fit.ph, test = "Chisq")
-ltx(fit.ph, dr = dr, caption = "Neonatal mortality, girls", label = "tab:neonatal")
+ltx(fit.ph, dr = dr, caption = "Neonatal mortality, girls.", label = "tab:neonatal")
 ```
@@
 ```{r neoanalboys, results = 'asis'}
-fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + I(year - 1923) + socBranch  + cold + emintemp,
+fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg + season,
                  data = neo[neo$sex == "boy", ])
 dr <- drop1(fit.ph, test = "Chisq")
-ltx(fit.ph, dr = dr, caption = "Neonatal mortality, boys", label = "tab:neonatal")
+ltx(fit.ph, dr = dr, caption = "Neonatal mortality, boys.", label = "tab:neonatalboys")
+```
+
+Here, however, boys are sensitive to temperatures below average, in addition to the 
+sensitivity for cold climate.
+
+## Postneonatal mortality, winter and spring
+
+This is the simplest part, because *B-P* predicts that the baseline distribution (given the log-cube
+transformation) is *exponential*, that is, the hazard function is *constant*.
+
+```{r postanalw, results='asis'}
+##infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
+##levels(infdat$subreg) <- c("ume", "ske")
+##infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
+##infdat$season <- as.factor(infdat$season)
+pneo <- age.window(infdat[infdat$season %in% c("winter", "spring"), ], c(28 / 365, 1))
+##
+pneo$enter <- g(pneo$enter * 365)
+pneo$exit <- g(pneo$exit * 365)
+##
+pneo$emintemp <- round(pneo$emintemp)
+##
+tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + 
+                    cold + emintemp + period + subreg, 
+                data = pneo, cuts = g(c(28, 365)))
+##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
+  ##           data = pneo, shape = 1)
+##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
+  ##           data = pneo, cuts = g(c(28, 365)))
+fit.tpch <- tpchreg(oe(event, exposure) ~ sex + socBranch + socStatus + illeg + parity + period + 
+                        cold + emintemp + subreg, 
+                    data = tpneo, time = age)
+dr <- drop1(fit.tpch, test = "Chisq")
+ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.", label = "tab:postanalw")
+```
+
+The result in Table \@ref(tab:postanalw) shows that *climate* (`emintemp`) is more important than 
+*weather* (`cold`). Moreover, no signs of interaction between weather or climate and the rest of 
+covariates (not shown).
+
+## Neonatal mortality, winter and spring
+
+It turns out that here we have some significant interactions, with `sex` involved.
+Maybe surprising. So we run separate analyses for the sexes.
+
+### Girls
+
+```{r neoanalw, results = 'asis'}
+neo <- age.window(infdat[infdat$season %in% c("winter", "spring"), ], c(0, 28 /365))
+neo$exit <- g(neo$exit * 365)
+neo$enter <- g(neo$enter * 365)
+fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg,
+                 data = neo[neo$sex == "girl", ])
+dr <- drop1(fit.ph, test = "Chisq")
+ltx(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, girls.", label = "tab:neonatalw")
+```
+
+Same here, *climate* is more important than *weather*.
+
+### Boys
+
+```{r neoanalboysw, results = 'asis'}
+fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg,
+                 data = neo[neo$sex == "boy", ])
+dr <- drop1(fit.ph, test = "Chisq")
+ltx(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, boys.", label = "tab:neonatalboysw")
 ```

### 2022-01-15 README.Rmd
first commit

---
title: Infant mortality, climate and weather
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Paper for presentation in Madrid, March 2022

The file *paper.pdf* is the result of this project.

### 2022-01-15 paper.Rmd
fixes
spell check

@@
 One important reason for studying neonatal and postneonatal mortality separately is the 
-findings by Bourgeouis-Pichat [@bp51a;  @bp51b] about endogeneous and exogeneous mortality and the log-cube transform,
+findings by Bourgeois-Pichat [@bp51a;  @bp51b] about endogenous and exogenous mortality and the log-cube transform,
 so we start by recollecting his results and formulate them in a modern statistical setting.
 
-# The Bourgeouis-Pichat model
+# The Bourgeois-Pichat model
 
-Bourgeouis-Picat (*B-P*) suggests that mortality
+Bourgeois-Pichat (*B-P*) suggests that mortality
 during the last eleven months (28 days) of infancy follows a *uniform* distribution
@@
 
-# Infant mortality, climate, and weather
+# Infant mortality, climate and weather
 
@@
 the property of proportional hazards are preserved under a strictly monotone increasing time transform.
-However, the estimates of baseline distribution characteristics will change.
+However, the estimates of baseline distribution characteristics will change, of course.
 
@@
 ##
-tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + period +
-                    cold + emintemp + season + subreg, 
+tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + 
+                    period + cold
+                + emintemp + season + subreg, 
                 data = pneo, cuts = g(c(28, 365)))

### 2022-01-16 paper.Rmd
play around ...

--- 2022-01-14 paper.Rmd
@@
 neo$enter <- g(neo$enter * 365)
-fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg + season,
+fit.ph <- coxph(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg + season,
                  data = neo[neo$sex == "girl", ])

### 2022-01-18 paper.Rmd
add time stamp

@@
 author: "Göran Broström and Tommy Bengtsson"
-date: "`r Sys.Date()`"
+date: "`r Sys.time()`"
 output:
@@
 ```{r neoanal, results = 'asis'}
+##library(survival)
 neo <- age.window(infdat, c(0, 28 /365))
@@
 neo$enter <- g(neo$enter * 365)
-fit.ph <- coxph(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg + season,
+fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg,
                  data = neo[neo$sex == "girl", ])
@@
 ```{r neoanalboys, results = 'asis'}
-fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg + season,
+fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg,
                  data = neo[neo$sex == "boy", ])

### 2022-01-19 paper.Rmd
cleaning ...

@@
 first half of the twentieth century. Two recent papers [@junkka; @karlsson] studied 
-neonatal mortality and climate variations in the same geographical area during the 
+neonatal mortality and temperature variations in a larger geographical area containing 
+the present one during the 
 years 1880--1950. Climate and mortality in general is a research area that has 
@@
 separately for neonatal  and postneonatal mortality, and for winter and summer seasons, 
-and in a joint analysis, the classification into endogenous and exogenous factors will be discussed.
+and the classification into endogenous and exogenous factors will be discussed.
 
 One important reason for studying neonatal and postneonatal mortality separately is the 
-findings by Bourgeois-Pichat [@bp51a;  @bp51b] about endogenous and exogenous mortality and the log-cube transform,
+empirical findings by Bourgeois-Pichat [@bp51a;  @bp51b] about endogenous and exogenous mortality and the log-cube transform,
 so we start by recollecting his results and formulate them in a modern statistical setting.
@@
 Bourgeois-Pichat (*B-P*) suggests that mortality
-during the last eleven months (28 days) of infancy follows a *uniform* distribution
+during the last eleven months of infancy follows a *uniform* distribution
 given a specific transformation of time (age), the 
-*log-cube transformation* $g$.
+*log-cube transformation* $g$, see equation \@ref(eq:g),
 
@@
 Furthermore, it is assumed that the *endogeneous* component is non-zero only during the first
-month of life.
+28 days of life.
 
@@
 The result is shown in Figure \@ref(fig:ggplot2), together with the original
-data.  The transform data is stretced  stretched to 365 for easier comparison.
+data.  The transform data is stretched  stretched to 365 for easier comparison.
 
@@
 
-We create the neonatal data set by censor all lives at the age 28:
+We create the neonatal data set by censor all lives at the age 28 days:
 
@@
 
-and then fit a piecewise constant hazards model with cutpoints at $g(0, 2, 4, 7, 11, 16, 22, 28)$, see 
+```{r defcuts}
+neocuts <- c(0, 2, 5, 8, 12, 17, 22, 25, 28)
+pcuts <- c(neocuts, 365)
+gneocuts <- g(neocuts)
+gpcuts <- g(pcuts)
+```
+
+and then fit a piecewise constant hazards model with cutpoints at `r gneocuts`, see 
 Figure \@ref(fig:pchneo).
@@
 ```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 3}
-neocuts <- g(c(0, 2, 4, 7, 11, 16, 22, 28))
-fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
-plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", axes = FALSE)
-axis(1)
+fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = gneocuts)
+plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "Age (days)", axes = FALSE, lwd = 1.5)
+axis(1, at = gneocuts, labels = neocuts)
 axis(2, las = 1)
@@
 par(las = 1)
-plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "g(days)", xlim = c(5, 39))
+plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "Age (days)", lwd = 1.5,
+     xlim = g(c(5, 28)), axes = FALSE)
+axis(1, at = gneocuts, labels = neocuts)
+axis(2)
+box()
 abline(h = 0, v = 0)
-abline(h = posthaz, col = "red", lty = 2)
+abline(h = posthaz, col = "red", lty = 2, lwd = 1.5)
 ```
@@
 
-```{r together, fig.cap = "Hazard function, infant mortality, transformed time."}
-allcuts <- c(neocuts, g(365))
+```{r together, fig.cap = "Hazard function, infant mortality, transformed time.", fig.height=6}
 ginfdat <- infdat
 ginfdat$exit <- g(ginfdat$exit)
-fitall <- pchreg(Surv(enter, exit, event) ~ 1, data = ginfdat, cuts = allcuts)
-plot(fitall, fn = "haz", col = "blue", main = "", xlab = "g(Days)", axes = FALSE,
-     ylim = c(0, 0.002))
-pcuts <- g(c(0, 28, 183, 365))
-axis(1, at = pcuts, labels = round(pcuts))
+fitall <- pchreg(Surv(enter, exit, event) ~ 1, data = ginfdat, cuts = gpcuts)
+plot(fitall, fn = "haz", col = "blue", main = "", xlab = "Age (days)", axes = FALSE,
+     lwd = 1.5)
+
+axis(1, at = gpcuts, labels = pcuts)
 axis(2, las = 1)
@@
 infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
-infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951))
+infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
 levels(infdat$subreg) <- c("ume", "ske")

### 2022-01-20 paper.Rmd
done B-P modelling

@@
 
+```{r excum, echo = FALSE}
+fitex <- pchreg(Surv(enter, exit, event) ~ 1, cuts = g(c(28, 365)), data = ginfdat)
+```
+
 ### Neonatal mortality
 
-Is the sum of two components, *external* and *internal* causes. According to the *B-P*
+Neonatal mortality is, according to Bourgeois-Pichat, the sum of two components, *external* and *internal* causes. According to the *B-P*
 theory, the external component is equal to the postneonatal mortality, which contains 
@@
 
-and then fit a piecewise constant hazards model with cutpoints at `r gneocuts`, see 
+and then fit a piecewise constant hazards model with cutpoints at g(`r neocuts`), see 
 Figure \@ref(fig:pchneo).
@@
 
-### Putting all together
+### Putting it all together
 
 Looking at the full infant mortality interval on the $g$-scale, we have a piecewise 
-constant hazards model with cutpoints at $g(0, 1, 2, 4, 7, 11, 16, 22, 28)$, see 
-Figure \@ref(fig:together) (where the value close to zero is truncated).
+constant hazards model with cutpoints at $g(0, 2, 5, 8, 12, 17, 22, 25, 28, 365)$, see 
+Figure \@ref(fig:together).
 
@@
 
+<!--
 Now we would like to see the content of Figure \@ref(fig:together) on the natural time 
@@
 
-```{r allnorm, fig.cap = "Hazard function, infant mortality, natural time scale."}
+```{r allnorm, fig.cap = "Hazard function, infant mortality, natural time scale.", eval = FALSE}
 y <- hazards(fitall, cum = FALSE, n.points = 3000)$x
@@
 plot(te, hte, type = "s", col = "blue", xlab = "Days", ylab = "", axes = FALSE,
-     ylim = c(0, 0.002)) 
-     ##xlim = c(0, 28))
-axis(1, at = ginv(pcuts))
+     ##ylim = c(0, 0.002)) 
+     xlim = c(0, 28))
+axis(1, at = pcuts)
 axis(2, las = 1)
@@
 
-# Data
+-->
 
-## Infant mortality, summary graphs
+The way to disentangle the endogenous and exogenous componets is to start with a plot 
+of the cumulative hazard function (Figure \@ref(fig:separ)).
+
+```{r separ, fig.cap = "Separation of exogenous and endogenous infant mortality (cumulative hazards, per cent).", fig.width = 5, fig.height = 4}
+plot(fitall, fn = "cum", col = "blue", main = "", xlab = "Age (days)", axes = FALSE,
+     lwd = 1.5, ylab = "Cumulative hazards")
+
+axis(1, at = c(0, g(28), g(365)), labels = c(0, 28, 365), pos = 0)
+axis(2, las = 1, pos = 0, at = c(0, 0.02, 0.04, 0.06), labels = c(0, 2, 4, 6))
+#box()
+abline(v = 0)
+texhaz <- g(365) * fitex$hazards[1, 1]
+lines(c(0, g(365)), c(0, texhaz), col = "red", lwd = 1.5)
+atte <- c(0.0000, 0.0763, texhaz)
+axis(4, at = atte, labels = 100 * round(atte, 4), las = 1, cex = 0.5,
+     pos = g(365))
+abline(v = g(365))
+text(100, 0.03, "Exogenous", col = "red", srt = 22, cex = 0.8)
+text(100, 0.053, "Endogenous plus exogenous", col = "blue", srt = 22, cex = 0.8)
+axis(3, at =  c(0, g(28), g(365)), labels = c(0, 28, 365))
+```
+
+So we can see from Figure \@ref(fig:separ) that of the total mortality, 
+`r round(5.27 / 7.63 * 100, 1)` per cent is exogenous and the rest is endogenous.
+AS always, this requires that we trust the Bourgeois-Pichat model.
+
+# Infant mortality, climate and weather
+
+## Data
+
+### Infant mortality, summary graphs
 
@@
 
-```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950."}
+```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 4}
 infdat <- readRDS("data/infdat3_year3.rds")
@@
 plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
-     xlab = "Week", ylab = "IMR")
+     xlab = "Week", ylab = "IMR", cex = 0.7, lwd = 1.5)
 axis(1, at = c(1, 10, 20, 30, 40, 52))
@@
 
-```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950."}
+```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 4}
 neo <- age.window(infdat, c(0, 28 / 365))
@@
 plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
-     xlab = "Week", ylab = "NMR")
+     xlab = "Week", ylab = "NMR", cex = 0.7, lwd = 1.5)
 axis(1, at = c(1, 10, 20, 30, 40, 52))
@@
 
-```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950."}
+```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 4}
 pneo <- age.window(infdat, c(28 / 365, 1))
@@
 plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
-     xlab = "Week", ylab = "PNMR")
+     xlab = "Week", ylab = "PNMR", cex = 0.7, lwd = 1.5)
 axis(1, at = c(1, 10, 20, 30, 40, 52))
@@
 plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
-     ylab = "IMR")
+     ylab = "IMR", lwd = 1.5, cex = 0.7)
+abline(h = c(0.05, 0.1, 0.15), lty = 3)
 abline(h = 0)
@@
 plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
-     ylab = "NMR")
+     ylab = "NMR", lwd = 1.5, cex = 0.7)
 abline(h = 0)
+abline(h = c(0.2, 0.4, 0.6), lty = 3)
 ```
 
-## Temperature
+### Temperature
 
@@
 
-```{r monavgtemp, fig.cap="Weekly max and min temperature averages, 1895--1950."}
-oldpar <- par(mfrow= c(1, 2))
+```{r monavgtemp, fig.cap="Weekly max and min temperature averages, 1895--1950.", fig.height = 4}
+oldpar <- par(mfrow = c(1, 2))
 ## Umeå:
@@
 plot(1:52, umaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
-     xlab = "Week", ylab = "Temperature (C)")
+     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
 text(17, 27, "Umeå")
@@
 abline(h = c(-20, -10, 10, 20), lty = 3)
-lines(1:52, uminavg, type = "b", col = "blue")
+lines(1:52, uminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
 text(8, 11, "max", col = "red")
@@
 plot(1:52, bmaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
-     xlab = "Week", ylab = "Temperature (C)")
+     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
 text(17, 27, "Skellefteå")
@@
 abline(h = c(-20, -10, 10, 20), lty = 3)
-lines(1:52, bminavg, type = "b", col = "blue")
+lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
 text(8, 11, "max", col = "red")
@@
 
-# Infant mortality, climate and weather
+Time trends of yearly average temperatures, see Figure
+
+```{r longterm}
+bmtemp <- rbind(umetemp, bjurtemp)
+umt <- with(bmtemp, tapply(meantemp, year, mean))
+plot(1894:1951, umt, type = "b", col = "blue", xlab = "Year",
+     ylab = "Mean temperature", lwd = 1.5, cex = 0.7, ylim = c(0, 7))
+abline(h = 0)
+abline(h = 1:6, lty = 3)
+```
+
+## Statistical modelling
 

### 2022-01-21 paper.Rmd
done B-P part

@@
 
-The BP procedure consists of three steps:
+The B-P procedure consists of three steps:
 
@@
 
-```{r ggplot2, fig.cap = "BB plots with and without time transform, Umeå-Skellefteå 1895-1950."}
+```{r ggplot2, fig.cap = "B-B plots with and without time transform, Umeå-Skellefteå 1895-1950.", fig.height = 4}
 nbths <- NROW(infdat)
@@
 lines(x, 1 - y, col = "red", lty = 2)
-legend("topleft", legend = c("Cumulative hazards", "CDF (Bourgeouis-Pichat)"), col = c("blue", "red"),
-       lty = c(1, 2))
+legend("topleft", legend = c("Cumulative hazards", "CDF (Bourgeois-Pichat)"), col = c("blue", "red"),
+       lty = c(1, 2), cex = 0.7)
 ```
@@
 
-The way to disentangle the endogenous and exogenous componets is to start with a plot 
-of the cumulative hazard function (Figure \@ref(fig:separ)).
+The way to disentangle the endogenous and exogenous components is to start with a plot 
+of the *cumulative* hazard function and add the exogenous 
+part from the postneonatal estimate, see Figure \@ref(fig:separ).
 
@@
 axis(1, at = c(0, g(28), g(365)), labels = c(0, 28, 365), pos = 0)
-axis(2, las = 1, pos = 0, at = c(0, 0.02, 0.04, 0.06), labels = c(0, 2, 4, 6))
+axis(2, las = 1, pos = 0, at = c(0, 0.0096, 0.0338), labels = c(0, 0.96, 3.38))
 #box()
@@
 axis(3, at =  c(0, g(28), g(365)), labels = c(0, 28, 365))
+lines(c(0, g(28)), c(0.0096, 0.0096), lty = 3)
+lines(c(0, g(28)), c(0.0338, 0.0338), lty = 3)
+abline(v = g(28))
 ```
 
-So we can see from Figure \@ref(fig:separ) that of the total mortality, 
-`r round(5.27 / 7.63 * 100, 1)` per cent is exogenous and the rest is endogenous.
-AS always, this requires that we trust the Bourgeois-Pichat model.
+So we can see from Figure \@ref(fig:separ) (vertical axis on the right hand side) that of the total mortality, 
+`r round(5.27 / 7.63 * 100)` per cent is exogenous and the rest is endogenous.
+Similarly, of the neonatal mortality, `r round(0.96 / 3.38 * 100)` per cent is
+exogenous and the rest is endogenous (vertical axis on the left hand side).
+As always, this requires that we trust the Bourgeois-Pichat model.
 
@@
 
-Remains to be written. Feels like two papers, one about Bourgeouis-Pichat and one
+Remains to be written. Feels like two papers, one about Bourgeois-Pichat and one
 about climate and weather. Maybe discuss exogeneity/endogeneity?

### 2022-01-23 paper.Rmd
truncate cold in postneonatal

@@
 
-```{r postanal, results='asis'}
+```{r postanalske, results='asis'}
 infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
@@
 ##
-tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + 
-                    period + cold
+tpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + socStatus + illeg + parity + 
+                    cold 
                 + emintemp + season + subreg, 
@@
   ##           data = pneo, cuts = g(c(28, 365)))
-fit.tpch <- tpchreg(oe(event, exposure) ~ sex + socBranch + socStatus + illeg + parity +
-                        period + cold + emintemp + season + subreg, 
-                    data = tpneo, time = age)
+tpneo$icold <- with(tpneo, cold * (cold < 0))
+fit.tpch <- tpchreg(oe(event, exposure) ~ icold + emintemp + season +  period + sex + socBranch + socStatus  + parity, 
+                    data = tpneo[tpneo$subreg == "ske", ], time = age)
 dr <- drop1(fit.tpch, test = "Chisq")
-ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality.", label = "tab:postanal")
+ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality, Skellefteå.", label = "tab:postanalske")
 ```
@@
 covariates (not shown).
+
+```{r postanalume, results = 'asis'}
+fit.tpch <- tpchreg(oe(event, exposure) ~ icold + period + sex + socBranch + socStatus + period  + parity +
+                        season, 
+                    data = tpneo[tpneo$subreg == "ume", ], time = age)
+dr <- drop1(fit.tpch, test = "Chisq")
+ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality, Umeå.", label = "tab:postanalume")
+```
 

### 2022-01-23 paper.Rmd
introduce small tables.

--- 2022-01-21 paper.Rmd
@@
 
-```{r postanal, results='asis'}
+```{r postanalske, results='asis'}
+source("R/ltx2.R")
 infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
@@
   ##           data = pneo, cuts = g(c(28, 365)))
-fit.tpch <- tpchreg(oe(event, exposure) ~ sex + socBranch + socStatus + illeg + parity +
-                        period + cold + emintemp + season + subreg, 
-                    data = tpneo, time = age)
+tpneo$excessCold <- with(tpneo, -cold * (cold < 0))
+fit.tpch <- tpchreg(oe(event, exposure) ~ excessCold + emintemp  +  period + sex + socBranch + socStatus  + 
+                        parity + season, 
+                    data = tpneo[tpneo$subreg == "ske", ], time = age)
 dr <- drop1(fit.tpch, test = "Chisq")
-ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality.", label = "tab:postanal")
+ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, Skellefteå.
+Adjusted for social status and branch, illegitimacy, parity, time period, season, and sub region.", label = "tab:postanalske", keep = 2)
 ```
 
-The result in Table \@ref(tab:postanal) shows that *climate* (`emintemp`) is more important than 
+The result in Table \@ref(tab:postanalske) shows that *climate* (`emintemp`) is more important than 
 *weather* (`cold`). Moreover, no signs of interaction between weather or climate and the rest of 
 covariates (not shown).
+
+```{r postanalume, results = 'asis'}
+fit.tpch <- tpchreg(oe(event, exposure) ~ excessCold + emintemp + period + sex + socBranch + socStatus + 
+                        parity + season, 
+                    data = tpneo[tpneo$subreg == "ume", ], time = age)
+dr <- drop1(fit.tpch, test = "Chisq")
+ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, Umeå. Adjusted for social status and branch, illegitimacy, parity, time period, season, and subregion.", label = "tab:postanalume", keep = 2)
+```
 
@@
 neo$enter <- g(neo$enter * 365)
-fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg,
+fit.ph <- coxreg(Surv(enter, exit, event) ~  cold + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
                  data = neo[neo$sex == "girl", ])
 dr <- drop1(fit.ph, test = "Chisq")
-ltx(fit.ph, dr = dr, caption = "Neonatal mortality, girls.", label = "tab:neonatal")
+ltx2(fit.ph, dr = dr, caption = "Neonatal mortality, girls. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatal", keep = 2)
 ```
@@
 ```{r neoanalboys, results = 'asis'}
-fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg,
+fit.ph <- coxreg(Surv(enter, exit, event) ~  cold + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
                  data = neo[neo$sex == "boy", ])
 dr <- drop1(fit.ph, test = "Chisq")
-ltx(fit.ph, dr = dr, caption = "Neonatal mortality, boys.", label = "tab:neonatalboys")
+ltx2(fit.ph, dr = dr, caption = "Neonatal mortality, boys. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatalboys", keep = 2)
 ```
@@
   ##           data = pneo, cuts = g(c(28, 365)))
-fit.tpch <- tpchreg(oe(event, exposure) ~ sex + socBranch + socStatus + illeg + parity + period + 
-                        cold + emintemp + subreg, 
+fit.tpch <- tpchreg(oe(event, exposure) ~ cold + emintemp + sex + socBranch + socStatus + illeg + parity + period + 
+                        subreg, 
                     data = tpneo, time = age)
 dr <- drop1(fit.tpch, test = "Chisq")
-ltx(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.", label = "tab:postanalw")
+ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", 
+     label = "tab:postanalw", keep = 2)
 ```
@@
 neo$enter <- g(neo$enter * 365)
-fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg,
+fit.ph <- coxreg(Surv(enter, exit, event) ~  cold + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
                  data = neo[neo$sex == "girl", ])
 dr <- drop1(fit.ph, test = "Chisq")
-ltx(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, girls.", label = "tab:neonatalw")
+ltx2(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, girls.
+Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatalw", keep = 2)
 ```
@@
 ```{r neoanalboysw, results = 'asis'}
-fit.ph <- coxreg(Surv(enter, exit, event) ~  socStatus + illeg + parity + period + socBranch  + cold + emintemp + subreg,
+fit.ph <- coxreg(Surv(enter, exit, event) ~  cold + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
                  data = neo[neo$sex == "boy", ])
 dr <- drop1(fit.ph, test = "Chisq")
-ltx(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, boys.", label = "tab:neonatalboysw")
+ltx2(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, boys. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", 
+     label = "tab:neonatalboysw", keep = 2)
 ```

### 2022-01-24 R/tbl.R
add parameter 'linesep'

@@ -1,7 +1,8 @@
-tbl <- function(tt, caption = "", fs = 11, rownames = FALSE, align = NULL){
+tbl <- function(tt, caption = "", fs = 11, rownames = FALSE, align = NULL,
+                linesep = c("", "", "", "", "\\addlinespace")){
     library(kableExtra)
     kbl(tt, caption = caption, booktabs = TRUE, row.names = rownames, 
-        format = "latex", align = align) %>%
+        format = "latex", align = align, linesep = linesep) %>%
         row_spec(row = 0, bold = TRUE) %>%
         kable_styling(full_width = FALSE, font_size = fs,
                     position = "center")

### 2022-01-24 paper.Rmd
major remake, B-P out

@@
 abstract: "The effect of extreme temperatures on infant mortality in the Umeå region 1895-1950 is studied in a biometric analysis setting.
-More precisely, the effect of climate and weather, measured by temperature, is investigated.
+More precisely, the effect of climate and weather, measured by temperature, on infant mortality is investigated.
 It turns out that climate is more important than weather, low average temperatures are more important than temporary dips in temperature.
@@
 One important reason for studying neonatal and postneonatal mortality separately is the 
-empirical findings by Bourgeois-Pichat [@bp51a;  @bp51b] about endogenous and exogenous mortality and the log-cube transform,
-so we start by recollecting his results and formulate them in a modern statistical setting.
+empirical findings by Bourgeois-Pichat [@bp51a;  @bp51b] about endogenous and exogenous 
+mortality and the log-cube transform.
 
-# The Bourgeois-Pichat model
+# Data
 
-Bourgeois-Pichat (*B-P*) suggests that mortality
-during the last eleven months of infancy follows a *uniform* distribution
-given a specific transformation of time (age), the 
-*log-cube transformation* $g$, see equation \@ref(eq:g),
+We have two sources of data which we combine into one data set suitable for our purpose.
+The first is demographic data obtained from the 
+*Centre for Demographic and Ageing Research* (CEDAR, https://cedar.umu.se),
+the second is daily temperature measurements obtained from the 
+*Swedish Meteorological and Hydrological Institute* (SMHI, https://www.smhi.se).
 
-\begin{equation*}\label{eq:g}
-g(t) = \log^3(1 + t), \quad 0 \le t \le 365,
-\end{equation*}
+## Infant mortality
 
-where $t$ is age measured in *days*. 
-Its graph is shown in Figure \@ref(fig:figgraph).
+Individual data with all births between 1 January 1895 and 31 December 1950 in two 
+coastal areas of north Sweden, Skellefteå (51560 births) and Umeå (31213 births).
+They were followed until death or age 365 days, whichever came first. 
+The following *static* characteristics were observed on each child:
 
-```{r figgraph, fig.cap = "The log-cube transform g.", fig.height = 4}
-##g <- function(x) (log(1 + x))^3  # Note: days!
-x <- seq(0, 365, length = 1000)
-y <- g(x)
-plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, 
-     col = "blue")
-xat <- c(0, 28, 183, 365)
-yat <- g(xat)
-axis(1, at = xat)
-axis(2, at = yat, labels = round(yat, 1), las = 1)
-box()
-lines(c(28, 28), c(0, g(28)), lty = 2, col = "darkgreen")
-lines(c(0, 28), c(g(28), g(28)), lty = 2, col = "darkgreen")
-##
-lines(c(183, 183), c(0, g(183)), lty = 2, col = "darkgreen")
-lines(c(0, 183), c(g(183), g(183)), lty = 2, col = "darkgreen")
-##
-lines(c(365, 365), c(0, g(365)), lty = 2, col = "darkgreen")
-lines(c(0, 365), c(g(365), g(365)), lty = 2, col = "darkgreen")
-abline(h = 0, v = 0)
-##abline(a = 0, b = 1, col = "red")
-```
+**birthdate** Date of birth.
 
-The *B-P* model specifies that, on the transformed age scale, the *exogeneous*
-mortality distribution, conditional on death before age 365 days,  is *uniform*, that is, the 
-survivor function of $T$ is *linear*,
+**sex** Girl or boy.
 
-\begin{equation}
-S(t) = 1 - \alpha t, \, 0 \le t \le g(365).
-\end{equation}
+**exit** Number of days under observation.
 
-Furthermore, it is assumed that the *endogeneous* component is non-zero only during the first
-28 days of life.
+**event** Logical, *TRUE* if a death is observed.
 
-Next we apply the model to our infant mortality data from the Umeå-Skellefteå region.
+**socBranch** Working branch of father (if any).
 
-## Infant mortality, a first look
+**socStatus** Social status of family, based on HISCLASS.
 
-```{r readdata}
-infdat <- readRDS("data/infdat3.rds")
-```
+**illeg** Mother unmarried?
 
-From the Umeå-Skellefteå region, we follow all persons born in the 
-region 1895--1950 until death, reaching the age of 1 year, or leaving follow-up for whatever 
-reason, whichever comes first. The first few lines (infants) are shown in Table 
-\@ref(tab:infdat3). Note that covariates are left out for the time being.
+**parity** Order among siblings.
 
-If `event == TRUE`, then the infant died at age (in days) given by `exit`, otherwise
-the infant survived (at least) 365 days and follow-up is lost at `exit == 365`. 
-So, in Table \@ref(tab:infdat3), infants No. 1--4 survived, 
-while infant No. 5 died at age 162 days. We assume here that all infants are followed from birth,
-that is, `enter == 0` in all cases (no left truncation).
-
-```{r infdat3}
-source("R/tbl.R")
-sinf <- infdat[, c("enter", "exit", "event")]
-x <- head(sinf, 4)
-y <- sinf[sinf$event == 1, ][1, ]
-z <- rbind(x, y)
-z$id <- 1:5
-z <- z[, c("id", "enter", "exit", "event")]
-tbl(z, caption = "A few infants in the data set.")
-```
-
-### The Bourgeois-Pichat procedure
-
-The B-P procedure consists of three steps:
-
-1.   Order the data set ascending by `exit`,
-
-2.   apply the log-cube transformation to the age data, and
-
-3.   plot the cumulative sum of the number of deaths vs. age.
-
-The result is shown in Figure \@ref(fig:ggplot2), together with the original
-data.  The transform data is stretched  stretched to 365 for easier comparison.
-
-```{r ggplot2, fig.cap = "B-B plots with and without time transform, Umeå-Skellefteå 1895-1950.", fig.height = 4}
-nbths <- NROW(infdat)
-##oldpar <- par(mfrow = c(2, 1))
-sur <- with(infdat, Surv(enter, exit, event))
-rs <- risksets(sur)
-ss <- cumsum(rs$n.events)
-## Natural:
-xt <- rs$risktimes
-att <- c(0, 28, 183, 365)
-plot(xt, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
-     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
-     ylab = "Cum. deaths")
-abline(h = 0)
-axis(2, las = 1)
-axis(1, at = att, labels = att)
-box()
-## g-transformed:
-lines(g(xt) / g(365) * 365, ss / nbths, 
-     col = "red", lwd = 2, xlab = "Age (days)", lty = 2)
-##abline(h = 0)
-##axis(2, las = 1)
-axis(3, at = g(att) / g(365) * 365, labels = att, col = "red", lty = 2)
-box()
-abline(v = 0)
-abline(v = g(c(28, 183, 365)) / g(365) * 365, col = "red", lty = 3)
-abline(v = c(28, 183, 365), col = "blue", lty = 3)
-##
-legend("bottomright", legend = c("Natural", "g-transform"), col = c("blue", "red"), lty = 1)
-abline(h = (ss / nbths) [28], lty = 3 )
-abline(h = (ss / nbths) [180], lty = 3 )
-##par(oldpar)
-```
-
-The curves are the empirical *cumulative distribution function (CDF)* expressed 
-on the two scales, that is, one minus the *Kaplan-Meier [@km58] estimates* of the 
-*survivor functions*.
-
-### Postneonatal mortality
-
-We begin by studying mortality after age 28 days on the $g$-scale, *B-P* predicts that it is approximately *uniform*. 
-The uniform distribution, however, is not very useful in a survival analysis context, so we will
-investigate a close alternative, the *exponential * distribution. Modifying the *B-P* plot by using successive 
-divisions by current *risk set size* instead of original sample size should result in an approximately linear graph if the 
-exponential assumption is valid.
-
-```{r pneo, fig.cap = "Relation between cumulative hazards and CDF for transformed postneonatal mortality data."}
-ginfdat <- age.window(infdat, c(28, 365))
-ginfdat$enter <- g(ginfdat$enter)
-ginfdat$exit <- g(ginfdat$exit)
-grs <- risksets(with(ginfdat, Surv(enter, exit, event)), members = FALSE)
-fit <- coxreg(Surv(enter, exit, event) ~ 1, data = ginfdat)
-ser <- plot(fit, fn = "sur", fig = FALSE)
-##plot(fit, col = "blue", axes = FALSE)
-cum <- fit$hazards[[1]]
-x <- cum[, 1]
-ch <- cumsum(cum[, 2])
-plot(x, ch, type = "s", axes = FALSE, col = "blue", xlab = "Age (days)", ylab = "")
-axis(1, at = c(g(28), g(183), g(365)), labels = c(28, 183, 365))
-axis(2, las = 1)
-abline(h = 0)
-box()
-x <- ser[[1]][[1]][, 1]
-y <- ser[[1]][[1]][, 2]
-lines(x, 1 - y, col = "red", lty = 2)
-legend("topleft", legend = c("Cumulative hazards", "CDF (Bourgeois-Pichat)"), col = c("blue", "red"),
-       lty = c(1, 2), cex = 0.7)
-```
-
-As is evident from Figure \@ref(fig:pneo), the difference between the the two curves is
-negligible, and both are well approximated by a straight line, so we will from now on work with the 
-assumption that the transformed postneonatal data are *exponentially distributed*.
-There is also a mathematical argument motivating the similarity: The cdf for an exponentially 
-distributed random variable $T$ is
-
-\begin{equation*}
-F_T(t) = 1 - \exp(-\lambda t), \quad t, \lambda > 0, 
-\end{equation*}
-
-and the first order approximation of this expression is
-
-\begin{equation*}
-F_T(t) \approx \lambda t, \quad t, \lambda > 0,
-\end{equation*}
-
-which is the *cumulative hazards* function of $T$,
-and the approximation is good for small values of $\lambda t$.
-
-So, the *hazard function* (which is the slope of the lines in Figure \@ref(fig:pneo))
-is simply estimated by the ratio of the *number of deaths* to the *total time of exposure*
-(an *occurrence/exposure* rate):
-
-```{r simpest, echo = TRUE}
-posthaz <- with(ginfdat, sum(event) / sum(exit - enter))
-```
-
-which evaluates to `r format(posthaz, digits = 3, scientific = TRUE)`. A small number, but remember
-that the time unit is defined on the scale defined by the function $g$ (close to *day*).
-
-```{r excum, echo = FALSE}
-fitex <- pchreg(Surv(enter, exit, event) ~ 1, cuts = g(c(28, 365)), data = ginfdat)
-```
-
-### Neonatal mortality
-
-Neonatal mortality is, according to Bourgeois-Pichat, the sum of two components, *external* and *internal* causes. According to the *B-P*
-theory, the external component is equal to the postneonatal mortality, which contains 
-only an external component.
-
-We create the neonatal data set by censor all lives at the age 28 days:
-
-```{r neon, echo = TRUE}
-neo <- age.window(infdat, c(0, 28))
-neo$enter <- g(neo$enter)
-neo$exit <- g(neo$exit)
-```
-
-```{r defcuts}
-neocuts <- c(0, 2, 5, 8, 12, 17, 22, 25, 28)
-pcuts <- c(neocuts, 365)
-gneocuts <- g(neocuts)
-gpcuts <- g(pcuts)
-```
-
-and then fit a piecewise constant hazards model with cutpoints at g(`r neocuts`), see 
-Figure \@ref(fig:pchneo).
-
-```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 3}
-fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = gneocuts)
-plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "Age (days)", axes = FALSE, lwd = 1.5)
-axis(1, at = gneocuts, labels = neocuts)
-axis(2, las = 1)
-box()
-abline(h = 0, v = 0)
-abline(h = posthaz, col = "red", lty = 2)
-```
-
-Let us take a closer look at what happens at the end of the neonatal interval, see 
-Figure \@ref(fig:pchneo2). It shows a smooth transition towards the postneonatal 
-(external) mortality.
-
-```{r pchneo2, fig.cap = "Neonatal hazard function with external component (dashed), right hand tail.", fig.height = 3}
-##fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
-par(las = 1)
-plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "Age (days)", lwd = 1.5,
-     xlim = g(c(5, 28)), axes = FALSE)
-axis(1, at = gneocuts, labels = neocuts)
-axis(2)
-box()
-abline(h = 0, v = 0)
-abline(h = posthaz, col = "red", lty = 2, lwd = 1.5)
-```
-
-### Putting it all together
-
-Looking at the full infant mortality interval on the $g$-scale, we have a piecewise 
-constant hazards model with cutpoints at $g(0, 2, 5, 8, 12, 17, 22, 25, 28, 365)$, see 
-Figure \@ref(fig:together).
-
-```{r together, fig.cap = "Hazard function, infant mortality, transformed time.", fig.height=6}
-ginfdat <- infdat
-ginfdat$exit <- g(ginfdat$exit)
-fitall <- pchreg(Surv(enter, exit, event) ~ 1, data = ginfdat, cuts = gpcuts)
-plot(fitall, fn = "haz", col = "blue", main = "", xlab = "Age (days)", axes = FALSE,
-     lwd = 1.5)
-
-axis(1, at = gpcuts, labels = pcuts)
-axis(2, las = 1)
-box()
-abline(h = 0, v = 0)
-```
-
-<!--
-Now we would like to see the content of Figure \@ref(fig:together) on the natural time 
-scale. If $Y = g(T)$, we have the relation between their respective hazard functions
-
-\begin{equation*}
-h_T(t) = \frac{d}{dt} g(t) h_Y\bigl(g^{-1}(t)\bigr), \quad 0 \le t \le 365,
-\end{equation*}
-
-and with $h_Y$ shown in Figure \@ref(fig:together), we get $h_T$ in Figure 
-\@ref(fig:allnorm) **(must be checked!)**.
-
-```{r allnorm, fig.cap = "Hazard function, infant mortality, natural time scale.", eval = FALSE}
-y <- hazards(fitall, cum = FALSE, n.points = 3000)$x
-hy <- hazards(fitall, cum = FALSE, n.points = 3000)$y
-te <- ginv(y)
-hte <- dg(te) * hy
-plot(te, hte, type = "s", col = "blue", xlab = "Days", ylab = "", axes = FALSE,
-     ##ylim = c(0, 0.002)) 
-     xlim = c(0, 28))
-axis(1, at = pcuts)
-axis(2, las = 1)
-box()
-abline(h = 0, v = 0)
-```
-
--->
-
-The way to disentangle the endogenous and exogenous components is to start with a plot 
-of the *cumulative* hazard function and add the exogenous 
-part from the postneonatal estimate, see Figure \@ref(fig:separ).
-
-```{r separ, fig.cap = "Separation of exogenous and endogenous infant mortality (cumulative hazards, per cent).", fig.width = 5, fig.height = 4}
-plot(fitall, fn = "cum", col = "blue", main = "", xlab = "Age (days)", axes = FALSE,
-     lwd = 1.5, ylab = "Cumulative hazards")
-
-axis(1, at = c(0, g(28), g(365)), labels = c(0, 28, 365), pos = 0)
-axis(2, las = 1, pos = 0, at = c(0, 0.0096, 0.0338), labels = c(0, 0.96, 3.38))
-#box()
-abline(v = 0)
-texhaz <- g(365) * fitex$hazards[1, 1]
-lines(c(0, g(365)), c(0, texhaz), col = "red", lwd = 1.5)
-atte <- c(0.0000, 0.0763, texhaz)
-axis(4, at = atte, labels = 100 * round(atte, 4), las = 1, cex = 0.5,
-     pos = g(365))
-abline(v = g(365))
-text(100, 0.03, "Exogenous", col = "red", srt = 22, cex = 0.8)
-text(100, 0.053, "Endogenous plus exogenous", col = "blue", srt = 22, cex = 0.8)
-axis(3, at =  c(0, g(28), g(365)), labels = c(0, 28, 365))
-lines(c(0, g(28)), c(0.0096, 0.0096), lty = 3)
-lines(c(0, g(28)), c(0.0338, 0.0338), lty = 3)
-abline(v = g(28))
-```
-
-So we can see from Figure \@ref(fig:separ) (vertical axis on the right hand side) that of the total mortality, 
-`r round(5.27 / 7.63 * 100)` per cent is exogenous and the rest is endogenous.
-Similarly, of the neonatal mortality, `r round(0.96 / 3.38 * 100)` per cent is
-exogenous and the rest is endogenous (vertical axis on the left hand side).
-As always, this requires that we trust the Bourgeois-Pichat model.
-
-# Infant mortality, climate and weather
-
-## Data
-
-### Infant mortality, summary graphs
+Some crude statistics about infant, neonatal, and postneonatal mortality are shown in 
+Figures.
 
@@
 
-### Temperature
+## Temperature
 
@@
 daily temperature data covering our time period, usually three measures per day, morning,
-noon, and evening. Weekly averages are calculated year by year, and deviations from 
+noon, and evening. In Table \@ref(tab:typweek), the Umeå data from the week 1--7 January,
+1923 is shown.
+
+```{r typweek}
+tum <- read.table("~/Forskning/Data/ume_temp.csv", skip = 10, sep = ";", header = TRUE)
+source("R/tbl.R")
+ww <- tum[70137:(70137 + 20), ]
+names(ww) <- c("Date", "Time", "Temperature", "Quality")
+tbl(ww, caption = "Raw temperature data from first week of 1923, Umeå.", fs = 10,
+    linesep = c("", "", "\\addlinespace"))
+```
+
+There are three measurements per day, or 21 per week. In the forthcoming analyses,
+the weekly data are summarized in a few measurements, see Table \@ref(tab:cond).
+
+```{r cond}
+sh <- umetemp[umetemp$year == 1923 & umetemp$week == 1, ]
+tbl(round(sh[, 1:8], 2), caption = "Weekly summarized temperature data: Umeå 1923, first week.", fs = 9)
+```
+
+Weekly averages are calculated year by year, and deviations from 
 the averages of the weekly averages are used as a time-varying *communal covariate*. See Figure 
@@
 
-## Statistical modelling
+## Mortality and temperature as a communal covariate
+
+The two data sets, mortality and weather, are combined ino one by treating temperature data 
+as a communal covariate and incorporate it as such in the mortality data set.
+
+# Statistical modelling
 
@@
 
-## Postneonatal mortality
+# Postneonatal mortality
 
@@
 
-## Neonatal mortality
+# Neonatal mortality
 
@@
 
-### Girls
+## Girls
 
@@
 
-### Boys
+## Boys
 
@@
 
-## Postneonatal mortality, winter and spring
+# Postneonatal mortality, winter and spring
 
@@
 
-## Neonatal mortality, winter and spring
+# Neonatal mortality, winter and spring
 

### 2022-02-02 getdata2.Rmd
updates

@@
 bjurtemp <- tempdat("bjur", temp_ivl)
+stentemp <- tempdat("sten", temp_ivl) # Added 27 Jan 2022
 ##stentemp <- tempdat("~/Forskning/Data/sten_temp.csv", c(1899, 1952))
@@
 saveRDS(bjurtemp, file = "data/bjurtemp.rds")
+saveRDS(stentemp, file = "data/stentemp.rds")  ## Added 27 Jan2022
 ```

### 2022-02-02 getdata3.Rmd
use lag=0 (no lag)
updates

--- 2021-12-28 getdata3.Rmd
@@
 
-Now (December 22, 2021) I have come to the conclusion that data from the data frame 
+**Revisions:**
+
+1.    (December 22, 2021) I have come to the conclusion that data from the data frame 
 *persons* is the best and easiest start.
 
-So we start with *skum::persons* instead of *skum::obs*, and age is to begin with in days,
+2.    (January 27. 2022) I found that teperature reports from *Stensele* are very similar
+to those from the very similar *Umeå* and *Skellefteå*. Conclusion: The Skellefteå inland is now included.
+
+We start with *skum::persons* instead of *skum::obs*, and age is to begin with in days,
 with stillbirths age *removed* and born alive but dead on birthday at age 0.25 days.
@@
                         rep("sten", 3))
-per <- per[(per$subreg != "sten"), ]
+##per <- per[(per$subreg != "sten"), ] # NOTE: Now Stensele station is included
 per$subreg <- factor(per$subreg)
@@
 per$parity <- per$paritet_g
-saveRDS(per, file = "data/infdat3.rds")
+per$paritet_g <- per$paritet_k <- per$doddat <- per$frsbosdat <- NULL
+per$fid <- per$fboort <- per$ab <- per$fyrkfd <- per$fyrkind <- NULL
+per$region <- per$kon <- per$fodhfrs <- per$frsbobdat <- per$frsbostyp <- NULL
+per$fbonmn <- per$fboind <- per$fyrkrel <- per$fyrketxt <- NULL
+per$foddat <- as.Date(as.character(per$foddat), format = "%Y%m%d")
+## NEW:
+per$socBranch <- as.character(per$socBranch)
+per$socBranch[is.na(per$socBranch)] <- "unknown"
+per$socBranch <- factor(per$socBranch)
+per$socStatus <- as.character(per$socStatus)
+per$socStatus[is.na(per$socStatus)] <- "unknown"
+per$socStatus <- factor(per$socStatus)
+per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", 
+               "subreg", "socBranch", "socStatus", "hisclass", "enter", "exit", "event")]
+#saveRDS(per, file = "data/infdat3.rds")
 ```
@@
 bjurtemp <- readRDS("data/bjurtemp.rds")
+stentemp <- readRDS("data/stentemp.rds")
 ##
 
-lagg <- 3 / 365  ## NOTE!!
-tempvars <-  c("heat", "aver", "cold", "emintemp", "emeantemp", "emaxtemp", 
+lagg <- 0 / 365 ##3 / 365  ## NOTE!!
+tempvars <-  c("heat", "heat.1", "aver", "aver.1", "cold", "cold.1", "emintemp", "emeantemp", "emaxtemp", 
                "week", "year") 
@@
 
+### Sten:
+bsten <- per[per$subreg == "sten", ]
+bsten <- make.communal(bsten, stentemp[, tempvars], 
+                       start = temp_start, period = 1/52, lag = lagg)
+
 bume$region <- "ume"
 bskel$region <- "ske"
-infdat <- rbind(bume, bskel)
-infdat <- infdat[order(infdat$id, infdat$enter), c("region", "id", "sex", "enter", "exit", "event", 
-                                                   "birthdate", "socStatus", "hisclass", "illeg",
-                                                   "parity", tempvars)]
-saveRDS(infdat, file = "data/infdat3_year3.rds")
+bsten$region <- "sten"
+infdat <- rbind(bume, bskel, bsten)
+infdat <- infdat[order(infdat$id, infdat$enter), c("subreg", "id", "sex", "enter", "exit", "event", 
+                                                   "birthdate", "socBranch", "socStatus", "illeg",
+                                                   "parity", "hisclass", tempvars)]
+## New addition: (28 Jan 2022)
+infdat <- rc(infdat) # rc from 'skum'
+n <- NROW(infdat)
+
+saveRDS(infdat, file = "data/infdat3.rds") # NOTE: Name change!!!
 
 ```
+
+Note the name change of the data file: *infdat3.rds*! And the the time lag is `r lagg`.

### 2022-02-02 paper.Rmd
updates

@@
 - \usepackage{a4wide}
-- \usepackage[figuresonly]{endfloat}
-- \renewcommand{\efloatseparator}{\mbox{}}
+#- \usepackage[figuresonly]{endfloat}
+#- \renewcommand{\efloatseparator}{\mbox{}}
 abstract: "The effect of extreme temperatures on infant mortality in the Umeå region 1895-1950 is studied in a biometric analysis setting.
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3,comment=NA)
+knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA, global.par=TRUE)
 library(knitr)
 library(eha)
+par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
 g <- function(x) (log(1 + x))^3 # x in days.
@@
 
-```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 4}
+```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
 infdat <- readRDS("data/infdat3_year3.rds")
@@
 y <- tid$event / tid$exposure
+par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
 plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
@@
 
-```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 4}
+```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
 neo <- age.window(infdat, c(0, 28 / 365))
@@
 y <- tneo$event / tneo$exposure
+par(cex.axis = 0.7)
+par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
 plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
@@
 
-```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 4}
+```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
 pneo <- age.window(infdat, c(28 / 365, 1))
@@
 y <- tpneo$event / tpneo$exposure
+par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
 plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
@@
 
-```{r longtrend, fig.cap = "Crude IMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
+```{r longtrend, fig.cap = "Crude IMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 3}
 trend <- toTpch(Surv(enter, exit, event) ~ year, data = infdat, cuts = c(0, 1))
@@
 par(las = 1)
+par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
 plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
@@
 
-Weekly averages are calculated year by year, and deviations from 
-the averages of the weekly averages are used as a time-varying *communal covariate*. See Figure 
-\@ref(fig:monavgtemp) for the monthly distributions.
+Weekly averages (`mintemp`, `maxtemp`, `meantemp`)  are calculated by week and year, and deviations from 
+the averages (`emintemp`, `emaxtemp`, `emeantemp`) of the weekly averages are used as time-varying *communal covariates*. 
+As an example, see Figure \@ref(fig:firstweek), where the variation around the average
+minimum temperature (`emintemp`) week 1 is shown.
 
-```{r monavgtemp, fig.cap="Weekly max and min temperature averages, 1895--1950.", fig.height = 4}
+```{r firstweek, fig.cap = "Minimum temperature the first week of each year.", fig.height = 3}
+w1 <- umetemp[umetemp$week == 1, ]
+emintemp <- w1$emintemp[1]
+mintemp <- w1$mintemp
+par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
+plot(w1$year, mintemp, type = "b", lty = 2, col = "blue", xlab = "Year", ylab = "Temperature", 
+     axes = FALSE, ylim = c(-35, 0), cex = 0.7)
+axis(1, at = c(1894, 1903, 1913, 1923, 1932, 1942, 1951))
+whh <- c(-30, -25, emintemp, -10, 0) 
+axis(2, at = whh, labels = round(whh))
+abline(h = emintemp, lwd = 1.5, col = "blue")
+abline (h = 0)
+abline(v = 1923, lty = 3)
+box()
+```
+
+Curiously, our randomly selected year 1923 turns out contain the warmest first week of all years, see 
+Figure \@ref(fig:firstweek).
+
+Figure 
+\@ref(fig:monavgtemp) shows the average monthly distribution over all years. The subregional patterns 
+and levels are very similar.
+
+```{r monavgtemp, fig.cap="Weekly max and min temperature averages, 1895--1950.", fig.height = 3}
 oldpar <- par(mfrow = c(1, 2))
@@
 uminavg <- umetemp$emintemp[1:52]
+par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
 plot(1:52, umaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
      xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
-text(17, 27, "Umeå")
+text(17, 27, "Umeå", cex = 0.7)
 axis(1, at = c(1, 10, 29, 43, 52))
@@
 lines(1:52, uminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
-text(8, 11, "max", col = "red")
-text(20, -9, "min", col = "blue")
+text(8, 11, "max", col = "red", cex = 0.7)
+text(20, -9, "min", col = "blue", cex = 0.7)
 box()
@@
      xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
-text(17, 27, "Skellefteå")
+text(17, 27, "Skellefteå", cex = 0.7)
 axis(1, at = c(1, 10, 29, 43, 52))
@@
 lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
-text(8, 11, "max", col = "red")
-text(20, -9, "min", col = "blue")
+text(8, 11, "max", col = "red", cex = 0.7)
+text(20, -9, "min", col = "blue", cex = 0.7)
 box()
@@
 
-Time trends of yearly average temperatures, see Figure
+Time trends of yearly average temperatures, see Figure \@ref(fig:longterm).
 
-```{r longterm}
+```{r longterm, fig.cap = "Yearly average temperatures, Umeå and Skellefteå."}
 bmtemp <- rbind(umetemp, bjurtemp)
 umt <- with(bmtemp, tapply(meantemp, year, mean))
+par(las = 1)
+par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
 plot(1894:1951, umt, type = "b", col = "blue", xlab = "Year",
@@
 
-## Mortality and temperature as a communal covariate
+## Temperature as communal covariates
 
-The two data sets, mortality and weather, are combined ino one by treating temperature data 
-as a communal covariate and incorporate it as such in the mortality data set.
+The two data sets, mortality and weather, are combined into one by treating temperature data 
+as a communal covariate and incorporate it as such in the mortality data set. The 
+function *make.communal* in the **R** [@cran] package *eha* [@eha; @ehar2] is used 
+for that purpose.
+
+```{r readdata3}
+levels(infdat$subreg) <- c("ume", "ske")
+##infdat$excessTemp <- cut(infdat$aver, c(-16, -2, 2, 9), include.lowest = TRUE) # NOTE!!!
+##infdat$excessTemp <- relevel(infdat$excessTemp, ref = "(-2,2]")
+infdat$excessTemp <- infdat$cold # CHECK THIS!!
+tbl(head(infdat))
+```
 
@@
 
-The analyses are performed on the log-cube scale with proportional hazards modelling. Note that
+The analyses are performed on the *log-cube* scale, following the hints of  
+Bourgeois-Pichat [@bp51a; @bp51b], with proportional hazards modelling. Note that
 the property of proportional hazards are preserved under a strictly monotone increasing time transform.
@@
 
-We start by performing separate analyses for neonatal and postneonatal mortality. Then we make comparisons.
+We start by performing separate analyses for neonatal and postneonatal mortality.
 
@@
 
-```{r postanalske, results='asis'}
+```{r postanal, results='asis'}
 source("R/ltx2.R")
@@
 tpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + socStatus + illeg + parity + 
-                    cold 
+                    excessTemp
                 + emintemp + season + subreg, 
@@
   ##           data = pneo, cuts = g(c(28, 365)))
-tpneo$excessCold <- with(tpneo, -cold * (cold < 0))
-fit.tpch <- tpchreg(oe(event, exposure) ~ excessCold + emintemp  +  period + sex + socBranch + socStatus  + 
-                        parity + season, 
-                    data = tpneo[tpneo$subreg == "ske", ], time = age)
+
+fit.tpch <- tpchreg(oe(event, exposure) ~ excessTemp + emintemp +  period + sex + socBranch + socStatus  + 
+                        illeg + parity, 
+                    data = tpneo, time = age)
 dr <- drop1(fit.tpch, test = "Chisq")
-ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, Skellefteå.
-Adjusted for social status and branch, illegitimacy, parity, time period, season, and sub region.", label = "tab:postanalske", keep = 2)
+ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality. Adjusted for social status and branch, illegitimacy, parity, time period, season, and subregion.", label = "tab:postanalske", keep = 2)
 ```
 
-The result in Table \@ref(tab:postanalske) shows that *climate* (`emintemp`) is more important than 
-*weather* (`cold`). Moreover, no signs of interaction between weather or climate and the rest of 
+The result in Table \@ref(tab:postanal) shows that *climate* (`emintemp`) is more important than 
+*weather* (`excessTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
 covariates (not shown).
-
-```{r postanalume, results = 'asis'}
-fit.tpch <- tpchreg(oe(event, exposure) ~ excessCold + emintemp + period + sex + socBranch + socStatus + 
-                        parity + season, 
-                    data = tpneo[tpneo$subreg == "ume", ], time = age)
-dr <- drop1(fit.tpch, test = "Chisq")
-ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, Umeå. Adjusted for social status and branch, illegitimacy, parity, time period, season, and subregion.", label = "tab:postanalume", keep = 2)
-```
 
@@
 neo$enter <- g(neo$enter * 365)
-fit.ph <- coxreg(Surv(enter, exit, event) ~  cold + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
-                 data = neo[neo$sex == "girl", ])
+fit.ph <- coxreg(Surv(enter, exit, event) ~  (excessTemp + emintemp) * sex + socStatus + illeg + parity + period + socBranch + subreg,
+                 data = neo)
 dr <- drop1(fit.ph, test = "Chisq")
-ltx2(fit.ph, dr = dr, caption = "Neonatal mortality, girls. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatal", keep = 2)
+ltx2(fit.ph, dr = dr, caption = "Neonatal mortality. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatal", keep = 2)
 ```
@@
 ```{r neoanalboys, results = 'asis'}
-fit.ph <- coxreg(Surv(enter, exit, event) ~  cold + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
-                 data = neo[neo$sex == "boy", ])
+fit.ph <- coxreg(Surv(enter, exit, event) ~  excessTemp + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
+                 data = neo[neo$sex == "boy" & neo$season %in% c("spring", "summer"), ])
 dr <- drop1(fit.ph, test = "Chisq")
@@
 ##
-tpneo <- toTpch(Surv(enter, exit, event) ~ sex + socBranch + socStatus + illeg + parity + 
-                    cold + emintemp + period + subreg, 
+tpneo <- toTpch(Surv(enter, exit, event) ~ excessTemp + emintemp + sex + socBranch + socStatus + illeg + parity + 
+                    period + subreg, 
                 data = pneo, cuts = g(c(28, 365)))
@@
   ##           data = pneo, cuts = g(c(28, 365)))
-fit.tpch <- tpchreg(oe(event, exposure) ~ cold + emintemp + sex + socBranch + socStatus + illeg + parity + period + 
+fit.tpch <- tpchreg(oe(event, exposure) ~ excessTemp + emintemp + sex + socBranch + socStatus + illeg + parity + period + 
                         subreg, 
@@
 neo$enter <- g(neo$enter * 365)
-fit.ph <- coxreg(Surv(enter, exit, event) ~  cold + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
-                 data = neo[neo$sex == "girl", ])
+fit.ph <- coxreg(Surv(enter, exit, event) ~  excessTemp + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
+                 data = neo[neo$sex == "girl" & neo$emeantemp < -0, ])
 dr <- drop1(fit.ph, test = "Chisq")
@@
 ```{r neoanalboysw, results = 'asis'}
-fit.ph <- coxreg(Surv(enter, exit, event) ~  cold + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
-                 data = neo[neo$sex == "boy", ])
+fit.ph <- coxreg(Surv(enter, exit, event) ~  excessTemp + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
+                 data = neo[neo$sex == "boy" & neo$season %in% c("winter", "spring"), ])
 dr <- drop1(fit.ph, test = "Chisq")

### 2022-02-02 R/tempdat.R

@@ -46,6 +46,11 @@ tempdat <- function(filename = "ume", period = c(1894, 1951)){
     temp$heat <- round(temp$maxtemp - temp$emaxtemp)
     temp$cold <- round(temp$mintemp - temp$emintemp)
     temp$aver <- round(temp$meantemp - temp$emeantemp)
+    # NEW 28 Jan 2022:
+    n <- NROW(temp)
+    temp$cold.1 <- c(NA, temp$cold[-n])
+    temp$aver.1 <- c(NA, temp$aver[-n])
+    temp$heat.1 <- c(NA, temp$heat[-n])
     ##
     temp
 }

### 2022-02-02 paper3.Rmd
first commit

--- 2022-02-02 paper.Rmd
@@
 ---
-title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
+title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950, including the Stensele station and the inland"
 author: "Göran Broström and Tommy Bengtsson"
@@
 ```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
-infdat <- readRDS("data/infdat3_year3.rds")
+infdat <- readRDS("data/infdat3.rds")
 tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))
@@
 bjurtemp <- readRDS("data/bjurtemp.rds")
+stentemp <- readRDS("data/stentemp.rds")
 ```
@@
 ```{r longterm, fig.cap = "Yearly average temperatures, Umeå and Skellefteå."}
-bmtemp <- rbind(umetemp, bjurtemp)
+bmtemp <- rbind(umetemp, bjurtemp, stentemp)
 umt <- with(bmtemp, tapply(meantemp, year, mean))
@@
 function *make.communal* in the **R** [@cran] package *eha* [@eha; @ehar2] is used 
-for that purpose.
+for that purpose.  Resulting data drame is partly shown in Table \@ref(tab:readdata3).
 
 ```{r readdata3}
-levels(infdat$subreg) <- c("ume", "ske")
+levels(infdat$subreg) <- c("ume", "ske", "inland")
 ##infdat$excessTemp <- cut(infdat$aver, c(-16, -2, 2, 9), include.lowest = TRUE) # NOTE!!!
 ##infdat$excessTemp <- relevel(infdat$excessTemp, ref = "(-2,2]")
-infdat$excessTemp <- infdat$cold # CHECK THIS!!
-tbl(head(infdat))
+##infdat$lowTemp <- (infdat$aver + infdat$aver.1) < -10  # CHECK THIS!!
+##infdat$highTemp <- infdat$aver + infdat$aver.1 > 10
+##infdat$lowTemp <- with(infdat, (aver < -3) & (aver.1 < -3))
+infdat$lowTemp <- (with(infdat, cold + cold.1 < -10))
+infdat$highTemp <- with(infdat, (aver > 3) & (aver.1 > 3))
+tbl(head(infdat[, c("enter", "exit", "event", "lowTemp", "highTemp", "aver", "emeantemp", "week", "year")]),
+    caption = "Data with communal covariates.", fs = 9)
 ```
@@
 
-We start by performing separate analyses for neonatal and postneonatal mortality.
+It turns out that extremely low temperature (`lowTemp`) is bad during all seasons except summer,
+and extremely high temperature (`highTemp`) is bad during summer, but good otherwise.
+So we group season into two categories, *summer* and *not summer*.
+In each case separate analyses for neonatal and postneonatal mortality are performed.
+
+```{r prepare}
+source("R/ltx2.R")
+infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
+infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
+##levels(infdat$subreg) <- c("ume", "ske")
+infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
+infdat$season <- as.factor(infdat$season)
+infdat$emeantemp <- round(infdat$emeantemp)
+levels(infdat$socBranch) <- c("office", "farming", "office", "worker", "worker") # NOTE!
+```
+
+# Neonatal mortality
+
+```{r introneo}
+neo <- age.window(infdat, c(0, 28/ 365))
+```
+
+The analysis is split into two parts by season, *winter + spring* is one, and *summer + fall* the other.
+
+## Winter and Spring
+
+This period refers to the months *December to May*. A Cox regression involves as
+interesting variables *highTemp*, an indicator of temperature at least four degrees
+above the expected for at least two weeks in a row, *emeantemp* the expected temperature
+the actual weeek, and *aver* the *excess temperature* the actual week.
+
+```{r neoanalw, results = 'asis'}
+##library(survival)
+winter <- neo[neo$season %in% c("winter", "spring"), ]
+winter$season <- factor(winter$season)
+##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
+##neo$enter <- g(neo$enter * 365)
+fit.cr <- coxreg(Surv(enter, exit, event) ~  highTemp + aver + emeantemp + 
+                     sex + season + subreg + illeg + parity + socBranch,
+                 data = winter)
+dr <- drop1(fit.cr, test = "Chisq")
+ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, winter and spring. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
+     label = "tab:neoanalw", keep = NULL)
+```
+
+## Summer and Fall
+
+This period refers to the months *April to October*. A Cox regression involves as
+interesting variables *highTemp*, an indicator of temperature at least four degrees
+above the expected for at least two weeks in a row, *emeantemp* the expected temperature
+the actual weeek, and *aver* the *excess temperature* the actual week.
+
+```{r neoanals, results = 'asis'}
+##library(survival)
+summer <- neo[neo$season %in% c("summer", "fall"), ]
+summer$season <- factor(summer$season)
+##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
+##neo$enter <- g(neo$enter * 365)
+fit.cr <- coxreg(Surv(enter, exit, event) ~  aver + emeantemp + 
+                     sex + season + subreg + illeg + parity + socBranch,
+                 data = summer)
+dr <- drop1(fit.cr, test = "Chisq")
+ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, summer and fall. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
+     label = "tab:neoanals", keep = NULL)
+```
 
@@
 
-This is the simplest part, because *B-P* predicts that the baseline distribution (given the log-cube
-transformation) is *exponential*, that is, the hazard function is *constant*.
+This is the simplest part, because @bp51a predicts that the baseline distribution (given the log-cube
+transformation **g**) is *exponential*, that is, the hazard function is *constant*.
+To give substance to this claim, a short introduction to the ideas of Bourgeois-Pichat is
+given.
 
-```{r postanal, results='asis'}
-source("R/ltx2.R")
-infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
-infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
-levels(infdat$subreg) <- c("ume", "ske")
-infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
-infdat$season <- as.factor(infdat$season)
+## Bourgeois-Pichat and the log-cube transform
+
+**FILL IN ...**
+
+```{r getpneo}
 pneo <- age.window(infdat, c(28 / 365, 1))
-##
 pneo$enter <- g(pneo$enter * 365)
@@
 ##
-pneo$emintemp <- round(pneo$emintemp)
-##
 tpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + socStatus + illeg + parity + 
-                    excessTemp
-                + emintemp + season + subreg, 
+                    lowTemp + highTemp + season + aver
+                + emeantemp + subreg, 
                 data = pneo, cuts = g(c(28, 365)))
-##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
-  ##           data = pneo, shape = 1)
-##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
-  ##           data = pneo, cuts = g(c(28, 365)))
-
-fit.tpch <- tpchreg(oe(event, exposure) ~ excessTemp + emintemp +  period + sex + socBranch + socStatus  + 
-                        illeg + parity, 
-                    data = tpneo, time = age)
-dr <- drop1(fit.tpch, test = "Chisq")
-ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality. Adjusted for social status and branch, illegitimacy, parity, time period, season, and subregion.", label = "tab:postanalske", keep = 2)
 ```
 
-The result in Table \@ref(tab:postanal) shows that *climate* (`emintemp`) is more important than 
+## Winter and Spring
+
+```{r postanalw, results='asis'}
+winter <- tpneo[tpneo$season %in% c("winter", "spring"), ]
+winter$season <- factor(winter$season)
+
+fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch + socStatus  + 
+                        illeg + parity + subreg, 
+                    data = winter, time = age)
+dr <- drop1(fit.tpch, test = "Chisq")
+ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.
+Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
+```
+
+The result in Table \@ref(tab:postanalw) shows that *climate* (`emintemp`) is more important than 
 *weather* (`excessTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
@@
 
-# Neonatal mortality
+## Summer and Fall
 
-It turns out that here we have some significant interactions, with `sex` involved.
-Maybe surprising. So we run separate analyses for the sexes.
+```{r postanals, results='asis'}
+summer <- tpneo[tpneo$season %in% c("summer", "fall"), ]
+winter$season <- factor(winter$season)
 
-## Girls
-
-```{r neoanal, results = 'asis'}
-##library(survival)
-neo <- age.window(infdat, c(0, 28 /365))
-neo$exit <- g(neo$exit * 365)
-neo$enter <- g(neo$enter * 365)
-fit.ph <- coxreg(Surv(enter, exit, event) ~  (excessTemp + emintemp) * sex + socStatus + illeg + parity + period + socBranch + subreg,
-                 data = neo)
-dr <- drop1(fit.ph, test = "Chisq")
-ltx2(fit.ph, dr = dr, caption = "Neonatal mortality. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatal", keep = 2)
+fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch + socStatus  + 
+                        illeg + parity + subreg, 
+                    data = summer, time = age)
+dr <- drop1(fit.tpch, test = "Chisq")
+ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.
+Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
 ```
 
-Same here, *climate* is more important than *weather*.
-
-## Boys
-
-```{r neoanalboys, results = 'asis'}
-fit.ph <- coxreg(Surv(enter, exit, event) ~  excessTemp + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
-                 data = neo[neo$sex == "boy" & neo$season %in% c("spring", "summer"), ])
-dr <- drop1(fit.ph, test = "Chisq")
-ltx2(fit.ph, dr = dr, caption = "Neonatal mortality, boys. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatalboys", keep = 2)
-```
-
-Here, however, boys are sensitive to temperatures below average, in addition to the 
-sensitivity for cold climate.
-
-# Postneonatal mortality, winter and spring
-
-This is the simplest part, because *B-P* predicts that the baseline distribution (given the log-cube
-transformation) is *exponential*, that is, the hazard function is *constant*.
-
-```{r postanalw, results='asis'}
-##infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
-##levels(infdat$subreg) <- c("ume", "ske")
-##infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
-##infdat$season <- as.factor(infdat$season)
-pneo <- age.window(infdat[infdat$season %in% c("winter", "spring"), ], c(28 / 365, 1))
-##
-pneo$enter <- g(pneo$enter * 365)
-pneo$exit <- g(pneo$exit * 365)
-##
-pneo$emintemp <- round(pneo$emintemp)
-##
-tpneo <- toTpch(Surv(enter, exit, event) ~ excessTemp + emintemp + sex + socBranch + socStatus + illeg + parity + 
-                    period + subreg, 
-                data = pneo, cuts = g(c(28, 365)))
-##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
-  ##           data = pneo, shape = 1)
-##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
-  ##           data = pneo, cuts = g(c(28, 365)))
-fit.tpch <- tpchreg(oe(event, exposure) ~ excessTemp + emintemp + sex + socBranch + socStatus + illeg + parity + period + 
-                        subreg, 
-                    data = tpneo, time = age)
-dr <- drop1(fit.tpch, test = "Chisq")
-ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", 
-     label = "tab:postanalw", keep = 2)
-```
-
-The result in Table \@ref(tab:postanalw) shows that *climate* (`emintemp`) is more important than 
-*weather* (`cold`). Moreover, no signs of interaction between weather or climate and the rest of 
+The result in Table \@ref(tab:postanals) shows that *climate* (`emintemp`) is more important than 
+*weather* (`excessTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
 covariates (not shown).
-
-# Neonatal mortality, winter and spring
-
-It turns out that here we have some significant interactions, with `sex` involved.
-Maybe surprising. So we run separate analyses for the sexes.
-
-## Girls
-
-```{r neoanalw, results = 'asis'}
-neo <- age.window(infdat[infdat$season %in% c("winter", "spring"), ], c(0, 28 /365))
-neo$exit <- g(neo$exit * 365)
-neo$enter <- g(neo$enter * 365)
-fit.ph <- coxreg(Surv(enter, exit, event) ~  excessTemp + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
-                 data = neo[neo$sex == "girl" & neo$emeantemp < -0, ])
-dr <- drop1(fit.ph, test = "Chisq")
-ltx2(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, girls.
-Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatalw", keep = 2)
-```
-
-Same here, *climate* is more important than *weather*.
-
-## Boys
-
-```{r neoanalboysw, results = 'asis'}
-fit.ph <- coxreg(Surv(enter, exit, event) ~  excessTemp + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
-                 data = neo[neo$sex == "boy" & neo$season %in% c("winter", "spring"), ])
-dr <- drop1(fit.ph, test = "Chisq")
-ltx2(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, boys. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", 
-     label = "tab:neonatalboysw", keep = 2)
-```
-
-Here, however, boys are sensitive to temperatures below average, in addition to the 
-sensitivity for cold climate.
 
@@
 
-Remains to be written. Feels like two papers, one about Bourgeois-Pichat and one
-about climate and weather. Maybe discuss exogeneity/endogeneity?
+Remains to be written.

### 2022-02-02 ts.Rmd
first commit

---
title: "Time series"
author: "Göran Broström"
date: "`Sys.time()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Temperature

## Umeå

```{r readdataume}
umetemp <- readRDS("data/umetemp.rds")
sketemp <- readRDS("data/bjurtemp.rds")
stentemp <- readRDS("data/stentemp.rds")
par(cex = 0.8)
for (yr in 1894:1951){
    with(umetemp[umetemp$year == yr, ], plot(week, aver, 
                                               type = "b", col = "blue",
                                               main = yr,
                                               xlab = "Week", lwd = 1.5,
                                               ylim = c(-15, 15), axes = FALSE))
    axis(1, at = c(1, seq(4,52, by = 4)))
    axis(2)
    box()
    ##with(umetemp[umetemp$year == yr, ], lines(week, emeantemp, type = "l"))
    abline (h = 0)
    abline(h = c(-5, -10, -15), v = c(1, seq(4, 52, by = 4)), lty = 3)
    with(sketemp[sketemp$year == yr, ], lines(week, aver, type = "b", col = "red"))
    with(stentemp[stentemp$year == yr, ], lines(week, aver, type = "b", col = "black"))
    legend("topright", legend = c("Umeå", "Bjuröklubb", "Stensele"), col = c("blue", "red", "black"),
           lty = 1, pch = "o", cex = 1)
}
```

# Deaths

## Umeå

```{r deathsume}
infdat <- readRDS("data/infdat3_year3.rds")
umeinf <- infdat[infdat$subreg == "ume", ]
```

## Skellefteå

```{r deathsske}
skeinf <- infdat[infdat$subreg == "bjur", ]
```

### 2022-02-02 R/ltx2.R

@@ -1,0 +1,307 @@
+ltx2 <- function(x, caption = NULL, label = NULL, dr = NULL,
+                 digits = max(options()$digits -4, 3),
+                 keep = NULL, ...){
+    
+    ## A variation of ltx.tpchreg (for the moment)
+    ## where some parameter estimates (corresponding to
+    ## variables, not levels) can be supressed.
+    ##
+    ## If keep == "all", look at exclude.
+    ## If keep != "all", don't look at exclude.
+
+    ## Use keep for the moment
+
+    ##if (!inherits(x, "coxreg")){
+      ##  stop("only for coxreg objects")
+    ##}
+
+    if (is.null(keep)) keep <- 1000 # Something BIG!
+    
+    if (!is.null(cl<- x$call)) {
+	##cat("Call:\n")
+	##dput(cl)
+	##cat("\n")
+	}
+
+    if (!is.null(x$fail)) {
+        if (x$fail != 0){
+            cat(" coxreg failed with: ")
+            stop(x$fail)
+        }
+    }
+
+    if (!length(x$coefficients)){
+        cat("Null log likelihood = $", x$loglik[2], "$\n")
+        return()
+    }
+
+    savedig <- options(digits = digits)
+    on.exit(options(savedig))
+
+    ltxCoef2(x, digits, dr, caption, keep) # Print regression coefficients
+### New 2020-11-26:    
+    if (inherits(x, "summary.tpchreg")){
+        ivl <- paste("(", min(x$cuts), ", ", max(x$cuts), "]", sep = "")
+        if (x$n.strata == 1){
+            cat("Restr. mean survival: & ", x$rmean, " & in", ivl, "\\\\ \n")
+        }else{
+            names(x$rmean) <- x$strata
+            cat("Restr. mean survival in & ", ivl, ": \\\\ \n")
+            for (ll in 1:(x$n.strata - 1)){
+                cat(" & ", formatC(x$strata[ll], format = "g", flag = "#"))
+            }
+            cat(" & ", x$strata[x$n.strata], "\\\\ \n")
+            for (ll in 1:(x$n.strata - 1)){
+                cat(" & ", formatC(x$rmean[ll], format = "g", flag = "#"))
+            }
+            cat(" & ", formatC(x$rmean[x$n.strata], format = "g", flag = "#"), "\\\\ \n")
+        }
+    }
+### End New 2020-11-26.    
+    cat("\\hline \n")
+
[…]

### 2022-02-03 paper3.Rmd
Bourgeois-Pichat is back

@@
 
-# Neonatal mortality
+# Results
+
+The results regarding neonatal mortality is much in accordance with the results
+found by @junkka. However, they used temperature in a "hockey-stick" regression with a breakpoint at
+14.5 degrees Celsius and a negative slope (decreasing risk) to the left and a positive slope
+(increasing risk) to the right. Instead, we are using the average weekly temperature
+for the 52 weeks of a year, for each week averaging over all the years in the study, as
+our "reference points" ("climate"), adding deviances up and down ("weather") as
+"short-term temperature stress". This is similar to the way prices and mortality were related
+in for instance @genus11, that is, a time series split into long time trend and short
+term variation.
+
+In accordance with the results of @bp51a and @bp51b, we separate the investigation into 
+two parts, *neonatal* and *postneonatal* mortality.
+
+## Neonatal mortality
 
@@
 
-The analysis is split into two parts by season, *winter + spring* is one, and *summer + fall* the other.
+The analyses are split into two parts by season, *winter + spring* is one, and *summer + fall* the other.
 
-## Winter and Spring
+### Winter and Spring
 
@@
 
-## Summer and Fall
+### Summer and Fall
 
@@
 
-# Postneonatal mortality
+## Postneonatal mortality
 
@@
 
-## Bourgeois-Pichat and the log-cube transform
+### Bourgeois-Pichat and the log-cube transform
 
-**FILL IN ...**
+The *log-cube transform g*, suggested by @bp51a, is defined as
+
+\begin{equation}
+g(t) = \log^3(1 + t), \quad 0 \le t \le 365,
+\end{equation}
+
+where $t$ is age in days. A graph of the transform is shown in Figure \@ref(fig:gplot).
+
+```{r gplot, fig.cap = "The g transform.",fig.height=3.2}
+x <- seq(0, 365, length = 1000)
+y <- g(x)
+plot(x, y, type = "l", ylab = "g(t)", xlab = "t (days)", col = "blue", axes = FALSE)
+abline(h = 0, v = 0)
+##
+lines(c(28, 28), c(0, g(28)), lty = 3)
+lines(c(0, 28), c(g(28), g(28)), lty = 3)
+##
+lines(c(183, 183), c(0, g(183)), lty = 3)
+lines(c(0, 183), c(g(183), g(183)), lty = 3)
+##
+lines(c(365, 365), c(0, g(365)), lty = 3)
+lines(c(0, 365), c(g(365), g(365)), lty = 3)
+##
+axis(1, at = c(0, 28, 183, 365))
+ycor <-  c(0, g(28), g(183), g(365))
+axis(2, at = ycor, labels = round(ycor), las = 1)
+box()
+```
+
+It was used as a tool for dividing infant mortality 
+into *endogenous* and *exogenous* causes, and a an important part in that venue was
+the observations (i) postneonatal mortality is purely exogenous, and (ii) the distribution 
+of exogenous mortality is *truncated uniform* (right truncated at $t = g(365)$) on the
+$g$-transformed time scale.
+
+Today, assumption (ii) is slightly outdated, for three reasons: (a) The uniform distribution
+is not a very practical tool in survival analysis, (b) replacing the uniform with an
+*exponential* not only fits better (usually), it also is a survival distribution that
+is extremely easy to work with in survival analysis, and (c) "all models are wrong, but some are useful [@box76]".
 
@@
 pneo <- age.window(infdat, c(28 / 365, 1))
-pneo$enter <- g(pneo$enter * 365)
-pneo$exit <- g(pneo$exit * 365)
+gpneo <- pneo
+gpneo$enter <- g(gpneo$enter * 365)
+gpneo$exit <- g(gpneo$exit * 365)
 ##
-tpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + socStatus + illeg + parity + 
+gtpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + socStatus + illeg + parity + 
                     lowTemp + highTemp + season + aver
                 + emeantemp + subreg, 
-                data = pneo, cuts = g(c(28, 365)))
+                data = gpneo, cuts = g(c(28, 365)))
 ```
 
-## Winter and Spring
+For now, let us look at the basic distribution of postneonatal life on the $g$ time
+scale. We do it nonparametrically by calculating the *Nelson-Aalen* estimator of the
+cumulative hazards function. See Figure \@ref(fig:nap), where the two curves are indistinguishable.
+
+```{r nap, fig.cap = "Nelson-Aalen estimator and exponential cumulative hazards of postneonatal survival on the g scale.",fig.height=3.2}
+fit <- coxreg(Surv(enter, exit, event) ~ 1, data = gpneo)
+plot(fit, xlam = "g(t)")
+fit.w <- tpchreg(oe(event, exposure) ~ 1, data = gtpneo, time = age)
+zz <- hazards(fit.w)
+lines(zz$x, zz$y, col = "red", lty = 2)
+legend("topleft", legend = c("Nelson-Aalen", "Exponential"), lty = 1:2, col = c("blue", "red"))
+```
+
+### Winter and Spring
 
 ```{r postanalw, results='asis'}
-winter <- tpneo[tpneo$season %in% c("winter", "spring"), ]
+winter <- gtpneo[gtpneo$season %in% c("winter", "spring"), ]
 winter$season <- factor(winter$season)
@@
 
-## Summer and Fall
+### Summer and Fall
 
 ```{r postanals, results='asis'}
-summer <- tpneo[tpneo$season %in% c("summer", "fall"), ]
+summer <- gtpneo[gtpneo$season %in% c("summer", "fall"), ]
 winter$season <- factor(winter$season)
@@
 
+# The biometric analysis of infant mortality
+
+A modern version of the Bourgeois-Pichat biometric modelling [@bp51a; @bp51b] is given here,
+*without* the "log-cube transform".
+
 # Conclusion

### 2022-02-04 paper3.Rmd
more on B-P and polish

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA, global.par=TRUE)
+knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA)
+knitr::opts_knit$set(global.par = TRUE)
+##
 library(knitr)
 library(eha)
-par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
+##par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
 g <- function(x) (log(1 + x))^3 # x in days.
@@
 ginv <- function(y) exp(y^(1/3)) - 1
+```
+
+```{r setpar}
+par(las = 1, cex.axis = 0.8, cex.lab = 0.8, cex = 0.8)
 ```
@@
 mortality and the log-cube transform.
+
+Figure \@ref(fig:map) shows the study area within Sweden, with the weather stations 
+marked. The map is taken from the paper by @junkka.
+
+```{r map, fig.height = 5, fig.cap = "Umeå, Skellefteå (orange) and its inland (green)."}
+include_graphics("myfigs/skum.png")
+```
 
@@
 
-Temperature data are collected from two weather stations, *Umeå* and *Bjuröklubb*
-(used with population data from Skellefteå coastal area). Both stations deliver
+Temperature data are collected from three weather stations, *Umeå*, *Bjuröklubb*
+(used with population data from Skellefteå coastal area), and *Stensele* (Inland). All stations deliver
 daily temperature data covering our time period, usually three measures per day, morning,
@@
 
-```{r monavgtemp, fig.cap="Weekly max and min temperature averages, 1895--1950.", fig.height = 3}
-oldpar <- par(mfrow = c(1, 2))
+```{r monavgtemp, fig.cap="Weekly max, mean, and min temperature averages, 1895--1950.", fig.height = 3}
+oldpar <- par(mfrow = c(1, 3))
 ## Umeå:
@@
 uminavg <- umetemp$emintemp[1:52]
-par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
-plot(1:52, umaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
+umeanavg <- umetemp$emeantemp[1:52]
+##par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
+plot(1:52, umaxavg, ylim = c(-30, 30), type = "b", col = "red", axes = FALSE,
      xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
@@
 axis(1, at = c(1, 10, 29, 43, 52))
-axis(2, las = 1)
+m12 <- c(-30, -15, 0, 14.5, 25)
+axis(2, las = 1, at = m12)
 abline(h = 0)
 abline(v = 29, lty = 3)
-abline(h = c(-20, -10, 10, 20), lty = 3)
+abline(h = m12, lty = 3)
 lines(1:52, uminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
+lines(1:52, umeanavg, type = "l", col = "black", cex = 0.6, lwd = 0.8)
 text(8, 11, "max", col = "red", cex = 0.7)
@@
 bminavg <- bjurtemp$emintemp[1:52]
-plot(1:52, bmaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
+bmeanavg <- bjurtemp$emeantemp[1:52]
+plot(1:52, bmaxavg, ylim = c(-30, 30), type = "b", col = "red", axes = FALSE,
      xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
@@
 axis(1, at = c(1, 10, 29, 43, 52))
-axis(2, las = 1)
+axis(2, las = 1, at = m12)
 abline(h = 0)
 abline(v = 29, lty = 3)
-abline(h = c(-20, -10, 10, 20), lty = 3)
+abline(h = m12, lty = 3)
 lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
+lines(1:52, bmeanavg, type = "l", col = "black", cex = 0.6, lwd = 0.8)
+text(8, 11, "max", col = "red", cex = 0.7)
+text(20, -9, "min", col = "blue", cex = 0.7)
+box()
+##
+## Inland:
+bmaxavg <- stentemp$emaxtemp[1:52]
+bminavg <- stentemp$emintemp[1:52]
+bmeanavg <- stentemp$emeantemp[1:52]
+plot(1:52, bmaxavg, ylim = c(-30, 30), type = "b", col = "red", axes = FALSE,
+     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
+text(17, 27, "Inland", cex = 0.7)
+axis(1, at = c(1, 10, 29, 43, 52))
+axis(2, las = 1, at = m12)
+abline(h = 0)
+abline(v = 29, lty = 3)
+abline(h = m12, lty = 3)
+lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
+lines(1:52, bmeanavg, type = "l", col = "black", cex = 0.6, lwd = 0.8)
 text(8, 11, "max", col = "red", cex = 0.7)
@@
 ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, winter and spring. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
-     label = "tab:neoanalw", keep = NULL)
+     label = "tab:neoanalw", keep = 3)
 ```
@@
 ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, summer and fall. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
-     label = "tab:neoanals", keep = NULL)
+     label = "tab:neoanals", keep = 2)
 ```
@@
 ##
-gtpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + socStatus + illeg + parity + 
+oj <- system.time(gtpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + illeg + parity + 
                     lowTemp + highTemp + season + aver
                 + emeantemp + subreg, 
-                data = gpneo, cuts = g(c(28, 365)))
+                data = gpneo, cuts = g(c(28, 365))))
 ```
@@
 ```{r nap, fig.cap = "Nelson-Aalen estimator and exponential cumulative hazards of postneonatal survival on the g scale.",fig.height=3.2}
+par(cex = 0.8)
 fit <- coxreg(Surv(enter, exit, event) ~ 1, data = gpneo)
-plot(fit, xlam = "g(t)")
+xx <- hazards(fit)
+plot(xx[[1]][, 1], cumsum(xx[[1]][, 2]), type = "s", xlab = "Age (days)",
+     col = "blue", ylab = "Cumulative hazards", axes = FALSE)
+axis(1, at = g(c(28, 183, 365)), labels = c(28, 183, 365))
+axis(2)
+box()
 fit.w <- tpchreg(oe(event, exposure) ~ 1, data = gtpneo, time = age)
@@
 ```
+
+The survival data set we work with here contains 102 thousand infants and 4.9 million records,
+an average of 48 records per infant, one record per calendar week in life. In total there are 
+`r sum(gpneo$event)` deaths, and a typical Cox regression execution time is slightly
+below one minute. A "naive" application of an exponential regression model to the same data
+is typically using around six minutes.
+
+However, with the exponential model (or generally, the *piecewise constant hazard* (pch)) in mind, data can be heavily reduced by *tabulation*, and the
+statistical *sufficiency principle* will guarantee that we get identical results.
+The tabulation itself takes about 20 seconds, and the analysis after that about 15 seconds, a
+total of 35 seconds. But the tabulation usually only needs to be done once. More about this later.
 
@@
 
-fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch + socStatus  + 
+fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
                         illeg + parity + subreg, 
@@
 ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.
-Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
+Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = 4)
 ```
@@
 ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.
-Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
+Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = 4)
 ```
@@
 
+```{r bpprim, fig.cap = "Postneonatal survival is Weibull vs. Nelson-Aalen.", echo = TRUE}
+fit.w <- phreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
+fit <- coxreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
+plot(fit, fn = "cum", col = "blue")
+xx <- hazards(fit.w)
+lines(xx$x, xx$y, col = "red", lty = 2)
+```
+
+Not very good fit. However, we need a *left-truncated Weibull distribution*, that is, assuming that
+*exogeneous infant mortality* is Weibull distributed. Look here (Figure \@ref(fig:ltruw)). A perfect fit, 
+just like the one under the $g$- transform. This is the way to go, *more will come tomorrow*!
+
+```{r ltruw, fig.cap = "Postneonatal survival is truncated Weibull vs. Nelson-Aalen.", echo = TRUE}
+fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
+fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pneo)
+plot(fit, fn = "cum", col = "blue")
+xx <- hazards(fit.w)
+lines(xx$x, xx$y - xx$y[1], col = "red", lty = 2)
+```
+
 # Conclusion

### 2022-02-05 getdata2.Rmd
add biometric and new temperatures for Inland

--- 2021-12-20 getdata2.Rmd
@@
 
-We have two stations, *Umeå* and *Bjuröklubb*.
+We have three stations, *Umeå*, *Bjuröklubb*, and *Stensele*.
 
@@
 bjurtemp <- tempdat("bjur", temp_ivl)
-##stentemp <- tempdat("~/Forskning/Data/sten_temp.csv", c(1899, 1952))
+
+##Changed 5 feb 2022:
+##stentemp <- tempdat("sten", temp_ivl) # Added 27 Jan 2022
+source("R/normal.R")
+stentemp <- normal(temp_ivl)
+## End change
+
 saveRDS(umetemp, file = "data/umetemp.rds")
 saveRDS(bjurtemp, file = "data/bjurtemp.rds")
+saveRDS(stentemp, file = "data/stentemp.rds")  ## Added 27 Jan2022
 ```

### 2022-02-05 getdata3.Rmd
add biometric and new temperatures for Inland

@@
 
-2.    (January 27. 2022) I found that teperature reports from *Stensele* are very similar
-to those from the very similar *Umeå* and *Skellefteå*. Conclusion: The Skellefteå inland is now included.
+2.    (January 27. 2022) I found that temperature reports from *Stensele* are very similar
+to those from the very similar *Umeå* and *Skellefteå*.
+
+3.    (February 5, 202). Conclusion: The Skellefteå inland is now included. But with 
+summer temperatures from *Bjuröklubb*.
 

### 2022-02-05 paper3.Rmd
add biometric and new temperatures for Inland

@@
 Figure \@ref(fig:map) shows the study area within Sweden, with the weather stations 
-marked. The map is taken from the paper by @junkka.
+marked. The map is taken (with permission) from the paper by @junkka.
 
 ```{r map, fig.height = 5, fig.cap = "Umeå, Skellefteå (orange) and its inland (green)."}
-include_graphics("myfigs/skum.png")
+include_graphics("myfigs/map2.png")
 ```
@@
 ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, winter and spring. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
-     label = "tab:neoanalw", keep = 3)
+     label = "tab:neoanalw", keep = NULL)
 ```
@@
 ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, summer and fall. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
-     label = "tab:neoanals", keep = 2)
+     label = "tab:neoanals", keep = NULL)
 ```
@@
 ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.
-Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = 4)
+Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
 ```
@@
 ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.
-Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = 4)
+Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
 ```
@@
 A modern version of the Bourgeois-Pichat biometric modelling [@bp51a; @bp51b] is given here,
-*without* the "log-cube transform".
+*without* the "log-cube transform". A first try is fit a *Weibull* distribution to the time
+to death for an infant alive at age 28 /365 years (28 days), that is, time is measured from
+age 28 days until death or time reaches 365 - 28 days, when follow-up is censored.
+In **R** language, with the package *eha* [@eha], we try
 
-```{r bpprim, fig.cap = "Postneonatal survival is Weibull vs. Nelson-Aalen.", echo = TRUE}
+```
+fit.w <- phreg(Surv(enter - 28/365, exit - 28/365, event) ~ 1, data = pneo)
+fit <- coxreg(Surv(enter - 28/365, exit - 28/365, event) ~ 1, data = pneo)
+```
+
+where `phreg` performs the Weibull regression (no covariates) and `coxreg` the Cox regression. In 
+Figure \@ref(fig:bpprim) the result is shown in the form of the two cumulative hazards estimates.
+
+```{r bpprim, fig.cap = "Postneonatal survival, Weibull vs. Nelson-Aalen."}
 fit.w <- phreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
@@
 
-Not very good fit. However, we need a *left-truncated Weibull distribution*, that is, assuming that
-*exogeneous infant mortality* is Weibull distributed. Look here (Figure \@ref(fig:ltruw)). A perfect fit, 
-just like the one under the $g$- transform. This is the way to go, *more will come tomorrow*!
+Not a very good fit. However, we need a *left-truncated Weibull distribution*, that is, assuming that
+*exogeneous infant mortality* is Weibull distributed.
 
-```{r ltruw, fig.cap = "Postneonatal survival is truncated Weibull vs. Nelson-Aalen.", echo = TRUE}
+```
 fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
 fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pneo)
-plot(fit, fn = "cum", col = "blue")
+```
+Note the difference: No subtraction of the 28 days from the time variables, which implies that the time origin
+is *birth* and *not* the 28th day of life. Thus our data are *left truncated* at 28 days, and the 
+Weibull attempt estimates neonatal exogeneous mortality through *extrapolation*. The Cox regression approach
+is a non-parametric procedure, and thus not capable of any extrapolated estimation, so neonatal
+mortality will simply be taken as zero.
+
+```{r ltruw, fig.cap = "Postneonatal survival, truncated Weibull vs. Nelson-Aalen."}
+fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
+fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pneo)
+oldpar <- par(mfrow = c(1, 2))
+##check.dist(fit, fit.w)
+cc <- hazards(fit)
+plot(cc[[1]][, 1], cumsum(cc[[1]][, 2]), type = "l", col = "blue", ylim = c(0, 0.07), axes = FALSE,
+     xlab = "Age", ylab = "Cumulative hazards")
+axis(1, at = c(28 / 365, 1), labels = c("28/365", "1"))
+axis(2, las = 1)
+box()
 xx <- hazards(fit.w)
+lines(xx$x, xx$y, col = "red", lty = 2)
+plot(cc[[1]][, 1], cumsum(cc[[1]][, 2]), type = "l", col = "blue", ylim = c(0, 0.07), axes = FALSE,
+     xlab = "Age", ylab = "Cumulative hazards")
+##plot(fit, fn = "cum", col = "blue", ylim = c(0, 0.07), axes = FALSE)
+axis(1, at = c(28 / 365, 1), labels = c("28/365", "1"))
+axis(2, las = 1)
+box()
 lines(xx$x, xx$y - xx$y[1], col = "red", lty = 2)
+par(oldpar)
 ```
+
+ Look here (Figure \@ref(fig:ltruw))! A perfect fit, 
+just like the one under the $g$- transform. But the Weibull curve
+does not start at zero (left hand panel), however it is parallell 
+to the Nelson-Aalen estimator, and the right hand panel shows what happens if we subtract 
+the start value from the Weibull curve and redraw.
+
+## Separate exogeneous and endogeneous Infant mortality
+
+So, finally, we show the full procedure how to perform the "biometric analysis" *without*
+the log-cube transform. It simply consists of two steps:
+
+1.    Run a Cox regression on the full data set.
+2.    Run a *left truncated* (at 28 days) Weibull regression.
+3.    Plot the two cumulative hazards functions in the same graph and read
+      their crossings of the vertical line at age = 1 year (365 days). Figure \@ref(fig:crossings).
+
+```{r crossings, fig.cap = "Biometric analysis.", fig.height = 4}
+fit.c <- coxreg(Surv(enter, exit, event) ~ 1, data = infdat)
+fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
+cc <- hazards(fit.c)
+plot(c(0, cc[[1]][, 1]), c(0, cumsum(cc[[1]][, 2])), type = "l", col = "blue", axes = FALSE,
+     xlab = "Age (days)", ylab = "Cumulative hazards", ylim = c(0, 0.08))
+axis(1, at = c(0, 28 / 365, 1), labels = c(0, 28, 365))
+axis(2, las = 1, at = round(c(0, max(cumsum(cc[[1]][, 2])), max(xx$y)), 3))
+##abline(v = 1)
+box()
+xx <- hazards(fit.w)
+lines(xx$x, xx$y, col = "red", lty = 2)
+abline(h = 0, v = 0)
+abline(v = 1, lty = 3)
+abline(v = 28 / 365, lty = 3)
+abline(h = c(max(cumsum(cc[[1]][, 2])), max(xx$y)), lty = 3)
+```
+
+So, total mortality is 0.076, exogeneous mortality is 0.063, and endogeneous mortality
+is the difference, 0.013, or 17 per cent of the total mortality.
+
+Now, try the original "biometric" approach and compare *(Exercise)*.
 
@@
 
-Remains to be written.
+Remains to be written, especially the discussion about temperature and mortality.

### 2022-02-07 README.Rmd
add time stamp
updates
updated, BP weq

@@
 
-The file *paper.pdf* is the result of this project.
+Now Bourgeois-Pichat and the log-cube transform are erased, but the present state of
+*paper4.pdf* chaotic. No text is updated, only some text is removed.
+
+The file *paper3.pdf* is the old result of the project. Note the time stamp on the front
+page, this report is *not* updated for some time.

### 2022-02-07 getdata2.Rmd
updated, BP weq

@@
 ##stentemp <- tempdat("sten", temp_ivl) # Added 27 Jan 2022
-source("R/normal.R")
+source("R/normal.R") # normal = Norsjö - Malå!!
 stentemp <- normal(temp_ivl)

### 2022-02-07 paper3.Rmd
updated, BP weq

@@
 
+The *summer* half-year is the weeks 14--39, about 1 April to 30 September, and the *winter*
+half-year is the rest, weeks 1-13 and 40--52, 1 January to 31 March and 1 October to 31 December.
+This is the division made in @karlsson, and we keep it for comparability reasons.
+
 ```{r prepare}
@@
 ##levels(infdat$subreg) <- c("ume", "ske")
-infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
-infdat$season <- as.factor(infdat$season)
+##infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
+##infdat$season <- as.factor(infdat$season)
+## New:
+infdat$summer <- infdat[infdat$week %in% 14:39, ]
 infdat$emeantemp <- round(infdat$emeantemp)
@@
 
-### Winter and Spring
+### Winter
 
-This period refers to the months *December to May*. A Cox regression involves as
+This period refers to the months *October to March*. A Cox regression involves as
 interesting variables *highTemp*, an indicator of temperature at least four degrees
@@
 ##library(survival)
-winter <- neo[neo$season %in% c("winter", "spring"), ]
-winter$season <- factor(winter$season)
+winter <- neo[!neo$summer, ]
+##winter$season <- factor(winter$season)
 ##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
@@
 fit.cr <- coxreg(Surv(enter, exit, event) ~  highTemp + aver + emeantemp + 
-                     sex + season + subreg + illeg + parity + socBranch,
+                     sex + subreg + illeg + parity + socBranch,
                  data = winter)
 dr <- drop1(fit.cr, test = "Chisq")
-ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, winter and spring. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
+ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
      label = "tab:neoanalw", keep = NULL)
@@
 
-### Summer and Fall
+### Summer
 
-This period refers to the months *April to October*. A Cox regression involves as
+This period refers to the months *April to September*. A Cox regression involves as
 interesting variables *highTemp*, an indicator of temperature at least four degrees
@@
 ##library(survival)
-summer <- neo[neo$season %in% c("summer", "fall"), ]
-summer$season <- factor(summer$season)
+summer <- neo[neo$summer, ]
+##summer$season <- factor(summer$season)
 ##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
@@
 fit.cr <- coxreg(Surv(enter, exit, event) ~  aver + emeantemp + 
-                     sex + season + subreg + illeg + parity + socBranch,
+                     sex + subreg + illeg + parity + socBranch,
                  data = summer)
 dr <- drop1(fit.cr, test = "Chisq")
-ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, summer and fall. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
+ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, April to September. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
      label = "tab:neoanals", keep = NULL)
@@
 oj <- system.time(gtpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + illeg + parity + 
-                    lowTemp + highTemp + season + aver
+                    lowTemp + highTemp + summer + aver
                 + emeantemp + subreg, 
@@
 
-### Winter and Spring
+### Winter
 
 ```{r postanalw, results='asis'}
-winter <- gtpneo[gtpneo$season %in% c("winter", "spring"), ]
-winter$season <- factor(winter$season)
+winter <- gtpneo[gtpneo$summer == 0, ]
+##winter$season <- factor(winter$season)
 
@@
 dr <- drop1(fit.tpch, test = "Chisq")
-ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.
-Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
+ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
 ```
@@
 
-### Summer and Fall
+### Summer
 
 ```{r postanals, results='asis'}
-summer <- gtpneo[gtpneo$season %in% c("summer", "fall"), ]
-winter$season <- factor(winter$season)
+summer <- gtpneo[gtpneo$summer == 1, ]
+##winter$season <- factor(winter$season)
 
@@
 dr <- drop1(fit.tpch, test = "Chisq")
-ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.
-Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
+ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, April to September.
+Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
 ```
@@
 ```
-fit.w <- phreg(Surv(enter - 28/365, exit - 28/365, event) ~ 1, data = pneo)
-fit <- coxreg(Surv(enter - 28/365, exit - 28/365, event) ~ 1, data = pneo)
+births <- readRDS("data/births.rds")
+pbirths <- age.window(births, c(28 / 365, 1)
+fit.w <- phreg(Surv(enter - 28/365, exit - 28/365, event) ~ 1, data = pbirths)
+fit <- coxreg(Surv(enter - 28/365, exit - 28/365, event) ~ 1, data = pbirths)
 ```
@@
 ```{r bpprim, fig.cap = "Postneonatal survival, Weibull vs. Nelson-Aalen."}
-fit.w <- phreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
-fit <- coxreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
+births <- readRDS("data/births.rds")
+pbirths <- age.window(births, c(28 / 365, 1))
+fit.w <- phreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pbirths)
+fit <- coxreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pbirths)
 plot(fit, fn = "cum", col = "blue")
@@
 ```
-fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
-fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pneo)
+fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pbirths)
+fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pbirths)
 ```
@@
 ```{r ltruw, fig.cap = "Postneonatal survival, truncated Weibull vs. Nelson-Aalen."}
-fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
-fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pneo)
+fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pbirths)
+fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pbirths)
 oldpar <- par(mfrow = c(1, 2))
@@
 
-## Separate exogeneous and endogeneous Infant mortality
+## The biometric analysis of  exogeneous and endogeneous infant mortality
 
@@
 ```{r crossings, fig.cap = "Biometric analysis.", fig.height = 4}
-fit.c <- coxreg(Surv(enter, exit, event) ~ 1, data = infdat)
-fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
+fit.c <- coxreg(Surv(enter, exit, event) ~ 1, data = births)
+fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pbirths)
 cc <- hazards(fit.c)

### 2022-02-07 getdata4.Rmd
first commit

---
title: "getdata4.Rmd"
author: "Göran Broström"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(skum) # Also loads 'eha'
age.cuts = c(0, 3, 7, 28, 56, 140, 183, 250, 365) / 365
```

# Introduction

Created 6 February 2022 after consultations.

** Run this (and only this) before running "paper4.Rmd"!**

We start with *skum::persons*  and age is to begin with in days,
with stillbirths  *removed* and born alive but dead on birthday at age 0.25 days.

For use of *eha::make.communal*, we need to temporarily change time unit to years.

Further, we eventually skip the use of eha::make.communal in  favor of *match*. Reason: 
eha::make.communal assumes age is measured in *years*. But this is no promise yet.

# Read skum::pers

```{r readpers}
per <- pers[(pers$frsbobtyp == 2) & (pers$df == 0), ]
per$frsbobtyp <- per$df <- per$f2id <- per$m2id <- NULL
per$frsbobdatmin <- per$frsbobdatmax <- per$utdatmax <- per$utdatmin <- NULL
per <- per[(per$foddat > 18950000), ]
dn <- per$doddat < 1
per$doddat[dn] <- per$foddat[dn] + 40000
## Fix foddat:
yr <- per$foddat %% 10000 == 0
per$foddat[yr] <- per$foddat[yr] + 101 # As early as possible
mon <- per$foddat %% 100 == 0
per$foddat[mon] <- per$foddat[mon] + 1 # Ditto
## Fix doddat:
yr <- per$doddat %% 10000 == 0
per$doddat[yr] <- per$doddat[yr] + 1231 # As late as possible
mon <- per$doddat %% 100 == 0
per$doddat[mon] <- per$doddat[mon] + 28 # Ditto (guard for February ...)
##
per$birthdate <- as.Date(as.character(per$foddat), format = "%Y%m%d")
per$deathdate <- as.Date(as.character(per$doddat), format = "%Y%m%d")
nabir <- is.na(per$birthdate)
per$foddat[nabir] <- per$foddat[nabir] - 2
per$doddat[nabir] <- per$doddat[nabir] - 2
##
per$birthdate <- as.Date(as.character(per$foddat), format = "%Y%m%d")
per$exit <- as.numeric(difftime(per$deathdate, per$birthdate))
per$deathdate <- as.Date(as.character(per$doddat), format = "%Y%m%d")
## Fix doddat:
per$doddat[is.na(per$deathdate)] <- per$doddat[is.na(per$deathdate)] - 1
per$deathdate <- as.Date(as.character(per$doddat), format = "%Y%m%d")
##
per$exit <- as.numeric(difftime(per$deathdate, per$birthdate, units = "days"))
oj <- per$exit < 0
per$doddat[oj] <- per$doddat[oj] + 10000
per$deathdate <-  as.Date(as.character(per$doddat), format = "%Y%m%d")
per$exit <- as.numeric(difftime(per$deathdate, per$birthdate, units = "days"))
per$event <- per$exit <= 365
per$exit[per$exit == 0] <- 0.25
per$exit[per$exit > 365] <- 365
```

Now we need *nofrs* for newborn, found in the *observation* file(s), so

```{r fromobs}
bobs <- obs[obs$starttyp == 2 & obs$id %in% per$id, ]
indx <- match(per$id, bobs$id)
per$socStatus <- bobs$socStatus[indx]
per$socBranch <- bobs$socBranch[indx]
per$hisclass <- bobs$hisclass[indx]
per$nofrs <- bobs$nofrs[indx]
per <- per[!is.na(per$nofrs), ]
per$subreg <- factor(per$nofrs)
levels(per$subreg) <- c(rep("ume", 2), 
                        rep("bjur", 4), 
                        rep("sten", 3))
##per <- per[(per$subreg != "sten"), ] # NOTE: Now Stensele station is included
per$subreg <- factor(per$subreg)
per$enter <- 0
per$sex <- factor(per$kon, labels = c("boy", "girl"))
per$birthdate <- toTime(per$birthdate)
per$illeg <- as.numeric(per$ab == 2)
per$parity <- per$paritet_g
per$paritet_g <- per$paritet_k <- per$doddat <- per$frsbosdat <- NULL
per$fid <- per$fboort <- per$ab <- per$fyrkfd <- per$fyrkind <- NULL
per$region <- per$kon <- per$fodhfrs <- per$frsbobdat <- per$frsbostyp <- NULL
per$fbonmn <- per$fboind <- per$fyrkrel <- per$fyrketxt <- NULL
per$foddat <- as.Date(as.character(per$foddat), format = "%Y%m%d")
## NEW:
per$socBranch <- as.character(per$socBranch)
per$socBranch[is.na(per$socBranch)] <- "none"
per$socBranch <- factor(per$socBranch)
per$socStatus <- as.character(per$socStatus)
per$socStatus[is.na(per$socStatus)] <- "unknown"
per$socStatus <- factor(per$socStatus)
per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", 
               "subreg", "socBranch", "socStatus", "hisclass", "enter", "exit", "event")]
#saveRDS(per, file = "data/infdat3.rds")
```

# Temperature data

We have three stations, *Umeå*, *Bjuröklubb*, and *Stensele*.

```{r gettemp}
source("R/tempdat.R")
temp_start <- 1894
temp_end <- 1951
temp_ivl <- c(temp_start, temp_end)
##
umetemp <- tempdat("ume", temp_ivl)
bjurtemp <- tempdat("bjur", temp_ivl)

##Changed 5 feb 2022:
##stentemp <- tempdat("sten", temp_ivl) # Added 27 Jan 2022
source("R/normal.R") # normal = Norsjö - Malå!!
stentemp <- normal(temp_ivl)
## End change

saveRDS(umetemp, file = "data/umetemp.rds")
saveRDS(bjurtemp, file = "data/bjurtemp.rds")
saveRDS(stentemp, file = "data/stentemp.rds")  ## Added 27 Jan2022
```

# Put on temperature data

Here we must change time unit to *year*.

```{r makecommunal}
per$enter <- per$enter / 365
per$exit <- per$exit / 365
## per$birthdate <- toTime(per$birthdate) #already done!
temp_start <- 1894
temp_end <- 1951
temp_ivl <- c(temp_start, temp_end)
##

lagg <- 0 / 365 ##3 / 365  ## NOTE!!
tempvars <-  c("heat", "heat.1", "aver", "aver.1", "cold", "cold.1", "emintemp", "emeantemp", "emaxtemp", 
               "week", "year") 
## Ume:
bume <- per[per$subreg == "ume", ]
bume <- make.communal(bume, umetemp[, tempvars], 
                      start = temp_start, period = 1/52, lag = lagg)

### Ske:
bskel <- per[per$subreg == "bjur", ]
bskel <- make.communal(bskel, bjurtemp[, tempvars], 
                       start = temp_start, period = 1/52, lag = lagg)

### Sten:
bsten <- per[per$subreg == "sten", ]
bsten <- make.communal(bsten, stentemp[, tempvars], 
                       start = temp_start, period = 1/52, lag = lagg)

bume$region <- "ume"
bskel$region <- "ske"
bsten$region <- "sten"
infdat <- rbind(bume, bskel, bsten)
infdat$birthmonth <- ceiling((infdat$birthdate - floor(infdat$birthdate)) * 12)
infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
infdat <- infdat[order(infdat$id, infdat$enter), c("subreg", "id", "sex", "enter", "exit", "event", 
                                                   "birthmonth", "period", "socBranch", "socStatus", "illeg",
                                                   "parity", "hisclass", tempvars)]
## New addition: (28 Jan 2022)
##infdat <- rc(infdat) # rc from 'skum'
##n <- NROW(infdat)

saveRDS(infdat, file = "data/infdat4.rds") # NOTE: Name change!!!

## Wait with this:
## tabdat4 <- toTpch(Surv(enter, exit, event) ~ subreg + sex + birthmonth + socBranch + illeg + parity 
```

Note the name change of the data file: *infdat3.rds*! And the the time lag is `r lagg`.

### 2022-02-07 paper4.Rmd
first commit

--- 2022-02-04 paper3.Rmd
@@
 ---
-title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950, including the Stensele station and the inland"
+title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
 author: "Göran Broström and Tommy Bengtsson"
@@
 #- \renewcommand{\efloatseparator}{\mbox{}}
-abstract: "The effect of extreme temperatures on infant mortality in the Umeå region 1895-1950 is studied in a biometric analysis setting.
-More precisely, the effect of climate and weather, measured by temperature, on infant mortality is investigated.
-It turns out that climate is more important than weather, low average temperatures are more important than temporary dips in temperature.
-The investigated geographical region is situated too far north for bad effects of high temperatures to show."
+abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895-1950 is studied in a biometric analysis setting.
+More precisely, the effect of climate and weather, measured by temperature, average and extremes, on infant mortality is investigated.
+It turns out that climate (average) is more important than weather (extremes), low average temperatures are more important than temporary dips in temperature."
 ---
@@
 library(eha)
-##par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
-g <- function(x) (log(1 + x))^3 # x in days.
-dg <- function(x) (3 * (log(1 + x))^2) / (1 + x)
-ginv <- function(y) exp(y^(1/3)) - 1
 ```
@@
 
+```{r readdata}
+births <- readRDS("data/births.rds")
+infdat  <- readRDS("data/infdat4.rds")
+```
 # Introduction
@@
 
-One important reason for studying neonatal and postneonatal mortality separately is the 
-empirical findings by Bourgeois-Pichat [@bp51a;  @bp51b] about endogenous and exogenous 
-mortality and the log-cube transform.
-
 Figure \@ref(fig:map) shows the study area within Sweden, with the weather stations 
-marked. The map is taken from the paper by @junkka.
+marked. The map is taken (with permission) from the paper by @junkka.
 
 ```{r map, fig.height = 5, fig.cap = "Umeå, Skellefteå (orange) and its inland (green)."}
-include_graphics("myfigs/skum.png")
+include_graphics("myfigs/map2.png")
 ```
@@
 Individual data with all births between 1 January 1895 and 31 December 1950 in two 
-coastal areas of north Sweden, Skellefteå (51560 births) and Umeå (31213 births).
-They were followed until death or age 365 days, whichever came first. 
+coastal and one inland areas of north Sweden, Skellefteå (51560 births) and Umeå (31213 births).
+They were followed until death or age one year, whichever came first. 
 The following *static* characteristics were observed on each child:
@@
 ```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
-infdat <- readRDS("data/infdat3.rds")
+##infdat <- readRDS("data/infdat4.rds") Already done
 tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))
@@
 
-The analyses are performed on the *log-cube* scale, following the hints of  
-Bourgeois-Pichat [@bp51a; @bp51b], with proportional hazards modelling. Note that
-the property of proportional hazards are preserved under a strictly monotone increasing time transform.
-However, the estimates of baseline distribution characteristics will change, of course.
-
 It turns out that extremely low temperature (`lowTemp`) is bad during all seasons except summer,
@@
 infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
-infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
+##infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
 ##levels(infdat$subreg) <- c("ume", "ske")
-infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
-infdat$season <- as.factor(infdat$season)
+##infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
+##infdat$season <- as.factor(infdat$season)
+## New:
+infdat$summer <- infdat$week %in% 14:39
 infdat$emeantemp <- round(infdat$emeantemp)
@@
 
-In accordance with the results of @bp51a and @bp51b, we separate the investigation into 
-two parts, *neonatal* and *postneonatal* mortality.
+We separate the investigation into two parts, *neonatal* and *postneonatal* mortality. But first,
+a joint analysis.
+
+```{r joint, cache = TRUE}
+library(survival)
+neo <- age.window(infdat, c(0, 28 / 365))
+neo$ageIvl <- "neo"
+pneo <- age.window(infdat, c(28/365, 1))
+pneo$ageIvl <- "postneo"
+all <- rbind(neo, pneo)
+all$ageIvl <- factor(all$ageIvl)
+##all$enter <- round(all$enter * 365)
+##all$exit <- round(all$exit * 365)
+fit.all <- coxph(Surv(enter, exit, event) ~ strata(ageIvl) * (aver + emeantemp + lowTemp + highTemp + socBranch + socStatus + sex + period),
+                 data = all)
+dr.all <- drop1(fit.all, test = "Chisq")
+
+inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + birthmonth + period + socBranch + socStatus + parity + 
+                     aver + emeantemp + lowTemp + highTemp + summer + ageIvl, data = all,
+                 cuts = c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365) / 365)
+tabneo <- inftab[inftab$ageIvl == "neo", ]
+tabpost <- inftab[inftab$ageIvl == "postneo", ]
+tabpost$age <- factor(tabpost$age)
+tabneo$age <- factor(tabneo$age)
+##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ aver + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
+  ##               data = tabneo, time = age))
+##system.time(tabpost.fit <- tpchreg(oe(event, exposure) ~ aver + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
+  ##               data = tabpost, time = age))
+```
+
+```{r shownonprop}
+tbl(dr.all, caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
+```
 
@@
 
-```{r introneo}
-neo <- age.window(infdat, c(0, 28/ 365))
-```
+The analyses are split into two parts by season, *winter* is one, and *summer* the other.
+Summer covers the months April to September, and winter the rest of the months.
 
-The analyses are split into two parts by season, *winter + spring* is one, and *summer + fall* the other.
+### Winter
 
-### Winter and Spring
-
-This period refers to the months *December to May*. A Cox regression involves as
+This period refers to the months *October to March*. A Cox regression involves as
 interesting variables *highTemp*, an indicator of temperature at least four degrees
@@
 fit.cr <- coxreg(Surv(enter, exit, event) ~  highTemp + aver + emeantemp + 
-                     sex + season + subreg + illeg + parity + socBranch,
+                     sex + subreg + illeg + parity + socBranch + period,
                  data = winter)
 dr <- drop1(fit.cr, test = "Chisq")
-ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, winter and spring. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
-     label = "tab:neoanalw", keep = 3)
+ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
+     label = "tab:neoanalw", keep = NULL)
 ```
 
-### Summer and Fall
+### Summer
 
-This period refers to the months *April to October*. A Cox regression involves as
+This period refers to the months *April to September*. A Cox regression involves as
 interesting variables *highTemp*, an indicator of temperature at least four degrees
@@
 ##library(survival)
-summer <- neo[neo$season %in% c("summer", "fall"), ]
-summer$season <- factor(summer$season)
+#summer <- neo[neo$summer, ]
+##summer$season <- factor(summer$season)
 ##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
 ##neo$enter <- g(neo$enter * 365)
-fit.cr <- coxreg(Surv(enter, exit, event) ~  aver + emeantemp + 
-                     sex + season + subreg + illeg + parity + socBranch,
-                 data = summer)
+fit.cr <- tpchreg(oe(event, exposure) ~  aver + emeantemp + 
+                     sex + subreg  + socBranch + period,
+                 data = inftab[inftab$summer, ], time = age)
 dr <- drop1(fit.cr, test = "Chisq")
-ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, summer and fall. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
-     label = "tab:neoanals", keep = 2)
+ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, April to September. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
+     label = "tab:neoanals", keep = NULL)
 ```
@@
 
-This is the simplest part, because @bp51a predicts that the baseline distribution (given the log-cube
-transformation **g**) is *exponential*, that is, the hazard function is *constant*.
-To give substance to this claim, a short introduction to the ideas of Bourgeois-Pichat is
-given.
-
-### Bourgeois-Pichat and the log-cube transform
-
-The *log-cube transform g*, suggested by @bp51a, is defined as
-
-\begin{equation}
-g(t) = \log^3(1 + t), \quad 0 \le t \le 365,
-\end{equation}
-
-where $t$ is age in days. A graph of the transform is shown in Figure \@ref(fig:gplot).
-
-```{r gplot, fig.cap = "The g transform.",fig.height=3.2}
-x <- seq(0, 365, length = 1000)
-y <- g(x)
-plot(x, y, type = "l", ylab = "g(t)", xlab = "t (days)", col = "blue", axes = FALSE)
-abline(h = 0, v = 0)
-##
-lines(c(28, 28), c(0, g(28)), lty = 3)
-lines(c(0, 28), c(g(28), g(28)), lty = 3)
-##
-lines(c(183, 183), c(0, g(183)), lty = 3)
-lines(c(0, 183), c(g(183), g(183)), lty = 3)
-##
-lines(c(365, 365), c(0, g(365)), lty = 3)
-lines(c(0, 365), c(g(365), g(365)), lty = 3)
-##
-axis(1, at = c(0, 28, 183, 365))
-ycor <-  c(0, g(28), g(183), g(365))
-axis(2, at = ycor, labels = round(ycor), las = 1)
-box()
-```
-
-It was used as a tool for dividing infant mortality 
-into *endogenous* and *exogenous* causes, and a an important part in that venue was
-the observations (i) postneonatal mortality is purely exogenous, and (ii) the distribution 
-of exogenous mortality is *truncated uniform* (right truncated at $t = g(365)$) on the
-$g$-transformed time scale.
-
-Today, assumption (ii) is slightly outdated, for three reasons: (a) The uniform distribution
-is not a very practical tool in survival analysis, (b) replacing the uniform with an
-*exponential* not only fits better (usually), it also is a survival distribution that
-is extremely easy to work with in survival analysis, and (c) "all models are wrong, but some are useful [@box76]".
-
-```{r getpneo}
-pneo <- age.window(infdat, c(28 / 365, 1))
-gpneo <- pneo
-gpneo$enter <- g(gpneo$enter * 365)
-gpneo$exit <- g(gpneo$exit * 365)
-##
-oj <- system.time(gtpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + illeg + parity + 
-                    lowTemp + highTemp + season + aver
-                + emeantemp + subreg, 
-                data = gpneo, cuts = g(c(28, 365))))
-```
-
-For now, let us look at the basic distribution of postneonatal life on the $g$ time
-scale. We do it nonparametrically by calculating the *Nelson-Aalen* estimator of the
-cumulative hazards function. See Figure \@ref(fig:nap), where the two curves are indistinguishable.
-
-```{r nap, fig.cap = "Nelson-Aalen estimator and exponential cumulative hazards of postneonatal survival on the g scale.",fig.height=3.2}
-par(cex = 0.8)
-fit <- coxreg(Surv(enter, exit, event) ~ 1, data = gpneo)
-xx <- hazards(fit)
-plot(xx[[1]][, 1], cumsum(xx[[1]][, 2]), type = "s", xlab = "Age (days)",
-     col = "blue", ylab = "Cumulative hazards", axes = FALSE)
-axis(1, at = g(c(28, 183, 365)), labels = c(28, 183, 365))
-axis(2)
-box()
-fit.w <- tpchreg(oe(event, exposure) ~ 1, data = gtpneo, time = age)
-zz <- hazards(fit.w)
-lines(zz$x, zz$y, col = "red", lty = 2)
-legend("topleft", legend = c("Nelson-Aalen", "Exponential"), lty = 1:2, col = c("blue", "red"))
-```
-
-The survival data set we work with here contains 102 thousand infants and 4.9 million records,
-an average of 48 records per infant, one record per calendar week in life. In total there are 
-`r sum(gpneo$event)` deaths, and a typical Cox regression execution time is slightly
-below one minute. A "naive" application of an exponential regression model to the same data
-is typically using around six minutes.
-
-However, with the exponential model (or generally, the *piecewise constant hazard* (pch)) in mind, data can be heavily reduced by *tabulation*, and the
-statistical *sufficiency principle* will guarantee that we get identical results.
-The tabulation itself takes about 20 seconds, and the analysis after that about 15 seconds, a
-total of 35 seconds. But the tabulation usually only needs to be done once. More about this later.
-
-### Winter and Spring
+### Winter
 
 ```{r postanalw, results='asis'}
-winter <- gtpneo[gtpneo$season %in% c("winter", "spring"), ]
-winter$season <- factor(winter$season)
+##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch + illeg + parity
+  ##              + subreg + period, data = infdat, cuts = c(0, 1))
+##winter <- gtpneo[gtpneo$summer == 0, ]
+##winter$season <- factor(winter$season)
 
 fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
-                        illeg + parity + subreg, 
-                    data = winter, time = age)
+                        parity + subreg + period + summer, data = tabpost[!tabpost$summer, ], time = age)
 dr <- drop1(fit.tpch, test = "Chisq")
-ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.
-Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = 4)
+ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
 ```
@@
 
-### Summer and Fall
+### Summer
 
 ```{r postanals, results='asis'}
-summer <- gtpneo[gtpneo$season %in% c("summer", "fall"), ]
-winter$season <- factor(winter$season)
+##summer <- gtpneo[gtpneo$summer == 1, ]
+##winter$season <- factor(winter$season)
 
 fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
-                        illeg + parity + subreg, 
-                    data = summer, time = age)
+                        parity + subreg + period, 
+                    data = tabpost[tabpost$summer, ], time = age)
 dr <- drop1(fit.tpch, test = "Chisq")
-ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring.
-Adjusted for social status and branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = 4)
+ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, April to September.
+Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
 ```
@@
 
-# The biometric analysis of infant mortality
-
-A modern version of the Bourgeois-Pichat biometric modelling [@bp51a; @bp51b] is given here,
-*without* the "log-cube transform".
-
-```{r bpprim, fig.cap = "Postneonatal survival is Weibull vs. Nelson-Aalen.", echo = TRUE}
-fit.w <- phreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
-fit <- coxreg(Surv(enter - 28 / 365, exit - 28 / 365, event) ~ 1, data = pneo)
-plot(fit, fn = "cum", col = "blue")
-xx <- hazards(fit.w)
-lines(xx$x, xx$y, col = "red", lty = 2)
-```
-
-Not very good fit. However, we need a *left-truncated Weibull distribution*, that is, assuming that
-*exogeneous infant mortality* is Weibull distributed. Look here (Figure \@ref(fig:ltruw)). A perfect fit, 
-just like the one under the $g$- transform. This is the way to go, *more will come tomorrow*!
-
-```{r ltruw, fig.cap = "Postneonatal survival is truncated Weibull vs. Nelson-Aalen.", echo = TRUE}
-fit.w <- phreg(Surv(enter, exit, event) ~ 1, data = pneo)
-fit <- coxreg(Surv(enter, exit, event) ~ 1, data = pneo)
-plot(fit, fn = "cum", col = "blue")
-xx <- hazards(fit.w)
-lines(xx$x, xx$y - xx$y[1], col = "red", lty = 2)
-```
-
 # Conclusion
 
-Remains to be written.
+Remains to be written, especially the discussion about temperature and mortality.

### 2022-02-08 getdata4.Rmd
fix errors and remove socStatus

@@
 per$socStatus <- as.character(per$socStatus)
-per$socStatus[is.na(per$socStatus)] <- "unknown"
+per$socStatus[is.na(per$socStatus)] <- "low"
 per$socStatus <- factor(per$socStatus)

### 2022-02-08 paper4.Rmd
fix errors and remove socStatus

@@
 - \usepackage{a4wide}
-#- \usepackage[figuresonly]{endfloat}
-#- \renewcommand{\efloatseparator}{\mbox{}}
+- \usepackage[tablesonly]{endfloat}
+- \renewcommand{\efloatseparator}{\mbox{}}
 abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895-1950 is studied in a biometric analysis setting.
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA)
+knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA, cache= TRUE) ## NOTE!!
 knitr::opts_knit$set(global.par = TRUE)
@@
 infdat$emeantemp <- round(infdat$emeantemp)
-levels(infdat$socBranch) <- c("office", "farming", "office", "worker", "worker") # NOTE!
+levels(infdat$socBranch) <- c("office", "farming", "none", "office", "worker") # NOTE!
 ```
@@
 
+Comments on other candidate covariates:
+
+*Birth month* is left out in the analyses despite that fact that
+it is an important factor in neonatal mortality. However, we include *time of year* in terms of 
+*winter* and *summer* as a time-varying covariate, and in the neonatal case it will coincide to a
+great extent with birth period. For the postneonal case the situation is different, but it turns out that
+for those infants who have survived the first month of life, birth month does not matter much.
 We separate the investigation into two parts, *neonatal* and *postneonatal* mortality. But first,
 a joint analysis.
+
+*Socioeconomic status* is divided into two factor covariates: *socBranch* and *socStatus*. The latter should be 
+seen relative to actual *socBranch*. We have information on whether the infant was *illegitimate* (mother unmarried),
+but we incorporate those cases in the category *none* of *socBranch* and *unknown* as category of *socStatus*.
 
@@
 ##all$exit <- round(all$exit * 365)
-fit.all <- coxph(Surv(enter, exit, event) ~ strata(ageIvl) * (aver + emeantemp + lowTemp + highTemp + socBranch + socStatus + sex + period),
-                 data = all)
-dr.all <- drop1(fit.all, test = "Chisq")
+##fit.all <- coxph(Surv(enter, exit, event) ~ strata(ageIvl) * (aver + emeantemp + lowTemp + highTemp + socBranch + ##socStatus + sex + period),
+  ##               data = all)
+##dr.all <- drop1(fit.all, test = "Chisq")
 
-inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + birthmonth + period + socBranch + socStatus + parity + 
+inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socBranch + socStatus + parity + 
                      aver + emeantemp + lowTemp + highTemp + summer + ageIvl, data = all,
@@
 tabneo$age <- factor(tabneo$age)
+fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * (aver + emeantemp + lowTemp + highTemp + socBranch + socStatus + sex + period) + emeantemp, data = inftab, time = age)
+dr.all <- drop1(fit.all, test = "Chisq")
 ##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ aver + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
@@
 ```{r shownonprop}
-tbl(dr.all, caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
+tbl(round(dr.all, 4), caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
 ```
+
+From Table \@ref(tab:shownonprop) we can see that the variables *emeantemp*, *lowTemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case.
+In terms of
+*infant mortality*, we can say that these caovariates have non-proportional effects. As an example, see Figure
 
@@
 Summer covers the months April to September, and winter the rest of the months.
+
+```{r formneo}
+formneo <- as.formula("oe(event, exposure) ~  lowTemp + highTemp + aver + emeantemp + 
+                     sex + subreg + parity + socStatus + period")
+```
 
@@
 ##library(survival)
-winter <- neo[!neo$summer, ]
+winter <- tabneo[!tabneo$summer, ]
 ##winter$season <- factor(winter$season)
@@
 ##neo$enter <- g(neo$enter * 365)
-fit.cr <- coxreg(Surv(enter, exit, event) ~  highTemp + aver + emeantemp + 
-                     sex + subreg + illeg + parity + socBranch + period,
-                 data = winter)
+fit.cr <- tpchreg(formneo, data = winter, time = age)
 dr <- drop1(fit.cr, test = "Chisq")
@@
 ##neo$enter <- g(neo$enter * 365)
-fit.cr <- tpchreg(oe(event, exposure) ~  aver + emeantemp + 
-                     sex + subreg  + socBranch + period,
-                 data = inftab[inftab$summer, ], time = age)
+fit.cr <- tpchreg(formneo, data = tabneo[tabneo$summer, ], time = age)
 dr <- drop1(fit.cr, test = "Chisq")
@@
 
-fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
-                        parity + subreg + period + summer, data = tabpost[!tabpost$summer, ], time = age)
+fit.tpch <- tpchreg(oe(event, exposure) ~ period * (lowTemp + highTemp + aver + emeantemp) + sex + socBranch  + parity + subreg + period, data = tabpost[!tabpost$summer, ], time = age)
 dr <- drop1(fit.tpch, test = "Chisq")
-ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
+tbl(round(dr, 4), caption = "Interaction tests with period, postneonatal mortality.", rownames = TRUE)
+##ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, October to March.
+Adjusted for social branch, sex, ##illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
 ```
 
-The result in Table \@ref(tab:postanalw) shows that *climate* (`emintemp`) is more important than 
-*weather* (`excessTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
-covariates (not shown).
+The result in Table \@ref(tab:postanalw) shows that the temperature effects vary with time period,
+so we continue with seaparate analyses for the time periods `r levels(tabpost$period)`.
+
+```{r setformw}
+form <- as.formula("oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socStatus + parity + subreg")
+```
+
+#### The time period 1895--1913
+
+```{r postw1, results = "asis"}
+fit.pw1 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]", ], time = age)
+dr <- drop1(fit.pw1, test = "Chisq")
+ltx2(fit.pw1, dr = dr, caption = "Postneonatal mortality, October to March 1895-1914. Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw1", keep = NULL)
+```
+
+Table \@ref(tab:postw1) shows that
+
+#### The time period 1914--1934
+
+```{r postw2, results = "asis"}
+fit.pw2 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]", ], time = age)
+dr2 <- drop1(fit.pw2, test = "Chisq")
+ltx2(fit.pw2, dr = dr2, caption = "Postneonatal mortality, October to March 1914-1934. Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw2", keep = NULL)
+```
+
+Table \@ref(tab:postw2) shows that
+
+#### The time period 1935--1950
+
+```{r postw3, results = "asis"}
+fit.pw3 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ], time = age)
+dr3 <- drop1(fit.pw3, test = "Chisq")
+ltx2(fit.pw3, dr = dr3, caption = "Postneonatal mortality, October to March 1935-1950. Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw3", keep = NULL)
+```
+
+Table \@ref(tab:postw3) shows that
 

### 2022-02-09 R/normal.R
first commit

@@ -1,0 +1,72 @@
+normal <- function(period = c(1894, 1951)){
+    ## "normal" stands for "norSJÖ + malÅ"
+
+    
+    bjur = "~/Forskning/Data/bjur_temp.csv"
+    sten = "~/Forskning/Data/sten_temp.csv"
+    stentemp <- read.table(sten, 
+                           header = TRUE,
+                           skip = 10, sep = ";")[, 1:3]
+    bjurtemp <- read.table(bjur,
+                           header = TRUE,
+                           skip = 10, sep = ";")[, 1:3]
+    cre <- function(temp){
+    ## We need only the first three columns
+        names(temp) <- c("date", "time", "temp")
+        temp$date <- as.Date(temp$date)
+        temp$year <- as.numeric(format(temp$date, "%Y"))
+        temp$month <- as.numeric(format(temp$date, "%m"))
+        temp <- temp[order(temp$date, temp$time), ]
+        temp$quarter <- quarters(temp$date)
+        temp
+    
+    
+        temp <- temp[temp$year %in% period[1]:period[2], ]
+        indx <- tapply(temp$date, temp$date)
+        temp$mintemp <- tapply(temp$temp, temp$dat, min)[indx]
+        temp$maxtemp <- tapply(temp$temp, temp$dat, max)[indx]
+        temp$meantemp <- tapply(temp$temp, temp$dat, mean)[indx]
+        temp <- temp[!duplicated(temp$date), 
+                     c("date", "year", "month", "quarter", "mintemp", "maxtemp", "meantemp")]
+    ## Create "week", 1:52 each year, week 52 with 8(9) days:
+        get_week <- function(days){ # days = vector of 365 or 366, length of year
+            n <- length(days)
+            tmp <- c(rep(1:52, each = 7), 52, 52)
+            tmp[1:n]
+        }
+        temp$week <- as.numeric(unlist(tapply(temp$year, temp$year, get_week)))
+        temp
+    }
+    bt <- cre(bjurtemp)
+    st <- cre(stentemp)
+    temp <- st[st$week %in% c(1:12, 38:52), ]
+    temp <- rbind(temp, bt[bt$week %in% 13:37, ])
+    temp <- temp[order(temp$date), ]
+    
+    ## Calculate weekly extremes of min and max temp
[…]

### 2022-02-09 getdata4.Rmd
change name of aver (extemp) and order

@@
 lagg <- 0 / 365 ##3 / 365  ## NOTE!!
-tempvars <-  c("heat", "heat.1", "aver", "aver.1", "cold", "cold.1", "emintemp", "emeantemp", "emaxtemp", 
+tempvars <-  c("heat", "heat.1", "extemp", "extemp.1", "cold", "cold.1", "emintemp", "emeantemp", "emaxtemp", 
                "week", "year") 

### 2022-02-09 paper4.Rmd
change name of aver (extemp) and order

@@
 levels(infdat$subreg) <- c("ume", "ske", "inland")
-##infdat$excessTemp <- cut(infdat$aver, c(-16, -2, 2, 9), include.lowest = TRUE) # NOTE!!!
+##infdat$excessTemp <- cut(infdat$extemp, c(-16, -2, 2, 9), include.lowest = TRUE) # NOTE!!!
 ##infdat$excessTemp <- relevel(infdat$excessTemp, ref = "(-2,2]")
-##infdat$lowTemp <- (infdat$aver + infdat$aver.1) < -10  # CHECK THIS!!
-##infdat$highTemp <- infdat$aver + infdat$aver.1 > 10
-##infdat$lowTemp <- with(infdat, (aver < -3) & (aver.1 < -3))
+##infdat$lowTemp <- (infdat$extemp + infdat$extemp.1) < -10  # CHECK THIS!!
+##infdat$highTemp <- infdat$extemp + infdat$extemp.1 > 10
+##infdat$lowTemp <- with(infdat, (extemp < -3) & (extemp.1 < -3))
 infdat$lowTemp <- (with(infdat, cold + cold.1 < -10))
-infdat$highTemp <- with(infdat, (aver > 3) & (aver.1 > 3))
-tbl(head(infdat[, c("enter", "exit", "event", "lowTemp", "highTemp", "aver", "emeantemp", "week", "year")]),
+infdat$highTemp <- with(infdat, (extemp > 3) & (extemp.1 > 3))
+tbl(head(infdat[, c("enter", "exit", "event", "lowTemp", "highTemp", "extemp", "emeantemp", "week", "year")]),
     caption = "Data with communal covariates.", fs = 9)
@@
 
-```{r joint, cache = TRUE}
+```{r managedata}
 library(survival)
@@
 ##all$exit <- round(all$exit * 365)
-##fit.all <- coxph(Surv(enter, exit, event) ~ strata(ageIvl) * (aver + emeantemp + lowTemp + highTemp + socBranch + ##socStatus + sex + period),
+##fit.all <- coxph(Surv(enter, exit, event) ~ strata(ageIvl) * (extemp + emeantemp + lowTemp + highTemp + socBranch + ##socStatus + sex + period),
   ##               data = all)
@@
 inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socBranch + socStatus + parity + 
-                     aver + emeantemp + lowTemp + highTemp + summer + ageIvl, data = all,
+                     extemp + emeantemp + lowTemp + highTemp + summer + ageIvl, data = all,
                  cuts = c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365) / 365)
@@
 tabneo$age <- factor(tabneo$age)
-fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * (aver + emeantemp + lowTemp + highTemp + socBranch + socStatus + sex + period) + emeantemp, data = inftab, time = age)
+```
+
+```{r joint}
+fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * 
+                       (extemp + emeantemp + lowTemp + highTemp + socBranch + period), 
+                   data = inftab, time = age)
 dr.all <- drop1(fit.all, test = "Chisq")
-##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ aver + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
+##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
   ##               data = tabneo, time = age))
-##system.time(tabpost.fit <- tpchreg(oe(event, exposure) ~ aver + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
+##system.time(tabpost.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
   ##               data = tabpost, time = age))
@@
 
-```{r shownonprop}
+```{r shownonprop, cache = FALSE}
 tbl(round(dr.all, 4), caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
@@
 ```{r formneo}
-formneo <- as.formula("oe(event, exposure) ~  lowTemp + highTemp + aver + emeantemp + 
-                     sex + subreg + parity + socStatus + period")
+formneo <- as.formula("oe(event, exposure) ~  lowTemp + highTemp + extemp + emeantemp + socBranch + period +
+                     subreg + sex + parity")
 ```
@@
 above the expected for at least two weeks in a row, *emeantemp* the expected temperature
-the actual weeek, and *aver* the *excess temperature* the actual week.
+the actual weeek, and *extemp* the *excess temperature* the actual week.
 
-```{r neoanalw, results = 'asis'}
+```{r neoanalwp, results = 'asis'}
 ##library(survival)
@@
 ##neo$enter <- g(neo$enter * 365)
-fit.cr <- tpchreg(formneo, data = winter, time = age)
-dr <- drop1(fit.cr, test = "Chisq")
-ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
+fit.neow <- tpchreg(formneo, data = winter, time = age)
+dr <- drop1(fit.neow, test = "Chisq")
+```
+
+```{r neoanalw, results = "asis", cache = FALSE}
+ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
      label = "tab:neoanalw", keep = NULL)
@@
 above the expected for at least two weeks in a row, *emeantemp* the expected temperature
-the actual weeek, and *aver* the *excess temperature* the actual week.
+the actual weeek, and *extemp* the *excess temperature* the actual week.
 
-```{r neoanals, results = 'asis'}
+```{r neoanalsp}
 ##library(survival)
@@
 ##neo$enter <- g(neo$enter * 365)
-fit.cr <- tpchreg(formneo, data = tabneo[tabneo$summer, ], time = age)
-dr <- drop1(fit.cr, test = "Chisq")
-ltx2(fit.cr, dr = dr, caption = "Neonatal mortality, April to September. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
+fit.neos <- tpchreg(formneo, data = tabneo[tabneo$summer, ], time = age)
+dr.neos <- drop1(fit.neos, test = "Chisq")
+```
+
+```{r neoanals, results = 'asis', cache = FALSE}
+ltx2(fit.neos, dr = dr.neos, caption = "Neonatal mortality, April to September. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
      label = "tab:neoanals", keep = NULL)
@@
 
-```{r postanalw, results='asis'}
-##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch + illeg + parity
+```{r postanalwp, results='asis'}
+##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socBranch + illeg + parity
   ##              + subreg + period, data = infdat, cuts = c(0, 1))
@@
 
-fit.tpch <- tpchreg(oe(event, exposure) ~ period * (lowTemp + highTemp + aver + emeantemp) + sex + socBranch  + parity + subreg + period, data = tabpost[!tabpost$summer, ], time = age)
-dr <- drop1(fit.tpch, test = "Chisq")
-tbl(round(dr, 4), caption = "Interaction tests with period, postneonatal mortality.", rownames = TRUE)
+fit.postw <- tpchreg(oe(event, exposure) ~ period * (lowTemp + highTemp + extemp + emeantemp), 
+                     data = tabpost[!tabpost$summer, ], time = age)
+dr.postw <- drop1(fit.postw, test = "Chisq")
+```
+
+```{r postanalw, results='asis', cache = FALSE}
+tbl(round(dr.postw, 4), caption = "Interaction tests with temperature and period, postneonatal mortality.", rownames = TRUE)
 ##ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, October to March.
@@
 ```{r setformw}
-form <- as.formula("oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socStatus + parity + subreg")
+form <- as.formula("oe(event, exposure) ~ lowTemp + highTemp + extemp + emeantemp + socStatus + subreg + sex + parity")
 ```
@@
 
-```{r postw1, results = "asis"}
-fit.pw1 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]", ], time = age)
-dr <- drop1(fit.pw1, test = "Chisq")
-ltx2(fit.pw1, dr = dr, caption = "Postneonatal mortality, October to March 1895-1914. Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw1", keep = NULL)
+```{r postw1p, results = "asis"}
+fit.postw1 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]", ], time = age)
+dr.postw1 <- drop1(fit.postw1, test = "Chisq")
+```
+
+```{r postw1, results = "asis", cache = FALSE}
+ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914.
+Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw1", keep = NULL)
 ```
@@
 
-```{r postw2, results = "asis"}
-fit.pw2 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]", ], time = age)
-dr2 <- drop1(fit.pw2, test = "Chisq")
-ltx2(fit.pw2, dr = dr2, caption = "Postneonatal mortality, October to March 1914-1934. Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw2", keep = NULL)
+```{r postw2p, results = "asis"}
+fit.postw2 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]", ], time = age)
+dr.postw2 <- drop1(fit.postw2, test = "Chisq")
+```
+
+```{r postw2, results = "asis", cache = FALSE}
+ltx2(fit.postw2, dr = dr.postw2, caption = "Postneonatal mortality, October to March 1914-1934.
+Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw2", keep = NULL)
 ```
@@
 
-```{r postw3, results = "asis"}
-fit.pw3 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ], time = age)
-dr3 <- drop1(fit.pw3, test = "Chisq")
-ltx2(fit.pw3, dr = dr3, caption = "Postneonatal mortality, October to March 1935-1950. Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw3", keep = NULL)
+```{r postw3p, results = "asis"}
+fit.postw3 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ], time = age)
+dr.postw3 <- drop1(fit.postw3, test = "Chisq")
+```
+
+```{r postw3, results = "asis", cache = FALSE}
+ltx2(fit.postw3, dr = dr.postw3, caption = "Postneonatal mortality, October to March 1935-1950.
+Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw3", keep = NULL)
 ```
@@
 
-```{r postanals, results='asis'}
+```{r postanalsp, results='asis'}
 ##summer <- gtpneo[gtpneo$summer == 1, ]
@@
 
-fit.tpch <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + aver + emeantemp + sex + socBranch  + 
-                        parity + subreg + period, 
+fit.posts <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + extemp + emeantemp + socBranch + period + 
+                        subreg + parity + sex, 
                     data = tabpost[tabpost$summer, ], time = age)
-dr <- drop1(fit.tpch, test = "Chisq")
-ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, April to September.
+dr <- drop1(fit.posts, test = "Chisq")
+```
+
+```{r postanals, results='asis', cache = FALSE}
+ltx2(fit.posts, dr = dr, caption = "Postneonatal mortality, April to September.
 Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)

### 2022-02-09 R/tempdat.R

@@ -45,11 +45,11 @@ tempdat <- function(filename = "ume", period = c(1894, 1951)){
     temp$emeantemp <- tapply(temp$meantemp, temp$week, mean)[indx]
     temp$heat <- round(temp$maxtemp - temp$emaxtemp)
     temp$cold <- round(temp$mintemp - temp$emintemp)
-    temp$aver <- round(temp$meantemp - temp$emeantemp)
+    temp$extemp <- round(temp$meantemp - temp$emeantemp)
     # NEW 28 Jan 2022:
     n <- NROW(temp)
     temp$cold.1 <- c(NA, temp$cold[-n])
-    temp$aver.1 <- c(NA, temp$aver[-n])
+    temp$extemp.1 <- c(NA, temp$extemp[-n])
     temp$heat.1 <- c(NA, temp$heat[-n])
     ##
     temp

### 2022-02-09 getdata4.Rmd
new births.rds

@@
                "subreg", "socBranch", "socStatus", "hisclass", "enter", "exit", "event")]
-#saveRDS(per, file = "data/infdat3.rds")
+saveRDS(per, file = "data/births4.rds")
 ```

### 2022-02-09 paper4.Rmd
new births.rds

@@
 More precisely, the effect of climate and weather, measured by temperature, average and extremes, on infant mortality is investigated.
-It turns out that climate (average) is more important than weather (extremes), low average temperatures are more important than temporary dips in temperature."
+It turns out that climate (average) is more important than weather (extremes), low average temperatures are more important than temporary dips in temperature, but effects are different in neonatal and postneonatal settings."
 ---
@@
 library(eha)
+starttid <- proc.time()
 ```
@@
 ```{r readdata}
-births <- readRDS("data/births.rds")
+births <- readRDS("data/births4.rds")
 infdat  <- readRDS("data/infdat4.rds")
@@
 Remains to be written, especially the discussion about temperature and mortality.
+
+Used processing time (seconds):
+
+```{r tidtagning}
+sluttid <- Sys.time()
+proc.time() - starttid
+```
+
+Done: `r sluttid`.

### 2022-02-09 paper4.Rmd
fix details and short tables.

@@
 The impact of ambient temperature variations on infant mortality is studied for 
-a northern Sweden coastal area, the Umeå and Skellefteå regions, during the 
+two northern Sweden areas, the Umeå coastal region and Skellefteå coastal and inland regions, during the 
 first half of the twentieth century. Two recent papers [@junkka; @karlsson] studied 
-neonatal mortality and temperature variations in a larger geographical area containing 
-the present one during the 
+neonatal mortality and temperature variations in this geographical area during the 
 years 1880--1950. Climate and mortality in general is a research area that has 
-generated great interest over the last years, see @tbgb10.
+generated great interest over the last years, see @tbgb10 for an example of our own efforts.
 
 The effect of seasonal variation and the occurrence of extreme monthly temperatures
-is studied and interacted with sex, social class, and legitimacy. Studies are performed 
-separately for neonatal  and postneonatal mortality, and for winter and summer seasons, 
-and the classification into endogenous and exogenous factors will be discussed.
+is studied and interacted with social class and time period. Studies are performed 
+separately for neonatal  and postneonatal mortality, and for winter and summer seasons.
 
@@
 
-**socStatus** Social status of family, based on HISCLASS.
-
 **illeg** Mother unmarried?
@@
 
-Figure \@ref(fig:mortstat) shows the average monthly crude infant mortality, and a clear seasonal pattern is visible.
+Figure \@ref(fig:mortstat) shows the average weekly crude infant mortality, and a clear seasonal pattern is visible.
 
@@
 
-The average monthly neonatal mortality is shown in Figure \@ref(fig:neodata).
+The average weekly neonatal mortality is shown in Figure \@ref(fig:neodata).
 
@@
 
-The average monthly postneonatal mortality is shown in Figure \@ref(fig:pneodata).
+The average weekly postneonatal mortality is shown in Figure \@ref(fig:pneodata).
 
@@
 
+Note that the estimates of neonatal and postneonatal mortality are non-standard here: 
+They are calculated as *occurrence-exposure rates*, that is, the number of deaths
+divided by *exposure time*. They are thus crude *hazard rates*, not probabilities, which is 
+more common in definitions of mortality.
+
 ## Temperature
@@
 the weekly data are summarized in a few measurements, see Table \@ref(tab:cond).
+Our rule for week numbering is that week No. 1 always start on January 1. Week No. 52
+ always ends at December 31, and so will be eight days long, except for leap years, when
+ it will be nine days long.
 
@@
 
+Covariates may affect neonatal and postneonatal mortality differently, and one way to 
+investigate that is to stratify the infant mortality data into two age intervals, one from
+birth to one month of age and the other from one month to one year of age. Then the
+interaction between the stratum variable and other variables of interest is investigated, 
+with this result:
+
 ```{r shownonprop, cache = FALSE}
-tbl(round(dr.all, 4), caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
+dr.all
+##tbl(round(dr.all, 4), caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
 ```
 
-From Table \@ref(tab:shownonprop) we can see that the variables *emeantemp*, *lowTemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case.
-In terms of
-*infant mortality*, we can say that these caovariates have non-proportional effects. As an example, see Figure
+We can see that the variables *emeantemp*, *lowTemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case. In terms of 
+*infant mortality*, we can say that these caovariates have non-proportional effects.
 
@@
 ```{r neoanalw, results = "asis", cache = FALSE}
-ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
+ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March.", 
      label = "tab:neoanalw", keep = NULL)
@@
 ```{r neoanals, results = 'asis', cache = FALSE}
-ltx2(fit.neos, dr = dr.neos, caption = "Neonatal mortality, April to September. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
+ltx2(fit.neos, dr = dr.neos, caption = "Neonatal mortality, April to September.", 
      label = "tab:neoanals", keep = NULL)
@@
 
-```{r postanalw, results='asis', cache = FALSE}
-tbl(round(dr.postw, 4), caption = "Interaction tests with temperature and period, postneonatal mortality.", rownames = TRUE)
-##ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, October to March.
-Adjusted for social branch, sex, ##illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
+Interaction tests with temperature and period, postneonatal mortality:
+
+```{r postanalw, cache = FALSE}
+round(dr.postw, 4)
+#tbl(round(dr.postw, 4), caption = "Interaction tests with temperature and period, postneonatal ##mortality.", rownames = TRUE)
 ```
 
-The result in Table \@ref(tab:postanalw) shows that the temperature effects vary with time period,
-so we continue with seaparate analyses for the time periods `r levels(tabpost$period)`.
+This shows that the temperature effects do not vary with time period,
+but we continue with separate analyses for the time periods `r levels(tabpost$period)`. Just
+to make sure, will be joined when appropriate.
 
 ```{r setformw}
-form <- as.formula("oe(event, exposure) ~ lowTemp + highTemp + extemp + emeantemp + socStatus + subreg + sex + parity")
+form <- as.formula("oe(event, exposure) ~ lowTemp + highTemp + extemp + emeantemp + socBranch + subreg + sex + parity")
 ```
@@
 ```{r postw1, results = "asis", cache = FALSE}
-ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914.
-Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw1", keep = NULL)
+ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914. Adjusted for social branch, sex, parity, and subregion.", label = "tab:postw1", keep = 5)
 ```
@@
 ```{r postw2, results = "asis", cache = FALSE}
-ltx2(fit.postw2, dr = dr.postw2, caption = "Postneonatal mortality, October to March 1914-1934.
-Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw2", keep = NULL)
+ltx2(fit.postw2, dr = dr.postw2, caption = "Postneonatal mortality, October to March 1914-1934. Adjusted for sex, parity, and subregion.", label = "tab:postw2", keep = 5)
 ```
@@
 ```{r postw3, results = "asis", cache = FALSE}
-ltx2(fit.postw3, dr = dr.postw3, caption = "Postneonatal mortality, October to March 1935-1950.
-Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw3", keep = NULL)
+ltx2(fit.postw3, dr = dr.postw3, caption = "Postneonatal mortality, October to March 1935-1950. Adjusted for sex, parity, and subregion.", label = "tab:postw3", keep = 5)
 ```
@@
 ```{r postanals, results='asis', cache = FALSE}
-ltx2(fit.posts, dr = dr, caption = "Postneonatal mortality, April to September.
-Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
+ltx2(fit.posts, dr = dr, caption = "Postneonatal mortality, April to September. Adjusted for sex, parity, time period, and subregion.", label = "tab:postanals", keep = 5)
 ```
 
-The result in Table \@ref(tab:postanals) shows that *climate* (`emintemp`) is more important than 
-*weather* (`excessTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
+The result in Table \@ref(tab:postanals) shows that *climate* (`emeantemp`) is more important than 
+*weather* (`exTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
 covariates (not shown).

### 2022-02-10 paper4.Rmd
add interaction checks for postneonatal ...

@@
 - \usepackage{a4wide}
-- \usepackage[tablesonly]{endfloat}
-- \renewcommand{\efloatseparator}{\mbox{}}
+#- \usepackage[tablesonly]{endfloat}
+#- \renewcommand{\efloatseparator}{\mbox{}}
 abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895-1950 is studied in a biometric analysis setting.
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA, cache= TRUE) ## NOTE!!
+knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA) ## NOTE!!
 knitr::opts_knit$set(global.par = TRUE)
@@
 
-```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
+```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
 ##infdat <- readRDS("data/infdat4.rds") Already done
@@
 
-```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
+```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
 neo <- age.window(infdat, c(0, 28 / 365))
@@
 
-```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
+```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
 pneo <- age.window(infdat, c(28 / 365, 1))
@@
 
-```{r longtrend, fig.cap = "Crude IMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 3}
+```{r longtrend, fig.cap = "Crude IMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
 trend <- toTpch(Surv(enter, exit, event) ~ year, data = infdat, cuts = c(0, 1))
@@
 
-```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
+```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4, cache = TRUE}
 ndat <- age.window(infdat, c(0, 28 / 365))
@@
 
-```{r monavgtemp, fig.cap="Weekly max, mean, and min temperature averages, 1895--1950.", fig.height = 3}
+```{r monavgtemp, fig.cap="Weekly max, mean, and min temperature averages, 1895--1950.", fig.height = 3, cache = TRUE}
 oldpar <- par(mfrow = c(1, 3))
@@
 
-```{r longterm, fig.cap = "Yearly average temperatures, Umeå and Skellefteå."}
+```{r longterm, fig.cap = "Yearly average temperatures, Umeå and Skellefteå.", cache = TRUE}
 bmtemp <- rbind(umetemp, bjurtemp, stentemp)
@@
 ```{r managedata}
-library(survival)
+if (!(file.exists("data/inftab.rds") && file.exists("data/tabneo.rds") && file.exists("data/tabpost.rds"))){
     neo <- age.window(infdat, c(0, 28 / 365))
@@
     inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socBranch + socStatus + parity + 
-                     extemp + emeantemp + lowTemp + highTemp + summer + ageIvl, data = all,
+                         extemp + extemp.1 + emeantemp + summer + ageIvl, data = all,
                      cuts = c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365) / 365)
@@
     tabneo$age <- factor(tabneo$age)
+    saveRDS(tabneo, file = "data/tabneo.rds")
+    saveRDS(tabpost, file = "data/tabpost.rds")
+    saveRDS(inftab, file = "data/inftab.rds")
+}else{
+    inftab <- readRDS("data/inftab.rds") #The rest are read later...
+}
 ```
 
-```{r joint}
+```{r joint, cache = TRUE}
 fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * 
-                       (extemp + emeantemp + lowTemp + highTemp + socBranch + period), 
+                       (extemp + extemp.1 + emeantemp + socBranch + period), 
                    data = inftab, time = age)
@@
 
-We can see that the variables *emeantemp*, *lowTemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case. In terms of 
+We can see that the variables *emeantemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case. In terms of 
 *infant mortality*, we can say that these caovariates have non-proportional effects.
@@
 ```{r formneo}
-formneo <- as.formula("oe(event, exposure) ~  lowTemp + highTemp + extemp + emeantemp + socBranch + period +
+formneo <- as.formula("oe(event, exposure) ~  extemp + extemp.1 + emeantemp + socBranch + period +
                      subreg + sex + parity")
+tabneo <- readRDS("data/tabneo.rds")
 ```
@@
 
-```{r neoanalwp, results = 'asis'}
+```{r neoanalwp, results = 'asis', cache = TRUE}
+
 ##library(survival)
@@
 ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March.", 
-     label = "tab:neoanalw", keep = NULL)
+     label = "tab:neoanalw", keep = 6)
 ```
@@
 
-```{r neoanalsp}
+```{r neoanalsp, cache = TRUE}
 ##library(survival)
@@
 ltx2(fit.neos, dr = dr.neos, caption = "Neonatal mortality, April to September.", 
-     label = "tab:neoanals", keep = NULL)
+     label = "tab:neoanals", keep = 6)
 ```
@@
 
+```{r setformw}
+form <- as.formula("oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + subreg + sex + parity")
+form.sb <-  as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * socBranch + subreg + sex + parity")
+form.sr <- as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * subreg + socBranch + sex + parity") 
+tabpost <- readRDS("data/tabpost.rds")
+```
+
 ### Winter
 
-```{r postanalwp, results='asis'}
+```{r postanalwp, results='asis', cache = TRUE}
 ##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socBranch + illeg + parity
@@
 
-fit.postw <- tpchreg(oe(event, exposure) ~ period * (lowTemp + highTemp + extemp + emeantemp), 
+fit.postw <- tpchreg(oe(event, exposure) ~ period * (extemp + extemp.1 + emeantemp), 
                      data = tabpost[!tabpost$summer, ], time = age)
@@
 
-This shows that the temperature effects do not vary with time period,
+This shows that the temperature effects do not vary much with time period,
 but we continue with separate analyses for the time periods `r levels(tabpost$period)`. Just
 to make sure, will be joined when appropriate.
-
-```{r setformw}
-form <- as.formula("oe(event, exposure) ~ lowTemp + highTemp + extemp + emeantemp + socBranch + subreg + sex + parity")
-```
 
@@
 ```{r postw1, results = "asis", cache = FALSE}
-ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914. Adjusted for social branch, sex, parity, and subregion.", label = "tab:postw1", keep = 5)
+ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914. Adjusted for sex and parity.", 
+     label = "tab:postw1", keep = 5)
 ```
 
-Table \@ref(tab:postw1) shows that
+Table \@ref(tab:postw1) shows that the effects of  extreme temperatures have the expected directions, extra high temperatures is *positive* 
+(remember that this concerns winter conditions), but not statistically significant. *Climate* (emeantemp), on the other hand shows to be very
+important, where increasing temperature by one degree lower mortality by 0.7 per cent. The normal short term fluctuations (extemp) are not so important, 
+but in the expected direction.
+
+The importance of social class (socBranch) in explaining the effects of climate and weather is negligible, but there is variation between
+the regions regarding *climate*:
+
+```{r postw1srpu, cache = TRUE}
+fit.postw1sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
+                        data = tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]" & tabpost$subreg == "ume", ], time = age)
+dr.postw1sru <- drop1(fit.postw1sru, test = "Chisq")
+```
+
+```{r postw1sru, results = "asis"}
+ltx2(fit.postw1sru, dr = dr.postw1sru, caption = "Postneonatal mortality, Umeå, October to March 1895-1913.",
+     label = "tab:postw1sru", keep = 4)
+```
+
+```{r postw1srps, cache = TRUE}
+fit.postw1srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
+                        data = tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]" & tabpost$subreg != "ume", ], time = age)
+dr.postw1srs <- drop1(fit.postw1srs, test = "Chisq")
+```
+
+```{r postw1srs, results = "asis"}
+ltx2(fit.postw1srs, dr = dr.postw1srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March 1895-1913.", 
+     label = "tab:postw1srs", keep = 4)
+```
+
+We can conclude from Tables \@ref(tab:postw1sru) and \@ref(tab:postw1srs) that the main difference between Umeå and Skellefteå is that
+*climate* has more severe effect in Skellefteå (with inland), while *weather* is less important.
 
@@
 
-Table \@ref(tab:postw2) shows that
+Table \@ref(tab:postw2) shows that the weather effect is lagged one week (*extemp.1*), and that climate continues to be very important.
+Let's look at regional differences.
+
+```{r postw2srpu, cache = TRUE}
+fit.postw2sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
+                        data = tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]" & tabpost$subreg == "ume", ], time = age)
+dr.postw2sru <- drop1(fit.postw2sru, test = "Chisq")
+```
+
+```{r postw2sru, results = "asis"}
+ltx2(fit.postw2sru, dr = dr.postw2sru, caption = "Postneonatal mortality, Umeå, October to March 1914-1934.",
+     label = "tab:postw2sru", keep = 4)
+```
+
+```{r postw2srps, cache = TRUE}
+fit.postw2srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
+                        data = tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]" & tabpost$subreg != "ume", ], time = age)
+dr.postw2srs <- drop1(fit.postw2srs, test = "Chisq")
+```
+
+```{r postw2srs, results = "asis"}
+ltx2(fit.postw2srs, dr = dr.postw2srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March 1914-1934.", 
+     label = "tab:postw2srs", keep = 4)
+```
 
@@
 ```{r postw3p, results = "asis"}
-fit.postw3 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ], time = age)
+fit.postw3 <- tpchreg(oe(event, exposure) ~ extemp.1 + emeantemp + socBranch + subreg + sex + parity, 
+                      data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ], time = age)
+##fit.postw3 <- tpchreg(form, data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ], time = age)
 dr.postw3 <- drop1(fit.postw3, test = "Chisq")
@@
 
+```{r postw3srpu, cache = TRUE}
+fit.postw3sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
+                        data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]" & tabpost$subreg == "ume", ], time = age)
+dr.postw3sru <- drop1(fit.postw3sru, test = "Chisq")
+```
+
+```{r postw3sru, results = "asis"}
+ltx2(fit.postw3sru, dr = dr.postw3sru, caption = "Postneonatal mortality, Umeå, October to March 1935-1950.",
+     label = "tab:postw3sru", keep = 4)
+```
+
+```{r postw3srps, cache = TRUE}
+fit.postw3srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
+                        data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]" & tabpost$subreg != "ume", ], time = age)
+dr.postw3srs <- drop1(fit.postw3srs, test = "Chisq")
+```
+
+```{r postw3srs, results = "asis"}
+ltx2(fit.postw3srs, dr = dr.postw3srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March 1895-1913.", 
+     label = "tab:postw3srs", keep = 4)
+```
+
 ### Summer
 
-```{r postanalsp, results='asis'}
+```{r postanalsp, results='asis', cache = TRUE}
 ##summer <- gtpneo[gtpneo$summer == 1, ]
@@
 
-fit.posts <- tpchreg(oe(event, exposure) ~ lowTemp + highTemp + extemp + emeantemp + socBranch + period + 
+fit.posts <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + period + 
                         subreg + parity + sex, 

### 2022-02-11 paper4.Rmd
add birth month, sex and temp

@@
     toc: yes
-    toc_depth: 3
+    toc_depth: 4
   bookdown::word_document2:
@@
 
+#### Temperature and sex
+
+We check the interaction between temperature and sex and period.
+
+```{r neosexper}
+fit <- tpchreg(oe(event, exposure) ~ sex * (extemp + extemp.1 + emeantemp), 
+               time = age, data = winter)
+dr <- drop1(fit, test = "Chisq")
+dr
+```
+
+Obviously no temperature and sex interactions
+
 ### Summer
@@
 
+#### By birth month
+
+```{r bybmonth, fig.cap = "Crude postneonatal relative (to January = 1) mortality by birth month.", cache = TRUE}
+pn <- age.window(infdat, c(28/365, 1))
+fit <- coxreg(Surv(enter, exit, event) ~ as.factor(birthmonth), data = pn)
+plot(exp(c(0, coef(fit))), type = "b", xlab = "Birth month", ylab = "Hazard ratios",
+     col = "blue")
+abline(h = 1)
+```
+
 ### Winter
@@
 to make sure, will be joined when appropriate.
+
+#### Sex and temperature
+
+```{r postanalwpsex, results='asis', cache = TRUE}
+##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socBranch + illeg + parity
+  ##              + subreg + period, data = infdat, cuts = c(0, 1))
+##winter <- gtpneo[gtpneo$summer == 0, ]
+##winter$season <- factor(winter$season)
+
+fit.postwsex <- tpchreg(oe(event, exposure) ~ sex * (extemp + extemp.1 + emeantemp), 
+                     data = tabpost[!tabpost$summer, ], time = age)
+dr.postwsex <- drop1(fit.postwsex, test = "Chisq")
+```
+
+Interaction tests with temperature and sex, postneonatal mortality:
+
+```{r postanalwsex, cache = FALSE}
+round(dr.postwsex, 4)
+#tbl(round(dr.postw, 4), caption = "Interaction tests with temperature and period, postneonatal ##mortality.", rownames = TRUE)
+```
+
+Obviously no interactions between temperature and sex.
 

### 2022-02-11 paper4.Rmd
fix

@@
 plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
-     xlab = "Week", ylab = "IMR", cex = 0.7, lwd = 1.5)
+     xlab = "Week", ylab = "O/E rate", cex = 0.7, lwd = 1.5)
 axis(1, at = c(1, 10, 20, 30, 40, 52))
@@
 plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
-     xlab = "Week", ylab = "NMR", cex = 0.7, lwd = 1.5)
+     xlab = "Week", ylab = "O/E rate", cex = 0.7, lwd = 1.5)
 axis(1, at = c(1, 10, 20, 30, 40, 52))
@@
 plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
-     xlab = "Week", ylab = "PNMR", cex = 0.7, lwd = 1.5)
+     xlab = "Week", ylab = "O/E rate", cex = 0.7, lwd = 1.5)
 axis(1, at = c(1, 10, 20, 30, 40, 52))
@@
 plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
-     ylab = "IMR", lwd = 1.5, cex = 0.7)
+     ylab = "O/E rate", lwd = 1.5, cex = 0.7)
 abline(h = c(0.05, 0.1, 0.15), lty = 3)
@@
 plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
-     ylab = "NMR", lwd = 1.5, cex = 0.7)
+     ylab = "O/E rate", lwd = 1.5, cex = 0.7)
+abline(h = 0)
+abline(h = c(0.2, 0.4, 0.6), lty = 3)
+```
+
+```{r pntrend, fig.cap = "Crude postneonatal mortality by year, Umeå-Skellefteå 1895--1950.", fig.height = 4, cache = TRUE}
+pndat <- age.window(infdat, c(28 / 365, 1))
+tpnda <- toTpch(Surv(enter, exit, event) ~ year, data = pndat, 
+                cuts = c(28 / 365, 1))
+y <- tpnda$event / tpnda$exposure
+par(las = 1)
+par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
+plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
+     ylab = "O/E rate", lwd = 1.5, cex = 0.7)
 abline(h = 0)
@@
 infdat$highTemp <- with(infdat, (extemp > 3) & (extemp.1 > 3))
-tbl(head(infdat[, c("enter", "exit", "event", "lowTemp", "highTemp", "extemp", "emeantemp", "week", "year")]),
+tbl(head(infdat[, c("enter", "exit", "event", "extemp", "extemp.1", "emeantemp", "week", "year")]),
     caption = "Data with communal covariates.", fs = 9)
@@
 
-The *summer* half-year is the weeks 14--39, about 1 April to 30 September, and the *winter*
+The *summer* half-year consists of the weeks 14--39, about 1 April to 30 September, and the *winter*
 half-year is the rest, weeks 1-13 and 40--52, 1 January to 31 March and 1 October to 31 December.
@@
 ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March.", 
-     label = "tab:neoanalw", keep = 6)
+     label = "tab:neoanalw", keep = 3)
 ```
@@
 ltx2(fit.neos, dr = dr.neos, caption = "Neonatal mortality, April to September.", 
-     label = "tab:neoanals", keep = 6)
+     label = "tab:neoanals", keep = 3)
 ```
@@
 #### By birth month
+
+See Figure \@ref(fig:bybmonth).
 
@@
 ```{r postw1, results = "asis", cache = FALSE}
-ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914. Adjusted for sex and parity.", 
-     label = "tab:postw1", keep = 5)
+ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914.", 
+     label = "tab:postw1", keep = NULL)
 ```
@@
 ltx2(fit.postw1sru, dr = dr.postw1sru, caption = "Postneonatal mortality, Umeå, October to March 1895-1913.",
-     label = "tab:postw1sru", keep = 4)
+     label = "tab:postw1sru", keep = 3)
 ```
@@
 ltx2(fit.postw1srs, dr = dr.postw1srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March 1895-1913.", 
-     label = "tab:postw1srs", keep = 4)
+     label = "tab:postw1srs", keep = 3)
 ```
@@
 ```{r postw2, results = "asis", cache = FALSE}
-ltx2(fit.postw2, dr = dr.postw2, caption = "Postneonatal mortality, October to March 1914-1934. Adjusted for sex, parity, and subregion.", label = "tab:postw2", keep = 5)
+ltx2(fit.postw2, dr = dr.postw2, caption = "Postneonatal mortality, October to March 1914-1934. Adjusted for sex, parity, and subregion.", label = "tab:postw2", keep = 3)
 ```
@@
 ltx2(fit.postw2sru, dr = dr.postw2sru, caption = "Postneonatal mortality, Umeå, October to March 1914-1934.",
-     label = "tab:postw2sru", keep = 4)
+     label = "tab:postw2sru", keep = 3)
 ```
@@
 ltx2(fit.postw2srs, dr = dr.postw2srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March 1914-1934.", 
-     label = "tab:postw2srs", keep = 4)
+     label = "tab:postw2srs", keep = 3)
 ```
@@
 ```{r postw3, results = "asis", cache = FALSE}
-ltx2(fit.postw3, dr = dr.postw3, caption = "Postneonatal mortality, October to March 1935-1950. Adjusted for sex, parity, and subregion.", label = "tab:postw3", keep = 5)
+ltx2(fit.postw3, dr = dr.postw3, caption = "Postneonatal mortality, October to March 1935-1950. Adjusted for sex, parity, and subregion.", label = "tab:postw3", keep = 3)
 ```
@@
 ltx2(fit.postw3sru, dr = dr.postw3sru, caption = "Postneonatal mortality, Umeå, October to March 1935-1950.",
-     label = "tab:postw3sru", keep = 4)
+     label = "tab:postw3sru", keep = 3)
 ```
@@
 ltx2(fit.postw3srs, dr = dr.postw3srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March 1895-1913.", 
-     label = "tab:postw3srs", keep = 4)
+     label = "tab:postw3srs", keep = 3)
 ```
@@
 ```{r postanals, results='asis', cache = FALSE}
-ltx2(fit.posts, dr = dr, caption = "Postneonatal mortality, April to September. Adjusted for sex, parity, time period, and subregion.", label = "tab:postanals", keep = 5)
+ltx2(fit.posts, dr = dr, caption = "Postneonatal mortality, April to September. Adjusted for sex, parity, time period, and subregion.", label = "tab:postanals", keep = 3)
 ```

### 2022-02-12 getdata4.Rmd
add socStatus and urban

@@
 
+Added *urban* on 12 February 2022.
+
 Created 6 February 2022 after consultations.
@@
 per <- pers[(pers$frsbobtyp == 2) & (pers$df == 0), ]
+per$urban <- per$fboort %in% c(274, 1208)
 per$frsbobtyp <- per$df <- per$f2id <- per$m2id <- NULL
@@
 per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", 
-               "subreg", "socBranch", "socStatus", "hisclass", "enter", "exit", "event")]
+               "subreg", "socBranch", "socStatus", "hisclass", "urban", "enter", "exit", "event")]
 saveRDS(per, file = "data/births4.rds")
@@
                                                    "birthmonth", "period", "socBranch", "socStatus", "illeg",
-                                                   "parity", "hisclass", tempvars)]
+                                                   "parity", "hisclass", "urban", tempvars)]
 ## New addition: (28 Jan 2022)

### 2022-02-12 paper4.Rmd
add socStatus and urban

@@
 - \usepackage{a4wide}
-#- \usepackage[tablesonly]{endfloat}
-#- \renewcommand{\efloatseparator}{\mbox{}}
+- \usepackage[tablesonly]{endfloat}
+- \renewcommand{\efloatseparator}{\mbox{}}
 abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895-1950 is studied in a biometric analysis setting.
@@
 **socBranch** Working branch of father (if any).
+
+**socStatus** Social status *within* the given working branch.
 
@@
 
-    inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socBranch + socStatus + parity + 
+    inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socBranch + socStatus + urban + parity + 
                          extemp + extemp.1 + emeantemp + summer + ageIvl, data = all,
@@
 fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * 
-                       (extemp + extemp.1 + emeantemp + socBranch + period), 
+                       (extemp + extemp.1 + emeantemp + socBranch + socStatus + period + urban), 
                    data = inftab, time = age)
@@
 ```{r formneo}
-formneo <- as.formula("oe(event, exposure) ~  extemp + extemp.1 + emeantemp + socBranch + period +
+formneo <- as.formula("oe(event, exposure) ~  extemp + extemp.1 + emeantemp + socBranch + socStatus + period + urban +
                      subreg + sex + parity")
@@
 ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March.", 
-     label = "tab:neoanalw", keep = 3)
+     label = "tab:neoanalw", keep = NULL)
 ```
@@
 ltx2(fit.neos, dr = dr.neos, caption = "Neonatal mortality, April to September.", 
-     label = "tab:neoanals", keep = 3)
+     label = "tab:neoanals", keep = NULL)
 ```
@@
 ```{r setformw}
-form <- as.formula("oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + subreg + sex + parity")
-form.sb <-  as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * socBranch + subreg + sex + parity")
-form.sr <- as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * subreg + socBranch + sex + parity") 
+form <- as.formula("oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + subreg + sex + parity")
+form.sb <-  as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * socBranch + socStatus + urban + subreg + sex + parity")
+form.sr <- as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * subreg + socBranch + socStatus + urban + sex + parity") 
 tabpost <- readRDS("data/tabpost.rds")
@@
 ```{r postanalwp, results='asis', cache = TRUE}
-##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socBranch + illeg + parity
+##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socBranch + socStatus + illeg + parity
   ##              + subreg + period, data = infdat, cuts = c(0, 1))
@@
 
-#### Sex and temperature
+#### Sex, urban,  and temperature
 
 ```{r postanalwpsex, results='asis', cache = TRUE}
-##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socBranch + illeg + parity
+##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socBranch + socStatus + illeg + parity
   ##              + subreg + period, data = infdat, cuts = c(0, 1))
@@
 
-fit.postwsex <- tpchreg(oe(event, exposure) ~ sex * (extemp + extemp.1 + emeantemp), 
+fit.postwsex <- tpchreg(oe(event, exposure) ~ (sex + urban) * (extemp + extemp.1 + emeantemp), 
                      data = tabpost[!tabpost$summer, ], time = age)
@@
 round(dr.postwsex, 4)
-#tbl(round(dr.postw, 4), caption = "Interaction tests with temperature and period, postneonatal ##mortality.", rownames = TRUE)
+#tbl(round(dr.postw, 4), caption = "Interaction tests with temperature and period, urban, postneonatal ##mortality.", rownames = TRUE)
 ```
 
-Obviously no interactions between temperature and sex.
+Obviously no interactions between temperature and sex, and urban.
 
@@
 ```{r postw1srpu, cache = TRUE}
-fit.postw1sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
+fit.postw1sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + sex + parity, 
                         data = tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]" & tabpost$subreg == "ume", ], time = age)
@@
 ltx2(fit.postw1sru, dr = dr.postw1sru, caption = "Postneonatal mortality, Umeå, October to March 1895-1913.",
-     label = "tab:postw1sru", keep = 3)
+     label = "tab:postw1sru", keep = NULL)
 ```
@@
 ```{r postw1srps, cache = TRUE}
-fit.postw1srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
+fit.postw1srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + sex + parity, 
                         data = tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]" & tabpost$subreg != "ume", ], time = age)
@@
 ltx2(fit.postw1srs, dr = dr.postw1srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March 1895-1913.", 
-     label = "tab:postw1srs", keep = 3)
+     label = "tab:postw1srs", keep = NULL)
 ```
@@
 ```{r postw2, results = "asis", cache = FALSE}
-ltx2(fit.postw2, dr = dr.postw2, caption = "Postneonatal mortality, October to March 1914-1934. Adjusted for sex, parity, and subregion.", label = "tab:postw2", keep = 3)
+ltx2(fit.postw2, dr = dr.postw2, caption = "Postneonatal mortality, October to March 1914-1934. Adjusted for sex, parity, and subregion.", label = "tab:postw2", keep = NULL)
 ```
@@
 ```{r postw2srpu, cache = TRUE}
-fit.postw2sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
+fit.postw2sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + sex + parity, 
                         data = tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]" & tabpost$subreg == "ume", ], time = age)
@@
 ltx2(fit.postw2sru, dr = dr.postw2sru, caption = "Postneonatal mortality, Umeå, October to March 1914-1934.",
-     label = "tab:postw2sru", keep = 3)
+     label = "tab:postw2sru", keep = NULL)
 ```
@@
 ```{r postw2srps, cache = TRUE}
-fit.postw2srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
+fit.postw2srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + sex + parity, 
                         data = tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]" & tabpost$subreg != "ume", ], time = age)
@@
 ltx2(fit.postw2srs, dr = dr.postw2srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March 1914-1934.", 
-     label = "tab:postw2srs", keep = 3)
+     label = "tab:postw2srs", keep = NULL)
 ```
@@
 ```{r postw3p, results = "asis"}
-fit.postw3 <- tpchreg(oe(event, exposure) ~ extemp.1 + emeantemp + socBranch + subreg + sex + parity, 
+fit.postw3 <- tpchreg(oe(event, exposure) ~ extemp.1 + emeantemp + socBranch + socStatus + urban + subreg + sex + parity, 
                       data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ], time = age)
@@
 ```{r postw3, results = "asis", cache = FALSE}
-ltx2(fit.postw3, dr = dr.postw3, caption = "Postneonatal mortality, October to March 1935-1950. Adjusted for sex, parity, and subregion.", label = "tab:postw3", keep = 3)
+ltx2(fit.postw3, dr = dr.postw3, caption = "Postneonatal mortality, October to March 1935-1950. Adjusted for sex, parity, and subregion.", label = "tab:postw3", keep = NULL)
 ```
@@
 ```{r postw3srpu, cache = TRUE}
-fit.postw3sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
+fit.postw3sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + sex + parity, 
                         data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]" & tabpost$subreg == "ume", ], time = age)
@@
 ltx2(fit.postw3sru, dr = dr.postw3sru, caption = "Postneonatal mortality, Umeå, October to March 1935-1950.",
-     label = "tab:postw3sru", keep = 3)
+     label = "tab:postw3sru", keep = NULL)
 ```
@@
 ```{r postw3srps, cache = TRUE}
-fit.postw3srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + sex + parity, 
+fit.postw3srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + sex + parity, 
                         data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]" & tabpost$subreg != "ume", ], time = age)
@@
 ltx2(fit.postw3srs, dr = dr.postw3srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March 1895-1913.", 
-     label = "tab:postw3srs", keep = 3)
+     label = "tab:postw3srs", keep = NULL)
 ```
@@
 
-fit.posts <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + period + 
+fit.posts <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + period + 
                         subreg + parity + sex, 
@@
 ```{r postanals, results='asis', cache = FALSE}
-ltx2(fit.posts, dr = dr, caption = "Postneonatal mortality, April to September. Adjusted for sex, parity, time period, and subregion.", label = "tab:postanals", keep = 3)
+ltx2(fit.posts, dr = dr, caption = "Postneonatal mortality, April to September. Adjusted for sex, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
 ```

### 2022-02-14 getdata4.Rmd
introduce 'socst' instead of 'socStatus' and 'socBranch'

@@
 # Introduction
+
+Add *socst* on 13 February 2022.
 
@@
 indx <- match(per$id, bobs$id)
+per$socst <- bobs$socst[indx]
+per$socst[is.na(per$socst)] <- 0
+per$socst <- factor(per$socst, labels = c("none", "highBusiness", "highOfficial", "farmer", "lowBusiness",
+                                          "lowOfficial", "qualWorker", "farmhand", "unqualWorker"))
 per$socStatus <- bobs$socStatus[indx]
@@
 per$socStatus <- factor(per$socStatus)
-per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", 
+per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", "socst",
                "subreg", "socBranch", "socStatus", "hisclass", "urban", "enter", "exit", "event")]
@@
 infdat <- infdat[order(infdat$id, infdat$enter), c("subreg", "id", "sex", "enter", "exit", "event", 
-                                                   "birthmonth", "period", "socBranch", "socStatus", "illeg",
+                                                   "birthmonth", "period", "socBranch", "socStatus", "socst", "illeg",
                                                    "parity", "hisclass", "urban", tempvars)]
@@
 
-Note the name change of the data file: *infdat3.rds*! And the the time lag is `r lagg`.
+Note the name change of the data file: *infdat4.rds*! And the the time lag is `r lagg`.

### 2022-02-14 paper4.Rmd
introduce 'socst' instead of 'socStatus' and 'socBranch'

@@
 
-    inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socBranch + socStatus + urban + parity + 
+    inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socst + urban + parity + 
                          extemp + extemp.1 + emeantemp + summer + ageIvl, data = all,
@@
 fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * 
-                       (extemp + extemp.1 + emeantemp + socBranch + socStatus + period + urban), 
+                       (extemp + extemp.1 + emeantemp + socst + period + urban), 
                    data = inftab, time = age)
 dr.all <- drop1(fit.all, test = "Chisq")
-##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
+##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socst, 
   ##               data = tabneo, time = age))
-##system.time(tabpost.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socBranch + socStatus, 
+##system.time(tabpost.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socst, 
   ##               data = tabpost, time = age))
@@
 ```{r formneo}
-formneo <- as.formula("oe(event, exposure) ~  extemp + extemp.1 + emeantemp + socBranch + socStatus + period + urban +
+formneo <- as.formula("oe(event, exposure) ~  extemp + extemp.1 + emeantemp + socst + period + urban +
                      subreg + sex + parity")
@@
 ```{r setformw}
-form <- as.formula("oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + subreg + sex + parity")
-form.sb <-  as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * socBranch + socStatus + urban + subreg + sex + parity")
-form.sr <- as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * subreg + socBranch + socStatus + urban + sex + parity") 
+form <- as.formula("oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + subreg + sex + parity")
+form.sb <-  as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * socst + urban + subreg + sex + parity")
+form.sr <- as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * subreg + socst + urban + sex + parity") 
 tabpost <- readRDS("data/tabpost.rds")
@@
 ```{r postanalwp, results='asis', cache = TRUE}
-##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socBranch + socStatus + illeg + parity
+##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socst + illeg + parity
   ##              + subreg + period, data = infdat, cuts = c(0, 1))
@@
 ```{r postanalwpsex, results='asis', cache = TRUE}
-##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socBranch + socStatus + illeg + parity
+##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socst + illeg + parity
   ##              + subreg + period, data = infdat, cuts = c(0, 1))
@@
 ```{r postw1srpu, cache = TRUE}
-fit.postw1sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + sex + parity, 
+fit.postw1sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
                         data = tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]" & tabpost$subreg == "ume", ], time = age)
@@
 ```{r postw1srps, cache = TRUE}
-fit.postw1srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + sex + parity, 
+fit.postw1srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
                         data = tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]" & tabpost$subreg != "ume", ], time = age)
@@
 ```{r postw2srpu, cache = TRUE}
-fit.postw2sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + sex + parity, 
+fit.postw2sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
                         data = tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]" & tabpost$subreg == "ume", ], time = age)
@@
 ```{r postw2srps, cache = TRUE}
-fit.postw2srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + sex + parity, 
+fit.postw2srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
                         data = tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]" & tabpost$subreg != "ume", ], time = age)
@@
 ```{r postw3p, results = "asis"}
-fit.postw3 <- tpchreg(oe(event, exposure) ~ extemp.1 + emeantemp + socBranch + socStatus + urban + subreg + sex + parity, 
+fit.postw3 <- tpchreg(oe(event, exposure) ~ extemp.1 + emeantemp + socst + urban + subreg + sex + parity, 
                       data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ], time = age)
@@
 ```{r postw3srpu, cache = TRUE}
-fit.postw3sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + sex + parity, 
+fit.postw3sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
                         data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]" & tabpost$subreg == "ume", ], time = age)
@@
 ```{r postw3srps, cache = TRUE}
-fit.postw3srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + sex + parity, 
+fit.postw3srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
                         data = tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]" & tabpost$subreg != "ume", ], time = age)
@@
 
-fit.posts <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socBranch + socStatus + urban + period + 
+fit.posts <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + period + 
                         subreg + parity + sex, 

### 2022-02-15 getdata4.Rmd
fix socst

--- 2022-02-12 getdata4.Rmd
@@
 indx <- match(per$id, bobs$id)
-per$socStatus <- bobs$socStatus[indx]
-per$socBranch <- bobs$socBranch[indx]
+per$socst <- bobs$socst[indx]
+per$socst[is.na(per$socst)] <- 9
+per$socst1 <- factor(per$socst, labels = c("highBusiness", "highOfficial", "farmer", "lowBusiness",
+                                          "lowOfficial", "qualWorker", "farmhand", "unqualWorker", "none"))
+per$socst <- factor(per$socst, labels = c("highWhiteC", "highWhiteC", "farmer", "lowWhiteC", "lowWhiteC", 
+                                           "qualWorker", "farmer", "worker", "none"))
+##per$socStatus <- bobs$socStatus[indx]
+##per$socBranch <- bobs$socBranch[indx]
 per$hisclass <- bobs$hisclass[indx]
@@
 ## NEW:
-per$socBranch <- as.character(per$socBranch)
-per$socBranch[is.na(per$socBranch)] <- "none"
-per$socBranch <- factor(per$socBranch)
-per$socStatus <- as.character(per$socStatus)
-per$socStatus[is.na(per$socStatus)] <- "low"
-per$socStatus <- factor(per$socStatus)
-per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", 
-               "subreg", "socBranch", "socStatus", "hisclass", "urban", "enter", "exit", "event")]
+##per$socBranch <- as.character(per$socBranch)
+##per$socBranch[is.na(per$socBranch)] <- "none"
+##per$socBranch <- factor(per$socBranch)
+##per$socStatus <- as.character(per$socStatus)
+##per$socStatus[is.na(per$socStatus)] <- "low"
+##per$socStatus <- factor(per$socStatus)
+per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", "socst",
+               "subreg", "hisclass", "urban", "enter", "exit", "event")]
 saveRDS(per, file = "data/births4.rds")
@@
 infdat <- infdat[order(infdat$id, infdat$enter), c("subreg", "id", "sex", "enter", "exit", "event", 
-                                                   "birthmonth", "period", "socBranch", "socStatus", "illeg",
+                                                   "birthmonth", "period", "socst", "illeg",
                                                    "parity", "hisclass", "urban", tempvars)]

### 2022-02-15 paper4.Rmd
fix socst

@@
 infdat$emeantemp <- round(infdat$emeantemp)
-levels(infdat$socBranch) <- c("office", "farming", "none", "office", "worker") # NOTE!
 ```

### 2022-02-15 R/intre.R
first commit

@@ -1,0 +1,21 @@
+intre <- function(who = "farming"){
+    tabpost <- readRDS("data/tabpost_br.rds")
+    library(eha)
+    ume <- tabpost[tabpost$subreg == "ume", ]
+    # Just to experiment:
+    ##fit <- tpchreg(oe(event, exposure) ~ (socBranch * socStatus):emeantemp, 
+      ##             data = ume, time = age)
+    res <- vector(mode = "list", length = 3)
+    i <- 0
+        fit <- glm(event ~ offset(log(exposure)) + emeantemp * extemp,
+                    data = ume, family = poisson()) 
+        fit1 <- glm(event ~ offset(log(exposure)) + extemp + extemp %in% emeantemp,
+                    data = ume, family = poisson())
+        fit2 <- glm(event ~ offset(log(exposure)) + emeantemp + (extemp + extemp.1) %in% emeantemp,
+                    data = ume, family = poisson())
+        
+        fit3 <- tpchreg(oe(event, exposure) ~ emeantemp + (extemp + extemp.1) %in% emeantemp, data = ume[ume$socBranch == who, ],
+                       time = age)
+    res <- anova(fit2, fit1, fit, test = "Chisq")
+    fit1
+}

### 2022-02-15 R/socfig.R
first commit

@@ -1,0 +1,23 @@
+socfig <- function(coeff, ran = c(-25, 5), main = "1935-1950"){
+    ## coeff is output from tpchreg(oe(event, exposure) ~ socst + socst:emeantemp 
+    ## controls+)
+    ny <- coeff[substr(names(coeff), 1, 5) == "socst"]
+    a <- c(0, ny[1:5])
+    b <- ny[6:11]
+    x <- seq(ran[1], ran[2], length = 100)
+    y <- matrix(nrow = 6, ncol = length(x))
+    for (j in 1:6){
+        ##y[j, ] <- exp(a[j] + b[j] * x)
+        y[j, ] <- exp(b[j] * x)
+    }
+    ymax <- max(y)
+    plot(x, y[1, ], type = "l", ylim = c(0, ymax), xlab = "Temperature", 
+         ylab = "Rel. hazards", lwd = 2, main = main)
+    for (i in 2:6){
+        lines(x, y[i, ], col = i, lty = i, lwd = 2)
+    }
+    nam <- c("WhiteCollar", "Farmer",  "lowWhiteC", "qualWorker", "Worker", "None")
+    legend("topright", legend = nam, col = 1:6, lty = 1:6, cex = 0.7)
+    points(0, 1, cex = 2)
+    abline(h = 1, v = 0, lty = 3)
+}

### 2022-02-15 paper4.Rmd
add plots of interactions

@@
 ```{r setformw}
-form <- as.formula("oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + subreg + sex + parity")
+form <- as.formula("oe(event, exposure) ~ socst + extemp + extemp.1 + emeantemp + urban + subreg + sex + parity")
+formex <- as.formula("oe(event, exposure) ~ socst + socst:extemp + extemp.1 + emeantemp + urban + subreg + sex + parity")
+formex1 <- as.formula("oe(event, exposure) ~ socst + socst:extemp.1 + extemp + emeantemp + urban + subreg + sex + parity")
+formem <- as.formula("oe(event, exposure) ~ socst + socst:I(emeantemp + extemp) + extemp.1 + urban + subreg + sex + parity")
 form.sb <-  as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * socst + urban + subreg + sex + parity")
@@
 
+The effects of temperature in different social classes are best shown graphically, see 
+Figure \@ref(fig:soc2).
+
+```{r soc1, fig.cap = "Interactions between social status and temperature, winter 1895--1913."}
+source("R/socfig.R")
+oldpar <- par(mfrow = c(1,3))
+tempdat <- tabpost[(!tabpost$summer) & tabpost$period == "(1895,1914]", ]
+fit <- tpchreg(formem, data = tempdat, time = age)
+fitex <- tpchreg(formex, data = tempdat, time = age)
+fitex1 <- tpchreg(formex1, data = tempdat, time = age)
+socfig(coef(fit), main = "Weekly mean temp")
+socfig(coef(fitex), main = "Excess temp", ran = c(-10, 10))
+socfig(coef(fitex1), main = "Lagged excess temp", ran = c(-10, 10))
+par(oldpar)
+```
+
 #### The time period 1914--1934
@@
      label = "tab:postw2srs", keep = NULL)
+```
+
+The effects of temperature in different social classes are best shown graphically, see 
+Figure \@ref(fig:soc2).
+
+```{r soc2, fig.cap = "Interactions between social status and temperature, winter 1914--1934."}
+source("R/socfig.R")
+oldpar <- par(mfrow = c(1,3))
+tempdat <- tabpost[(!tabpost$summer) & tabpost$period == "(1914,1935]", ]
+fit <- tpchreg(formem, data = tempdat, time = age)
+fitex <- tpchreg(formex, data = tempdat, time = age)
+fitex1 <- tpchreg(formex1, data = tempdat, time = age)
+socfig(coef(fit), main = "Weekly mean temp")
+socfig(coef(fitex), main = "Excess temp", ran = c(-10, 10))
+socfig(coef(fitex1), main = "Lagged excess temp", ran = c(-10, 10))
+par(oldpar)
 ```
@@
 
+The effects of temperature in different social classes are best shown graphically, see 
+Figure \@ref(fig:soc2).
+
+```{r soc3, fig.cap = "Interactions between social status and temperature, winter 1935--1950."}
+source("R/socfig.R")
+oldpar <- par(mfrow = c(1,3))
+tempdat <- tabpost[(!tabpost$summer) & tabpost$period == "(1935,1951]", ]
+fit <- tpchreg(formem, data = tempdat, time = age)
+fitex <- tpchreg(formex, data = tempdat, time = age)
+fitex1 <- tpchreg(formex1, data = tempdat, time = age)
+socfig(coef(fit), main = "Weekly mean temp")
+socfig(coef(fitex), main = "Excess temp", ran = c(-10, 10))
+socfig(coef(fitex1), main = "Lagged excess temp", ran = c(-10, 10))
+par(oldpar)
+```
+
 ### Summer

### 2022-02-18 getdata.Rmd
modifications ...
clean complete, fresh start

--- 2021-11-27 getdata.Rmd
@@
 ---
-title: "Get population and temperature data"
+title: "getdata4.Rmd"
 author: "Göran Broström"
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
-library(knitr)
-library(eha)
-library(skum)
-options(show.signif.stars = FALSE)
-g <- function(x) (log(1 + 365 * x))^3
+knitr::opts_chunk$set(echo = TRUE)
+library(skum) # Also loads 'eha'
+age.cuts = c(0, 3, 7, 28, 56, 140, 183, 250, 365) / 365
 ```
 
-# Population data
+# Introduction
 
-We use the observation file rather than the  population file, so we can get 
-correct parish information.
+*Note* that this is the old getdata4.Rmd, developed after cleaning.
 
-```{r readobs}
-bb <- obs$id[obs$starttyp == 2]
-births <- obs[obs$id %in% bb, 
-    c("id", "sex", "birthdate", "nofrs", "sluttyp", "socStatus", "hisclass", 
-      "enter", "exit", "ort", "ortnmn")]
-births$event <- as.numeric(births$sluttyp == 2)
-births <- age.window(births, c(0, 1))
-births <- births[order(births$id, births$enter), ]
-##indx <- tapply(births$id, births$id)
-##births$mexit <- tapply(births$exit, births$id, max)[indx]
-##births$mevent <- tapply(births$event, births$id, max)[indx]
-##births <- births[!duplicated(births$id), ]
-### Deaths:
-did <- unique(births$id[births$event == 1])
-deaths <- births[births$id %in% did, ]
-indx <- tapply(deaths$id, deaths$id)
-deaths$menter <- tapply(deaths$enter, deaths$id, min)[indx]
-deaths <- deaths[!duplicated(deaths$id), ]
-deaths$enter <- deaths$menter
-deaths$menter <- NULL
-deaths$enter <- 0 ## satisfied....
-### Survivors:
-survs <- births[!(births$id %in% did), ]
-indx <- tapply(survs$id, survs$id)
-survs$mexit <- tapply(survs$exit, survs$id, max)[indx]
-survs <- survs[!duplicated(survs$id), ]
-survs$exit <- survs$mexit
-survs$mexit <- NULL
-survs$enter <- 0 ## Ok ...
-### Now, give birth to NEW births:
-births <- rbind(deaths, survs)
-births <- births[births$birthdate > 1895, ]
+Add *socst* on 13 February 2022.
 
-indx <- match(births$id, pers$id)
-births$ab <- pers$ab[indx]
-##births$df <- pers$df[indx]
-births$illeg <- factor(births$ab == 2, labels = c("no", "yes"))
-births$ab <- NULL
-births$sluttyp <- NULL
-births$parity <- pers$paritet_g[indx]
-summary(births)                                
+Added *urban* on 12 February 2022.
+
+Created 6 February 2022 after consultations.
+
+** Run this (and only this) before running "paper4.Rmd"!**
+
+We start with *skum::persons*  and age is to begin with in days,
+with stillbirths  *removed* and born alive but dead on birthday at age 0.25 days.
+
+For use of *eha::make.communal*, we need to temporarily change time unit to years.
+
+Further, we eventually skip the use of eha::make.communal in  favor of *match*. Reason: 
+eha::make.communal assumes age is measured in *years*. But this is no promise yet.
+
+# Read skum::pers
+
+```{r readpers}
+per <- pers[(pers$frsbobtyp == 2) & (pers$df == 0), ]
+per$urban <- per$fboort %in% c(274, 1208)
+per$frsbobtyp <- per$df <- per$f2id <- per$m2id <- NULL
+per$frsbobdatmin <- per$frsbobdatmax <- per$utdatmax <- per$utdatmin <- NULL
+per <- per[(per$foddat > 18950000), ]
+dn <- per$doddat < 1
+per$doddat[dn] <- per$foddat[dn] + 40000
+## Fix foddat:
+yr <- per$foddat %% 10000 == 0
+per$foddat[yr] <- per$foddat[yr] + 101 # As early as possible
+mon <- per$foddat %% 100 == 0
+per$foddat[mon] <- per$foddat[mon] + 1 # Ditto
+## Fix doddat:
+yr <- per$doddat %% 10000 == 0
+per$doddat[yr] <- per$doddat[yr] + 1231 # As late as possible
+mon <- per$doddat %% 100 == 0
+per$doddat[mon] <- per$doddat[mon] + 28 # Ditto (guard for February ...)
+##
+per$birthdate <- as.Date(as.character(per$foddat), format = "%Y%m%d")
+per$deathdate <- as.Date(as.character(per$doddat), format = "%Y%m%d")
+nabir <- is.na(per$birthdate)
+per$foddat[nabir] <- per$foddat[nabir] - 2
+per$doddat[nabir] <- per$doddat[nabir] - 2
+##
+per$birthdate <- as.Date(as.character(per$foddat), format = "%Y%m%d")
+per$exit <- as.numeric(difftime(per$deathdate, per$birthdate))
+per$deathdate <- as.Date(as.character(per$doddat), format = "%Y%m%d")
+## Fix doddat:
+per$doddat[is.na(per$deathdate)] <- per$doddat[is.na(per$deathdate)] - 1
+per$deathdate <- as.Date(as.character(per$doddat), format = "%Y%m%d")
+##
+per$exit <- as.numeric(difftime(per$deathdate, per$birthdate, units = "days"))
+oj <- per$exit < 0
+per$doddat[oj] <- per$doddat[oj] + 10000
+per$deathdate <-  as.Date(as.character(per$doddat), format = "%Y%m%d")
+per$exit <- as.numeric(difftime(per$deathdate, per$birthdate, units = "days"))
+per$event <- per$exit <= 365
+per$exit[per$exit == 0] <- 0.25
+per$exit[per$exit > 365] <- 365
 ```
 
-We put on 'subreg':
+Now we need *nofrs* for newborn, found in the *observation* file(s), so
 
-```{r subreg}
-births$subreg <- factor(births$nofrs)
-levels(births$subreg) <- c(rep("ume", 2), rep("bjur", 4), rep("sten", 3))
-## Remove "sten":
-births <- births[births$subreg != "sten", ]
-births$subreg <- factor(births$subreg) # Remove level "sten"
-```
-
-# Neonatal mortality, an overview
-
-```{r neo, fig.cap = "Neonatal mortality by year, all regions."}
-neo <- age.window(births, c(0, 1 / 13))
-surv <- as.matrix(neo[, c("enter", "exit", "event", "birthdate")])
-pall <- perstat(surv, period = 1900:1951)$intensity
-plot(1900:1950, pall, type = "b", ylim = c(0, 0.5), col = "blue", xlab = "Year",
-     ylab = "Mortality")
-abline(h = 0)
-abline(h = 0.3, lty = 3)
-```
-
-## By region
-
-```{r neoby, fig.cap = "Neonatal mortality by year  and region."}
-usurv <- as.matrix(neo[neo$subreg == "ume", c("enter", "exit", "event", "birthdate")])
-pume <-  perstat(usurv, period = 1900:1951)$intensity
-plot(1900:1950, pall, type = "b", ylim = c(0, 0.7), col = "blue", xlab = "Year",
-     ylab = "Mortality")
-bsurv <- as.matrix(neo[neo$subreg == "bjur", c("enter", "exit", "event", "birthdate")])
-pjur <-  perstat(bsurv, period = 1900:1951)$intensity
-lines(1900:1950, pjur, type = "b", ylim = c(0, 0.5), col = "red")
-##ssurv <- as.matrix(neo[neo$subreg == "sten", c("enter", "exit", "event", "birthdate")])
-##sjur <-  perstat(ssurv, period = 1900:1951)$intensity
-##lines(1900:1950, sjur, type = "b", ylim = c(0, 0.5), col = "black")
-abline(h = 0)
-abline(h = 0.3, lty = 3)
-```
-
-# Post-neonatal mortality, an overview
-
-```{r pnneo, fig.cap = "Post-eonatal mortality by year, all regions."}
-pneo <- age.window(births, c(1 / 13, 1))
-surv <- as.matrix(pneo[, c("enter", "exit", "event", "birthdate")])
-pall <- perstat(surv, period = 1900:1951)$intensity
-plot(1900:1950, pall, type = "b", ylim = c(0, 0.1), col = "blue",
-     xlab = "Year", ylab = "Mortality")
-abline(h = 0)
-abline(h = 0.3, lty = 3)
-```
-
-## By region
-
-```{r pneoby, fig.cap = "Post-neonatal mortality by year and regions."}
-usurv <- as.matrix(pneo[pneo$subreg == "ume", c("enter", "exit", "event", "birthdate")])
-pume <-  perstat(usurv, period = 1900:1951)$intensity
-plot(1900:1950, pall, type = "b", ylim = c(0, 0.1), col = "blue",
-     xlab = "Year", ylab = "Mortality")
-bsurv <- as.matrix(pneo[pneo$subreg == "bjur", c("enter", "exit", "event", "birthdate")])
-pjur <-  perstat(bsurv, period = 1900:1951)$intensity
-lines(1900:1950, pjur, type = "b", ylim = c(0, 0.5), col = "red")
-##ssurv <- as.matrix(pneo[pneo$subreg == "sten", c("enter", "exit", "event", "birthdate")])
-##sjur <-  perstat(ssurv, period = 1900:1951)$intensity
-##lines(1900:1950, sjur, type = "b", ylim = c(0, 0.5), col = "black")
-abline(h = 0)
-abline(h = 0.3, lty = 3)
+```{r fromobs}
+bobs <- obs[obs$starttyp == 2 & obs$id %in% per$id, ]
+indx <- match(per$id, bobs$id)
+per$socst <- bobs$socst[indx]
+per$socst[is.na(per$socst)] <- 9
+per$socst1 <- factor(per$socst, labels = c("highBusiness", "highOfficial", "farmer", "lowBusiness",
+                                          "lowOfficial", "qualWorker", "farmhand", "unqualWorker", "none"))
+per$socst <- factor(per$socst, labels = c("highWhiteC", "highWhiteC", "farmer", "lowWhiteC", "lowWhiteC", 
+                                           "qualWorker", "farmer", "worker", "none"))
+##per$socStatus <- bobs$socStatus[indx]
+##per$socBranch <- bobs$socBranch[indx]
+per$hisclass <- bobs$hisclass[indx]
+per$nofrs <- bobs$nofrs[indx]
+per <- per[!is.na(per$nofrs), ]
+per$subreg <- factor(per$nofrs)
+levels(per$subreg) <- c(rep("ume", 2), 
+                        rep("bjur", 4), 
+                        rep("sten", 3))
+##per <- per[(per$subreg != "sten"), ] # NOTE: Now Stensele station is included
+per$subreg <- factor(per$subreg)
+per$enter <- 0
+per$sex <- factor(per$kon, labels = c("boy", "girl"))
+per$birthdate <- toTime(per$birthdate)
+per$illeg <- as.numeric(per$ab == 2)
+per$parity <- per$paritet_g
+per$paritet_g <- per$paritet_k <- per$doddat <- per$frsbosdat <- NULL
+per$fid <- per$fboort <- per$ab <- per$fyrkfd <- per$fyrkind <- NULL
+per$region <- per$kon <- per$fodhfrs <- per$frsbobdat <- per$frsbostyp <- NULL
+per$fbonmn <- per$fboind <- per$fyrkrel <- per$fyrketxt <- NULL
+per$foddat <- as.Date(as.character(per$foddat), format = "%Y%m%d")
+## NEW:
+##per$socBranch <- as.character(per$socBranch)
+##per$socBranch[is.na(per$socBranch)] <- "none"
+##per$socBranch <- factor(per$socBranch)
+##per$socStatus <- as.character(per$socStatus)
+##per$socStatus[is.na(per$socStatus)] <- "low"
+##per$socStatus <- factor(per$socStatus)
+per <- per[, c("id", "sex", "birthdate", "mid", "foddat", "parity", "illeg", "socst",
+               "subreg", "hisclass", "urban", "enter", "exit", "event")]
+saveRDS(per, file = "data/births4.rds")
 ```
@@
 source("R/tempdat.R")
-umetemp <- tempdat("ume", c(1895, 1950))
-bjurtemp <- tempdat("bjur", c(1895, 1950))
-##stentemp <- tempdat("~/Forskning/Data/sten_temp.csv", c(1899, 1952))
+temp_start <- 1894
+temp_end <- 1951
+temp_ivl <- c(temp_start, temp_end)
+##
+umetemp <- tempdat("ume", temp_ivl)
+bjurtemp <- tempdat("bjur", temp_ivl)
+
+##Changed 5 feb 2022:
+##stentemp <- tempdat("sten", temp_ivl) # Added 27 Jan 2022
+source("R/normal.R") # normal = Norsjö - Malå!!
+stentemp <- normal(temp_ivl)
+## End change
+
 saveRDS(umetemp, file = "data/umetemp.rds")
 saveRDS(bjurtemp, file = "data/bjurtemp.rds")
+saveRDS(stentemp, file = "data/stentemp.rds")  ## Added 27 Jan2022
 ```
 
-# Make communal
+# Put on temperature data
+
+Here we must change time unit to *year*.
 
 ```{r makecommunal}
-##source("R/make_comm.R")
-lagg <- 3 / 365  ## NOTE!!
+per$enter <- per$enter / 365
+per$exit <- per$exit / 365
+## per$birthdate <- toTime(per$birthdate) #already done!
+temp_start <- 1894
+temp_end <- 1951
+temp_ivl <- c(temp_start, temp_end)
+##
+
+lagg <- 0 / 365 ##3 / 365  ## NOTE!!
+tempvars <-  c("heat", "heat.1", "extemp", "extemp.1", "cold", "cold.1", "emintemp", "emeantemp", "emaxtemp", 
+               "week", "year") 
 ## Ume:
-bume <- births[births$subreg == "ume", ]
-bume <- make.communal(bume, umetemp["heat"], start = 1895, period = 1/52, lag = lagg)
-bume <- bume[bume$enter < bume$exit, ]
-bume <- make.communal(bume, umetemp["cold"], start = 1895, period = 1 / 52, lag = lagg)
-bume <- bume[bume$enter < bume$exit, ]
-bume <- make.communal(bume, umetemp["week"], start = 1895, period = 1 / 52, lag = lagg)
-bume <- bume[bume$enter < bume$exit, ]
+bume <- per[per$subreg == "ume", ]
+bume <- make.communal(bume, umetemp[, tempvars], 
+                      start = temp_start, period = 1/52, lag = lagg)
+
 ### Ske:
-bskel <- births[births$subreg == "bjur", ]
-bskel <- make.communal(bskel, bjurtemp["heat"], start = 1895, period = 1/52, lag = lagg)
-bskel <- bskel[bskel$enter < bskel$exit, ]
-bskel <- make.communal(bskel, bjurtemp["cold"], start = 1895, period = 1 / 52, lag = lagg)
-bskel <- bskel[bskel$enter < bskel$exit, ]
-bskel <- make.communal(bskel, bjurtemp["week"], start = 1895, period = 1 / 52, lag = lagg)
-bskel <- bskel[bskel$enter < bskel$exit, ]
-##bsten <- births[births$subreg == "sten", ]
-##bsten <- make_comm(bsten, stentemp, start = 1899, period = 1/52, lag = lagg)
+bskel <- per[per$subreg == "bjur", ]
+bskel <- make.communal(bskel, bjurtemp[, tempvars], 
+                       start = temp_start, period = 1/52, lag = lagg)
+
+### Sten:
+bsten <- per[per$subreg == "sten", ]
+bsten <- make.communal(bsten, stentemp[, tempvars], 
+                       start = temp_start, period = 1/52, lag = lagg)
+
 bume$region <- "ume"
 bskel$region <- "ske"
-infdat <- rbind(bume, bskel)
-infdat <- infdat[order(infdat$id, infdat$enter), c("region", "id", "sex", "enter", "exit", "event", 
-                                                   "birthdate", "socStatus", "hisclass", "illeg",
-                                                   "parity", "heat", "cold", "week")]
-saveRDS(infdat, file = "data/infdat.rds")
+bsten$region <- "sten"
+infdat <- rbind(bume, bskel, bsten)
+infdat$birthmonth <- ceiling((infdat$birthdate - floor(infdat$birthdate)) * 12)
+infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
+infdat <- infdat[order(infdat$id, infdat$enter), c("subreg", "id", "sex", "enter", "exit", "event", 
+                                                   "birthmonth", "period", "socst", "illeg",
+                                                   "parity", "hisclass", "urban", tempvars)]
+## New addition: (28 Jan 2022)
+##infdat <- rc(infdat) # rc from 'skum'
+##n <- NROW(infdat)
+
+saveRDS(infdat, file = "data/infdat4.rds") # NOTE: Name change!!!
+
+## Wait with this:
+## tabdat4 <- toTpch(Surv(enter, exit, event) ~ subreg + sex + birthmonth + socBranch + illeg + parity 
 ```
 
-Done!
+Note the name change of the data file: *infdat4.rds*! And the the time lag is `r lagg`.

### 2022-02-18 paper.Rmd
clean complete, fresh start

@@
 - \usepackage{a4wide}
-#- \usepackage[figuresonly]{endfloat}
-#- \renewcommand{\efloatseparator}{\mbox{}}
-abstract: "The effect of extreme temperatures on infant mortality in the Umeå region 1895-1950 is studied in a biometric analysis setting.
-More precisely, the effect of climate and weather, measured by temperature, on infant mortality is investigated.
-It turns out that climate is more important than weather, low average temperatures are more important than temporary dips in temperature.
-The investigated geographical region is situated too far north for bad effects of high temperatures to show."
+- \usepackage[tablesonly]{endfloat}
+- \renewcommand{\efloatseparator}{\mbox{}}
+abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895-1950 is studied in a biometric analysis setting.
+More precisely, the effect of climate and weather, measured by temperature, average and extremes, on infant mortality is investigated.
+It turns out that climate (average) is more important than weather (extremes), low average temperatures are more important than temporary dips in temperature, but effects are different in neonatal and postneonatal settings."
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA, global.par=TRUE)
+knitr::opts_chunk$set(echo = FALSE, fig.path='figs/',fig.height=3, comment=NA) ## NOTE!!
+knitr::opts_knit$set(global.par = TRUE)
+##
 library(knitr)
 library(eha)
-par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
-g <- function(x) (log(1 + x))^3 # x in days.
-dg <- function(x) (3 * (log(1 + x))^2) / (1 + x)
-ginv <- function(y) exp(y^(1/3)) - 1
+starttid <- proc.time()
 ```
 
+```{r setpar}
+par(las = 1, cex.axis = 0.8, cex.lab = 0.8, cex = 0.8)
+```
+
+```{r readdata}
+births <- readRDS("data/births4.rds")
+infdat  <- readRDS("data/infdat4.rds")
+```
 # Introduction
@@
 The impact of ambient temperature variations on infant mortality is studied for 
-a northern Sweden coastal area, the Umeå and Skellefteå regions, during the 
+two northern Sweden areas, the Umeå coastal region and Skellefteå coastal and inland regions, during the 
 first half of the twentieth century. Two recent papers [@junkka; @karlsson] studied 
-neonatal mortality and temperature variations in a larger geographical area containing 
-the present one during the 
+neonatal mortality and temperature variations in this geographical area during the 
 years 1880--1950. Climate and mortality in general is a research area that has 
-generated great interest over the last years, see @tbgb10.
+generated great interest over the last years, see @tbgb10 for an example of our own efforts.
 
 The effect of seasonal variation and the occurrence of extreme monthly temperatures
-is studied and interacted with sex, social class, and legitimacy. Studies are performed 
-separately for neonatal  and postneonatal mortality, and for winter and summer seasons, 
-and the classification into endogenous and exogenous factors will be discussed.
+is studied and interacted with social class and time period. Studies are performed 
+separately for neonatal  and postneonatal mortality, and for winter and summer seasons.
 
-One important reason for studying neonatal and postneonatal mortality separately is the 
-empirical findings by Bourgeois-Pichat [@bp51a;  @bp51b] about endogenous and exogenous 
-mortality and the log-cube transform.
+Figure \@ref(fig:map) shows the study area within Sweden, with the weather stations 
+marked. The map is taken (with permission) from the paper by @junkka.
+
+```{r map, fig.height = 5, fig.cap = "Umeå, Skellefteå (orange) and its inland (green)."}
+include_graphics("myfigs/map2.png")
+```
 
@@
 
-**socStatus** Social status of family, based on HISCLASS.
+**socStatus** Social status *within* the given working branch.
 
@@
 
-Figure \@ref(fig:mortstat) shows the average monthly crude infant mortality, and a clear seasonal pattern is visible.
+Figure \@ref(fig:mortstat) shows the average weekly crude infant mortality, and a clear seasonal pattern is visible.
 
-```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
-infdat <- readRDS("data/infdat3_year3.rds")
+```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
+##infdat <- readRDS("data/infdat4.rds") Already done
 tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))
@@
 
-The average monthly neonatal mortality is shown in Figure \@ref(fig:neodata).
+The average weekly neonatal mortality is shown in Figure \@ref(fig:neodata).
 
-```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
+```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
 neo <- age.window(infdat, c(0, 28 / 365))
@@
 
-The average monthly postneonatal mortality is shown in Figure \@ref(fig:pneodata).
+The average weekly postneonatal mortality is shown in Figure \@ref(fig:pneodata).
 
-```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3}
+```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
 pneo <- age.window(infdat, c(28 / 365, 1))
@@
 plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
-     ylab = "NMR", lwd = 1.5, cex = 0.7)
+     ylab = "O/E rate", lwd = 1.5, cex = 0.7)
 abline(h = 0)
@@
 ```
+
+```{r pntrend, fig.cap = "Crude postneonatal mortality by year, Umeå-Skellefteå 1895--1950.", fig.height = 4, cache = TRUE}
+pndat <- age.window(infdat, c(28 / 365, 1))
+tpnda <- toTpch(Surv(enter, exit, event) ~ year, data = pndat, 
+                cuts = c(28 / 365, 1))
+y <- tpnda$event / tpnda$exposure
+par(las = 1)
+par(las = 1, cex.axis = 0.7, cex.lab = 0.7)
+plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
+     ylab = "O/E rate", lwd = 1.5, cex = 0.7)
+abline(h = 0)
+abline(h = c(0.2, 0.4, 0.6), lty = 3)
+```
+
+Note that the estimates of neonatal and postneonatal mortality are non-standard here: 
+They are calculated as *occurrence-exposure rates*, that is, the number of deaths
+divided by *exposure time*. They are thus crude *hazard rates*, not probabilities, which is 
+more common in definitions of mortality.
 
@@
 bjurtemp <- readRDS("data/bjurtemp.rds")
+stentemp <- readRDS("data/stentemp.rds")
 ```
 
-Temperature data are collected from two weather stations, *Umeå* and *Bjuröklubb*
-(used with population data from Skellefteå coastal area). Both stations deliver
+Temperature data are collected from three weather stations, *Umeå*, *Bjuröklubb*
+(used with population data from Skellefteå coastal area), and *Stensele* (Inland). All stations deliver
 daily temperature data covering our time period, usually three measures per day, morning,
@@
 
-```{r monavgtemp, fig.cap="Weekly max and min temperature averages, 1895--1950.", fig.height = 3}
-oldpar <- par(mfrow = c(1, 2))
+```{r monavgtemp, fig.cap="Weekly max, mean, and min temperature averages, 1895--1950.", fig.height = 3, cache = TRUE}
+oldpar <- par(mfrow = c(1, 3))
 ## Umeå:
@@
 
-```{r longterm, fig.cap = "Yearly average temperatures, Umeå and Skellefteå."}
-bmtemp <- rbind(umetemp, bjurtemp)
+```{r longterm, fig.cap = "Yearly average temperatures, Umeå and Skellefteå.", cache = TRUE}
+bmtemp <- rbind(umetemp, bjurtemp, stentemp)
 umt <- with(bmtemp, tapply(meantemp, year, mean))
@@
 function *make.communal* in the **R** [@cran] package *eha* [@eha; @ehar2] is used 
-for that purpose.
+for that purpose.  Resulting data drame is partly shown in Table \@ref(tab:readdata3).
 
 ```{r readdata3}
-levels(infdat$subreg) <- c("ume", "ske")
-##infdat$excessTemp <- cut(infdat$aver, c(-16, -2, 2, 9), include.lowest = TRUE) # NOTE!!!
+levels(infdat$subreg) <- c("ume", "ske", "inland")
+##infdat$excessTemp <- cut(infdat$extemp, c(-16, -2, 2, 9), include.lowest = TRUE) # NOTE!!!
 ##infdat$excessTemp <- relevel(infdat$excessTemp, ref = "(-2,2]")
-infdat$excessTemp <- infdat$cold # CHECK THIS!!
-tbl(head(infdat))
+##infdat$lowTemp <- (infdat$extemp + infdat$extemp.1) < -10  # CHECK THIS!!
+##infdat$highTemp <- infdat$extemp + infdat$extemp.1 > 10
+##infdat$lowTemp <- with(infdat, (extemp < -3) & (extemp.1 < -3))
+infdat$lowTemp <- (with(infdat, cold + cold.1 < -10))
+infdat$highTemp <- with(infdat, (extemp > 3) & (extemp.1 > 3))
+tbl(head(infdat[, c("enter", "exit", "event", "extemp", "extemp.1", "emeantemp", "week", "year")]),
+    caption = "Data with communal covariates.", fs = 9)
 ```
@@
 
-The analyses are performed on the *log-cube* scale, following the hints of  
-Bourgeois-Pichat [@bp51a; @bp51b], with proportional hazards modelling. Note that
-the property of proportional hazards are preserved under a strictly monotone increasing time transform.
-However, the estimates of baseline distribution characteristics will change, of course.
+It turns out that extremely low temperature (`lowTemp`) is bad during all seasons except summer,
+and extremely high temperature (`highTemp`) is bad during summer, but good otherwise.
+So we group season into two categories, *summer* and *not summer*.
+In each case separate analyses for neonatal and postneonatal mortality are performed.
 
-We start by performing separate analyses for neonatal and postneonatal mortality.
+The *summer* half-year consists of the weeks 14--39, about 1 April to 30 September, and the *winter*
+half-year is the rest, weeks 1-13 and 40--52, 1 January to 31 March and 1 October to 31 December.
+This is the division made in @karlsson, and we keep it for comparability reasons.
 
-# Postneonatal mortality
-
-This is the simplest part, because *B-P* predicts that the baseline distribution (given the log-cube
-transformation) is *exponential*, that is, the hazard function is *constant*.
-
-```{r postanal, results='asis'}
+```{r prepare}
 source("R/ltx2.R")
 infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
-infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
-levels(infdat$subreg) <- c("ume", "ske")
-infdat$season <- cut(infdat$week, c(-1, 9.5, 22.5, 35.5, 48.5, 52.5), labels = c("winter", "spring", "summer", "fall", "winter"))
-infdat$season <- as.factor(infdat$season)
-pneo <- age.window(infdat, c(28 / 365, 1))
-##
-pneo$enter <- g(pneo$enter * 365)
-pneo$exit <- g(pneo$exit * 365)
-##
-pneo$emintemp <- round(pneo$emintemp)
-##
-tpneo <- toTpch(Surv(enter, exit, event) ~ period + sex + socBranch + socStatus + illeg + parity + 
-                    excessTemp
-                + emintemp + season + subreg, 
-                data = pneo, cuts = g(c(28, 365)))
-##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
-  ##           data = pneo, shape = 1)
-##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
-  ##           data = pneo, cuts = g(c(28, 365)))
-
-fit.tpch <- tpchreg(oe(event, exposure) ~ excessTemp + emintemp +  period + sex + socBranch + socStatus  + 
-                        illeg + parity, 
-                    data = tpneo, time = age)
-dr <- drop1(fit.tpch, test = "Chisq")
-ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality. Adjusted for social status and branch, illegitimacy, parity, time period, season, and subregion.", label = "tab:postanalske", keep = 2)
-```
-
-The result in Table \@ref(tab:postanal) shows that *climate* (`emintemp`) is more important than 
-*weather* (`excessTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
-covariates (not shown).
-
-# Neonatal mortality
-
-It turns out that here we have some significant interactions, with `sex` involved.
-Maybe surprising. So we run separate analyses for the sexes.
-
-## Girls
-
-```{r neoanal, results = 'asis'}
-##library(survival)
-neo <- age.window(infdat, c(0, 28 /365))
-neo$exit <- g(neo$exit * 365)
-neo$enter <- g(neo$enter * 365)
-fit.ph <- coxreg(Surv(enter, exit, event) ~  (excessTemp + emintemp) * sex + socStatus + illeg + parity + period + socBranch + subreg,
-                 data = neo)
-dr <- drop1(fit.ph, test = "Chisq")
-ltx2(fit.ph, dr = dr, caption = "Neonatal mortality. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatal", keep = 2)
-```
-
-Same here, *climate* is more important than *weather*.
-
-## Boys
-
-```{r neoanalboys, results = 'asis'}
-fit.ph <- coxreg(Surv(enter, exit, event) ~  excessTemp + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
-                 data = neo[neo$sex == "boy" & neo$season %in% c("spring", "summer"), ])
-dr <- drop1(fit.ph, test = "Chisq")
-ltx2(fit.ph, dr = dr, caption = "Neonatal mortality, boys. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatalboys", keep = 2)
-```
-
-Here, however, boys are sensitive to temperatures below average, in addition to the 
-sensitivity for cold climate.
-
-# Postneonatal mortality, winter and spring
-
-This is the simplest part, because *B-P* predicts that the baseline distribution (given the log-cube
-transformation) is *exponential*, that is, the hazard function is *constant*.
-
-```{r postanalw, results='asis'}
-##infdat$parity <- cut(infdat$parity, c(-2, 1.5, 4.5, 19), labels = c("1", "2-4", "5+"))
+##infdat$period <- cut(infdat$birthdate, c(1895, 1914, 1935, 1951), dig.lab = 5)
 ##levels(infdat$subreg) <- c("ume", "ske")
@@
 ##infdat$season <- as.factor(infdat$season)
-pneo <- age.window(infdat[infdat$season %in% c("winter", "spring"), ], c(28 / 365, 1))
-##
-pneo$enter <- g(pneo$enter * 365)
-pneo$exit <- g(pneo$exit * 365)
-##
-pneo$emintemp <- round(pneo$emintemp)
-##
-tpneo <- toTpch(Surv(enter, exit, event) ~ excessTemp + emintemp + sex + socBranch + socStatus + illeg + parity + 
-                    period + subreg, 
-                data = pneo, cuts = g(c(28, 365)))
-##fit2 <- phreg(Surv(enter - g(28), exit - g(28), event) ~ sex + socBranch + socStatus + cold + emintemp, 
-  ##           data = pneo, shape = 1)
-##fit.pch <- pchreg(Surv(enter, exit, event) ~ sex + socBranch + socStatus + cold + emintemp, 
-  ##           data = pneo, cuts = g(c(28, 365)))
-fit.tpch <- tpchreg(oe(event, exposure) ~ excessTemp + emintemp + sex + socBranch + socStatus + illeg + parity + period + 
-                        subreg, 
-                    data = tpneo, time = age)
-dr <- drop1(fit.tpch, test = "Chisq")
-ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, winter and spring. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", 
-     label = "tab:postanalw", keep = 2)
+## New:
+infdat$summer <- infdat$week %in% 14:39
+infdat$emeantemp <- round(infdat$emeantemp)
 ```
 
-The result in Table \@ref(tab:postanalw) shows that *climate* (`emintemp`) is more important than 
-*weather* (`cold`). Moreover, no signs of interaction between weather or climate and the rest of 
+# Results
+
+The results regarding neonatal mortality is much in accordance with the results
+found by @junkka. However, they used temperature in a "hockey-stick" regression with a breakpoint at
+14.5 degrees Celsius and a negative slope (decreasing risk) to the left and a positive slope
+(increasing risk) to the right. Instead, we are using the average weekly temperature
+for the 52 weeks of a year, for each week averaging over all the years in the study, as
+our "reference points" ("climate"), adding deviances up and down ("weather") as
+"short-term temperature stress". This is similar to the way prices and mortality were related
+in for instance @genus11, that is, a time series split into long time trend and short
+term variation.
+
+Comments on other candidate covariates:
+
+*Birth month* is left out in the analyses despite that fact that
+it is an important factor in neonatal mortality. However, we include *time of year* in terms of 
+*winter* and *summer* as a time-varying covariate, and in the neonatal case it will coincide to a
+great extent with birth period. For the postneonal case the situation is different, but it turns out that
+for those infants who have survived the first month of life, birth month does not matter much.
+We separate the investigation into two parts, *neonatal* and *postneonatal* mortality. But first,
+a joint analysis.
+
+*Socioeconomic status* is divided into two factor covariates: *socBranch* and *socStatus*. The latter should be 
+seen relative to actual *socBranch*. We have information on whether the infant was *illegitimate* (mother unmarried),
+but we incorporate those cases in the category *none* of *socBranch* and *unknown* as category of *socStatus*.
+
+```{r managedata}
+if (!(file.exists("data/inftab.rds") && file.exists("data/tabneo.rds") && file.exists("data/tabpost.rds"))){
+    neo <- age.window(infdat, c(0, 28 / 365))
+    neo$ageIvl <- "neo"
+    pneo <- age.window(infdat, c(28/365, 1))
+    pneo$ageIvl <- "postneo"
+    all <- rbind(neo, pneo)
+    all$ageIvl <- factor(all$ageIvl)
+    ##all$enter <- round(all$enter * 365)
+    ##all$exit <- round(all$exit * 365)
+    ##fit.all <- coxph(Surv(enter, exit, event) ~ strata(ageIvl) * (extemp + emeantemp + lowTemp + highTemp + socBranch + ##socStatus + sex + period),
+    ##               data = all)
+    ##dr.all <- drop1(fit.all, test = "Chisq")
+
+    inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socst + urban + parity + 
+                         extemp + extemp.1 + emeantemp + summer + ageIvl, data = all,
+                     cuts = c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365) / 365)
+    tabneo <- inftab[inftab$ageIvl == "neo", ]
+    tabpost <- inftab[inftab$ageIvl == "postneo", ]
+    tabpost$age <- factor(tabpost$age)
+    tabneo$age <- factor(tabneo$age)
+    saveRDS(tabneo, file = "data/tabneo.rds")
+    saveRDS(tabpost, file = "data/tabpost.rds")
+    saveRDS(inftab, file = "data/inftab.rds")
+}else{
+    inftab <- readRDS("data/inftab.rds") #The rest are read later...
+}
+```
+
+```{r joint, cache = TRUE}
+fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * 
+                       (extemp + extemp.1 + emeantemp + socst + period + urban), 
+                   data = inftab, time = age)
+dr.all <- drop1(fit.all, test = "Chisq")
+##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socst, 
+  ##               data = tabneo, time = age))
+##system.time(tabpost.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socst, 
+  ##               data = tabpost, time = age))
+```
+
+Covariates may affect neonatal and postneonatal mortality differently, and one way to 
+investigate that is to stratify the infant mortality data into two age intervals, one from
+birth to one month of age and the other from one month to one year of age. Then the
+interaction between the stratum variable and other variables of interest is investigated, 
+with this result:
+
+```{r shownonprop, cache = FALSE}
+dr.all
+##tbl(round(dr.all, 4), caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
+```
+
+We can see that the variables *emeantemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case. In terms of 
+*infant mortality*, we can say that these caovariates have non-proportional effects.
+
+## Neonatal mortality
+
+The analyses are split into two parts by season, *winter* is one, and *summer* the other.
+Summer covers the months April to September, and winter the rest of the months.
+
+```{r formneo}
+formneo <- as.formula("oe(event, exposure) ~  extemp + extemp.1 + emeantemp + socst + period + urban +
+                     subreg + sex + parity")
+tabneo <- readRDS("data/tabneo.rds")
+```
+
+### Winter
+
+This period refers to the months *October to March*. A Cox regression involves as
+interesting variables *highTemp*, an indicator of temperature at least four degrees
+above the expected for at least two weeks in a row, *emeantemp* the expected temperature
+the actual weeek, and *extemp* the *excess temperature* the actual week.
+
+```{r neoanalwp, results = 'asis', cache = TRUE}
+
+##library(survival)
+winter <- tabneo[!tabneo$summer, ]
+##winter$season <- factor(winter$season)
+##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
+##neo$enter <- g(neo$enter * 365)
+fit.neow <- tpchreg(formneo, data = winter, time = age)
+dr <- drop1(fit.neow, test = "Chisq")
+```
+
+```{r neoanalw, results = "asis", cache = FALSE}
+ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March.", 
+     label = "tab:neoanalw", keep = NULL)
+```
+
+#### Temperature and sex
+
+We check the interaction between temperature and sex and period.
+
+```{r neosexper}
+fit <- tpchreg(oe(event, exposure) ~ sex * (extemp + extemp.1 + emeantemp), 
+               time = age, data = winter)
+dr <- drop1(fit, test = "Chisq")
+dr
+```
+
+Obviously no temperature and sex interactions
+
+### Summer
+
+This period refers to the months *April to September*. A Cox regression involves as
+interesting variables *highTemp*, an indicator of temperature at least four degrees
+above the expected for at least two weeks in a row, *emeantemp* the expected temperature
+the actual weeek, and *extemp* the *excess temperature* the actual week.
+
+```{r neoanalsp, cache = TRUE}
+##library(survival)
+#summer <- neo[neo$summer, ]
+##summer$season <- factor(summer$season)
+##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
+##neo$enter <- g(neo$enter * 365)
+fit.neos <- tpchreg(formneo, data = tabneo[tabneo$summer, ], time = age)
+dr.neos <- drop1(fit.neos, test = "Chisq")
+```
+
+```{r neoanals, results = 'asis', cache = FALSE}
+ltx2(fit.neos, dr = dr.neos, caption = "Neonatal mortality, April to September.", 
+     label = "tab:neoanals", keep = NULL)
+```
+
+## Postneonatal mortality
+
+```{r setformw}
+form <- as.formula("oe(event, exposure) ~ socst + extemp + extemp.1 + emeantemp + urban + subreg + sex + parity")
+formex <- as.formula("oe(event, exposure) ~ socst + socst:extemp + extemp.1 + emeantemp + urban + subreg + sex + parity")
+formex1 <- as.formula("oe(event, exposure) ~ socst + socst:extemp.1 + extemp + emeantemp + urban + subreg + sex + parity")
+formem <- as.formula("oe(event, exposure) ~ socst + socst:emeantemp + extemp + extemp.1 + urban + subreg + sex + parity")
+form.sb <-  as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * socst + urban + subreg + sex + parity")
+form.sr <- as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * subreg + socst + urban + sex + parity") 
+tabpost <- readRDS("data/tabpost.rds")
+```
+
+#### By birth month
+
+See Figure \@ref(fig:bybmonth).
+
+```{r bybmonth, fig.cap = "Crude postneonatal relative (to January = 1) mortality by birth month.", cache = TRUE}
+pn <- age.window(infdat, c(28/365, 1))
+fit <- coxreg(Surv(enter, exit, event) ~ as.factor(birthmonth), data = pn)
+plot(exp(c(0, coef(fit))), type = "b", xlab = "Birth month", ylab = "Hazard ratios",
+     col = "blue")
+abline(h = 1)
+```
+
+### Winter
+
+```{r postanalwp, results='asis', cache = TRUE}
+##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socst + illeg + parity
+  ##              + subreg + period, data = infdat, cuts = c(0, 1))
+##winter <- gtpneo[gtpneo$summer == 0, ]
+##winter$season <- factor(winter$season)
+
+fit.postw <- tpchreg(oe(event, exposure) ~ period * (extemp + extemp.1 + emeantemp), 
+                     data = tabpost[!tabpost$summer, ], time = age)
+dr.postw <- drop1(fit.postw, test = "Chisq")
+```
+
+Interaction tests with temperature and period, postneonatal mortality:
+
+```{r postanalw, cache = FALSE}
+round(dr.postw, 4)
+#tbl(round(dr.postw, 4), caption = "Interaction tests with temperature and period, postneonatal ##mortality.", rownames = TRUE)
+```
+
+This shows that the temperature effects do not vary much with time period,
+but we continue with separate analyses for the time periods `r levels(tabpost$period)`. Just
+to make sure, will be joined when appropriate.
+
+#### Sex, urban,  and temperature
+
+```{r postanalwpsex, results='asis', cache = TRUE}
+##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socst + illeg + parity
+  ##              + subreg + period, data = infdat, cuts = c(0, 1))
+##winter <- gtpneo[gtpneo$summer == 0, ]
+##winter$season <- factor(winter$season)
+
+fit.postwsex <- tpchreg(oe(event, exposure) ~ (sex + urban) * (extemp + extemp.1 + emeantemp), 
+                     data = tabpost[!tabpost$summer, ], time = age)
+dr.postwsex <- drop1(fit.postwsex, test = "Chisq")
+```
+
+Interaction tests with temperature and sex, postneonatal mortality:
+
+```{r postanalwsex, cache = FALSE}
+round(dr.postwsex, 4)
+#tbl(round(dr.postw, 4), caption = "Interaction tests with temperature and period, urban, postneonatal ##mortality.", rownames = TRUE)
+```
+
+Obviously no interactions between temperature and sex, and urban.
+
+#### The full time period
+
+```{r postw1p, results = "asis"}
+fit.postw1 <- tpchreg(form, data = tabpost[(!tabpost$summer), ], time = age)
+dr.postw1 <- drop1(fit.postw1, test = "Chisq")
+```
+
+```{r postw1, results = "asis", cache = FALSE}
+ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914.", 
+     label = "tab:postw1", keep = NULL)
+```
+
+Table \@ref(tab:postw1) shows that the effects of  extreme temperatures have the expected directions, extra high temperatures is *positive* 
+(remember that this concerns winter conditions), but not statistically significant. *Climate* (emeantemp), on the other hand shows to be very
+important, where increasing temperature by one degree lower mortality by 0.7 per cent. The normal short term fluctuations (extemp) are not so important, 
+but in the expected direction.
+
+The importance of social class (socBranch) in explaining the effects of climate and weather is negligible, but there is variation between
+the regions regarding *climate*:
+
+```{r postw1srpu, cache = TRUE}
+fit.postw1sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
+                        data = tabpost[(!tabpost$summer) & tabpost$subreg == "ume", ], time = age)
+dr.postw1sru <- drop1(fit.postw1sru, test = "Chisq")
+```
+
+```{r postw1sru, results = "asis"}
+ltx2(fit.postw1sru, dr = dr.postw1sru, caption = "Postneonatal mortality, Umeå.",
+     label = "tab:postw1sru", keep = NULL)
+```
+
+```{r postw1srps, cache = TRUE}
+fit.postw1srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
+                        data = tabpost[(!tabpost$summer) & tabpost$subreg != "ume", ], time = age)
+dr.postw1srs <- drop1(fit.postw1srs, test = "Chisq")
+```
+
+```{r postw1srs, results = "asis"}
+ltx2(fit.postw1srs, dr = dr.postw1srs, caption = "Postneonatal mortality, Skellefteå with inland.", 
+     label = "tab:postw1srs", keep = NULL)
+```
+
+We can conclude from Tables \@ref(tab:postw1sru) and \@ref(tab:postw1srs) that the main difference between Umeå and Skellefteå is that
+*climate* has more severe effect in Skellefteå (with inland), while *weather* is less important.
+
+The effects of temperature in different social classes are best shown graphically, see 
+Figure \@ref(fig:socw1).
+
+```{r socw1, fig.cap = "Interactions between social status and temperature, winter."}
+source("R/socfig.R")
+oldpar <- par(mfrow = c(1,3))
+tempdat <- tabpost[(!tabpost$summer), ]
+fit <- tpchreg(formem, data = tempdat, time = age)
+fitex <- tpchreg(formex, data = tempdat, time = age)
+fitex1 <- tpchreg(formex1, data = tempdat, time = age)
+socfig(coef(fit), main = "Weekly mean temp")
+socfig(coef(fitex), main = "Excess temp", ran = c(-10, 10))
+socfig(coef(fitex1), main = "Lagged excess temp", ran = c(-10, 10))
+par(oldpar)
+```
+
+### Summer
+
+```{r postanalsp, results='asis', cache = TRUE}
+##summer <- gtpneo[gtpneo$summer == 1, ]
+##summer$season <- factor(summer$season)
+
+fit.posts <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + period + 
+                        subreg + parity + sex, 
+                    data = tabpost[tabpost$summer, ], time = age)
+dr <- drop1(fit.posts, test = "Chisq")
+```
+
+```{r postanals, results='asis', cache = FALSE}
+ltx2(fit.posts, dr = dr, caption = "Postneonatal mortality, April to September. Adjusted for sex, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
+```
+
+The result in Table \@ref(tab:postanals) shows that *climate* (`emeantemp`) is more important than 
+*weather* (`exTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
 covariates (not shown).
 
-# Neonatal mortality, winter and spring
+#### The full time period
 
-It turns out that here we have some significant interactions, with `sex` involved.
-Maybe surprising. So we run separate analyses for the sexes.
-
-## Girls
-
-```{r neoanalw, results = 'asis'}
-neo <- age.window(infdat[infdat$season %in% c("winter", "spring"), ], c(0, 28 /365))
-neo$exit <- g(neo$exit * 365)
-neo$enter <- g(neo$enter * 365)
-fit.ph <- coxreg(Surv(enter, exit, event) ~  excessTemp + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
-                 data = neo[neo$sex == "girl" & neo$emeantemp < -0, ])
-dr <- drop1(fit.ph, test = "Chisq")
-ltx2(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, girls.
-Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", label = "tab:neonatalw", keep = 2)
+```{r posts1p, results = "asis"}
+fit.posts1 <- tpchreg(form, data = tabpost[(tabpost$summer), ], time = age)
+dr.posts1 <- drop1(fit.posts1, test = "Chisq")
 ```
 
-Same here, *climate* is more important than *weather*.
-
-## Boys
-
-```{r neoanalboysw, results = 'asis'}
-fit.ph <- coxreg(Surv(enter, exit, event) ~  excessTemp + emintemp + socStatus + illeg + parity + period + socBranch + subreg,
-                 data = neo[neo$sex == "boy" & neo$season %in% c("winter", "spring"), ])
-dr <- drop1(fit.ph, test = "Chisq")
-ltx2(fit.ph, dr = dr, caption = "Neonatal mortality winter and spring, boys. Adjusted for social status and branch, illegitimacy, parity, time period, and sub region.", 
-     label = "tab:neonatalboysw", keep = 2)
+```{r posts1, results = "asis", cache = FALSE}
+ltx2(fit.posts1, dr = dr.posts1, caption = "Postneonatal mortality, October to March.", 
+     label = "tab:posts1", keep = NULL)
 ```
 
-Here, however, boys are sensitive to temperatures below average, in addition to the 
-sensitivity for cold climate.
+Table \@ref(tab:posts1) shows that the effects of  extreme temperatures have the expected directions, extra high temperatures is *positive* 
+(remember that this concerns summer conditions), but not statistically significant. *Climate* (emeantemp), on the other hand shows to be very
+important, where increasing temperature by one degree lower mortality by 0.7 per cent. The normal short term fluctuations (extemp) are not so important, 
+but in the expected direction.
+
+The importance of social class (socBranch) in explaining the effects of climate and weather is negligible, but there is variation between
+the regions regarding *climate*:
+
+```{r posts1srpu, cache = TRUE}
+fit.posts1sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
+                        data = tabpost[(tabpost$summer) & tabpost$subreg == "ume", ], time = age)
+dr.posts1sru <- drop1(fit.posts1sru, test = "Chisq")
+```
+
+```{r posts1sru, results = "asis"}
+ltx2(fit.posts1sru, dr = dr.posts1sru, caption = "Postneonatal mortality, Umeå, October to March.",
+     label = "tab:posts1sru", keep = NULL)
+```
+
+```{r posts1srps, cache = TRUE}
+fit.posts1srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
+                        data = tabpost[(tabpost$summer) & tabpost$subreg != "ume", ], time = age)
+dr.posts1srs <- drop1(fit.posts1srs, test = "Chisq")
+```
+
+```{r posts1srs, results = "asis"}
+ltx2(fit.posts1srs, dr = dr.posts1srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March.", 
+     label = "tab:posts1srs", keep = NULL)
+```
+
+We can conclude from Tables \@ref(tab:posts1sru) and \@ref(tab:posts1srs) that the main difference between Umeå and Skellefteå is that
+*climate* has more severe effect in Skellefteå (with inland), while *weather* is less important.
+
+The effects of temperature in different social classes are best shown graphically, see 
+Figure \@ref(fig:soc2).
+
+```{r soc1, fig.cap = "Interactions between social status and temperature, summer."}
+source("R/socfig.R")
+oldpar <- par(mfrow = c(1,3))
+tempdat <- tabpost[(tabpost$summer), ]
+fit <- tpchreg(formem, data = tempdat, time = age)
+fitex <- tpchreg(formex, data = tempdat, time = age)
+fitex1 <- tpchreg(formex1, data = tempdat, time = age)
+socfig(coef(fit), main = "Weekly mean temp")
+socfig(coef(fitex), main = "Excess temp", ran = c(-10, 10))
+socfig(coef(fitex1), main = "Lagged excess temp", ran = c(-10, 10))
+par(oldpar)
+```
 
@@
 
-Remains to be written. Feels like two papers, one about Bourgeois-Pichat and one
-about climate and weather. Maybe discuss exogeneity/endogeneity?
+Remains to be written, especially the discussion about temperature and mortality.
+
+Used processing time (seconds):
+
+```{r tidtagning}
+sluttid <- Sys.time()
+proc.time() - starttid
+```
+
+Done: `r sluttid`.

### 2022-02-18 R/socfig.R

@@ -1,4 +1,4 @@
-socfig <- function(coeff, ran = c(-25, 5), main = "1935-1950"){
+socfig <- function(coeff, ran = c(-25, 5), main = "1935-1950", pval = 0.03){
     ## coeff is output from tpchreg(oe(event, exposure) ~ socst + socst:emeantemp 
     ## controls+)
     ny <- coeff[substr(names(coeff), 1, 5) == "socst"]
@@ -20,4 +20,5 @@ socfig <- function(coeff, ran = c(-25, 5), main = "1935-1950"){
     legend("topright", legend = nam, col = 1:6, lty = 1:6, cex = 0.7)
     points(0, 1, cex = 2)
     abline(h = 1, v = 0, lty = 3)
+    text(0, ymax - 1, paste("p = ", round(pval, 3)), col = ifelse(pval < 0.05, "red", "black"))
 }

### 2022-02-24 getdata.Rmd
final commit

@@
 lagg <- 0 / 365 ##3 / 365  ## NOTE!!
-tempvars <-  c("heat", "heat.1", "extemp", "extemp.1", "cold", "cold.1", "emintemp", "emeantemp", "emaxtemp", 
+tempvars <-  c("heat", "heat.1", "extemp", "extemp.1", "cold", "cold.1", "emintemp", "emeantemp", "emeantemp.1",  "emaxtemp", 
                "week", "year") 

### 2022-02-24 paper.Rmd
final commit

@@
 ---
-title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
+title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895--1950"
 author: "Göran Broström and Tommy Bengtsson"
@@
 - \renewcommand{\efloatseparator}{\mbox{}}
-abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895-1950 is studied in a biometric analysis setting.
+abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895--1950 is studied in a biometric analysis setting.
 More precisely, the effect of climate and weather, measured by temperature, average and extremes, on infant mortality is investigated.
-It turns out that climate (average) is more important than weather (extremes), low average temperatures are more important than temporary dips in temperature, but effects are different in neonatal and postneonatal settings."
+The main question is to what extent temperature, directly or indirectly via humidity, affect the survival of infants and to what extent it depends on the resources of the family.
+It turns out that the normal seasonality (the average temperature) is generally more important than the extremes (the deviations), low average temperatures are more important than temporary dips in temperature, but effects are different in neonatal and postneonatal settings.
+It is at normal temperatures that we expect the resources of the family to be of most importance, and this is also what we find.
+In extremely cold winters, also the children of the higher classes suffered from increased postneonatal mortality."
 ---
@@
 infdat <- readRDS("data/infdat4.rds")
+infdat$marginTemp <- round(infdat$extemp.1)
+infdat$expTemp <- round(infdat$emeantemp.1)
 ```
@@
 neonatal mortality and temperature variations in this geographical area during the 
-years 1880--1950. Climate and mortality in general is a research area that has 
-generated great interest over the last years, see @tbgb10 for an example of our own efforts.
+years 1880--1950.
+@rock2014 studies associations between temperature and age-stratified mortality,
+Skellefteå 1749--1859.
+Effects of temperature on mortality in general is a research area that has generated great interest over the last years, partly due to new weather conditions related to climate change.
 
-The effect of seasonal variation and the occurrence of extreme monthly temperatures
-is studied and interacted with social class and time period. Studies are performed 
-separately for neonatal  and postneonatal mortality, and for winter and summer seasons.
+In our previous study, we analysed the role of climate for mortality in different ages in southern Sweden 1766--1860 combining time-series data on temperature and longitudinal individual level data
+[@tbgb10].
+We found that low winter temperatures led to an increase in mortality among adults but not children in ages 1-15 years [@tbgb10].
+To understand how hard winters and warm summers affected different parts of the population, we used methods similar to the ones we have developed for our analyses of the impact of food prices.
+This means that the focus was on annual variations on temperature and mortality.
+The main focus was on demographic vulnerability to changes in external factors, such as food process and temperature, as a measurement of living standards.
+Well-off groups could afford buying food and firewood even of prices went up, while those living close to their margins suffered to such an extent that family members might die [@tb04].
+Still, child mortality was fairly equal between social classes until the latter part of the nineteenth century to widen in the beginning of the twentieth century, for both post-neonatal infant and child mortality [@dk21].
+For neonatal mortality, no systematic class differences existed in any period.
+While infant was equal between social classes up until the end of the nineteenth century, it differed between urban and rural areas.
+In fact, in 1850, infant mortality in urban areas was 50 percent higher than in rural areas and it was not until the 1920s that mortality rates converged [@scb69].
+
+Periods with low temperature might not only cause problems of keeping body heat, indoors and outdoors, but also lead to an increase in the spread of virus.
+The reason is that low temperature is associated with low humidity.
+A study of England and Wales in 1921--27, for example, shows the close connections between seasonality in temperature, humidity, and diagnosed cases of smallpox, which was a mild disease by then [@rl28].
+This means that you are more likely to be infected by viral diseases, such as measles, whooping-cough, and influenza, in low temperatures.
+Since the outcome of such diseases are not dependent on the nutritional status of the host [@rr83], we anticipate that they will not have a social gradient.
+Other diseases that are nutritionally dependent, such as pneumonia, TB, and gastro-intestinal diseases, might, however, have such a gradient.
+It means that low winter temperatures might give rise to a social gradient in mortality, but it may as well be socially neutral.
+
+In this study, we differentiate between neonatal and postneonatal mortality in an effort to understand the role of resources to protect infants rom both normal and extreme ambient temperatures.
+We analyse the effects of seasonal variation and the occurrence of extreme weekly temperatures and its interaction with social class and time period.
+Analyses are performed separately for neonatal and post neonatal mortality, and for winter and summer seasons.
+The reason is that mortality in the first four weeks of life is associate with different factors than mortality later in infancy.
+For this reason, Bourgeois-Pichat differentiate between endogenous and exogenous factors, arguing that mortality shortly after birth is due to the maternal health, fetal factors, and the delivery itself, whereas it afterwards is due to exposure to external factors following the birth-weight development [@bp51a;@bp51b].
+The main question is to what extent temperature, directly or indirectly via humidity, affect the survival of infants and to what extent it depends on the resources of the family.
 
@@
 
-```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
+```{r pneodata, fig.cap="Crude postneonatal mortality by week of year, cache=FALSE, Umeå/Skellefteå 1895--1950.", fig.height=3}
 pneo <- age.window(infdat, c(28 / 365, 1))
@@
 
-```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4, cache = TRUE}
+```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
 ndat <- age.window(infdat, c(0, 28 / 365))
@@
 
-```{r pntrend, fig.cap = "Crude postneonatal mortality by year, Umeå-Skellefteå 1895--1950.", fig.height = 4, cache = TRUE}
+```{r pntrend, fig.cap = "Crude postneonatal mortality by year, Umeå-Skellefteå 1895--1950.", fig.height = 3, cache = TRUE}
 pndat <- age.window(infdat, c(28 / 365, 1))
@@
 function *make.communal* in the **R** [@cran] package *eha* [@eha; @ehar2] is used 
-for that purpose.  Resulting data drame is partly shown in Table \@ref(tab:readdata3).
+for that purpose.  Resulting data frame is partly shown in Table \@ref(tab:readdata3).
 
@@
 infdat$highTemp <- with(infdat, (extemp > 3) & (extemp.1 > 3))
-tbl(head(infdat[, c("enter", "exit", "event", "extemp", "extemp.1", "emeantemp", "week", "year")]),
-    caption = "Data with communal covariates.", fs = 9)
+##infdat$weather <- infdat$extemp.1
+##infdat$climate <- round(infdat$emeantemp)
+tbl(head(infdat[, c("enter", "exit", "event", "marginTemp", "expTemp", "week", "year")]),
+    caption = "Data with communal covariates.", fs = 11)
 ```
+
+We will use the temperature variables *meantemp* and *emeantemp* under the names *marginTemp* 
+and *expTemp*, respectively, lagged one week. That is, we assume that the effect of 
+temperatures on mortality show itself the following week.
 
@@
 
-It turns out that extremely low temperature (`lowTemp`) is bad during all seasons except summer,
-and extremely high temperature (`highTemp`) is bad during summer, but good otherwise.
-So we group season into two categories, *summer* and *not summer*.
+It turns out that extremely low temperature is bad during all seasons,
+and extremely high temperature is bad during summer, but good otherwise.
+So we group season into two categories, *summer* and *winter*.
 In each case separate analyses for neonatal and postneonatal mortality are performed.
@@
 infdat$summer <- infdat$week %in% 14:39
-infdat$emeantemp <- round(infdat$emeantemp)
+##infdat$eexpTemp <- round(infdat$emeantemp)
 ```
+
+## Variables
+
+These are the covariates considered in our analyses. Central are the two first, 
+already discussed, but of interest is also their effect (on mortality) covariation,
+primarily with *social status* (`socst`).
+
+*   **expTemp** (time varying) expected temperature given week number.
+*   **marginTemp** (time varying) Observed week temperature minus expTemp.
+*   **socst** Social status:
+     + *highWhiteC* Upper white collar
+     + *farmer*
+     + *lowWhiteC* Lower white collar
+     + *qualWorker* Qualified worker
+     + *worker* Unqualified worker
+     + *none* No notation (for instance unmarried mother)
+*   **sex** Boy or girl
+*   **urban** Small town (Umeå or Skellefteå) or not (logical)
+*   **parity** Birth order
+*   **year**  (time varying)
+
+Comments on other candidate covariates: 
+*Birth month* is left out in the analyses despite that fact that
+it is an important factor in neonatal mortality. However, we include *time of year* in terms of 
+*winter* and *summer* as a time-varying covariate, and in the neonatal case it will coincide to a
+great extent with birth period. For the postneonal case the situation is different, but it turns out that
+for those infants who have survived the first month of life, birth month does not matter much.
+We separate the investigation into two parts, *neonatal* and *postneonatal* mortality. But first,
+a joint analysis.
+
+ We have information on whether the infant was *illegitimate* (mother unmarried);
+those cases are incorporated in the category *none* of **socst**.
 
@@
 for the 52 weeks of a year, for each week averaging over all the years in the study, as
-our "reference points" ("climate"), adding deviances up and down ("weather") as
+our "reference points" ("expTemp"), adding deviances up and down ("marginTemp") as
 "short-term temperature stress". This is similar to the way prices and mortality were related
@@
 
-Comments on other candidate covariates:
+```{r managedata}
+neo <- age.window(infdat, c(0, 28 / 365))
+ pneo <- age.window(infdat, c(28/365, 1))
+```
 
-*Birth month* is left out in the analyses despite that fact that
-it is an important factor in neonatal mortality. However, we include *time of year* in terms of 
-*winter* and *summer* as a time-varying covariate, and in the neonatal case it will coincide to a
-great extent with birth period. For the postneonal case the situation is different, but it turns out that
-for those infants who have survived the first month of life, birth month does not matter much.
-We separate the investigation into two parts, *neonatal* and *postneonatal* mortality. But first,
-a joint analysis.
+```{r nonenow, eval=FALSE}
+if (!(file.exists("data/inftab.rds") && file.exists("data/tabneo.rds") && file.exists("data/tabpost.rds"))){
 
-*Socioeconomic status* is divided into two factor covariates: *socBranch* and *socStatus*. The latter should be 
-seen relative to actual *socBranch*. We have information on whether the infant was *illegitimate* (mother unmarried),
-but we incorporate those cases in the category *none* of *socBranch* and *unknown* as category of *socStatus*.
+    neo$ageIvl <- "neo"
 
-```{r managedata}
-if (!(file.exists("data/inftab.rds") && file.exists("data/tabneo.rds") && file.exists("data/tabpost.rds"))){
-    neo <- age.window(infdat, c(0, 28 / 365))
-    neo$ageIvl <- "neo"
-    pneo <- age.window(infdat, c(28/365, 1))
     pneo$ageIvl <- "postneo"
@@
 
-    inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socst + urban + parity + 
-                         extemp + extemp.1 + emeantemp + summer + ageIvl, data = all,
+    inftab <- toTpch(Surv(enter, exit, event) ~ subreg + sex + period + socst + urban + parity +  marginTemp + expTemp + summer + ageIvl, data = all,
                      cuts = c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365) / 365)
@@
 
-```{r joint, cache = TRUE}
+```{r joint, cache=FALSE}
+if (!file.exists("Rdata/dr.all.rds")){
     fit.all <- tpchreg(oe(event, exposure) ~  strata(ageIvl) * 
-                       (extemp + extemp.1 + emeantemp + socst + period + urban), 
+                           (extemp.1 + emeantemp.1 + socst + period + urban), 
                        data = inftab, time = age)
     dr.all <- drop1(fit.all, test = "Chisq")
+    saveRDS(dr.all, file = "Rdata/dr.all.rds")
+}
 ##system.time(tabneo.fit <- tpchreg(oe(event, exposure) ~ extemp + emeantemp + highTemp + lowTemp + socst, 
@@
 interaction between the stratum variable and other variables of interest is investigated, 
-with this result:
+with the result shown in Table \@ref(tab:shownonprop).
 
-```{r shownonprop, cache = FALSE}
-dr.all
-##tbl(round(dr.all, 4), caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
+```{r shownonprop, cache = FALSE, results='asis'}
+dr.all <- readRDS("Rdata/dr.all.rds")
+rownames(dr.all)[2] <- "strata(ageIvl):marginTemp"
+rownames(dr.all)[3] <- "strata(ageIvl):expTemp"
+
+tbl(round(dr.all, 4), caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
 ```
 
-We can see that the variables *emeantemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case. In terms of 
-*infant mortality*, we can say that these caovariates have non-proportional effects.
+We can see that the variables *expTemp*, *socst* and *period* all have different effects 
+on mortality in the neonatal case compared to the postneonatal case. In terms of 
+*infant mortality*, we can say that these covariates have *strong non-proportional* effects.
+The response to *marginTemp* variations are probably also different in the neonatal and 
+postneonatal periods ($p$-value = 0.09).
+
+Figure \@ref(fig:ph) shows the crude infant mortality, calculated over the complete sample.
+
+```{r ph, fig.cap = "Crude infant mortality."}
+##daycuts <- c(0, 3, 7, 14, 28, 56, 84, 145, 207, 286, 365)
+##daycuts <- c(0, 3, 7, 28, 56, 140, 183, 250, 365)
+b1 <- births
+b1$enter <- b1$enter / 365
+b1$exit <- b1$exit / 365
+daycuts <- c(seq(0, 28, by = 2), seq(49, 245, by = 28), 285, 324, 365)
+bb <- toTpch(Surv(enter, exit, event) ~ 1, data = b1, cuts = daycuts / 365)
+fit.ph1 <- tpchreg(oe(event, exposure) ~ 1, data = bb, time = age)
+plot(fit.ph1, fn = "haz", axes = FALSE, xlab = "Days", main = "", lwd = 1.5, col = "blue")
+dax <- c(0, 28, 91, 183, 365)
+axis(1, at = dax / 365, labels = dax)
+axis(2, las = 1)
+box()
+abline(v = 28 / 365, lty = 3)
+abline( h = 0, v = 0)
+```
+
+A calmer development for the postneonatal mortality, see Figure \@ref(fig:phneo).
+
+```{r phneo, fig.cap = "Crude postneonatal mortality."}
+b2 <- age.window(b1, c(28 / 365, 1))
+pcuts <- c(28, seq(49, 245, by = 28), 285, 324, 365)
+tp <- toTpch(Surv(enter, exit, event) ~ 1, data = b2, cuts = pcuts / 365)
+
+fit.ph <- tpchreg(oe(event, exposure) ~ 1, data = tp, time = age)
+plot(fit.ph, fn = "haz", axes = FALSE, xlab = "Days", main = "", lwd = 1.5, col = "blue")
+axis(1, at = pcuts / 365, labels = pcuts)
+axis(2, las = 1)
+box()
+##abline(v = 28 / 365, lty = 3)
+abline( h = 0, v = 0)
+```
+
+Figure \@ref(fig:phneo) suggests that some smooth parametric model could fit well,
+for instance the *left truncated Weibull*  distribution. Let's see (Figure \@ref(fig:weib)).
+
+```{r weib, fig.cap = "A perfect Weibull hazard fit."}
+##births <- readRDS("data/births4.rds")
+postbirths <- age.window(births, c(28, 365))
+postbirths$exit <- postbirths$exit / 365
+postbirths$enter <- postbirths$enter / 365
+fit <- phreg(Surv(enter, exit, event) ~ 1, data = postbirths)
+plot(fit, fn = "haz", lwd = 1.5, col = "blue", main = "", xlab = "Days", axes = FALSE)
+axis(1, at = dax / 365, labels = dax)
+axis(2, las = 1)
+abline(h = 0, v = 28 / 365)
+box()
+hh <- hazards(fit.ph, cum = FALSE)
+lines(hh$x, hh$y, col = "red")
+legend("topright", legend = c("Truncated Weibull", "Piecewise constant"), col = c("blue", "red"), lwd = c(1.5,1))
+```
+
+A perfect fit!
+
+In the following we split the analyses into the neonatal and the postneonatal periods, with
+a focus on the latter. But we start with a short comment on the link to *biometric* analysis
+à la @bp51a.
+
+## The biometric finding
+
+What happens if we extend the truncated Weibull back into neonatal territory? The answer
+is given in Figure \@ref(fig:neowei).
+
+```{r neowei, fig.cap = "Weibull hazard functions, total and exogenous neonatal mortality.", fig.height = 4}
+b3 <- age.window(b1, c(0, 28 / 365))
+fit.w0 <- phreg(Surv(enter, exit, event) ~ 1, data = b3)
+plot(fit.w0, fn = "haz", axes = FALSE, xlab = "Days", main = "", lwd = 1.5, col = "red", ylim = c(0, 3))
+#fit.ph <- tpchreg(oe(event, exposure) ~ 1, data = inftab, time = age)
+##plot(fit.ph1, fn = "haz", axes = FALSE, xlab = "Days", main = "", lwd = 1.5, col = "red",
+  ##   xlim = c(0, 28 / 365))
+dor <- seq(0, 28, by = 2)
+axis(1, at = dor / 365, labels = dor)
+axis(2, las = 1)
+box()
+##abline(v = 28 / 365, lty = 3)
+abline( h = 0, v = 0)
+pars <- exp(fit$coefficients)
+x <- seq(0, 28/365, length = 1000)
+y <- hweibull(x, shape = pars[2], scale = pars[1])
+lines(x, y, col = "blue", lwd = 1.5)
+legend("topright", legend = c("Weibull postneonatal (exogenous), extrapolated", "Weibull neonatal (total)"), col = c("blue", "red"), lwd = c(1.5, 1))
+```
+
+Not very good fit in this territory. The difference here is that to the exogenous mortality, endogenous mortality is added.
+The difference between the piecewise constant and the Weibull in the figure would be 
+called the *endogeneous infant mortality* [@bp51a; @bp51b].
 
@@
 
-The analyses are split into two parts by season, *winter* is one, and *summer* the other.
-Summer covers the months April to September, and winter the rest of the months.
+The neonatal mortality in our geographical area and time period is thoroughly analysed
+in two papers by @karlsson and @junkka and we only present a brief confirmatory analysis.
 
-```{r formneo}
-formneo <- as.formula("oe(event, exposure) ~  extemp + extemp.1 + emeantemp + socst + period + urban +
-                     subreg + sex + parity")
-tabneo <- readRDS("data/tabneo.rds")
-```
+No strong signs of interactions with marginTemp and expTemp data, so these factors strike
+almost evenly over different subpopulations. Moreover, the effects are stronger during winter,
+with a decrease in mortality for each degree temperature increase, almost the same amount 
+for both, around 1.2 per cent.
+
+Separate analyses for winter and summer, see Tables \@ref(tab:neowinter) and 
+\@ref(tab:neosummer) follow.
 
@@
 
-This period refers to the months *October to March*. A Cox regression involves as
-interesting variables *highTemp*, an indicator of temperature at least four degrees
-above the expected for at least two weeks in a row, *emeantemp* the expected temperature
-the actual weeek, and *extemp* the *excess temperature* the actual week.
-
-```{r neoanalwp, results = 'asis', cache = TRUE}
-
-##library(survival)
-winter <- tabneo[!tabneo$summer, ]
-##winter$season <- factor(winter$season)
-##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
-##neo$enter <- g(neo$enter * 365)
-fit.neow <- tpchreg(formneo, data = winter, time = age)
-dr <- drop1(fit.neow, test = "Chisq")
+```{r neowinter, results = "asis"}
+##neo$marginTemp <- neo$extemp.1
+##neo$expTemp <- neo$emeantemp.1
+fit.neow <- coxreg(Surv(enter, exit, event) ~ expTemp + marginTemp + socst + 
+                      sex + urban + parity + I(year - 1920),
+                  data = neo[!neo$summer, ])
+drw <- drop1(fit.neow, test = "Chisq")
+ltx(fit.neow, dr = drw, caption = "Neonatal mortality, winter.", label = "tab:neowinter")
 ```
 
-```{r neoanalw, results = "asis", cache = FALSE}
-ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March.", 
-     label = "tab:neoanalw", keep = NULL)
+The result of a Cox regression of neonatal mortality during winter is shown in Table \@ref(tab:neowinter).
+There is a (statistically) strong effect of *expTemp*, an increase of the expected temperature
+lowers neonatal mortality by around 1.3 per cent per degree. Almost the same effect is shown for *marginTemp*.
+
+The interaction between social status and expTemp is shown in Figure \@ref(fig:plotwcsn). Nothing interesting is
+going on, the effect of expTemp on neonatal mortality is the same in all social classes ("no interaction" between
+expTemp and socst).
+
+```{r plotwcsn, fig.cap = "Socioeconomic status and expTemp, winter. Neonatal mortality.", fig.height = 3}
+source("R/socfig.R")
+winter <- neo[!neo$summer, ]
+fitw <- coxreg(Surv(enter, exit, event) ~ socst:expTemp + socst, data = winter)
+fitw0 <- coxreg(Surv(enter, exit, event) ~ socst + expTemp, data = winter)
+p <- anova(fitw, fitw0)[[4]][2]
+rang <- range(winter$expTemp)
+socfig(coef(fitw), main = "", pval = p, ran = rang, ymax = 4)
 ```
 
-#### Temperature and sex
+The interaction between social status and marginTemp is shown in Figure \@ref(fig:plotwwsn). A little
+bit more action here, but the conclusion stays the same: No interaction between marginTemp and social class,
+bad marginTemp hits all by the same strength.
 
-We check the interaction between temperature and sex and period.
-
-```{r neosexper}
-fit <- tpchreg(oe(event, exposure) ~ sex * (extemp + extemp.1 + emeantemp), 
-               time = age, data = winter)
-dr <- drop1(fit, test = "Chisq")
-dr
+```{r plotwwsn, fig.cap = "Socioeconomic status and marginTemp, winter. Neonatal mortality.", fig.height = 3}
+source("R/socfig.R")
+fitww <- coxreg(Surv(enter, exit, event) ~ socst:marginTemp + socst, data = winter)
+fitww0 <- coxreg(Surv(enter, exit, event) ~ socst + marginTemp, data = winter)
+p <- anova(fitww, fitww0)[[4]][2]
+rang <- range(winter$marginTemp)
+socfig(coef(fitww), main = "", pval = p, ran = rang, ymax = 4)
 ```
-
-Obviously no temperature and sex interactions
 
@@
 
-This period refers to the months *April to September*. A Cox regression involves as
-interesting variables *highTemp*, an indicator of temperature at least four degrees
-above the expected for at least two weeks in a row, *emeantemp* the expected temperature
-the actual weeek, and *extemp* the *excess temperature* the actual week.
+The summer result of a Cox regression is shown in Table \@ref(tab:neosummer).
+Compared to winter, the *expTemp* effect is in the same direction but weaker, while 
+the effect of *marginTemp* is zero.
 
-```{r neoanalsp, cache = TRUE}
-##library(survival)
-#summer <- neo[neo$summer, ]
-##summer$season <- factor(summer$season)
-##neo$exit <- g(neo$exit * 365) # Skip the log-cube transformation
-##neo$enter <- g(neo$enter * 365)
-fit.neos <- tpchreg(formneo, data = tabneo[tabneo$summer, ], time = age)
-dr.neos <- drop1(fit.neos, test = "Chisq")
+```{r neosummer, results = "asis"}
+##neo$marginTemp <- neo$extemp.1
+##neo$expTemp <- neo$emeantemp.1
+fit.neos <- coxreg(Surv(enter, exit, event) ~ expTemp + marginTemp + socst + 
+                      sex + urban + parity + I(year - 1920),
+                  data = neo[neo$summer, ])
+##data = neo[neo$week %in% 20:32, ])
+drs <- drop1(fit.neos, test = "Chisq")
+ltx(fit.neos, dr = drs, caption = "Neonatal mortality, summer.", label = "tab:neosummer")
 ```
 
-```{r neoanals, results = 'asis', cache = FALSE}
-ltx2(fit.neos, dr = dr.neos, caption = "Neonatal mortality, April to September.", 
-     label = "tab:neoanals", keep = NULL)
+During summer, expTemp and marginTemp effects are clearly smaller, and regarding marginTemp, the opposite 
+direction: Excess heat increases the death risk, however not very much and not statistically significant.
+
+The interaction between social status and *expTemp* is shown in Figure \@ref(fig:plotscsn). 
+Nothing of interest is going on.
+
+```{r plotscsn, fig.cap = "Socioeconomic status and expTemp, summer. Neonatal mortality.", fig.height = 3}
+source("R/socfig.R")
+summer <- neo[neo$summer, ]
+fits <- coxreg(Surv(enter, exit, event) ~ socst:expTemp + socst, data = summer)
+fits0 <- coxreg(Surv(enter, exit, event) ~ socst + expTemp, data = summer)
+p <- anova(fits, fits0)[[4]][2]
+rang <- range(summer$expTemp)
+socfig(coef(fits), main = "", pval = p, ran = rang, ymax = 4)
+```
+
+The interaction between social status and *marginTemp* is shown in Figure \@ref(fig:plotswsn). The same conclusion as 
+for *expTemp*, nothing of interest shows up.
+
+```{r plotswsn, fig.cap = "Socioeconomic status and marginTemp, summer. Neonatal mortality.", fig.height = 3}
+source("R/socfig.R")
+fitsw <- coxreg(Surv(enter, exit, event) ~ socst:marginTemp + socst, data = summer)
+fitsw0 <- coxreg(Surv(enter, exit, event) ~ socst + marginTemp, data = summer)
+p <- anova(fitsw, fitsw0)[[4]][2]
+rang <- range(summer$marginTemp)
+socfig(coef(fitsw), main = "", pval = p, ran = rang, ymax = 4)
 ```
@@
 ```{r setformw}
-form <- as.formula("oe(event, exposure) ~ socst + extemp + extemp.1 + emeantemp + urban + subreg + sex + parity")
-formex <- as.formula("oe(event, exposure) ~ socst + socst:extemp + extemp.1 + emeantemp + urban + subreg + sex + parity")
-formex1 <- as.formula("oe(event, exposure) ~ socst + socst:extemp.1 + extemp + emeantemp + urban + subreg + sex + parity")
-formem <- as.formula("oe(event, exposure) ~ socst + socst:emeantemp + extemp + extemp.1 + urban + subreg + sex + parity")
-form.sb <-  as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * socst + urban + subreg + sex + parity")
-form.sr <- as.formula("oe(event, exposure) ~ (extemp + extemp.1 + emeantemp) * subreg + socst + urban + sex + parity") 
+form <- as.formula("oe(event, exposure) ~ socst + extemp.1 + eexpTemp.1 + urban + subreg + sex + parity")
+
+formex1 <- as.formula("oe(event, exposure) ~ socst + socst:extemp.1 + eexpTemp.1 + urban + subreg + sex + parity")
+formem <- as.formula("oe(event, exposure) ~ socst + socst:eexpTemp.1 + extemp.1 + urban + subreg + sex + parity")
+form.sb <-  as.formula("oe(event, exposure) ~ extemp.1 * socst + urban + subreg + sex + parity")
+form.sr <- as.formula("oe(event, exposure) ~ (extemp.1 + eexpTemp.1) * subreg + socst + urban + sex + parity") 
 tabpost <- readRDS("data/tabpost.rds")
-```
-
-#### By birth month
-
-See Figure \@ref(fig:bybmonth).
-
-```{r bybmonth, fig.cap = "Crude postneonatal relative (to January = 1) mortality by birth month.", cache = TRUE}
-pn <- age.window(infdat, c(28/365, 1))
-fit <- coxreg(Surv(enter, exit, event) ~ as.factor(birthmonth), data = pn)
-plot(exp(c(0, coef(fit))), type = "b", xlab = "Birth month", ylab = "Hazard ratios",
-     col = "blue")
-abline(h = 1)
 ```
@@
 
-```{r postanalwp, results='asis', cache = TRUE}
-##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socst + illeg + parity
-  ##              + subreg + period, data = infdat, cuts = c(0, 1))
-##winter <- gtpneo[gtpneo$summer == 0, ]
-##winter$season <- factor(winter$season)
+See Table \@ref(tab:allpneo) for winter results. Variation in normal temperature (expTemp) has a large and 
+statistically significant effect on postneonatal mortality. The same goes for the marginal 
+temperature effect, however not so pregnant as fot the expTemp.
 
-fit.postw <- tpchreg(oe(event, exposure) ~ period * (extemp + extemp.1 + emeantemp), 
-                     data = tabpost[!tabpost$summer, ], time = age)
-dr.postw <- drop1(fit.postw, test = "Chisq")
+```{r allpneo, results = "asis", cache = TRUE}
+##pneo$expTemp <- pneo$eexpTemp.1
+##pneo$marginTemp <- pneo$extemp.1
+winter <- pneo[!pneo$summer, ]
+fit.pneo <- coxreg(Surv(enter, exit, event) ~ expTemp + marginTemp + sex + socst +
+                        urban + parity + I(year - 1920), data = winter)
+dr <- drop1(fit.pneo, test = "Chisq")
+ltx(fit.pneo, dr = dr, caption = "Postneonatal mortality, winter.", label = "tab:allpneo")
 ```
 
-Interaction tests with temperature and period, postneonatal mortality:
+The interaction between social status and expTemp is shown in Figure \@ref(fig:plotwcs). All except the 
+upper white collar have almost the same (negative) slope, the upper class do not have much problem with normal 
+temperature variations.
 
-```{r postanalw, cache = FALSE}
-round(dr.postw, 4)
-#tbl(round(dr.postw, 4), caption = "Interaction tests with temperature and period, postneonatal ##mortality.", rownames = TRUE)
+```{r plotwcs, fig.cap = "Socioeconomic status and expTemp, winter. Postneonatal mortality.", fig.height = 3}
+source("R/socfig.R")
+fitw <- coxreg(Surv(enter, exit, event) ~ socst:expTemp + socst, data = winter)
+fitw0 <- coxreg(Surv(enter, exit, event) ~ socst + expTemp, data = winter)
+p <- anova(fitw, fitw0)[[4]][2]
+rang <- range(winter$expTemp)
+socfig(coef(fitw), main = "", pval = p, ran = rang, ymax = 8)
 ```
 
-This shows that the temperature effects do not vary much with time period,
-but we continue with separate analyses for the time periods `r levels(tabpost$period)`. Just
-to make sure, will be joined when appropriate.
+The interaction between social status and marginTemp is shown in Figure \@ref(fig:plotwws). Here the situation
+is different for the upper class: It follows essentially the same path (on a lower level) as the rest.
 
-#### Sex, urban,  and temperature
-
-```{r postanalwpsex, results='asis', cache = TRUE}
-##tpneo <- toTpch(Surv(enter, exit, event) ~ lowTemp + highTemp + extemp + emeantemp + sex + socst + illeg + parity
-  ##              + subreg + period, data = infdat, cuts = c(0, 1))
-##winter <- gtpneo[gtpneo$summer == 0, ]
-##winter$season <- factor(winter$season)
-
-fit.postwsex <- tpchreg(oe(event, exposure) ~ (sex + urban) * (extemp + extemp.1 + emeantemp), 
-                     data = tabpost[!tabpost$summer, ], time = age)
-dr.postwsex <- drop1(fit.postwsex, test = "Chisq")
-```
-
-Interaction tests with temperature and sex, postneonatal mortality:
-
-```{r postanalwsex, cache = FALSE}
-round(dr.postwsex, 4)
-#tbl(round(dr.postw, 4), caption = "Interaction tests with temperature and period, urban, postneonatal ##mortality.", rownames = TRUE)
-```
-
-Obviously no interactions between temperature and sex, and urban.
-
-#### The full time period
-
-```{r postw1p, results = "asis"}
-fit.postw1 <- tpchreg(form, data = tabpost[(!tabpost$summer), ], time = age)
-dr.postw1 <- drop1(fit.postw1, test = "Chisq")
-```
-
-```{r postw1, results = "asis", cache = FALSE}
-ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914.", 
-     label = "tab:postw1", keep = NULL)
-```
-
-Table \@ref(tab:postw1) shows that the effects of  extreme temperatures have the expected directions, extra high temperatures is *positive* 
-(remember that this concerns winter conditions), but not statistically significant. *Climate* (emeantemp), on the other hand shows to be very
-important, where increasing temperature by one degree lower mortality by 0.7 per cent. The normal short term fluctuations (extemp) are not so important, 
-but in the expected direction.
-
-The importance of social class (socBranch) in explaining the effects of climate and weather is negligible, but there is variation between
-the regions regarding *climate*:
-
-```{r postw1srpu, cache = TRUE}
-fit.postw1sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
-                        data = tabpost[(!tabpost$summer) & tabpost$subreg == "ume", ], time = age)
-dr.postw1sru <- drop1(fit.postw1sru, test = "Chisq")
-```
-
-```{r postw1sru, results = "asis"}
-ltx2(fit.postw1sru, dr = dr.postw1sru, caption = "Postneonatal mortality, Umeå.",
-     label = "tab:postw1sru", keep = NULL)
-```
-
-```{r postw1srps, cache = TRUE}
-fit.postw1srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
-                        data = tabpost[(!tabpost$summer) & tabpost$subreg != "ume", ], time = age)
-dr.postw1srs <- drop1(fit.postw1srs, test = "Chisq")
-```
-
-```{r postw1srs, results = "asis"}
-ltx2(fit.postw1srs, dr = dr.postw1srs, caption = "Postneonatal mortality, Skellefteå with inland.", 
-     label = "tab:postw1srs", keep = NULL)
-```
-
-We can conclude from Tables \@ref(tab:postw1sru) and \@ref(tab:postw1srs) that the main difference between Umeå and Skellefteå is that
-*climate* has more severe effect in Skellefteå (with inland), while *weather* is less important.
-
-The effects of temperature in different social classes are best shown graphically, see 
-Figure \@ref(fig:socw1).
-
-```{r socw1, fig.cap = "Interactions between social status and temperature, winter."}
+```{r plotwws, fig.cap = "Socioeconomic status and marginTemp, winter. Postneonatal mortality.", fig.height = 3}
 source("R/socfig.R")
-oldpar <- par(mfrow = c(1,3))
-tempdat <- tabpost[(!tabpost$summer), ]
-fit <- tpchreg(formem, data = tempdat, time = age)
-fitex <- tpchreg(formex, data = tempdat, time = age)
-fitex1 <- tpchreg(formex1, data = tempdat, time = age)
-socfig(coef(fit), main = "Weekly mean temp")
-socfig(coef(fitex), main = "Excess temp", ran = c(-10, 10))
-socfig(coef(fitex1), main = "Lagged excess temp", ran = c(-10, 10))
-par(oldpar)
+fitww <- coxreg(Surv(enter, exit, event) ~ socst:marginTemp + socst, data = winter)
+fitww0 <- coxreg(Surv(enter, exit, event) ~ socst + marginTemp, data = winter)
+p <- anova(fitww, fitww0)[[4]][2]
+rang <- range(winter$marginTemp)
+socfig(coef(fitww), main = "", pval = p, ran = rang, ymax = 8)
 ```
@@
 
-```{r postanalsp, results='asis', cache = TRUE}
-##summer <- gtpneo[gtpneo$summer == 1, ]
-##summer$season <- factor(summer$season)
+See Table \@ref(tab:allpneos) for summer results. The normal temperature variation 
+still has a strong importance, while temporary variations on the margin have a  
+slightly lower effect on the postneonatal mortality.
 
-fit.posts <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + period + 
-                        subreg + parity + sex, 
-                    data = tabpost[tabpost$summer, ], time = age)
-dr <- drop1(fit.posts, test = "Chisq")
+```{r allpneos, results = "asis", cache = TRUE}
+summer <- pneo[pneo$summer, ]
+fit.pneos <- coxreg(Surv(enter, exit, event) ~ expTemp + marginTemp + sex + socst + urban +
+                         parity + I(year - 1920), data = summer)
+drs <- drop1(fit.pneos, test = "Chisq")
+ltx(fit.pneos, dr = drs, caption = "Postneonatal mortality, summer.", label = "tab:allpneos")
 ```
 
-```{r postanals, results='asis', cache = FALSE}
-ltx2(fit.posts, dr = dr, caption = "Postneonatal mortality, April to September. Adjusted for sex, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
+The interaction between social status and expTemp is shown in Figure \@ref(fig:plotscs).
+The impression is that the upper class suffer more of higher expected temperature (that is, 
+the months July and August) than farmers and workers, with lower white collars and qualified workers somwhere in between.
+
+```{r plotscs, fig.cap = "Socioeconomic status and expTemp, summer. Postneonatal mortality.", cache = TRUE, fig.height = 3}
+source("R/socfig.R")
+fits <- coxreg(Surv(enter, exit, event) ~ socst:expTemp + socst, data = summer)
+fits0 <- coxreg(Surv(enter, exit, event) ~ socst + expTemp, data = summer)
+p <- anova(fits, fits0)[[4]][2]
+rang <- range(summer$expTemp)
+socfig(coef(fits), main = "", pval = p, ran = rang, ymax = 8)
 ```
 
-The result in Table \@ref(tab:postanals) shows that *climate* (`emeantemp`) is more important than 
-*weather* (`exTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
-covariates (not shown).
+The interaction between social status and marginTemp is shown in Figure \@ref(fig:plotsws). 
+The pattern on the margin very much resembles that of the expected variations.
 
-#### The full time period
-
-```{r posts1p, results = "asis"}
-fit.posts1 <- tpchreg(form, data = tabpost[(tabpost$summer), ], time = age)
-dr.posts1 <- drop1(fit.posts1, test = "Chisq")
-```
-
-```{r posts1, results = "asis", cache = FALSE}
-ltx2(fit.posts1, dr = dr.posts1, caption = "Postneonatal mortality, October to March.", 
-     label = "tab:posts1", keep = NULL)
-```
-
-Table \@ref(tab:posts1) shows that the effects of  extreme temperatures have the expected directions, extra high temperatures is *positive* 
-(remember that this concerns summer conditions), but not statistically significant. *Climate* (emeantemp), on the other hand shows to be very
-important, where increasing temperature by one degree lower mortality by 0.7 per cent. The normal short term fluctuations (extemp) are not so important, 
-but in the expected direction.
-
-The importance of social class (socBranch) in explaining the effects of climate and weather is negligible, but there is variation between
-the regions regarding *climate*:
-
-```{r posts1srpu, cache = TRUE}
-fit.posts1sru <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
-                        data = tabpost[(tabpost$summer) & tabpost$subreg == "ume", ], time = age)
-dr.posts1sru <- drop1(fit.posts1sru, test = "Chisq")
-```
-
-```{r posts1sru, results = "asis"}
-ltx2(fit.posts1sru, dr = dr.posts1sru, caption = "Postneonatal mortality, Umeå, October to March.",
-     label = "tab:posts1sru", keep = NULL)
-```
-
-```{r posts1srps, cache = TRUE}
-fit.posts1srs <- tpchreg(oe(event, exposure) ~ extemp + extemp.1 + emeantemp + socst + urban + sex + parity, 
-                        data = tabpost[(tabpost$summer) & tabpost$subreg != "ume", ], time = age)
-dr.posts1srs <- drop1(fit.posts1srs, test = "Chisq")
-```
-
-```{r posts1srs, results = "asis"}
-ltx2(fit.posts1srs, dr = dr.posts1srs, caption = "Postneonatal mortality, Skellefteå with inland, October to March.", 
-     label = "tab:posts1srs", keep = NULL)
-```
-
-We can conclude from Tables \@ref(tab:posts1sru) and \@ref(tab:posts1srs) that the main difference between Umeå and Skellefteå is that
-*climate* has more severe effect in Skellefteå (with inland), while *weather* is less important.
-
-The effects of temperature in different social classes are best shown graphically, see 
-Figure \@ref(fig:soc2).
-
-```{r soc1, fig.cap = "Interactions between social status and temperature, summer."}
+```{r plotsws, fig.cap = "Socioeconomic status and marginTemp, summer. Postneonatal mortality.", cache = TRUE, fig.height = 3}
 source("R/socfig.R")
-oldpar <- par(mfrow = c(1,3))
-tempdat <- tabpost[(tabpost$summer), ]
-fit <- tpchreg(formem, data = tempdat, time = age)
-fitex <- tpchreg(formex, data = tempdat, time = age)
-fitex1 <- tpchreg(formex1, data = tempdat, time = age)
-socfig(coef(fit), main = "Weekly mean temp")
-socfig(coef(fitex), main = "Excess temp", ran = c(-10, 10))
-socfig(coef(fitex1), main = "Lagged excess temp", ran = c(-10, 10))
-par(oldpar)
+fitsw <- coxreg(Surv(enter, exit, event) ~ socst:marginTemp + socst, data = summer)
+fitsw0 <- coxreg(Surv(enter, exit, event) ~ socst + marginTemp, data = summer)
+p <- anova(fitsw, fitsw0)[[4]][2]
+rang <- range(summer$marginTemp)
+socfig(coef(fitsw), main = "", pval = p, ran = rang, ymax = 8 )
 ```
@@
 
-Remains to be written, especially the discussion about temperature and mortality.
+In this study, we differentiate between neonatal and post-neonatal mortality in an effort to understand the role of resources to protect infants from both normal and extreme ambient temperatures in northern Sweden 1895-1950.
+Our results show that, like in other parts of Sweden, continuously declining infant mortality.
+They show that the health penalty if living in a city has disappeared among infants [@scb69].
+They also show the typical u-shape mortality pattern of birth-order and higher mortality for boys than girls.
+The neonatal mortality was about the same for all social groups in the winter and slightly lower for children of higher social classes in the summer.
+This is much as expected since mortality shortly after birth often is not related to social and economic resources but to specific maternal factors, what Bourgeois-Pichat calls endogenous factors.
+Still, external factors, such as ambient temperature, affects mortality in first month of life.
+It is not so much extremely low or high temperatures that are important but the seasonality, the annual variation, in temperature and particularly the low temperature during the winter season.
+The way temperature affect different social groups are small but slightly larger in extreme weather conditions.
+
+For postneonatal mortality, we find a similar decline, a u-shaped patter for birth-order, and higher mortality for boys than for girls, as for neonatal mortality.
+We find much larger social differences in mortality than during the first months of life, similar to what has been found for southern Sweden [@dk21].
+In fact, during the summer mortality among children of lower white collar, farmers, and workers are twice as high comparative to children of higher classes.
+
+During the winter, the differences are even larger.
+The effects of temperature are also much larger and, again, the effect the normal seasonal variation that mattered the most, especially for mortality during the summer season.
+Thus, postneonatal mortality was much more sensitive than neonatal mortality, both to parental and external factors.
+The ways that temperature affects different social groups are also much more pronounced than during the first month of life.
+Summer temperatures have no impact on mortality among children of higher classes, neither normal variation nor extreme situations.
+Other social groups have higher mortality during cold summer weeks.
+Turning to the effects of winter temperatures, while the normal seasonally, that is the average weekly temperature, has a strong influence on postneonatal mortality among all other social classes, it has no impact at all on white-collar workers.
+Extremely cold winters, however, also affects children of white-collar workers.
+Still, the differences are not statistically significant possibly due to the fact that higher white collar was a small group in this period.
+
+Taken together, we find that postneonatal mortality is far more sensitive both to individual, family, and external factors, such as temperature, than mortality in first month of life.
+During normal temperature variations, children of higher classes did not suffer from increased mortality, only during extremely cold winters.
+The fact that children of higher classes do not suffer from increased mortality during normal winter suggests that the effect of low temperature is not due to low humidity, which would affect all classes similarly, but to temperature itself.
+
+# Session info
 

### 2022-02-24 R/normal.R

@@ -67,6 +67,7 @@ normal <- function(period = c(1894, 1951)){
     temp$cold.1 <- c(NA, temp$cold[-n])
     temp$extemp.1 <- c(NA, temp$extemp[-n])
     temp$heat.1 <- c(NA, temp$heat[-n])
+    temp$emeantemp.1 <- c(NA, temp$emeantemp[-n])
     ##
     temp
 }

### 2022-02-24 R/socfig.R

@@ -1,4 +1,4 @@
-socfig <- function(coeff, ran = c(-25, 5), main = "1935-1950", pval = 0.03){
+socfig <- function(coeff, ran = c(-25, 5), main = "1935-1950", pval = 0.03, ymax = NULL){
     ## coeff is output from tpchreg(oe(event, exposure) ~ socst + socst:emeantemp 
     ## controls+)
     ny <- coeff[substr(names(coeff), 1, 5) == "socst"]
@@ -7,18 +7,24 @@ socfig <- function(coeff, ran = c(-25, 5), main = "1935-1950", pval = 0.03){
     x <- seq(ran[1], ran[2], length = 100)
     y <- matrix(nrow = 6, ncol = length(x))
     for (j in 1:6){
-        ##y[j, ] <- exp(a[j] + b[j] * x)
-        y[j, ] <- exp(b[j] * x)
+        y[j, ] <- exp(a[j] + b[j] * x)
+        ##y[j, ] <- exp(b[j] * x)
     }
-    ymax <- max(y)
+    if (is.null(ymax)){
+        ymax <- max(y)
+    }
     plot(x, y[1, ], type = "l", ylim = c(0, ymax), xlab = "Temperature", 
          ylab = "Rel. hazards", lwd = 2, main = main)
-    for (i in 2:6){
+    ##for (i in 2:6){
+    for (i in 1:5){                                          ### !! NOTE !!
         lines(x, y[i, ], col = i, lty = i, lwd = 2)
     }
-    nam <- c("WhiteCollar", "Farmer",  "lowWhiteC", "qualWorker", "Worker", "None")
-    legend("topright", legend = nam, col = 1:6, lty = 1:6, cex = 0.7)
+    ##
+    ##nam <- c("WhiteCollar", "Farmer",  "lowWhiteC", "qualWorker", "Worker", "None")
+    ##legend("topright", legend = nam, col = 1:6, lty = 1:6, cex = 0.7)
+    nam <- c("WhiteCollar", "Farmer",  "lowWhiteC", "qualWorker", "Worker")
+    legend("topright", legend = nam, col = 1:5, lty = 1:5, cex = 0.8)
     points(0, 1, cex = 2)
     abline(h = 1, v = 0, lty = 3)
-    text(0, ymax - 1, paste("p = ", round(pval, 3)), col = ifelse(pval < 0.05, "red", "black"))
[…]

### 2022-02-24 R/tempdat.R

@@ -51,6 +51,7 @@ tempdat <- function(filename = "ume", period = c(1894, 1951)){
     temp$cold.1 <- c(NA, temp$cold[-n])
     temp$extemp.1 <- c(NA, temp$extemp[-n])
     temp$heat.1 <- c(NA, temp$heat[-n])
+    temp$emeantemp.1 <- c(NA, temp$emeantemp[-n])
     ##
     temp
 }

### 2022-04-06 getdata.Rmd
fresh start on revision

--- 2022-02-18 getdata.Rmd
@@
 ---
-title: "getdata4.Rmd"
+title: "getdata.Rmd"
 author: "Göran Broström"
@@
 
-** Run this (and only this) before running "paper4.Rmd"!**
+** Run this (and only this) before running "paper.Rmd"!**
 
@@
 lagg <- 0 / 365 ##3 / 365  ## NOTE!!
-tempvars <-  c("heat", "heat.1", "extemp", "extemp.1", "cold", "cold.1", "emintemp", "emeantemp", "emaxtemp", 
-               "week", "year") 
+##tempvars <-  c("heat", "heat.1", "extemp", "extemp.1", "cold", "cold.1", "emintemp", "emeantemp", "emeantemp.1",  "emaxtemp", 
+  ##             "week", "year")
+tempvars <- c("expTemp", paste0("expTemp.", 1:4), "margTemp", paste0("margTemp.", 1:4), "week", "year")
 ## Ume:
@@
 
-saveRDS(infdat, file = "data/infdat4.rds") # NOTE: Name change!!!
+saveRDS(infdat, file = "Rdata/infdat4.rds") # NOTE: Name change!!!
 

### 2022-04-06 paper.Rmd
fresh start on revision

@@
 ---
-title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895--1950"
+title: "Infant mortality and temperature, northern Sweden 1895--1950"
 author: "Göran Broström and Tommy Bengtsson"
@@
 - \renewcommand{\efloatseparator}{\mbox{}}
-abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895--1950 is studied in a biometric analysis setting.
+abstract: "The effect of extreme temperatures on infant mortality in the Umeå and Skellefteå regions 1895--1950 is studied.
 More precisely, the effect of climate and weather, measured by temperature, average and extremes, on infant mortality is investigated.
@@
 births <- readRDS("data/births4.rds")
-infdat <- readRDS("data/infdat4.rds")
-infdat$marginTemp <- round(infdat$extemp.1)
-infdat$expTemp <- round(infdat$emeantemp.1)
+infdat <- readRDS("Rdata/infdat4.rds")
+infdat$margTemp <- round(infdat$margTemp)
+infdat$margTemp.1 <- round(infdat$margTemp.1)
+infdat$margTemp.2 <- round(infdat$margTemp.2)
+infdat$margTemp.3 <- round(infdat$margTemp.3)
+infdat$margTemp.4 <- round(infdat$margTemp.4)
+infdat$expTemp <- round(infdat$expTemp)
+infdat$expTemp.1 <- round(infdat$expTemp.1)
+infdat$expTemp.2 <- round(infdat$expTemp.2)
+infdat$expTemp.3 <- round(infdat$expTemp.3)
+infdat$expTemp.4 <- round(infdat$expTemp.4)
 ```
@@
 ##infdat$lowTemp <- with(infdat, (extemp < -3) & (extemp.1 < -3))
-infdat$lowTemp <- (with(infdat, cold + cold.1 < -10))
-infdat$highTemp <- with(infdat, (extemp > 3) & (extemp.1 > 3))
+##infdat$lowTemp <- (with(infdat, cold + cold.1 < -10))
+##infdat$highTemp <- with(infdat, (extemp > 3) & (extemp.1 > 3))
 ##infdat$weather <- infdat$extemp.1
 ##infdat$climate <- round(infdat$emeantemp)
-tbl(head(infdat[, c("enter", "exit", "event", "marginTemp", "expTemp", "week", "year")]),
+tbl(head(infdat[, c("enter", "exit", "event", "margTemp.1", "expTemp.1", "week", "year")]),
     caption = "Data with communal covariates.", fs = 11)
@@
 
-In the following we split the analyses into the neonatal and the postneonatal periods, with
-a focus on the latter. But we start with a short comment on the link to *biometric* analysis
-à la @bp51a.
-
-## The biometric finding
-
-What happens if we extend the truncated Weibull back into neonatal territory? The answer
-is given in Figure \@ref(fig:neowei).
-
-```{r neowei, fig.cap = "Weibull hazard functions, total and exogenous neonatal mortality.", fig.height = 4}
-b3 <- age.window(b1, c(0, 28 / 365))
-fit.w0 <- phreg(Surv(enter, exit, event) ~ 1, data = b3)
-plot(fit.w0, fn = "haz", axes = FALSE, xlab = "Days", main = "", lwd = 1.5, col = "red", ylim = c(0, 3))
-#fit.ph <- tpchreg(oe(event, exposure) ~ 1, data = inftab, time = age)
-##plot(fit.ph1, fn = "haz", axes = FALSE, xlab = "Days", main = "", lwd = 1.5, col = "red",
-  ##   xlim = c(0, 28 / 365))
-dor <- seq(0, 28, by = 2)
-axis(1, at = dor / 365, labels = dor)
-axis(2, las = 1)
-box()
-##abline(v = 28 / 365, lty = 3)
-abline( h = 0, v = 0)
-pars <- exp(fit$coefficients)
-x <- seq(0, 28/365, length = 1000)
-y <- hweibull(x, shape = pars[2], scale = pars[1])
-lines(x, y, col = "blue", lwd = 1.5)
-legend("topright", legend = c("Weibull postneonatal (exogenous), extrapolated", "Weibull neonatal (total)"), col = c("blue", "red"), lwd = c(1.5, 1))
-```
-
-Not very good fit in this territory. The difference here is that to the exogenous mortality, endogenous mortality is added.
-The difference between the piecewise constant and the Weibull in the figure would be 
-called the *endogeneous infant mortality* [@bp51a; @bp51b].
-
 ## Neonatal mortality
@@
 ##neo$expTemp <- neo$emeantemp.1
-fit.neow <- coxreg(Surv(enter, exit, event) ~ expTemp + marginTemp + socst + 
+fit.neow <- coxreg(Surv(enter, exit, event) ~ expTemp + expTemp.1 + expTemp.2 + expTemp.3 + expTemp.4 + 
+                       margTemp + margTemp.1 + margTemp.2 + margTemp.3 + margTemp.4 + socst + 
                       sex + urban + parity + I(year - 1920),
@@
 source("R/socfig.R")
-fitww <- coxreg(Surv(enter, exit, event) ~ socst:marginTemp + socst, data = winter)
-fitww0 <- coxreg(Surv(enter, exit, event) ~ socst + marginTemp, data = winter)
+fitww <- coxreg(Surv(enter, exit, event) ~ socst:margTemp + socst, data = winter)
+fitww0 <- coxreg(Surv(enter, exit, event) ~ socst + margTemp, data = winter)
 p <- anova(fitww, fitww0)[[4]][2]
-rang <- range(winter$marginTemp)
+rang <- range(winter$margTemp)
 socfig(coef(fitww), main = "", pval = p, ran = rang, ymax = 4)
@@
 ```{r neosummer, results = "asis"}
-##neo$marginTemp <- neo$extemp.1
+##neo$margTemp <- neo$extemp.1
 ##neo$expTemp <- neo$emeantemp.1
-fit.neos <- coxreg(Surv(enter, exit, event) ~ expTemp + marginTemp + socst + 
+fit.neos <- coxreg(Surv(enter, exit, event) ~ expTemp + margTemp + socst + 
                       sex + urban + parity + I(year - 1920),
@@
 
-During summer, expTemp and marginTemp effects are clearly smaller, and regarding marginTemp, the opposite 
+During summer, expTemp and margTemp effects are clearly smaller, and regarding margTemp, the opposite 
 direction: Excess heat increases the death risk, however not very much and not statistically significant.
@@
 
-The interaction between social status and *marginTemp* is shown in Figure \@ref(fig:plotswsn). The same conclusion as 
+The interaction between social status and *margTemp* is shown in Figure \@ref(fig:plotswsn). The same conclusion as 
 for *expTemp*, nothing of interest shows up.
 
-```{r plotswsn, fig.cap = "Socioeconomic status and marginTemp, summer. Neonatal mortality.", fig.height = 3}
+```{r plotswsn, fig.cap = "Socioeconomic status and margTemp, summer. Neonatal mortality.", fig.height = 3}
 source("R/socfig.R")
-fitsw <- coxreg(Surv(enter, exit, event) ~ socst:marginTemp + socst, data = summer)
-fitsw0 <- coxreg(Surv(enter, exit, event) ~ socst + marginTemp, data = summer)
+fitsw <- coxreg(Surv(enter, exit, event) ~ socst:margTemp + socst, data = summer)
+fitsw0 <- coxreg(Surv(enter, exit, event) ~ socst + margTemp, data = summer)
 p <- anova(fitsw, fitsw0)[[4]][2]
-rang <- range(summer$marginTemp)
+rang <- range(summer$margTemp)
 socfig(coef(fitsw), main = "", pval = p, ran = rang, ymax = 4)
@@
 ##pneo$expTemp <- pneo$eexpTemp.1
-##pneo$marginTemp <- pneo$extemp.1
+##pneo$margTemp <- pneo$extemp.1
 winter <- pneo[!pneo$summer, ]
-fit.pneo <- coxreg(Surv(enter, exit, event) ~ expTemp + marginTemp + sex + socst +
+fit.pneo <- coxreg(Surv(enter, exit, event) ~ expTemp + margTemp + sex + socst +
                         urban + parity + I(year - 1920), data = winter)
@@
 
-The interaction between social status and marginTemp is shown in Figure \@ref(fig:plotwws). Here the situation
+The interaction between social status and margTemp is shown in Figure \@ref(fig:plotwws). Here the situation
 is different for the upper class: It follows essentially the same path (on a lower level) as the rest.
 
-```{r plotwws, fig.cap = "Socioeconomic status and marginTemp, winter. Postneonatal mortality.", fig.height = 3}
+```{r plotwws, fig.cap = "Socioeconomic status and margTemp, winter. Postneonatal mortality.", fig.height = 3}
 source("R/socfig.R")
-fitww <- coxreg(Surv(enter, exit, event) ~ socst:marginTemp + socst, data = winter)
-fitww0 <- coxreg(Surv(enter, exit, event) ~ socst + marginTemp, data = winter)
+fitww <- coxreg(Surv(enter, exit, event) ~ socst:margTemp + socst, data = winter)
+fitww0 <- coxreg(Surv(enter, exit, event) ~ socst + margTemp, data = winter)
 p <- anova(fitww, fitww0)[[4]][2]
-rang <- range(winter$marginTemp)
+rang <- range(winter$margTemp)
 socfig(coef(fitww), main = "", pval = p, ran = rang, ymax = 8)
@@
 summer <- pneo[pneo$summer, ]
-fit.pneos <- coxreg(Surv(enter, exit, event) ~ expTemp + marginTemp + sex + socst + urban +
+fit.pneos <- coxreg(Surv(enter, exit, event) ~ expTemp + margTemp + sex + socst + urban +
                          parity + I(year - 1920), data = summer)
@@
 
-The interaction between social status and marginTemp is shown in Figure \@ref(fig:plotsws). 
+The interaction between social status and margTemp is shown in Figure \@ref(fig:plotsws). 
 The pattern on the margin very much resembles that of the expected variations.
 
-```{r plotsws, fig.cap = "Socioeconomic status and marginTemp, summer. Postneonatal mortality.", cache = TRUE, fig.height = 3}
+```{r plotsws, fig.cap = "Socioeconomic status and margTemp, summer. Postneonatal mortality.", cache = TRUE, fig.height = 3}
 source("R/socfig.R")
-fitsw <- coxreg(Surv(enter, exit, event) ~ socst:marginTemp + socst, data = summer)
-fitsw0 <- coxreg(Surv(enter, exit, event) ~ socst + marginTemp, data = summer)
+fitsw <- coxreg(Surv(enter, exit, event) ~ socst:margTemp + socst, data = summer)
+fitsw0 <- coxreg(Surv(enter, exit, event) ~ socst + margTemp, data = summer)
 p <- anova(fitsw, fitsw0)[[4]][2]
-rang <- range(summer$marginTemp)
+rang <- range(summer$margTemp)
 socfig(coef(fitsw), main = "", pval = p, ran = rang, ymax = 8 )

### 2022-04-06 R/normal.R

@@ -58,16 +58,26 @@ normal <- function(period = c(1894, 1951)){
     indx <- tapply(temp$week, temp$week)
     temp$emintemp <- tapply(temp$mintemp, temp$week, mean)[indx]
     temp$emaxtemp <- tapply(temp$maxtemp, temp$week, mean)[indx]
-    temp$emeantemp <- tapply(temp$meantemp, temp$week, mean)[indx]
+    temp$expTemp <- tapply(temp$meantemp, temp$week, mean)[indx]
     temp$heat <- round(temp$maxtemp - temp$emaxtemp)
     temp$cold <- round(temp$mintemp - temp$emintemp)
-    temp$extemp <- round(temp$meantemp - temp$emeantemp)
+    temp$margTemp <- round(temp$meantemp - temp$expTemp)
     # NEW 28 Jan 2022:
     n <- NROW(temp)
     temp$cold.1 <- c(NA, temp$cold[-n])
-    temp$extemp.1 <- c(NA, temp$extemp[-n])
+    temp$margTemp.1 <- c(NA, temp$margTemp[-n])
     temp$heat.1 <- c(NA, temp$heat[-n])
-    temp$emeantemp.1 <- c(NA, temp$emeantemp[-n])
+    temp$expTemp.1 <- c(NA, temp$expTemp[-n])
+    ##
+    ## New 28 March 2022:
+    temp$margTemp.2 <- c(NA, temp$margTemp.1[-n])
+    temp$expTemp.2 <- c(NA, temp$expTemp.1[-n])
+    ## New 28 March 2022:
+    temp$margTemp.3 <- c(NA, temp$margTemp.2[-n])
+    temp$expTemp.3 <- c(NA, temp$expTemp.2[-n])
+    ## New 28 March 2022:
+    temp$margTemp.4 <- c(NA, temp$margTemp.3[-n])
+    temp$expTemp.4 <- c(NA, temp$expTemp.3[-n])
     ##
     temp
 }

### 2022-04-06 R/tempdat.R

@@ -42,16 +42,27 @@ tempdat <- function(filename = "ume", period = c(1894, 1951)){
     indx <- tapply(temp$week, temp$week)
     temp$emintemp <- tapply(temp$mintemp, temp$week, mean)[indx]
     temp$emaxtemp <- tapply(temp$maxtemp, temp$week, mean)[indx]
-    temp$emeantemp <- tapply(temp$meantemp, temp$week, mean)[indx]
+    temp$expTemp <- tapply(temp$meantemp, temp$week, mean)[indx]
     temp$heat <- round(temp$maxtemp - temp$emaxtemp)
     temp$cold <- round(temp$mintemp - temp$emintemp)
-    temp$extemp <- round(temp$meantemp - temp$emeantemp)
+    temp$margTemp <- round(temp$meantemp - temp$expTemp)
     # NEW 28 Jan 2022:
     n <- NROW(temp)
     temp$cold.1 <- c(NA, temp$cold[-n])
-    temp$extemp.1 <- c(NA, temp$extemp[-n])
+    temp$margTemp.1 <- c(NA, temp$margTemp[-n])
     temp$heat.1 <- c(NA, temp$heat[-n])
-    temp$emeantemp.1 <- c(NA, temp$emeantemp[-n])
+    temp$expTemp.1 <- c(NA, temp$expTemp[-n])
     ##
+    ## New 28 March 2022:
+    temp$margTemp.2 <- c(NA, temp$margTemp.1[-n])
+    temp$expTemp.2 <- c(NA, temp$expTemp.1[-n])
+    ## New 28 March 2022:
+    temp$margTemp.3 <- c(NA, temp$margTemp.2[-n])
+    temp$expTemp.3 <- c(NA, temp$expTemp.2[-n])
+    ## New 28 March 2022:
+    temp$margTemp.4 <- c(NA, temp$margTemp.3[-n])
+    temp$expTemp.4 <- c(NA, temp$expTemp.3[-n])
+    ##
+    
     temp
 }

### 2022-04-06 R/ploteff.R
first commit

@@ -1,0 +1,10 @@
+ploteff <- function(){
+    library(effects)
+    library(eha)
+    post <- readRDS("data/tabpost.rds")
+    fit <- glm(event ~ offset(log(exposure)) + age + urban + emeantemp:urban, data = post[!post$summer, ],
+               family = poisson())
+    fit.tp <- tpchreg(oe(event, exposure) ~ urban + emeantemp:urban, data = post[!post$summer, ], time = age)
+    ##plot(predictorEffect(mod = fit, predictor = "emeantemp"))
+    list(fit = fit, fit.tp = fit.tp)
+}
